---
title: "00 - Site Map (Holzmaar)"
author: "Holzmaar Research Team"
date: "2026"
output:
  html_notebook:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

# Overview

This notebook creates individual publication-quality maps for the Holzmaar paleoclimate study:

**Panel A:** Regional context map of the Westeifel Volcanic Field (Germany) - Shows topography, rivers, and major towns (Köln, Koblenz, Mainz) - Includes inset map of Germany with Holzmaar highlighted - Elevation data overlay (real or synthetic)

**Germany Inset:** Shows the location of the study area within Germany - Displays Bundesland (state) boundaries - Marks Holzmaar location with red triangle

**Panel B:** Local Holzmaar bathymetric map - Lake shoreline polygon from digitized bathymetric survey - Isobaths (depth contours) with color gradient - Core locations marked with red diamonds - Scale bar in meters

**Output Format:** All maps are saved as individual PNG files at 300 dpi

## Getting Exact Bathymetry Data

### ⚡ Quickest Start: OpenStreetMap Data

Generate map files instantly from OSM:

``` r
source("../data/SourceData/Map/FETCH_OSM_DATA.R")
# Generates: Holzmaar_shoreline_xy.csv, Holzmaar_isobaths_xy.csv, Holzmaar_core_locations_xy.csv
# Then move these files to ../data/SourceData/ and re-run this notebook
```

This gives you: - OSM lake shoreline (accurate where mapped) - Synthetic isobaths based on crater lake morphology - Core location markers

### Publication Quality: Digitize from Literature

For exact bathymetry in your manuscript:

1.  See `../data/SourceData/Map/DIGITIZATION_WORKFLOW.md` for a complete guide to:
    - Finding and downloading the published bathymetric map PDF (Zolitschka 1998 or García et al. 2024)
    - Using WebPlotDigitizer to extract contours
    - Exporting as calibrated CSV coordinates
2.  Save the digitized coordinates with these exact names in `../data/SourceData/`:
    - `Holzmaar_shoreline_xy.csv` (lake polygon)
    - `Holzmaar_isobaths_xy.csv` (depth contours)
    - `Holzmaar_core_locations_xy.csv` (core site coordinates)
3.  Re-run this notebook — Panel B will automatically render with your exact geometry

### Optional: Add Regional DEM Context

To add elevation data to Panel A: - Run the script `../data/SourceData/Map/DEM_ACQUISITION.R` - This fetches free SRTM data via the `elevatr` package - Or download high-resolution DEM from GeoBasis RLP (German state data portal)

**If exact files are missing,** the notebook shows a synthetic placeholder so it remains runnable.

------------------------------------------------------------------------

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(pacman)

# Keep a clean package state for reproducibility.
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load(
  ggplot2,
  sf,
  rnaturalearth,
  rnaturalearthdata,
  readr,
  ggrepel,
  dplyr,
  patchwork,
  viridis,
  maptiles,
  terra,
  tmap,
  elevatr
)

path <- getwd()
setwd(path)

# Ensure figs directory exists
if (!dir.exists("../figs")) {
  dir.create("../figs", showWarnings = FALSE)
  cat("✓ Created ../figs directory\n")
}

cat("R Version:", R.version$version.string, "\n")
```

------------------------------------------------------------------------

# Panel A: Regional Context (Eifel Region Topographic Map with Cities & Rivers)

```{r panel_a, message=FALSE, warning=FALSE}
# Holzmaar coordinates (DMS: Degrees, Minutes, Seconds format)
# 6° 53' 0" E = 6.88333°E
# 50° 7' 0" N = 50.11667°N
holzmaar_lon <- 6 + 53/60  # 6.88333
holzmaar_lat <- 50 + 7/60  # 50.11667

cat("═════════════════════════════════════════════════════════\n")
cat("PANEL A: Eifel Region Topographic Map\n")
cat("═════════════════════════════════════════════════════════\n")
cat("Holzmaar coordinates: ", holzmaar_lat, "°N, ", holzmaar_lon, "°E\n\n", sep = "")

# Define study area bounding box (wider Eifel region)
bbox <- st_bbox(c(
  xmin = 5.8, xmax = 9.5,
  ymin = 49.3, ymax = 51.8
), crs = st_crs(4326))

# Get Natural Earth data
cat("Loading geographic data from Natural Earth...\n")

# Countries
germany <- ne_countries(scale = "medium", country = "Germany", returnclass = "sf")

# Neighboring countries for context
countries <- ne_countries(scale = "medium", returnclass = "sf")
europe <- countries %>%
  filter(st_intersects(geometry, st_as_sfc(bbox), sparse = FALSE))

# Rivers (Natural Earth 10m scale)
cat("  Loading rivers data...\n")
rivers_all <- ne_download(scale = 10, type = "rivers_lake_centerlines", category = "physical", 
                          returnclass = "sf", load = TRUE)

# Crop rivers to study area
rivers <- rivers_all %>%
  st_crop(bbox)

# Try to fetch elevation data using more reliable methods
cat("  Attempting to fetch REAL elevation data...\n")
elevation_raster <- NULL

# Method 1: geodata::elevation_global() - Copernicus DEM (most reliable for Europe)
tryCatch({
  cat("    Trying geodata::elevation_global() - Copernicus 30m resolution...\n")
  cat("    (This downloads ~50-100MB, may take 2-5 minutes)\n")
  
  # Need to install geodata first if not available
  if (!require(geodata, quietly = TRUE)) {
    cat("    Installing geodata package...\n")
    install.packages("geodata", quiet = TRUE)
    library(geodata)
  }
  
  # Download elevation data for Eifel region
  elevation_raster <- geodata::elevation_global(
    lon = 7.5,      # Center of study area
    lat = 50.3,
    res = 0.5,      # 30 arc-seconds (~1km)
    path = tempdir()
  )
  
  # Crop to exact bbox
  elevation_raster <- terra::crop(elevation_raster, terra::ext(5.8, 9.5, 49.3, 51.8))
  
  cat("    ✓ SUCCESS: Downloaded REAL Copernicus DEM 30m elevation data!\n")
  cat("    Elevation range: ", min(values(elevation_raster), na.rm=TRUE), "-", 
      max(values(elevation_raster), na.rm=TRUE), "m\n")
  
}, error = function(e) {
  cat("    ✗ geodata::elevation_global() failed:", conditionMessage(e), "\n")
})

# Method 2: Try downloading GEBCO data (alternative)
if (is.null(elevation_raster)) {
  tryCatch({
    cat("    Trying geodata::elevation_global() with GEBCO...\n")
    
    elevation_raster <<- geodata::elevation_global(
      lon = 7.5,
      lat = 50.3,
      res = 1,
      path = tempdir()
    )
    
    # Crop to bbox
    elevation_raster <<- terra::crop(elevation_raster, terra::ext(5.8, 9.5, 49.3, 51.8))
    
    cat("    ✓ SUCCESS: Downloaded GEBCO elevation data!\n")
  }, error = function(e) {
    cat("    ✗ GEBCO download failed:", conditionMessage(e), "\n")
  })
}

# Method 3: Use elevatr as fallback (more reliable than direct downloads)
if (is.null(elevation_raster)) {
  tryCatch({
    cat("    Trying elevatr::get_elev_raster() with OpenTopography...\n")
    
    elevation_raster <<- elevatr::get_elev_raster(
      locations = bbox,
      z = 10,
      src = "ol"
    )
    
    cat("    ✓ SUCCESS: Downloaded elevation data from OpenTopography!\n")
  }, error = function(e) {
    cat("    ✗ elevatr OpenTopography failed:", conditionMessage(e), "\n")
  })
}

# Method 4: Final fallback to AWS
if (is.null(elevation_raster)) {
  tryCatch({
    cat("    Trying elevatr::get_elev_raster() with AWS...\n")
    
    elevation_raster <<- elevatr::get_elev_raster(
      locations = bbox,
      z = 10,
      src = "aws"
    )
    
    cat("    ✓ SUCCESS: Downloaded elevation data from AWS!\n")
  }, error = function(e) {
    cat("    ✗ AWS download failed:", conditionMessage(e), "\n")
  })
}

# If still no data, provide helpful diagnostic
if (is.null(elevation_raster)) {
  cat("\n")
  cat("⚠ WARNING: All real elevation data sources failed to download.\n")
  cat("═════════════════════════════════════════════════════════════\n")
  cat("POSSIBLE ISSUES:\n")
  cat("• Internet connection problems or firewall blocking downloads\n")
  cat("• Insufficient disk space in tempdir() (need >500MB)\n")
  cat("• Geodata servers temporarily unavailable\n")
  cat("\nTROUBLESHOOTING:\n")
  cat("1. Update packages: update.packages(c('terra','geodata','elevatr'))\n")
  cat("2. Restart R session\n")
  cat("3. Check disk space: disk.free(tempdir())\n")
  cat("4. Try again - sometimes network issues are temporary\n")
  cat("═════════════════════════════════════════════════════════════\n\n")
  
  cat("Creating SYNTHETIC elevation model as fallback...\n")
  
  x <- seq(5.8, 9.5, length.out = 150)
  y <- seq(49.3, 52, length.out = 130)
  grid <- expand.grid(x = x, y = y)
  
  grid$elev <- 400 + 
    100 * sin((grid$x - 6.9) * 2.5) + 
    80 * cos((grid$y - 50.2) * 2) +
    50 * sin((grid$x - grid$y) * 1.8) +
    rnorm(nrow(grid), 0, 8)
  
  elevation_raster <- rast(
    nrows = 130,
    ncols = 150,
    xmin = 5.8, xmax = 9.5,
    ymin = 49.3, ymax = 52.3,
    vals = pmax(grid$elev, 200),
    crs = "EPSG:4326"
  )
  
  cat("⚠ Using SYNTHETIC topography (not real data).\n")
  cat("   Map will display with simulated elevation.\n\n")
}

cat("  ✓ Elevation data ready for mapping\n\n")

# Create major cities data for the region
cat("Creating major cities layer...\n")
cities_data <- data.frame(
  name = c("Köln", "Koblenz", "Mainz", "Holzmaar"),
  lon = c(6.96, 7.61, 8.27, holzmaar_lon),
  lat = c(50.94, 50.36, 49.97, holzmaar_lat),
  is_holzmaar = c(FALSE, FALSE, FALSE, TRUE)
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = st_crs(4326))

cat("  ✓ Cities layer created\n\n")

# Create the tmap with elevation data as base
cat("Building topographic map with elevation data...\n")

panel_a <- tm_shape(elevation_raster, bbox = bbox) +
  tm_raster(
    palette = "terrain",
    style = "quantile",
    n = 12,
    title = "Elevation (m)",
    alpha = 0.85,
    colorNA = "#CCCCCC",
    show.legend = FALSE
  ) +
  
  # Add rivers on top
  tm_shape(rivers, bbox = bbox) +
  tm_lines(col = "#1A5F99", lwd = 1.2, alpha = 0.8) +
  
  # Add cities
  tm_shape(cities_data %>% filter(!is_holzmaar), bbox = bbox) +
  tm_symbols(
    shape = 20,  # Circle
    size = 0.7,
    fill = "darkblue",
    col = "black",
    col.alpha = 0.9,
    border.lwd = 1.3
  ) +
  tm_text(text = "name", size = 1.15, col = "darkblue", fontface = "bold",
          xmod = 0.75, ymod = 0.65, shadow = TRUE) +
  
  # Add Holzmaar (highlighted)
  tm_shape(cities_data %>% filter(is_holzmaar), bbox = bbox) +
  tm_symbols(
    shape = 24,  # Triangle (pointing up)
    size = 1.1,
    fill = "red",
    col = "darkred",
    col.alpha = 0.95,
    border.lwd = 1.6
  ) +
  tm_text(text = "name", size = 1.35, col = "darkred", fontface = "bold",
          xmod = 0.75, ymod = -0.65, shadow = TRUE) +
  
  # Scale bar
  tm_scale_bar(
    breaks = c(0, 25, 50),
    position = c(0.65, 0.15),
    text.size = 1.0,
    bg.color = "white",
    bg.alpha = 0.85
  ) +
  
  # Compass / North arrow
  tm_compass(
    type = "arrow",
    position = c(0.75, 0.93),
    size = 1.3,
    color.dark = "#333333",
    color.light = "white"
  ) +
  
  # Coordinate grid with degree lines
  tm_graticules(
    labels.show = TRUE,
    labels.size = 0.85,
    n.x = 8,  # Show lines at regular intervals (whole degrees)
    n.y = 5,  # Show lines at regular intervals (whole degrees)
    lwd = 0.6,
    alpha = 0.6,
    col = "#666666",
    ticks = TRUE
  ) +
  
  # Title and legend
 # tm_title("A) Eifel Region - Topographic Map", size = 1.7, fontface = "bold") +
  
  # Layout
  tm_layout(
    legend.show = FALSE,
    frame = TRUE,
    frame.lwd = 1.5,
    bg.color = "white",
    inner.margins = c(0.02, 0.02, 0.02, 0.02),
    outer.margins = c(0.02, 0.02, 0.02, 0.02),
    fontfamily = "sans"
  )

# Switch to static plot mode
tmap_mode("plot")

print(panel_a)

cat("✓ Panel A created successfully\n")
cat("  - Shows Germany and surrounding countries\n")
cat("  - Rivers highlighted in blue\n")
cat("  - Major cities: Köln, Koblenz, Mainz (dark blue circles)\n")
cat("  - Holzmaar marked with red triangle\n")
cat("  - Scale bar (0-50-100 km) and north arrow included\n")
cat("═════════════════════════════════════════════════════════\n\n")

# Create simple inset map of Germany
cat("Creating inset map of Germany...\n")

germany_full <- ne_countries(scale = "medium", country = "Germany", returnclass = "sf")
study_box <- st_as_sfc(bbox)

# Create Holzmaar point for inset map
holzmaar_point <- st_point(c(holzmaar_lon, holzmaar_lat)) %>%
  st_sfc(crs = st_crs(4326)) %>%
  st_as_sf()

# Load Bundesland (state) boundaries using ne_download
states <- ne_download(scale = 10, type = "admin_1_states_provinces", category = "cultural", 
                      returnclass = "sf", load = TRUE)
# Clean geometry - show all states in Germany on the map
states_germany <- states %>%
  st_make_valid() %>%
  st_intersection(germany_full)

inset_map <- tm_shape(germany_full) +
  tm_fill(fill = "lightgray", alpha = 0.4) +
  tm_borders(lwd = 0.8, col = "gray40") +
  tm_shape(states_germany) +
  tm_borders(lwd = 0.5, col = "gray60", lty = "dashed") +
  tm_shape(study_box) +
  tm_borders(col = "red", lwd = 3) +
  tm_shape(holzmaar_point) +
  tm_symbols(shape = 24, size = 0.8, fill = "red", col = "darkred", border.lwd = 1.2) +
  tm_layout(
    frame = TRUE,
    frame.lwd = 1.2,
    bg.color = "white",
    inner.margins = c(0.05, 0.05, 0.05, 0.05),
    outer.margins = c(0, 0, 0, 0),
    #title = "Location in Germany",
    title.size = 0.8
  )

print(inset_map)
cat("  ✓ Inset map created\n\n")

# ===================================================================
# SAVE INSET MAP AND PANEL A (Both as Individual PNG Files)
# ===================================================================
cat("Saving individual maps as PNG files...\n")

# Save inset map as PNG
tmap_save(
  inset_map,
  filename = "../figs/Figure_01_Inset_Germany_Map.png",
  width = 4,
  height = 5,
  dpi = 300
)
cat("  ✓ Saved: Figure_01_Inset_Germany_Map.png\n")

# Save Panel A as PNG
tmap_save(
  panel_a,
  filename = "../figs/Figure_01_Regional_Germany_Map.png",
  width = 8,
  height = 7,
  dpi = 300
)
cat("  ✓ Saved: Figure_01_Regional_Germany_Map.png\n\n")
```

------------------------------------------------------------------------

# Exact Input Format

```{r input_format}
cat("Expected column format:\n")
cat("  Holzmaar_shoreline_xy.csv: x_m, y_m\n")
cat("  Holzmaar_isobaths_xy.csv: contour_id, depth_m, x_m, y_m, vertex_order\n")
cat("  Holzmaar_core_locations_xy.csv: core_id, x_m, y_m\n")
cat("\nAll coordinates should be in local projected meters (same CRS/source).\n")
```

------------------------------------------------------------------------

# Panel B: Holzmaar Bathymetry

```{r panel_b}
# Paths to data files
shoreline_file <- "../data/SourceData/Map/countour.csv"
isoline_dir <- "../data/SourceData/Map/"
coring_file <- "../data/SourceData/Map/coring.csv"

if (file.exists(shoreline_file)) {
  cat("=== BATHYMETRIC MAP WITH OPENSTREETMAP BACKGROUND ===\n\n")
  
  # ===================================================================
  # STEP 1: Load shoreline
  # ===================================================================
  
  shoreline_raw <- readr::read_csv(shoreline_file, col_names = c("x_m", "y_m"), show_col_types = FALSE) %>%
    filter(!is.na(x_m), !is.na(y_m))
  
  cat("Shoreline points loaded:", nrow(shoreline_raw), "\n")
  
  # Sort shoreline circularly
  center_x <- mean(shoreline_raw$x_m)
  center_y <- mean(shoreline_raw$y_m)
  
  shoreline_sorted <- shoreline_raw %>%
    select(x_m, y_m)
  # Use original order from file - points were digitized in sequence around shoreline
  
  # Close the shoreline loop first (add first point at end)
  shoreline_closed <- bind_rows(shoreline_sorted, slice(shoreline_sorted, 1))
  
  # Smooth shoreline with spline interpolation
  # Using parametric splines for closed curves
  t_shore <- seq(0, 1, length.out = nrow(shoreline_closed))
  t_shore_new <- seq(0, 1, length.out = nrow(shoreline_closed) * 2.5)  # 2.5x interpolation
  
  x_shore_spline <- spline(t_shore, shoreline_closed$x_m, xout = t_shore_new, method = "natural")
  y_shore_spline <- spline(t_shore, shoreline_closed$y_m, xout = t_shore_new, method = "natural")
  
  shoreline_sorted <- data.frame(
    x_m = x_shore_spline$y,
    y_m = y_shore_spline$y
  )
  
  cat("Shoreline smoothed:", nrow(shoreline_sorted), "points\n")
  
  # ===================================================================
  # STEP 2: Load all isoline files
  # ===================================================================
  # Each file is one radial cross-section at a specific depth
  # File number = depth (isoline20.csv = 20m depth, isoline1.csv = 2m depth, etc.)
  # Reconstruct circular contours by sorting points by angle for each depth
  
  isoline_files <- list.files(isoline_dir, pattern = "^[Ii]soline\\d+\\.csv$", full.names = FALSE)
  
  cat("Found", length(isoline_files), "isoline files\n")
  cat("Loading bathymetric contours (keeping original point order)...\n\n")
  
  # Load all points from all isoline files
  # KEEP ORIGINAL ORDER - points naturally form contours
  all_isolines <- data.frame(x_m = numeric(), y_m = numeric(), isoline_file = numeric())
  
  for (iso_file in isoline_files) {
    # Extract depth from filename: isoline20.csv → depth 20, isoline1.csv → depth 1
    iso_num <- as.numeric(gsub("[^0-9]", "", iso_file))
    
    iso_data <- tryCatch({
      read_csv(file.path(isoline_dir, iso_file), col_names = c("x_m", "y_m"), 
               show_col_types = FALSE) %>%
        filter(!is.na(x_m), !is.na(y_m)) %>%
        mutate(isoline_file = iso_num)
    }, error = function(e) {
      cat("  ✗ Error reading", iso_file, "\n")
      NULL
    })
    
    if (!is.null(iso_data)) {
      cat("  Isoline", iso_num, ":", nrow(iso_data), "points\n")
      all_isolines <- bind_rows(all_isolines, iso_data)
    }
  }
  
  # Create closed polygons by adding first point to end
  # This closes each contour without re-arranging points
  isoline_polygons <- all_isolines %>%
    group_by(isoline_file) %>%
    bind_rows(., slice(., 1)) %>%
    ungroup()
  
  # ===================================================================
  # Smooth contours using spline interpolation
  # ===================================================================
  # Interpolate additional points along each contour for smooth curves
  
  cat("Smoothing contours with spline interpolation...\n")
  
  smooth_contours <- data.frame()
  
  for (depth_val in unique(isoline_polygons$isoline_file)) {
    contour_points <- isoline_polygons %>%
      filter(isoline_file == depth_val) %>%
      select(x_m, y_m)
    
    if (nrow(contour_points) > 3) {
      # Create parameter t for parametric curve (0 to 1)
      t <- seq(0, 1, length.out = nrow(contour_points))
      t_new <- seq(0, 1, length.out = nrow(contour_points) * 3)  # 3x more interpolated points
      
      # Spline interpolation for x and y coordinates separately
      x_spline <- spline(t, contour_points$x_m, xout = t_new, method = "natural")
      y_spline <- spline(t, contour_points$y_m, xout = t_new, method = "natural")
      
      smoothed <- data.frame(
        x_m = x_spline$y,
        y_m = y_spline$y,
        isoline_file = depth_val
      )
      
      smooth_contours <- bind_rows(smooth_contours, smoothed)
    } else {
      # Keep original if too few points
      smooth_contours <- bind_rows(smooth_contours, 
                                   mutate(contour_points, isoline_file = depth_val))
    }
  }
  
  # Update isoline_polygons with smoothed contours
  isoline_polygons <- smooth_contours %>%
    select(x_m, y_m, isoline_file)
  
  cat("✓ Contours smoothed\n")
  cat("\nTotal points in contours:", nrow(all_isolines), "\n")
  cat("Depth levels:", n_distinct(all_isolines$isoline_file), "\n")
  cat("Points per depth level (after smoothing):\n")
  print(isoline_polygons %>% group_by(isoline_file) %>% summarise(n = n(), .groups = "drop"))
  cat("\n")
  
  # ===================================================================
  # STEP 3: Load coring points
  # ===================================================================
  
  cores <- NULL
  if (file.exists(coring_file)) {
    cores <- readr::read_csv(coring_file, col_names = c("x_m", "y_m"), show_col_types = FALSE) %>%
      filter(!is.na(x_m), !is.na(y_m))
    cat("Coring points loaded:", nrow(cores), "\n\n")
  }
  
  # ===================================================================
  # STEP 4: Create the bathymetric map
  # ===================================================================
  
  x_min <- min(shoreline_sorted$x_m, na.rm = TRUE)
  x_max <- max(shoreline_sorted$x_m, na.rm = TRUE)
  y_min <- min(shoreline_sorted$y_m, na.rm = TRUE)
  y_max <- max(shoreline_sorted$y_m, na.rm = TRUE)
  
  x_range <- x_max - x_min
  y_range <- y_max - y_min
  
# Expand plot area below the lake to accommodate scale bar and its labels
  y_min_expanded <- y_min - 0.15 * y_range
  y_max_expanded <- y_max + 0.05 * y_range
  
  
  # Scale bar: 100 m at bottom right (positioned lower)
  bar_width <- 100
  bar_x0 <- x_max - bar_width - 0.1 * x_range
  bar_x1 <- bar_x0 + bar_width
  bar_y <- y_min - 0.05 * y_range
  
  # Build the map with filled polygons
  panel_b <- ggplot() +
    # Shoreline fill (light blue background)
    geom_polygon(
      data = shoreline_sorted,
      aes(x = x_m, y = y_m),
      fill = "#E8F4F8",
      color = NA
    ) +
    # Filled isoline polygons - color gradient (light=shallow, dark=deep)
    geom_polygon(
      data = isoline_polygons,
      aes(x = x_m, y = y_m, fill = isoline_file, group = isoline_file),
      color = NA,
      alpha = 0.85
    ) +
    # Shoreline outline (bold black)
    geom_path(
      data = shoreline_sorted,
      aes(x = x_m, y = y_m),
      color = "black",
      linewidth = 1.0
    ) +
    # Color scale for depth: light blue (shallow) to dark blue (deep)
    scale_fill_gradient(
      name = "Depth (m)",
      low = "#D4E6F1",
      high = "#1B4965",
      guide = guide_colorbar(
        title.position = "top",
        title.hjust = 0.5,
        barwidth = 1.2,
        barheight = 4,
        label.theme = element_text(size = 12)
      )
    )
  
  # Add coring points if available
  if (!is.null(cores)) {
    panel_b <- panel_b +
      geom_point(
        data = cores,
        aes(x = x_m, y = y_m),
        color = "red",
        size = 3.5,
        shape = 18,
        alpha = 0.8
      )
  }
  
  # Add scale bar
  panel_b <- panel_b +
   annotate("segment", x = bar_x0, xend = bar_x1, y = bar_y, yend = bar_y, linewidth = 1.5) +
    annotate("segment", x = bar_x0, xend = bar_x0, y = bar_y - 8, yend = bar_y + 8, linewidth = 1.5) +
    annotate("segment", x = bar_x1, xend = bar_x1, y = bar_y - 8, yend = bar_y + 8, linewidth = 1.5) +
    annotate("text", x = bar_x0, y = bar_y - 16, label = "0", size = 4, hjust = 0.5, vjust = 1) +
    annotate("text", x = bar_x1, y = bar_y - 16, label = "100 m", size = 4, hjust = 0.5, vjust = 1) +
    ylim(y_min_expanded, y_max_expanded) +
    coord_equal(expand = FALSE) +
    #labs(title = "B") +
    theme_void(base_size = 12) +
    theme(
      #plot.title = element_text(face = "bold", size = 28, hjust = 0.02),
      panel.border = element_rect(color = NA, fill = NA, linewidth = 0),
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),  # Increased bottom margin for scale bar
      legend.position = "right",
      legend.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.key.size = unit(1.2, "cm"),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.box.background = element_rect(fill = "transparent", color = NA)
    )
  
  cat("✓ Bathymetric map created successfully\n")
  cat("✓", nrow(all_isolines), "isoline points plotted\n")
  if (!is.null(cores)) cat("✓", nrow(cores), "coring points marked\n")
  
  # Save Panel B as PNG
  cat("\nSaving Panel B (Bathymetric Map) as PNG...\n")
  ggsave(
    filename = "../figs/Figure_01_Bathymetric_Map.png",
    plot = panel_b,
    width = 8,
    height = 9.5,
    dpi = 300,
    device = "png"
  )
  cat("  ✓ Saved: Figure_01_Bathymetric_Map.png\n\n")
  
  print(panel_b)

} else {
  cat("✗ Shoreline file not found at:", shoreline_file, "\n")
  cat("Creating synthetic placeholder...\n\n")
  
  # Synthetic fallback
  theta <- seq(0, 2 * pi, length.out = 500)
  r <- 130 + 12 * sin(2 * theta) + 9 * cos(3 * theta)
  shoreline <- data.frame(
    x_m = 170 + r * cos(theta),
    y_m = 160 + 0.88 * r * sin(theta)
  )

  depth_levels <- seq(2, 20, by = 2)
  isobaths <- lapply(depth_levels, function(d) {
    shrink <- 1 - (d / 28)
    data.frame(
      contour_id = paste0("c", d),
      depth_m = -d,
      x_m = 170 + shrink * (r - 5) * cos(theta) + 2 * sin(theta),
      y_m = 160 + 0.88 * shrink * (r - 5) * sin(theta)
    )
  }) %>% bind_rows()

  contour_labels <- isobaths %>%
    group_by(depth_m) %>%
    slice(40) %>%
    ungroup() %>%
    select(x_m, y_m, depth_m)

  panel_b <- ggplot() +
    geom_polygon(data = shoreline, aes(x = x_m, y = y_m), fill = "gray96", color = NA) +
    geom_path(data = isobaths, aes(x = x_m, y = y_m, group = contour_id), color = "gray20", linewidth = 0.45) +
    geom_path(data = shoreline, aes(x = x_m, y = y_m), color = "black", linewidth = 1.0) +
    geom_text(data = contour_labels, aes(x = x_m, y = y_m, label = abs(depth_m)), color = "gray20", size = 4.2, fontface = "italic") +
    coord_equal(expand = TRUE) +
    theme_void(base_size = 12) +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    )
  
  # Save synthetic Panel B as PNG
  cat("\nSaving Panel B (Synthetic Bathymetric Map) as PNG...\n")
  ggsave(
    filename = "../figs/Figure_01_Bathymetric_Map_Synthetic.png",
    plot = panel_b,
    width = 8,
    height = 9.5,
    dpi = 300,
    device = "png"
  )
  cat("  ✓ Saved: Figure_01_Bathymetric_Map_Synthetic.png (placeholder)\\n\\n")
  
  print(panel_b)
}
```

------------------------------------------------------------------------

# Summary

All individual maps have been created and saved as PNG files in the `../figs/` directory:

- **Figure_01_Regional_Germany_Map.png** — Topographic map of the Eifel region (Panel A)
- **Figure_01_Inset_Germany_Map.png** — Germany map with state boundaries and Holzmaar location
- **Figure_01_Bathymetric_Map.png** — Bathymetric map of Holzmaar lake (Panel B)

Each map is saved individually at 300 dpi for publication quality.
