---
title: "XRF Data Normalization and Transformation"
author: "Holzmaar Research Team"
date: "2025"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: inline
---

# Overview

This notebook performs standardized XRF (X-ray Fluorescence) data processing following published protocols (Bertrand et al. 2024; Jansen et al. 1998).

## Processing Steps:
1. Load ITRAX XRF measurements and HSI calibrated data
2. Clean data (remove invalid measurements, NaNs)
3. Normalize XRF counts to percentages
4. Apply Centered Log Ratio (CLR) transformation for compositional data
5. Apply Additive Log Ratio (ALR) transformation for subset analysis

## Input Data:
- `SourceData/HZM19_LG_section_XRF.xlsx` - ITRAX XRF raw measurements
- `HZM_HSI-XRF_1mm_calibrated_napari_2025-01-13_V3.xlsx` - HSI calibrated data

## Output Data:
- `XRF_HZM_norm.xlsx` - Normalized XRF data (% counts)
- `XRF_HZM_clr.xlsx` - CLR-transformed XRF data
- `XRF_HZM_alr.xlsx` - ALR-transformed selected elements
- `HZM_HR_cleaned.xlsx` - Cleaned high-resolution dataset

---

# Setup

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(pacman)

# Unload any previously loaded packages to ensure clean environment
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

# Load required packages
pacman::p_load(
  "readxl",      # Read Excel files
  "dplyr",       # Data manipulation
  "tidyr",       # Data tidying
  "openxlsx"     # Write Excel files
)

# Set working directory
path <- getwd()
setwd(path)

# Document R version for reproducibility
cat("R Version:", R.version$version.string, "\n")
```

---

# Step 1: Load Raw Data

```{r load_data, message=FALSE}
# Load ITRAX XRF measurements
# This dataset contains elemental counts (in counts per second) from ITRAX scanner
HZM_XRF <- read_excel("../data/SourceData/HZM_xrf_raw_measurements.xlsx")

# Load high-resolution combined HSI-XRF dataset
# HSI data calibrated using spectrophotometer measurements
HZM_HR_all <- read_excel("../data/SourceData/HZM_hsi_xrf_calibrated_1mm.xlsx")

cat("XRF data dimensions:", nrow(HZM_XRF), "rows ×", ncol(HZM_XRF), "columns\n")
cat("HR data dimensions:", nrow(HZM_HR_all), "rows ×", ncol(HZM_HR_all), "columns\n")
```

---

# Step 2: Data Cleaning

## 2.1 Select Relevant Elements

```{r clean_select}
# Select XRF elemental data (removes Ar/air background and non-elemental columns)
# Columns 13-17, 19-43, 60-61, 66-70 contain counted elements
data_XRF <- select(HZM_XRF, 13:17, 19:43, 60:61, 66:70)

# Select HR data corresponding to XRF elements and auxiliary data
# Columns 1-9: calibrated elements; 23-27, 29-53: pigments and scaling; 
# 70-71, 75-79: depth/age model
data_HR <- select(HZM_HR_all, 1:9, 23:27, 29:53, 70:71, 75:79)

cat("XRF selected columns:", ncol(data_XRF), "\n")
cat("HR selected columns:", ncol(data_HR), "\n")
```

## 2.2 Remove Missing Values

```{r clean_na}
# Remove rows with any NaN/NA values
data_XRF_cl <- na.omit(data_XRF)
data_HR_cl <- na.omit(data_HR)

cat("XRF after NaN removal:", nrow(data_XRF_cl), "rows\n")
cat("HR after NaN removal:", nrow(data_HR_cl), "rows\n")
cat("Rows removed from XRF:", nrow(data_XRF) - nrow(data_XRF_cl), "\n")
```

## 2.3 Save Cleaned HR Dataset

```{r save_hr_cleaned}
write.xlsx(data_HR_cl, "../data/HZM_HR_cleaned.xlsx", rowNames = FALSE)
cat("Cleaned HR data saved to: ../data/HZM_HR_cleaned.xlsx\n")
```

## 2.4 Extract XRF-Only Elements

```{r extract_xrf}
# Select only elemental data (first 30 columns contain XRF elements)
data_XRF_only <- select(data_XRF_cl, 1:30)

cat("XRF element columns:", ncol(data_XRF_only), "\n")
```

## 2.5 Filter High-Zero Columns

```{r filter_zeros}
# Remove elements with >5% zero values
# Zero values indicate detection limit or measurement failure
# These columns are unsuitable for log-ratio transformations

pre_clean_XRF <- data_XRF_only %>% 
  select_if(~ mean(. == 0) <= 0.05)

cat("After filtering high-zero columns:", ncol(pre_clean_XRF), "elements remain\n")
cat("Columns removed:", ncol(data_XRF_only) - ncol(pre_clean_XRF), "\n")
```

---

# Step 3: Normalization to Count Percentages

## 3.1 Normalize to % Counts

```{r normalize}
# Convert elemental counts to percentages
# This removes matrix effects from different sample compositions
# Formula: Element % = (Element cps / Sum of all elements cps) × 100
# Reference: Bertrand et al. (2024); Jansen et al. (1998)

HZM_norm <- pre_clean_XRF / rowSums(pre_clean_XRF) * 100

# Verify normalization (each row should sum to ~100)
row_sums <- rowSums(HZM_norm)
cat("Min row sum:", round(min(row_sums), 2), "%\n")
cat("Max row sum:", round(max(row_sums), 2), "%\n")
cat("Mean row sum:", round(mean(row_sums), 2), "%\n")
```

## 3.2 Save Normalized Data

```{r save_normalized}
# Combine normalized data with depth and age-depth model
HZM_normD <- bind_cols(
  "Composite depth (mm)" = data_XRF_cl$depth_VT22,
  HZM_norm,
  data_XRF_cl[31:37]  # Depth/age model columns
)

write.xlsx(HZM_normD, "../data/XRF_HZM_norm.xlsx", rowNames = FALSE)
cat("Normalized XRF data saved to: ../data/XRF_HZM_norm.xlsx\n")
```

---

# Step 4: Centered Log Ratio (CLR) Transformation

## Background

CLR transformation accounts for non-linear matrix effects in compositional data.
- Suitable for correlation analysis
- Improves statistical properties
- Requires all positive values (no zeros)

Reference: Bertrand et al. (2024)

## 4.1 Replace Remaining Zero Values

```{r clr_prepare}
# For elements with occasional zero values (<5%):
# Replace with half of minimum non-zero value in that element

replace_zeros <- function(x) {
  # Find minimum non-zero value
  min_value <- min(x[x != 0])
  
  # If all zeros, set minimum to small value
  if (is.na(min_value)) {
    min_value <- 1e-10
  }
  
  # Replace zeros with half of minimum
  x[x == 0] <- min_value / 2
  
  return(x)
}

# Apply replacement to all elements
data_clr_ready <- as.data.frame(sapply(pre_clean_XRF, replace_zeros))

# Check for remaining zeros
zero_count <- sum(apply(data_clr_ready, 2, function(x) sum(x == 0)))
cat("Zero values remaining:", zero_count, "\n")
```

## 4.2 Apply CLR Transformation

```{r clr_transform}
# CLR = log(Xi / Geometric_Mean)
# Where Geometric_Mean = (X1 × X2 × ... × Xn)^(1/n)

normalize_by_geo_mean <- function(x) {
  # Calculate geometric mean for the observation
  geo_mean <- prod(x)^(1/length(x))
  
  # Return ratios to geometric mean
  x / geo_mean
}

# Apply to each row
clr_normalized <- as.data.frame(apply(data_clr_ready, 1, normalize_by_geo_mean))
clr_normalized <- as.data.frame(t(clr_normalized))

# Take natural logarithm
clr <- log(clr_normalized)

cat("CLR transformation applied\n")
cat("CLR data dimensions:", nrow(clr), "×", ncol(clr), "\n")
```

## 4.3 Save CLR Data

```{r save_clr}
HZM_clrD <- bind_cols(
  "Composite depth (mm)" = data_XRF_cl$depth_VT22,
  clr,
  data_XRF_cl[31:37]
)

write.xlsx(HZM_clrD, "../data/XRF_HZM_clr.xlsx", rowNames = FALSE)
cat("CLR-transformed data saved to: ../data/XRF_HZM_clr.xlsx\n")
```

---

# Step 5: Additive Log Ratio (ALR) Transformation

## Background

ALR transformation for subset of elements selected for downstream PCA/clustering analysis.
- Allows combination with other datasets
- Uses selected elements only (columns 2-8)
- Necessary before merging with external proxies

## 5.1 Select Elements for ALR

```{r alr_select}
# Select subset of elements for combined dataset analysis
data_for_alr <- select(pre_clean_XRF, 2:8)

cat("Elements selected for ALR:", ncol(data_for_alr), "\n")
```

## 5.2 Normalize Data

```{r alr_normalize}
pre_alr_norm <- data_for_alr / rowSums(data_for_alr) * 100

# Verify normalization
cat("Min row sum:", round(min(rowSums(pre_alr_norm)), 2), "%\n")
cat("Max row sum:", round(max(rowSums(pre_alr_norm)), 2), "%\n")
```

## 5.3 Apply ALR Function

```{r alr_prep}
# ALR = log(Xi / (100 - Xi))
# Where Xi is the abundance percentage

prep <- function(x) {
  # Calculate ALR for each element
  x / (100 - x)
}

prep_data <- as.data.frame(apply(pre_alr_norm, 1, prep))
trans_data <- as.data.frame(t(prep_data))
alr <- log(trans_data)

cat("ALR transformation applied\n")
cat("ALR data dimensions:", nrow(alr), "×", ncol(alr), "\n")
```

## 5.4 Add Cr Ratio and Save

```{r alr_save}
# Calculate Cr coherent/incoherent ratio as additional proxy
lnCr <- log(data_XRF_cl$`Cr inc` / data_XRF_cl$`Cr coh`)

# Combine into final dataset
HZM_alr_selected <- bind_cols(
  "Composite depth (mm)" = data_XRF_cl$depth_VT22,
  alr,
  "ln(Inc/Coh)" = lnCr,
  data_XRF_cl[33:37]  # Age-depth model columns
)

write.xlsx(HZM_alr_selected, "../data/XRF_HZM_alr.xlsx", rowNames = FALSE)
cat("ALR-transformed data saved to: ../data/XRF_HZM_alr.xlsx\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```

---

# Summary

**Outputs generated:**
- ✓ XRF_HZM_norm.xlsx (Normalized, % counts)
- ✓ XRF_HZM_clr.xlsx (CLR-transformed)
- ✓ XRF_HZM_alr.xlsx (ALR-transformed for merged analysis)
- ✓ HZM_HR_cleaned.xlsx (Cleaned high-resolution data)

**Next steps:**
- Use normalized data (XRF_HZM_norm.xlsx) for proxy correlations
- Use CLR data (XRF_HZM_clr.xlsx) for PCA on XRF alone
- Use ALR data (XRF_HZM_alr.xlsx) for combined multi-proxy analysis

---

*Last updated: 2025-03-11*
*Notebook version: 1.0*
