---
title: "Proxy-proxy calibration"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Install and load packages

```{r message=FALSE, warning=FALSE, include=FALSE}
#install.packages("pacman")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("readxl","dplyr","tidyr","openxlsx","car","rstatix", "DT", "data.table", "ipred", "EnvStats")#"ipred", "magicaxis", "mvnormtest", ,,"car", )
```

```{r}
Scripts  <- file.path("../notebooks/Functions", fsep="/")
#Out.Dir  <- file.path(Main.Path,"output" , fsep="/")

# Needed Scripts
source(file.path(Scripts, "series5_functions.R" , fsep="/" ))
source(file.path(Scripts, "user_functions.r", fsep="/" ))
source(file.path(Scripts, "utility_functions.r", fsep="/" ))
```

```{r}
HSI_raw <- read_excel("../data/SourceData/HSI_composite.xlsx")
HZM_calib <- read_excel("../data/SourceData/HZM_calib.xlsx", sheet = "for_LM")
Data <- select(HZM_calib, 1:2, 5:6)
colnames(Data) <- c('Chl', 'Bphe', 'RABD670', 'RABD845')
```

# Normality tests

As we want to use linear regression to calibrate out HSI data, we need to ensure that the data used for calibration are normally distributed. Therefore, we first use Shapiro-Wilk test to test the normality of the data.

```{r}
apply(Data,2,shapiro.test)
```

As the p-values are below 0.05, the null hypothesis of a normal distribution is falsified = the data are not normally distributed. Thus, we can try to transform the data (standardization, log transformation, square or cube root transformation.

We can also display the data distribution using histograms of Q-Q plots for further exploration.

```{r}
par(mfrow=c(2,4)) 
hist(Data$Chl, col='green4')
hist(Data$RABD670, col='forestgreen')
hist(Data$Bphe, col='purple')
hist(Data$RABD845, col='darkred')
qqnorm(Data$Chl)
qqline(Data$Chl)
qqnorm(Data$RABD670)
qqline(Data$RABD670)
qqnorm(Data$Bphe)
qqline(Data$Bphe)
qqnorm(Data$RABD845)
qqline(Data$RABD845)
```

## Data transformation to reach normality

### Square root

```{r}
root2_Data <- Data^(1/2)
apply(root2_Data,2,shapiro.test)
```

```{r}
par(mfrow=c(2,4)) 
hist(root2_Data$Chl, col='green4')
hist(root2_Data$RABD670, col='forestgreen')
hist(root2_Data$Bphe, col='purple')
hist(root2_Data$RABD845, col='darkred')
qqnorm(root2_Data$Chl)
qqline(root2_Data$Chl)
qqnorm(root2_Data$RABD670)
qqline(root2_Data$RABD670)
qqnorm(root2_Data$Bphe)
qqline(root2_Data$Bphe)
qqnorm(root2_Data$RABD845)
qqline(root2_Data$RABD845)
```

This transformation is not improving the normality

### Log transformation

```{r}
log_Data <- log(Data)
apply(log_Data,2,shapiro.test)
```

```{r}
log_Data <- log(Data)

par(mfrow=c(2,4)) 
hist(log_Data$Chl, col='green4')
hist(log_Data$RABD670, col='forestgreen')
hist(log_Data$Bphe, col='purple')
hist(log_Data$RABD845, col='darkred')
qqnorm(log_Data$Chl)
qqline(log_Data$Chl)
qqnorm(log_Data$RABD670)
qqline(log_Data$RABD670)
qqnorm(log_Data$Bphe)
qqline(log_Data$Bphe)
qqnorm(log_Data$RABD845)
qqline(log_Data$RABD845)
```

None of the transformation helps to reach normality, thus we need to identify the outliers and remove them from out calibration dataset

## Outliers identification and removal

```{r}
z_scores <- as.data.frame(sapply(Data, function(Data) (abs(Data-mean(Data))/sd(Data))))
#only keep rows in dataframe with all z-scores less than absolute value of 3 
no_outliers <- z_scores[!rowSums(z_scores>3), ]
```

```{r}
apply(no_outliers,2,shapiro.test)
```

There are no outliers in the data using the z-score outlier detection.

We check in Q-Q plots for the normality and outliers

```{r}
par(mfrow=c(2,2)) 

qqPlot(Data$Chl)
qqPlot(Data$Bphe)
qqPlot(Data$RABD670)
qqPlot(Data$RABD845)
```

From the Q-Q plots, there might be 3-7 outliers in each of the dataset. Thus we test for those using Rosner test:

```{r}
# identify_outliers(
#   data = Data,
#   variable = "Chl" #"Bhpe" #"RABD670" #"RABD845"
# )

library("EnvStats")
# Rosner test
testChl <- rosnerTest(Data$Chl, k = 7)
testBphe <- rosnerTest(Data$Bphe, k = 7)
test670 <- rosnerTest(Data$RABD670, k = 7)
test845 <- rosnerTest(Data$RABD845,k = 7)

results <- cbind(testChl, testBphe, test670, test845)
datatable(results)
```

Base on the Rosner's test, there are no outliers found and the data are normally distributed. Further, as the HSI data were used to select the calibration points, and this selection was designed to be a normally distributed, we assume that if the measured data are correlated with the normally distributed HSI selection, we can conduct linear regression to model the concentration values in the entire HSI dataset.

Thus, we lastly examine the correlation of the RADBs with the spectrophotometer data to confirm the data suitability for LRM.

```{r}
cor(Data)
```

# Linear regression model

```{r}
type <- c('Standard')
```

## LRM of the Chlorophylls

```{r}
xlim <- c(min(Data[,3]),max(Data[,3]))
ylim <- c(min(Data[,1]),max(Data[,1]))
ylab <- c("TChl [μg/g wet sed.]")
xlab <- c("RABD (670 nm) - wet sed.")


# Function to do and plot the simple linear regression
modelChla <- plotreg(Xvect = Data[,3],
                 Yvect = Data[,1],
                  type = type,
                  xlim = xlim,
                  ylim = ylim,
                  xlab = xlab,
                  ylab = ylab,
                  plot = TRUE)


# check residuals for normal distribution
stdresChla <- standardize.data(modelChla$residuals)
stdresChla <- as.vector(stdresChla[,1])

```

## LRM of the Bacteriopheophytin-a

```{r}
#plot options
#####################################
xlim <- c(min(Data[,4]),max(Data[,4]))
ylim <- c(min(Data[,2]),max(Data[,2]))
ylab <- c("Bphe a [μg/g wet sed.]")
xlab <- c("RABD (845 nm) - wet sed.")

# Function to do and plot the simple linear regression
modelBphe <- plotreg(Xvect = Data[,4],
                 Yvect = Data[,2],
                  type = type,
                  xlim = xlim,
                  ylim = ylim,
                  xlab = xlab,
                  ylab = ylab,
                  plot = TRUE)

# check residuals for normal distribution
stdresBphe <- standardize.data(modelBphe$residuals)
stdresBphe <- as.vector(stdresBphe[,1])



```

```{r}
LRMeq <- as.data.frame(cbind("Chla" = as.numeric(modelChla$coefficients), "Bphe" = as.numeric(modelBphe$coefficients)))
#Eq <- as.data.frame(transpose(LRMeq))
colnames(LRMeq) <- c("Chla", "Bphe")
row.names(LRMeq) <- c("intercept", "slope")
 

modelBphe
modelChla

```

## Calibration Coefficients Summary

```{r}
cat("=== LINEAR REGRESSION MODEL EQUATIONS ===\n\n")
cat("Total Chlorophyll (TChl):\n")
cat("TChl [μg/g] =", LRMeq[1,1], "+", LRMeq[2,1], "* RABD(670 nm)\n")
cat("Intercept (baseline/noise):", LRMeq[1,1], "\n")
cat("Slope:", LRMeq[2,1], "\n\n")

cat("Bacteriopheophytine (Bphe a):\n")
cat("Bphe a [μg/g] =", LRMeq[1,2], "+", LRMeq[2,2], "* RABD(845 nm)\n")
cat("Intercept (baseline/noise):", LRMeq[1,2], "\n")
cat("Slope:", LRMeq[2,2], "\n")
```

## Analysis of Low-Concentration Data Points (Signal vs Noise)

Since negative intercepts are not plausible, we define signal thresholds based on the actual low-concentration calibration data points.

### Low-Concentration Points for TChl

```{r}
# Identify and analyze low-concentration TChl measurements
Chl_data <- data.frame(
  measured_Chl = as.numeric(unlist(Data[,1])),
  RABD670 = as.numeric(unlist(Data[,3])),
  predicted_Chl = as.numeric(LRMeq[1,1] + LRMeq[2,1] * unlist(Data[,3])),
  residual = as.numeric(modelChla$residuals)
)

# Sort by measured concentration and show the lowest values
Chl_low <- Chl_data[order(Chl_data$measured_Chl), ]
Chl_low_quartile <- Chl_low[1:ceiling(nrow(Chl_low)/4), ]  # Lower 25% of concentrations

cat("=== LOWEST 25% OF TChl CONCENTRATIONS ===\n")
datatable(Chl_low_quartile)

cat("\nSummary statistics for low-concentration TChl samples:\n")
cat("Min measured value:", min(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Max measured value (25th percentile):", max(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Median measured value:", median(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Mean residual:", mean(Chl_low_quartile$residual), "\n")
cat("SD of residuals:", sd(Chl_low_quartile$residual), "\n")
cat("Suggested noise threshold (median of low values):", median(Chl_low_quartile$measured_Chl), "μg/g\n")
```

### Low-Concentration Points for Bphe

```{r}
# Identify and analyze low-concentration Bphe measurements
Bphe_data <- data.frame(
  measured_Bphe = as.numeric(unlist(Data[,2])),
  RABD845 = as.numeric(unlist(Data[,4])),
  predicted_Bphe = as.numeric(LRMeq[1,2] + LRMeq[2,2] * unlist(Data[,4])),
  residual = as.numeric(modelBphe$residuals)
)

# Sort by measured concentration and show the lowest values
Bphe_low <- Bphe_data[order(Bphe_data$measured_Bphe), ]
Bphe_low_quartile <- Bphe_low[1:ceiling(nrow(Bphe_low)/4), ]  # Lower 25% of concentrations

cat("=== LOWEST 25% OF Bphe CONCENTRATIONS ===\n")
datatable(Bphe_low_quartile)

cat("\nSummary statistics for low-concentration Bphe samples:\n")
cat("Min measured value:", min(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Max measured value (25th percentile):", max(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Median measured value:", median(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Mean residual:", mean(Bphe_low_quartile$residual), "\n")
cat("SD of residuals:", sd(Bphe_low_quartile$residual), "\n")
cat("Suggested noise threshold (median of low values):", median(Bphe_low_quartile$measured_Bphe), "μg/g\n")
```

# Detection Limits - LOD and LOQ

The limit of detection (LOD) and limit of quantification (LOQ) define the sensitivity and reliability of the calibration. These are calculated from the linear regression residuals and represent the minimum concentrations that can be reliably detected and quantified.

## LOD and LOQ for Total Chlorophyll (TChl)

```{r}
# Calculate LOD and LOQ for Chlorophyll
# LOD = 3.3 * SD(residuals) / slope
# LOQ = 10 * SD(residuals) / slope

SD_Chl <- sd(modelChla$residuals)
slope_Chl <- as.numeric(modelChla$coefficients[2])

LOD_Chl <- (3.3 * SD_Chl) / abs(slope_Chl)
LOQ_Chl <- (10 * SD_Chl) / abs(slope_Chl)

cat("Chlorophyll (TChl) - Detection Limits:\n")
cat("SD of residuals:", SD_Chl, "\n")
cat("Slope:", slope_Chl, "\n")
cat("LOD (μg/g wet sed.):", LOD_Chl, "\n")
cat("LOQ (μg/g wet sed.):", LOQ_Chl, "\n\n")
```

## LOD and LOQ for Bacteriopheophytine (Bphe)

```{r}
# Calculate LOD and LOQ for Bacteriopheophytine
SD_Bphe <- sd(modelBphe$residuals)
slope_Bphe <- as.numeric(modelBphe$coefficients[2])

LOD_Bphe <- (3.3 * SD_Bphe) / abs(slope_Bphe)
LOQ_Bphe <- (10 * SD_Bphe) / abs(slope_Bphe)

cat("Bacteriopheophytine (Bphe a) - Detection Limits:\n")
cat("SD of residuals:", SD_Bphe, "\n")
cat("Slope:", slope_Bphe, "\n")
cat("LOD (μg/g wet sed.):", LOD_Bphe, "\n")
cat("LOQ (μg/g wet sed.):", LOQ_Bphe, "\n\n")
```

## Summary Table of Detection Limits

```{r}
detection_limits <- data.frame(
  Compound = c("TChl", "Bphe a"),
  LOD_ugpg = c(LOD_Chl, LOD_Bphe),
  LOQ_ugpg = c(LOQ_Chl, LOQ_Bphe),
  SD_residuals = c(SD_Chl, SD_Bphe),
  Slope = c(slope_Chl, slope_Bphe)
)

datatable(detection_limits)
```

# Apply the LRM to the HSI dataset

```{r}
calibChl <- function(x) {
   LRMeq[2,1] * x + LRMeq[1,1] 
}

HSI_raw$Chla_calibrated <- as.numeric(lapply(HSI_raw$RABD670, calibChl))

calibBphe <- function(x) {
   LRMeq[2,2] * x + LRMeq[1,2] 
}

HSI_raw$Bphe_calibrated <- as.numeric(lapply(HSI_raw$RABD845, calibBphe))

HSI_calibrated <- as.data.frame(HSI_raw)
```

Saving the data

```{r}
write.xlsx(HSI_raw, "../data/HZM_HSI_calibrated.xlsx", rowNames = FALSE)
```

```{r}
# Add age model (depth_VT, mean_VT) to HSI_calibrated using interpolation/extrapolation
HZM_HR_cleaned <- read_excel("../data/HZM_HR_cleaned.xlsx")

depth_target <- as.numeric(HSI_calibrated$`Sediment.Y..mm.`)
depth_source <- as.numeric(HZM_HR_cleaned$Depth_mm)
depth_vt_col <- if ("depth_VT" %in% names(HZM_HR_cleaned)) "depth_VT" else "depth_VT22"
mean_vt_col <- if ("mean_VT" %in% names(HZM_HR_cleaned)) "mean_VT" else "mean_VT22"

depth_vt <- as.numeric(HZM_HR_cleaned[[depth_vt_col]])
mean_vt <- as.numeric(HZM_HR_cleaned[[mean_vt_col]])

ord <- order(depth_source)

valid_depth_vt <- !is.na(depth_source[ord]) & !is.na(depth_vt[ord])
valid_mean_vt <- !is.na(depth_source[ord]) & !is.na(mean_vt[ord])

if (sum(valid_depth_vt) < 2 || sum(valid_mean_vt) < 2) {
   stop("Not enough non-NA values in depth_VT/mean_VT to interpolate.")
}

HSI_calibrated$depth_VT <- approx(
   x = depth_source[ord][valid_depth_vt],
   y = depth_vt[ord][valid_depth_vt],
   xout = depth_target,
   rule = 2
)$y

HSI_calibrated$mean_VT <- approx(
   x = depth_source[ord][valid_mean_vt],
   y = mean_vt[ord][valid_mean_vt],
   xout = depth_target,
   rule = 2
)$y

write.xlsx(HSI_calibrated, "../data/HZM_HSI_calibrated.xlsx", rowNames = FALSE)
```

```{r}
# Quick plot: mean_VT vs depth_VT
plot(
   HSI_calibrated$depth_VT,
   HSI_calibrated$mean_VT,
   type = "l",
   xlab = "depth_VT",
   ylab = "mean_VT",
   main = "mean_VT vs depth_VT"
)
```
