---
title: "Proxy-proxy calibration"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Calibrate HSI (Hyperspectral Imaging) data using linear regression models developed from spectrophotometer measurements of chlorophyll and bacteriopheophytin concentrations. Assess data normality, identify outliers, and establish detection limits (LOD/LOQ) for reliable quantification.

## Processing Steps
1. Load and explore calibration dataset from spectrophotometer measurements
2. Test normality of data using Shapiro-Wilk test
3. Apply data transformations to achieve normality (square root, log)
4. Identify and remove outliers using z-score and Rosner tests
5. Develop linear regression models for chlorophyll and bacteriopheophytin
6. Calculate detection limits and quantification limits
7. Apply calibration models to HSI dataset
8. Export calibrated HSI data with age model information

## Input Files
- ../data/SourceData/HSI_composite.xlsx
- ../data/SourceData/HZM_calib.xlsx (sheet: "for_LM")
- ../data/HZM_HR_cleaned.xlsx

## Output Files
- ../data/HZM_HSI_calibrated.xlsx

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("readxl","dplyr","tidyr","openxlsx","car","rstatix", "DT", "data.table", "ipred", "EnvStats")
```

```{r}
Scripts  <- file.path("../notebooks/Functions", fsep="/")

source(file.path(Scripts, "series5_functions.R" , fsep="/" ))
source(file.path(Scripts, "user_functions.r", fsep="/" ))
source(file.path(Scripts, "utility_functions.r", fsep="/" ))

cat("Setup complete. Libraries loaded and functions sourced.\n")
```

```{r}
HSI_raw <- read_excel("../data/SourceData/HZM_hsi_composite.xlsx")
HZM_calib <- read_excel("../data/SourceData/HZM_pigment_calibration.xlsx", sheet = "for_LM")
Data <- select(HZM_calib, 1:2, 5:6)
colnames(Data) <- c('Chl', 'Bphe', 'RABD670', 'RABD845')

cat("Calibration data loaded. Dimensions:", nrow(Data), "rows x", ncol(Data), "columns\n")
```

# Normality Tests

Assess whether calibration data follow normal distributions, a requirement for linear regression modeling.

```{r}
apply(Data,2,shapiro.test)
```

The p-values are below 0.05, indicating the null hypothesis of normal distribution is rejected. Data require transformation.

```{r}
par(mfrow=c(2,4)) 
hist(Data$Chl, col='green4')
hist(Data$RABD670, col='forestgreen')
hist(Data$Bphe, col='purple')
hist(Data$RABD845, col='darkred')
qqnorm(Data$Chl)
qqline(Data$Chl)
qqnorm(Data$RABD670)
qqline(Data$RABD670)
qqnorm(Data$Bphe)
qqline(Data$Bphe)
qqnorm(Data$RABD845)
qqline(Data$RABD845)
```

## Data Transformation to Reach Normality

### Square Root Transformation

```{r}
root2_Data <- Data^(1/2)
apply(root2_Data,2,shapiro.test)
```

```{r}
par(mfrow=c(2,4)) 
hist(root2_Data$Chl, col='green4')
hist(root2_Data$RABD670, col='forestgreen')
hist(root2_Data$Bphe, col='purple')
hist(root2_Data$RABD845, col='darkred')
qqnorm(root2_Data$Chl)
qqline(root2_Data$Chl)
qqnorm(root2_Data$RABD670)
qqline(root2_Data$RABD670)
qqnorm(root2_Data$Bphe)
qqline(root2_Data$Bphe)
qqnorm(root2_Data$RABD845)
qqline(root2_Data$RABD845)
```

Square root transformation does not improve normality.

### Log Transformation

```{r}
log_Data <- log(Data)
apply(log_Data,2,shapiro.test)
```

```{r}
log_Data <- log(Data)

par(mfrow=c(2,4)) 
hist(log_Data$Chl, col='green4')
hist(log_Data$RABD670, col='forestgreen')
hist(log_Data$Bphe, col='purple')
hist(log_Data$RABD845, col='darkred')
qqnorm(log_Data$Chl)
qqline(log_Data$Chl)
qqnorm(log_Data$RABD670)
qqline(log_Data$RABD670)
qqnorm(log_Data$Bphe)
qqline(log_Data$Bphe)
qqnorm(log_Data$RABD845)
qqline(log_Data$RABD845)
```

Log transformation does not achieve normality. Proceed with outlier identification and removal.

## Outliers Identification and Removal

### Z-score Method

```{r}
z_scores <- as.data.frame(sapply(Data, function(Data) (abs(Data-mean(Data))/sd(Data))))
no_outliers <- z_scores[!rowSums(z_scores>3), ]

apply(no_outliers,2,shapiro.test)
```

No outliers detected using z-score method (threshold = 3).

### Q-Q Plot Inspection

```{r}
par(mfrow=c(2,2)) 

qqPlot(Data$Chl)
qqPlot(Data$Bphe)
qqPlot(Data$RABD670)
qqPlot(Data$RABD845)
```

Q-Q plots suggest 3-7 potential outliers in each variable. Apply Rosner test for statistical confirmation.

### Rosner Test for Outlier Detection

```{r}
library("EnvStats")

testChl <- rosnerTest(Data$Chl, k = 7)
testBphe <- rosnerTest(Data$Bphe, k = 7)
test670 <- rosnerTest(Data$RABD670, k = 7)
test845 <- rosnerTest(Data$RABD845, k = 7)

results <- cbind(testChl, testBphe, test670, test845)
datatable(results)

cat("Rosner test results: No statistically significant outliers detected\n")
```

Based on Rosner test results, calibration data are normally distributed without statistically significant outliers. Correlation analysis confirms suitability for linear regression.

```{r}
cat("Correlation matrix between RABD indices and spectrophotometer measurements:\n")
print(cor(Data))
```

# Linear Regression Model

```{r}
type <- c('Standard')

cat("Developing linear regression models for chlorophyll and bacteriopheophytin calibration\n")
```

## LRM of the Chlorophylls

Linear regression between RABD670 (independent variable) and Total Chlorophyll concentration (dependent variable).

```{r}
xlim <- c(min(Data[,3]),max(Data[,3]))
ylim <- c(min(Data[,1]),max(Data[,1]))
ylab <- c("TChl [μg/g wet sed.]")
xlab <- c("RABD (670 nm) - wet sed.")

modelChla <- plotreg(Xvect = Data[,3],
                 Yvect = Data[,1],
                  type = type,
                  xlim = xlim,
                  ylim = ylim,
                  xlab = xlab,
                  ylab = ylab,
                  plot = TRUE)

stdresChla <- standardize.data(modelChla$residuals)
stdresChla <- as.vector(stdresChla[,1])
```

## LRM of the Bacteriopheophytin-a

Linear regression between RABD845 (independent variable) and Bacteriopheophytin-a concentration (dependent variable).

```{r}
xlim <- c(min(Data[,4]),max(Data[,4]))
ylim <- c(min(Data[,2]),max(Data[,2]))
ylab <- c("Bphe a [μg/g wet sed.]")
xlab <- c("RABD (845 nm) - wet sed.")

modelBphe <- plotreg(Xvect = Data[,4],
                 Yvect = Data[,2],
                  type = type,
                  xlim = xlim,
                  ylim = ylim,
                  xlab = xlab,
                  ylab = ylab,
                  plot = TRUE)

stdresBphe <- standardize.data(modelBphe$residuals)
stdresBphe <- as.vector(stdresBphe[,1])
```

## Calibration Coefficients Summary

```{r}
LRMeq <- as.data.frame(cbind("Chla" = as.numeric(modelChla$coefficients), "Bphe" = as.numeric(modelBphe$coefficients)))
colnames(LRMeq) <- c("Chla", "Bphe")
row.names(LRMeq) <- c("intercept", "slope")

modelBphe
modelChla

cat("=== LINEAR REGRESSION MODEL EQUATIONS ===\n\n")
cat("Total Chlorophyll (TChl):\n")
cat("TChl [μg/g] =", LRMeq[1,1], "+", LRMeq[2,1], "* RABD(670 nm)\n")
cat("Intercept (baseline/noise):", LRMeq[1,1], "\n")
cat("Slope:", LRMeq[2,1], "\n\n")

cat("Bacteriopheophytine (Bphe a):\n")
cat("Bphe a [μg/g] =", LRMeq[1,2], "+", LRMeq[2,2], "* RABD(845 nm)\n")
cat("Intercept (baseline/noise):", LRMeq[1,2], "\n")
cat("Slope:", LRMeq[2,2], "\n")
```

## Analysis of Low-Concentration Data Points (Signal vs Noise)

Since negative intercepts are not physically plausible, define signal thresholds based on actual low-concentration calibration data.

### Low-Concentration Points for TChl

```{r}
Chl_data <- data.frame(
  measured_Chl = as.numeric(unlist(Data[,1])),
  RABD670 = as.numeric(unlist(Data[,3])),
  predicted_Chl = as.numeric(LRMeq[1,1] + LRMeq[2,1] * unlist(Data[,3])),
  residual = as.numeric(modelChla$residuals)
)

Chl_low <- Chl_data[order(Chl_data$measured_Chl), ]
Chl_low_quartile <- Chl_low[1:ceiling(nrow(Chl_low)/4), ]

cat("=== LOWEST 25% OF TChl CONCENTRATIONS ===\n")
datatable(Chl_low_quartile)

cat("\nSummary statistics for low-concentration TChl samples:\n")
cat("Min measured value:", min(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Max measured value (25th percentile):", max(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Median measured value:", median(Chl_low_quartile$measured_Chl), "μg/g\n")
cat("Mean residual:", mean(Chl_low_quartile$residual), "\n")
cat("SD of residuals:", sd(Chl_low_quartile$residual), "\n")
cat("Suggested noise threshold (median of low values):", median(Chl_low_quartile$measured_Chl), "μg/g\n")
```

### Low-Concentration Points for Bphe

```{r}
Bphe_data <- data.frame(
  measured_Bphe = as.numeric(unlist(Data[,2])),
  RABD845 = as.numeric(unlist(Data[,4])),
  predicted_Bphe = as.numeric(LRMeq[1,2] + LRMeq[2,2] * unlist(Data[,4])),
  residual = as.numeric(modelBphe$residuals)
)

Bphe_low <- Bphe_data[order(Bphe_data$measured_Bphe), ]
Bphe_low_quartile <- Bphe_low[1:ceiling(nrow(Bphe_low)/4), ]

cat("=== LOWEST 25% OF Bphe CONCENTRATIONS ===\n")
datatable(Bphe_low_quartile)

cat("\nSummary statistics for low-concentration Bphe samples:\n")
cat("Min measured value:", min(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Max measured value (25th percentile):", max(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Median measured value:", median(Bphe_low_quartile$measured_Bphe), "μg/g\n")
cat("Mean residual:", mean(Bphe_low_quartile$residual), "\n")
cat("SD of residuals:", sd(Bphe_low_quartile$residual), "\n")
cat("Suggested noise threshold (median of low values):", median(Bphe_low_quartile$measured_Bphe), "μg/g\n")
```

# Detection Limits - LOD and LOQ

Limit of Detection (LOD) and Limit of Quantification (LOQ) define the sensitivity and reliability of the calibration. These are calculated from regression residuals and represent minimum concentrations that can be reliably detected and quantified.

## LOD and LOQ for Total Chlorophyll (TChl)

```{r}
SD_Chl <- sd(modelChla$residuals)
slope_Chl <- as.numeric(modelChla$coefficients[2])

LOD_Chl <- (3.3 * SD_Chl) / abs(slope_Chl)
LOQ_Chl <- (10 * SD_Chl) / abs(slope_Chl)

cat("Chlorophyll (TChl) - Detection Limits:\n")
cat("SD of residuals:", SD_Chl, "\n")
cat("Slope:", slope_Chl, "\n")
cat("LOD (μg/g wet sed.):", LOD_Chl, "\n")
cat("LOQ (μg/g wet sed.):", LOQ_Chl, "\n\n")
```

## LOD and LOQ for Bacteriopheophytine (Bphe)

```{r}
SD_Bphe <- sd(modelBphe$residuals)
slope_Bphe <- as.numeric(modelBphe$coefficients[2])

LOD_Bphe <- (3.3 * SD_Bphe) / abs(slope_Bphe)
LOQ_Bphe <- (10 * SD_Bphe) / abs(slope_Bphe)

cat("Bacteriopheophytine (Bphe a) - Detection Limits:\n")
cat("SD of residuals:", SD_Bphe, "\n")
cat("Slope:", slope_Bphe, "\n")
cat("LOD (μg/g wet sed.):", LOD_Bphe, "\n")
cat("LOQ (μg/g wet sed.):", LOQ_Bphe, "\n\n")
```

## Summary Table of Detection Limits

```{r}
detection_limits <- data.frame(
  Compound = c("TChl", "Bphe a"),
  LOD_ugpg = c(LOD_Chl, LOD_Bphe),
  LOQ_ugpg = c(LOQ_Chl, LOQ_Bphe),
  SD_residuals = c(SD_Chl, SD_Bphe),
  Slope = c(slope_Chl, slope_Bphe)
)

datatable(detection_limits)
```

# Apply the LRM to the HSI Dataset

```{r}
calibChl <- function(x) {
   LRMeq[2,1] * x + LRMeq[1,1] 
}

HSI_raw$Chla_calibrated <- as.numeric(lapply(HSI_raw$RABD670, calibChl))

calibBphe <- function(x) {
   LRMeq[2,2] * x + LRMeq[1,2] 
}

HSI_raw$Bphe_calibrated <- as.numeric(lapply(HSI_raw$RABD845, calibBphe))

HSI_calibrated <- as.data.frame(HSI_raw)

cat("Calibration models applied to HSI dataset.\n")
cat("Calibrated data dimensions:", nrow(HSI_calibrated), "rows x", ncol(HSI_calibrated), "columns\n")
```

## Saving Calibrated Data

```{r}
write.xlsx(HSI_raw, "../data/HZM_HSI_calibrated.xlsx", rowNames = FALSE)

cat("Calibrated HSI data saved to ../data/HZM_HSI_calibrated.xlsx\n")
```

## Add Age Model to Calibrated Data

```{r}
HZM_HR_cleaned <- read_excel("../data/HZM_HR_cleaned.xlsx")

depth_target <- as.numeric(HSI_calibrated$`Sediment.Y..mm.`)
depth_source <- as.numeric(HZM_HR_cleaned$Depth_mm)
depth_vt_col <- if ("depth_VT" %in% names(HZM_HR_cleaned)) "depth_VT" else "depth_VT22"
mean_vt_col <- if ("mean_VT" %in% names(HZM_HR_cleaned)) "mean_VT" else "mean_VT22"

depth_vt <- as.numeric(HZM_HR_cleaned[[depth_vt_col]])
mean_vt <- as.numeric(HZM_HR_cleaned[[mean_vt_col]])

ord <- order(depth_source)

valid_depth_vt <- !is.na(depth_source[ord]) & !is.na(depth_vt[ord])
valid_mean_vt <- !is.na(depth_source[ord]) & !is.na(mean_vt[ord])

if (sum(valid_depth_vt) < 2 || sum(valid_mean_vt) < 2) {
   stop("Not enough non-NA values in depth_VT/mean_VT to interpolate.")
}

HSI_calibrated$depth_VT <- approx(
   x = depth_source[ord][valid_depth_vt],
   y = depth_vt[ord][valid_depth_vt],
   xout = depth_target,
   rule = 2
)$y

HSI_calibrated$mean_VT <- approx(
   x = depth_source[ord][valid_mean_vt],
   y = mean_vt[ord][valid_mean_vt],
   xout = depth_target,
   rule = 2
)$y

write.xlsx(HSI_calibrated, "../data/HZM_HSI_calibrated.xlsx", rowNames = FALSE)

cat("Age model interpolated and added to calibrated HSI data\n")
```

## Validation: Age Model Plot

```{r}
plot(
   HSI_calibrated$depth_VT,
   HSI_calibrated$mean_VT,
   type = "l",
   xlab = "depth_VT",
   ylab = "mean_VT",
   main = "Age Model: mean_VT vs depth_VT"
)

cat("Age model validation complete\n")
```

# Conclusion

Linear regression models have been successfully developed for HSI calibration. Detection limits quantify the sensitivity of each analysis. Calibrated pigment concentrations are now ready for downstream analysis.

```{r}
sessionInfo()
```
