---
title: "Interpolation"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Installing and loading packages

```{r message=FALSE, warning=FALSE, include=FALSE}
#install.packages("pacman")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("tibble","readxl","dplyr","tidyr","openxlsx","car","rstatix", "DT", "data.table")
```

## Calling functions

```{r message=FALSE, warning=FALSE, include=FALSE}
Scripts  <- file.path("../notebooks/Functions", fsep="/")
source(file.path(Scripts, "utility_functions.r", fsep="/" ))
```

## Preparing the data frames into a suitable format

As the XRF data are from a long core composite and we look only at the Late-Glacial section, the starting composite depth is not 0 but 10989 mm (10.9m). Our concatenated hyperspectral data are on a scale from 0mm. Therefore, we need to adjust the depth to the XRF depth.

```{r}
HSI_data <- read_excel("../data/HZM_HSI_calibrated.xlsx", col_types = "numeric")
NewScale <- read_excel("../data/SourceData/HZM19_LG_section_XRF.xlsx") 
CleanedScale <- read_excel("../data/XRF_HZM_alr.xlsx")


# Creating a new columns with converted HSI depths to the XRF depths (0 -> 10989)
Source <- add_column(HSI_data, "Depth_LG_comp"= as.numeric(HSI_data$Sediment.Y..mm. + as.numeric(NewScale[1,"comp_depth_mm"])), .before = HSI_data$Sediment.Y..mm.) 

#Saving as a csv file
write.table(Source,file="../data/HSI_HR_data.csv", sep=";",row.names = FALSE)

# Exctracting the XRF depth scale to a separate file (optional)
Destin <- data.frame("depth"= as.integer(NewScale$comp_depth_mm))

#Saving as a csv file
write.table(Destin,file="../data/XRF_scale.csv",sep=";", row.names = FALSE)

# Exctracting the XRF depth scale to a separate file (optional)
Destin2 <- data.frame("depth"= as.integer(CleanedScale$`Composite depth (mm)`))

#Saving as a csv file
write.table(Destin2,file="../data/XRF_scale_cleaned.csv",sep=";", row.names = FALSE)
```

# Loading the prepared XRF and HSI data for interpolation

```{r}
Source <- read.csv(file= "../data/HSI_HR_data.csv", header=TRUE,sep=";")
Destin <- read.csv(file="../data/XRF_scale.csv", header=TRUE,sep=";") 
Destin2 <-read.csv(file="../data/XRF_scale_cleaned.csv", header=TRUE,sep=";") 
```

# Interpolating the data

```{r}
Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin2
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin2[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("Depth_mm",Names[2:length(Names)])
```

# Saving the interpolated HSI data

```{r}
# write output file
write.table(Data,file="../data/HSI_interpolated_cleaned.csv",sep=";",row.names = FALSE)

write.xlsx(Data, "../data/HSI_interpolated_cleaned.xlsx", rowNames = FALSE)
```

# Validation

Plots the old and the new data for comparison

```{r}
for(j in 2:ncol(Source)){
layout(matrix(c(1,2),1,2, byrow = TRUE))
plot(Source[,j],-Source[,1], type="l", main="Original"    ,xlab=colnames(Source[j]),ylab=colnames(  Data[1]))
plot(  Data[,j],  -Data[,1], type="l", main="interpolated",xlab=colnames(  Data[j]),ylab=colnames(  Data[1]))
}


#par(.def_par)

```
