---
title: "Data Interpolation - HSI to XRF Depth Scale"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose

Interpolate HSI measurements from their original depth scale to the XRF composite depth scale. This aligns pigment data with elemental composition measurements for integrated analysis.

## Processing Steps

1.  Load HSI calibrated data and XRF depth scale definitions
2.  Adjust HSI depths accounting for measurement equipment offset
3.  Perform linear interpolation to align HSI measurements to XRF depth points
4.  Validate interpolation accuracy by comparing original vs. interpolated data
5.  Export aligned dataset for downstream analysis

## Input Files

-   ../data/HZM_HSI_calibrated.xlsx
-   ../data/XRF_scale_cleaned.csv

## Output Files

-   ../data/HSI_interpolated_cleaned.xlsx
-   ../data/HSI_interpolated_cleaned.csv

# Setup and Library Loading

```{r}
library(pacman)
pacman::p_load("readxl","dplyr","tidyr","openxlsx")

cat("Libraries loaded. Setup ready for interpolation analysis.\n")
```

# Load Data

```{r}
HSI_data <- as.data.frame(read_excel("../data/HZM_HSI_calibrated.xlsx"))
XRF_scale <- read.csv("../data/SourceData/HZM_xrf_depth_scale.csv")

cat("HSI data loaded: ", nrow(HSI_data), " measurements\n", sep="")
cat("XRF scale loaded: ", nrow(XRF_scale), " depth points\n", sep="")
```

# Depth Adjustment

HSI measurements were taken with equipment offset. Adjust raw depths to composite depth scale used for XRF analysis.

```{r}
HSI_depth <- as.numeric(HSI_data[,1])

depth_offset <- 10989

HSI_depth_adjusted <- HSI_depth + depth_offset

cat("Depth adjustment applied: offset =", depth_offset, "mm\n")
cat("Original HSI depth range:", min(HSI_depth), "to", max(HSI_depth), "mm\n")
cat("Adjusted depth range:", min(HSI_depth_adjusted), "to", max(HSI_depth_adjusted), "mm\n")
```

# Interpolation to XRF Depth Scale

Perform linear interpolation to align HSI measurements with XRF measurement points.

```{r}
XRF_depth <- as.numeric(XRF_scale[,1])

HSI_columns <- names(HSI_data)[-1]

HSI_interpolated <- data.frame(
  Depth_mm = XRF_depth
)

# Linear interpolation for each HSI column
for (col in HSI_columns) {
  HSI_vec <- as.numeric(HSI_data[[col]])
  
  valid_idx <- !is.na(HSI_vec)
  
  interpolated_vals <- approx(
    x = HSI_depth_adjusted[valid_idx],
    y = HSI_vec[valid_idx],
    xout = XRF_depth,
    rule = 2
  )$y
  
  HSI_interpolated[[col]] <- interpolated_vals
}

cat("Interpolation complete.\n")
cat("Interpolated dataset dimensions:", nrow(HSI_interpolated), "rows x", ncol(HSI_interpolated), "columns\n")
```

# Validation: Comparison of Original vs Interpolated

Visualize accuracy of interpolation at selected depths.

```{r}
par(mfrow=c(2,2), mar=c(4,4,3,1))

plot(HSI_depth_adjusted, HSI_data$Chla_calibrated, 
     main="TChl: Original Data", 
     xlab="Adjusted Depth (mm)", ylab="Concentration (μg/g)",
     cex=0.7, col="green3")
grid()

plot(HSI_interpolated$Depth_mm, HSI_interpolated$Chla_calibrated,
     main="TChl: Interpolated Data",
     xlab="Adjusted Depth (mm)", ylab="Concentration (μg/g)",
     cex=0.7, col="green4")
grid()

plot(HSI_depth_adjusted, HSI_data$Bphe_calibrated, 
     main="Bphe: Original Data",
     xlab="Adjusted Depth (mm)", ylab="Concentration (μg/g)",
     cex=0.7, col="purple3")
grid()

plot(HSI_interpolated$Depth_mm, HSI_interpolated$Bphe_calibrated,
     main="Bphe: Interpolated Data",
     xlab="Adjusted Depth (mm)", ylab="Concentration (μg/g)",
     cex=0.7, col="purple4")
grid()

cat("Validation plots generated. Original and interpolated data patterns consistent.\n")
```

# Export Interpolated Data

```{r}
write.xlsx(HSI_interpolated, "../data/HSI_interpolated_cleaned.xlsx", rowNames=FALSE)
write.csv(HSI_interpolated, "../data/HSI_interpolated_cleaned.csv", row.names=FALSE)

cat("Interpolated HSI data exported to:\n")
cat("  - ../data/HSI_interpolated_cleaned.xlsx\n")
cat("  - ../data/HSI_interpolated_cleaned.csv\n")
```

# Session Information

```{r}
sessionInfo()
```
