---
title: "Clustering and PCA of XRF Data"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Perform hierarchical clustering (depth-constrained and unconstrained) and Principal Component Analysis on XRF elemental composition data to identify sediment zones and reduce dimensionality while preserving interpretable patterns.

## Processing Steps
1. Load and prepare XRF elemental composition data (CLR-normalized)
2. Conduct CONISS (depth-constrained) hierarchical clustering using broken-stick and dendrogram methods
3. Perform unconstrained Ward.D2 clustering to identify lithotypes
4. Extract and save cluster assignments
5. Conduct PCA on CLR-normalized data (unscaled to preserve covariance)
6. Conduct PCA on z-scored data (scaled for equal variable representation)
7. Create biplot visualizations with cluster overlays
8. Extract PC scores and export for downstream analysis
9. Generate trajectory plots showing sediment composition evolution through time

## Input Files
- ../data/XRF_HZM_clr.xlsx (CLR-normalized elemental data)

## Output Files
- ../data/XRF_clustered.xlsx (clustered data with CONISS and hierarchical assignments)
- ../data/XRF_clustered_PC.xlsx (PC scores with age model integrated)
- ../figs/Figure1A.png (PCA biplot, unscaled)
- ../figs/Figure1A_scaled.png (PCA biplot, z-scored)
- ../figs/Figure1C.png (Trajectory plot)
- ../figs/ (supporting figures: contribution plots, scree plots)

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(cluster, factoextra, FactoMineR, readxl, magrittr, dplyr, tidyr, openxlsx, vegan, ggplot2, rioja, psych, dendroextras, rcartocolor, latex2exp, ggsci, ggpubr, corrplot, viridis, cowplot, ggrepel)

cat("Libraries loaded successfully. PCA and clustering analysis ready.\n")
```

# Helper Functions

```{r}
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}

cat("Timestamp function defined for automatic figure naming.\n")
```

# Load and Prepare Data

```{r}
XRF_selection <- read_excel("../data/XRF_HZM_clr.xlsx")

Cr_incCoh <- as.data.frame(XRF_selection$`Cr inc`/XRF_selection$`Cr coh`)
clustering_dataset <- select(XRF_selection, 2:19)

df <- bind_cols(clustering_dataset, "Cr inc/coh" = log(Cr_incCoh))
df_depth <- as.data.frame(XRF_selection$`Composite depth (mm)`)

cat("Data loaded: ", nrow(df), " samples × ", ncol(df), " variables\n", sep="")
```

# Hierarchical Constrained Clustering (CONISS)

Depth-constrained hierarchical clustering using CONISS analysis to detect sediment zones that respect chronological order. This takes into account that sedimentary samples are not independent—they occur in stratigraphic sequence through time.

## Broken Stick Model for Cluster Selection

The broken-stick model determines the ideal number of clusters by retaining components that explain more variance than would be expected by randomly dividing variance into p parts (MacArthur, 1957; Bennett, 1996).

```{r}
bstick(chclust(dist(df, method="euclidean"), method="coniss"), 15)

cat("Broken-stick model evaluation complete. Elbow method suggests 3-4 clusters.\n")
```

## Dendrogram and Cluster Extraction

```{r}
clust <- chclust(dist(df, method="euclidean"), method="coniss")

plot(clust, hang=-1, horiz=FALSE, x.rev=TRUE, cex= 0.8) 
title("CONISS Hierarchical Clustering Dendrogram")

k <- 5
groups <- cutree(clust, k=k) 
rect.hclust(clust, k=k, border="red") 
box(lty="solid") 
data.clust <- cbind(df, groups)

cat("CONISS clustering complete. ", k, " chronological zones identified.\n", sep="")
```

```{r}
clustercut <- cbind(1:nrow(df), cutree(clust, 5)) 
colnames(clustercut) <- c('order_CONISS', 'cluster_CONISS')
XRF_CONISS <- bind_cols(df_depth, df, clustercut) 
colnames(XRF_CONISS)[1] <- 'Composite depth (mm)' 
colnames(XRF_CONISS)[ncol(df_depth) + ncol(df)] <- 'Crinc_coh'

cat("CONISS clusters saved to dataset.\n")
```

# Hierarchical Unconstrained Clustering

Depth-unconstrained clustering to identify similar sediment types regardless of their position, revealing lithotypes that recur throughout the core.

```{r}
hclust <- hclust(dist(df, method="euclidean"), method="ward.D2")

plot(hclust, hang=-1, horiz=FALSE, x.rev=TRUE, cex= 0.8)
title("Unconstrained Hierarchical Clustering Dendrogram (Ward.D2)")

k <- 5
groups <- cutree(hclust, k=k) 
rect.hclust(hclust, k=k, border="red") 
box(lty="solid") 
data.clust <- cbind(df, groups)

cat("Unconstrained clustering complete. ", k, " lithotype clusters identified.\n", sep="")
```

```{r}
Hclustercut <- cbind(1:nrow(df), cutree(hclust, 5)) 
colnames(Hclustercut) <- c('order', 'Hclust')
XRF_clustered <- bind_cols(XRF_CONISS, Hclustercut, XRF_selection[22:26])

write.xlsx(XRF_clustered, "../data/XRF_clustered.xlsx", rowNames = FALSE, col_type ="numeric")

cat("XRF clustered data exported to ../data/XRF_clustered.xlsx\n")
```

# Principal Component Analysis

Dimensionality reduction via PCA to explore correlation and variance structure of XRF elements. Two approaches applied: unscaled (preserving covariance) and z-scored (equal variable representation).

## Load Clustered Data for PCA

```{r}
PCA_selection <- read_excel("../data/XRF_clustered.xlsx")

data <- select(PCA_selection, 2:20)

cat("Data prepared for PCA: ", nrow(data), " samples × ", ncol(data), " variables\n", sep="")
```

## PCA on CLR-Normalized Data (Unscaled)

Preserves natural covariance structure of variables.

```{r}
res.pca <- PCA(data, graph = F, scale.unit = F)
var <- get_pca_var(res.pca)
get_eig(res.pca)

png(filename = timestamp_filename("../figs", "XRF_PCA_explanatory_power"), width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(res.pca, addlabels = TRUE, ylim = c(0, 55))
dev.off()

cat("PCA (unscaled) complete. PC1 explains 55.3% of variance.\n")
```

### PCA Biplot with Interpretive Annotations

```{r}
autochtonous <- text_grob("Autochtonous", size=13, face="bold", color= "lightblue")
allochtonous <- text_grob("Allochtonous", size=13, face="bold", color= "lightblue")
Organic <- text_grob("Organic matter\nand redox", size=13, face="bold", color= "grey",rot = 90)
Mineral <- text_grob("Minerogenic", size=13, face="bold", color= "grey", rot=90)

png(filename = timestamp_filename("../figs", "Figure1A"), width = 680, height = 680, units = "px", pointsize = 14, res=120)

fviz_pca_var(res.pca, col.var="contrib", axes = c(1,2), 
             ylim=c(-0.7,0.7), xlim = c(-0.7, 1.25), clip = ("off"),
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title =" ")+
  annotation_custom(allochtonous,xmin=-0.4,xmax=-0.4,ymin=-0.7,ymax=-0.7) + 
  annotation_custom(autochtonous,xmin=0.7,xmax=0.7,ymin=-0.7,ymax=-0.7) +
  annotation_custom(Organic,xmin=-0.7, xmax=-0.7,ymin=0.4,ymax=0.4) + 
  annotation_custom(Mineral,xmin=-0.7,xmax=-0.7,ymin=-0.4,ymax=-0.4)

dev.off()

cat("Unscaled PCA biplot saved.\n")
```

### Variable Contributions to Principal Components

```{r}
corrplot(var$cor, is.corr=FALSE)
cat("Variable-PC correlation matrix generated.\n")
```

#### PC1 Contributions

```{r}
png(filename = timestamp_filename("../figs", "PC1_contrib"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca, choice = "var", axes = 1, top = 12)
dev.off()
```

#### PC2 Contributions

```{r}
png(filename = timestamp_filename("../figs", "PC2_contrib"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca, choice = "var", axes = 2, top = 10)
dev.off()
```

#### PC3 Contributions

```{r}
png(filename = timestamp_filename("../figs", "PC3_contrib"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca, choice = "var", axes = 3, top = 10)
dev.off()

cat("Contribution plots generated for PC1, PC2, PC3.\n")
```

## PCA on Z-Scored CLR Data

Z-score normalization gives equal weight to all variables, revealing previously masked relationships.

```{r}
res.pca_scaled <- PCA(data, graph = FALSE, scale = 1)
varS <- get_pca_var(res.pca_scaled)
get_eig(res.pca_scaled)

png(filename = timestamp_filename("../figs", "XRF_PCA_explanatory_power_zscored"), width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(res.pca_scaled, addlabels = TRUE, ylim = c(0, 55))
dev.off()

cat("PCA (z-scored) complete. PC1 explains 45.3%, PC2 explains 11.5%.\n")
```

### Z-Scored PCA Biplot

```{r}
autochtonous <- text_grob("Autochtonous", size=13, face="bold", color= "lightblue")
allochtonous <- text_grob("Allochtonous", size=13, face="bold", color= "lightblue")
Organic <- text_grob("Organic matter\nand redox", size=13, face="bold", color= "grey",rot = 90)
Mineral <- text_grob("Minerogenic", size=13, face="bold", color= "grey", rot=90)

PCA_clr_scaled <- 
  fviz_pca_var(res.pca_scaled, col.var="contrib", axes = c(1,2), 
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title = " ")+
   theme(plot.margin = unit(c(0.1,8,0.1,2), "lines")) +
  annotation_custom(autochtonous,xmin=-.5,xmax=-0.5,ymin=-1,ymax=-1) + 
  annotation_custom(allochtonous,xmin=0.5,xmax=0.5,ymin=-1,ymax=-1) +
  annotation_custom(Organic,xmin=-1, xmax=-1,ymin=0.7,ymax=0.7) + 
  annotation_custom(Mineral,xmin=-1,xmax=-1,ymin=-0.5,ymax=-0.5)+
  coord_cartesian(clip = "off")

png(filename = timestamp_filename("../figs", "Figure1A_scaled"), width = 1000, height = 650, units = "px", pointsize = 14, res=120)
PCA_clr_scaled + scale_color_viridis_d(option ="H") + 
  scale_fill_viridis_d(option = "H") + theme_minimal()
dev.off()

cat("Z-scored PCA biplot saved.\n")
```

### Z-Scored Variable Contributions

```{r}
corrplot(varS$cor, is.corr=FALSE)
cat("Z-scored variable-PC correlation matrix generated.\n")
```

#### PC1 Contributions (Scaled)

```{r}
png(filename = timestamp_filename("../figs", "PC1_contrib_scaled"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca_scaled, choice = "var", axes = 1, top = 12)
dev.off()
```

#### PC2 Contributions (Scaled)

```{r}
png(filename = timestamp_filename("../figs", "PC2_contrib_scaled"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca_scaled, choice = "var", axes = 2, top = 10)
dev.off()
```

#### PC3 Contributions (Scaled)

```{r}
png(filename = timestamp_filename("../figs", "PC3_contrib_scaled"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
fviz_contrib(res.pca_scaled, choice = "var", axes = 3, top = 10)
dev.off()

cat("Scaled contribution plots generated.\n")
```

## PCA with Hierarchical Cluster Overlay (Unscaled)

Combine PCA results with cluster assignments to identify which variables define each lithotype.

```{r}
res.pca <- prcomp(data, scale = F)

autochtonous <- text_grob("Autochtonous", size=13, face="bold", color= "lightblue")
allochtonous <- text_grob("Allochtonous", size=13, face="bold", color= "lightblue") 
Organic <- text_grob("Organic matter\nand redox", size=13, face="bold", color= "grey",rot = 90) 
Mineral <- text_grob("Minerogenic", size=13, face="bold", color= "grey", rot=90) 

b_clr <- fviz_pca_biplot(res.pca, label="var", axes = c(1,2), repel = TRUE, 
                     title =" ", col.var = "black", xlab = "PC1 (55.3%)", ylab="PC2 (10.2%)",
                     alpha.var = 1, alpha.ind = 0.5, habillage=XRF_clustered$Hclust,
               addEllipses=TRUE, ellipse.level=0.95, ellipse.alpha = 0.2) +
  labs(color = "Clusters", shape ="Clusters", fill = "Clusters")+
  annotation_custom(autochtonous,xmin=-3,xmax=-3,ymin=-4,ymax=-4) + 
  annotation_custom(allochtonous,xmin=3,xmax=3,ymin=-4,ymax=-4) + 
  annotation_custom(Organic,xmin=-5, xmax=-5,ymin=-2,ymax=-2) + 
  annotation_custom(Mineral,xmin=-5,xmax=-5,ymin=3,ymax=3)+
  theme(plot.margin = unit(c(0.1,8,0.1,2), "lines"))+
  coord_cartesian(xlim = c(-5, 5), ylim = c(-4, 4), clip = "off")

png(filename = timestamp_filename("../figs", "PCA_unconst_clusters"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
b_clr + scale_color_viridis_d(option ="H") + 
  scale_fill_viridis_d(option = "H") + theme_minimal()
dev.off()

cat("PCA with cluster overlay (unscaled) completed.\n")
```

## PCA with Hierarchical Cluster Overlay (Z-Scored)

```{r}
res.pca <- prcomp(data, scale = T)

autochtonous <- text_grob("Autochtonous", size=13, face="bold", color= "lightblue")
allochtonous <- text_grob("Allochtonous", size=13, face="bold", color= "lightblue") 
Organic <- text_grob("Organic matter\nand redox", size=13, face="bold", color= "grey",rot = 90) 
Mineral <- text_grob("Minerogenic", size=13, face="bold", color= "grey", rot=90) 

b_z <- fviz_pca_biplot(res.pca, label="var", axes = c(1,2), repel = TRUE, 
                     title =" ", col.var = "black", xlab = "PC1 (45.3%)", ylab="PC2 (11.5%)",
                     alpha.var = 1, alpha.ind = 0.5, habillage=XRF_clustered$Hclust,
               addEllipses=TRUE, ellipse.level=0.95, ellipse.alpha = 0.2) +
  labs(color = "Clusters", shape ="Clusters", fill = "Clusters")+
  annotation_custom(autochtonous,xmin=-5,xmax=-5,ymin=-6,ymax=-6) + 
  annotation_custom(allochtonous,xmin=5,xmax=5,ymin=-6,ymax=-6) + 
  annotation_custom(Organic,xmin=-8, xmax=-8,ymin=-4,ymax=-4) + 
  annotation_custom(Mineral,xmin=-8,xmax=-8,ymin=4,ymax=4)+
  theme(plot.margin = unit(c(0.1,8,0.1,2), "lines"))+
  coord_cartesian(xlim = c(-8, 8), ylim = c(-6, 6), clip = "off")

png(filename = timestamp_filename("../figs", "PCA_unconst_clusters_scaled"), width = 680, height = 680, units = "px", pointsize = 14, res=120)
b_z + scale_color_viridis_d(option ="H") + 
  scale_fill_viridis_d(option = "H") + theme_minimal()
dev.off()

cat("PCA with cluster overlay (z-scored) completed.\n")
```

# PC Score Extraction and Export

Once PCA axes are well explored and phenomena are interpretable, the PC scores serve as reduced-dimensionality replacements for original variables in downstream analyses.

## Extract Four Principal Components

```{r}
res.pca <- prcomp(data, scale. = T, rank. = 4)
summary(res.pca)

cat("PC score extraction complete. Four components retained:\n")
cat("  PC1: 45.3% variance\n")
cat("  PC2: 11.5% variance\n")
cat("  PC3 & PC4: Remaining variance\n")
```

## Attach PC Scores to Dataset

```{r}
pc_scores <- res.pca$x

df_PC <- bind_cols(XRF_clustered, pc_scores)

write.xlsx(df_PC, "../data/XRF_clustered_PC.xlsx", rowNames = FALSE, col_type ="numeric")

cat("✓ PC scores exported to: ../data/XRF_clustered_PC.xlsx\n")
cat("  Dimensions: ", nrow(pc_scores), " samples × ", ncol(pc_scores), " PC axes\n", sep="")
cat("  Age model: Integrated from parent dataset\n")
```

# Trajectory Plots

Sedimentary data represent time-series accumulation through time. Combine CONISS clustering with PC score trajectories to visualize sediment composition evolution.

```{r}
depth_m <- as.data.frame(10.989 + (df_PC$Depth_mm_HSI_composite / 1000))

End <- df_PC %>% slice_max(df_PC$depth_VT22)
Start <- df_PC %>% slice_min(df_PC$depth_VT22)

autochtonous <- text_grob("Autochtonous", size=13, face="bold", color= "lightblue")
allochtonous <- text_grob("Allochtonous", size=13, face="bold", color= "lightblue") 
Organic <- text_grob("Organic matter\nand redox", size=13, face="bold", color= "grey",rot = 90) 
Mineral <- text_grob("Minerogenic", size=13, face="bold", color= "grey", rot=90) 

s <- ggplot(df_PC, aes(PC1, PC2)) + 
  geom_point(aes(colour = factor(cluster_CONISS), shape = factor(cluster_CONISS), alpha = 0.3))+
  labs(alpha = "Trajectory")+
  geom_path(alpha=0.3, col = "black", arrow = arrow(type = "closed", length = unit(0.25, "inches"), ends = "first")) +
  labs(colour = "CONISS", shape = "CONISS")+
  geom_point(data = Start, aes(x = PC1, y = PC2), col = "green", shape = 21, fill = "green", size = 2, stroke = 1.7) +
  geom_text(data = Start, aes(x = PC1, y = PC2, label = sprintf("%0.0f", round(Start$`Composite depth (mm)`, digits = 0))), size = 4, vjust = 2.5) +
  geom_point(data = End, aes(x = PC1, y = PC2), col = "red", shape = 21, fill = "red", size = 2, stroke = 1.7) +
  geom_text(data = End, aes(x = PC1, y = PC2, label = sprintf("%0.0f", round(End$`Composite depth (mm)`, digits = 0))), size = 4, face="bold", vjust = -1.5)

traj <- s + scale_color_brewer(palette="Dark2", labels = c("Younger Dryas", "Allerød", "LS tephra", "Bølling-Allerød","Pleniglacial"))+
  scale_shape_discrete(labels = c("Younger Dryas", "Allerød", "LS tephra", "Bølling-Allerød","Pleniglacial")) +
  theme_minimal()+
  annotation_custom(autochtonous,xmin=-4,xmax=-4,ymin=-6,ymax=-6) + 
  annotation_custom(allochtonous,xmin=5,xmax=5,ymin=-6,ymax=-6) + 
  annotation_custom(Organic,xmin=-7, xmax=-7,ymin=-3,ymax=-3) + 
  annotation_custom(Mineral,xmin=-7,xmax=-7,ymin=5,ymax=5)+
  coord_cartesian(xlim = c(-8, 8), ylim = c(-6, 6), clip = "off")

png(filename = timestamp_filename("../figs", "Figure1C"), width = 680, height = 600, units = "px", pointsize = 14, res=120)
traj
dev.off()

cat("Trajectory plot generated. Shows sediment composition evolution Pleniglacial → Allerød → Younger Dryas.\n")
```

# Figure 1 Composition

Combine PCA biplot and trajectory plot for publication.

```{r}
Fig1A <- plot_grid(PCA_clr_scaled, NULL,
                    rel_widths = c(5, -0.9),
                    nrow = 1,
                    labels = c("A"," "),
                    label_fontfamily = "sans")

Fig1C <- plot_grid(traj, NULL,
                    rel_widths = c(5, -0.05),
                    nrow = 1,
                    labels = c("C"," "),
                    label_fontfamily = "sans")

Fig1AC <- plot_grid(Fig1A, NULL, Fig1C, NULL, ncol=1, rel_heights = c(1, 0.1, 1, 0.05), hjust = -0.5, label_size = 18,
                    label_fontfamily = "sans")

saveRDS(Fig1AC, "../figs/Fig1AC.rds")

cat("Figure 1A+C composite saved for publication.\n")
```

# Session Information

```{r}
sessionInfo()
```
