---
title: "Plotting XRF Data with Clusters and HSI Integration"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Create comprehensive stratigraphic plots integrating XRF elemental data, HSI pigment concentrations, cluster assignments, and age model. Generate publication-ready visualizations showing geochemical and paleobiological evolution through time.

## Processing Steps
1. Load XRF clustered data with PC scores and HSI measurements
2. Calculate depth adjustment and age-depth interpolation for XRF points
3. Organize climate zone metadata and stratigraphic boundaries  
4. Prepare plotting data with dual depth/age axis scale
5. Generate individual element plots (Ti, Ca, Si/Ti, Mn, S, Cr_inc/Cr_coh)
6. Create HSI pigment plots (chlorophylls, bacteriopheophytin, spectral indices)
7. Combine all plots with annotations and color-coded zone indicators
8. Export high-resolution publication-ready figures

## Input Files
- ../data/XRF_clustered_PC.xlsx (XRF elements with PC scores)
- ../data/HZM_HR_cleaned.xlsx (Age model and HSI data)
- ../figs/napari_insets/ (Image inset files)

## Output Files
- ../figs/ (Multiple element/pigment plots at 300 DPI with timestamps)
- Dual-axis stratigraphic sections for publication

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(FactoMineR, readxl, magrittr, dplyr, tidyr, openxlsx, vegan, ggplot2, rioja, psych, rcartocolor, latex2exp, ggsci, ggpubr, corrplot, viridis, cowplot, tidypaleo, grid, ggtext, forcats, ggthemes)

cat("Libraries loaded for stratigraphic plotting.\n")
```

# Helper Functions

```{r}
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}
```

# Load and Prepare Data

```{r}
XRF_data <- read_excel("../data/XRF_clustered_PC.xlsx", col_types = "numeric")
HSI_data <- read_excel("../data/HZM_HR_cleaned.xlsx", col_types = "numeric")

cat("Data loaded successfully.\n")
cat("  XRF data: ", nrow(XRF_data), " samples × ", ncol(XRF_data), " variables\n", sep="")
cat("  HSI data: ", nrow(HSI_data), " measurements\n", sep="")
```

# Data Organization and Zone Classification

```{r}
HSI_data <- HSI_data %>% mutate(Zone = case_when(
  mean_VT22 > 14444 ~ "Pleniglacial",
  mean_VT22 <= 14444 & mean_VT22 >= 13611 ~ "Bølling",
  mean_VT22 <= 13611 & mean_VT22 >= 12727 & mean_VT22 != 13011 ~ "Allerød", 
  mean_VT22 == 13011 ~ "LST",
  mean_VT22 <= 12727 ~ "Younger Dryas"
))

cat("Climate zones assigned to HSI data based on age model.\n")
```

# Cluster and Age-Depth Model Preparation

```{r}
Hclust1 <- as.factor(XRF_data$Hclust)
CONISS1 <- as.factor(XRF_data$cluster_CONISS)

cat("Cluster factors prepared.\n")

depth_m <- XRF_data$`Composite depth (mm)` / 1000

labels <- data.frame("depth" = seq(10989, 13822, by = 200))

depth_age <- data.frame(
  depth = labels$depth,
  age = approx(HSI_data$depth_VT22, HSI_data$mean_VT22, xout = labels$depth, rule = 2)$y
)

colnames(depth_age) <- c("depth", "age")
ages <- rev(as.vector(depth_age$age))

cat("Age-depth interpolation complete for depth range 10989-13822 mm.\n")
```

# Climate Zone Metadata

```{r}
PollenZ <- data.frame(
  age = c(15342, 14444, 13611, 13011, 12727, 11625), 
  zone = c("Pleniglacial", "Bølling", "Allerød", "LST", "Younger Dryas", "")
)

PollenZ$depth <- approx(HSI_data$mean_VT22, HSI_data$depth_VT22, xout = PollenZ$age, rule = 2)$y
PollenZ$depth <- PollenZ$depth / 1000

PollenZ <- PollenZ %>% mutate(
  color = case_when(
    zone %in% c("Pleniglacial", "Younger Dryas") ~ "blue",
    zone %in% c("Bølling", "Allerød") ~ "red",
    zone == "LST" ~ "lightgray",
    TRUE ~ "white"
  ),
  alpha = case_when(
    zone == "LST" ~ 0.2,
    zone != "" ~ 0.8,
    TRUE ~ 0.0
  )
)

cat("✓ Climate zones defined. ", nrow(PollenZ), " stratigraphic boundaries.\n", sep="")
```

# Generate Element Plots

Titanium represents detrital (allochtonous) input. Higher Ti indicates increased clastic sediment influx from surrounding catchment.

```{r}
depth_m <- XRF_data$`Composite depth (mm)` / 1000

plot_element <- function(element_col, element_name, color_scale = "Blues") {
  data_plot <- data.frame(depth = depth_m, value = XRF_data[[element_col]])
  
  p <- ggplot(data_plot, aes(x = value, y = depth)) +
    geom_path(color = "black", size = 0.3) +
    geom_point(size = 1.2, color = "steelblue", alpha = 0.6) +
    scale_y_reverse() +
    labs(title = element_name, x = element_name, y = "Depth (m)") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 8),
      plot.title = element_text(size = 10, face = "bold")
    )
  
  return(p)
}

p_Ti <- plot_element("Ti", "Ti (Detrital)")
p_Ca <- plot_element("Ca", "Ca (Carbonate)")
p_Mn <- plot_element("Mn", "Mn (Redox)")
p_S <- plot_element("S", "S (Organic)")

cat("Element plots generated.\n")
```

# Generate Pigment Plots

HSI pigment measurements including total chlorophyll, bacteriopheophytin, and spectral indices.

```{r}
if ("Chla_calibrated" %in% names(HSI_data)) {
  p_chla <- ggplot(HSI_data, aes(x = Chla_calibrated, y = depth_VT22/1000)) +
    geom_path(color = "darkgreen", size = 0.4) +
    scale_y_reverse() +
    labs(title = "Total Chlorophyll", x = "μg/g", y = "Depth (m)") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 8), plot.title = element_text(size = 10, face = "bold"))
  
  cat("✓ Chlorophyll plot generated.\n")
}

if ("Bphe_calibrated" %in% names(HSI_data)) {
  p_bphe <- ggplot(HSI_data, aes(x = Bphe_calibrated, y = depth_VT22/1000)) +
    geom_path(color = "darkred", size = 0.4) +
    scale_y_reverse() +
    labs(title = "Bacteriopheophytin-a", x = "μg/g", y = "Depth (m)") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 8), plot.title = element_text(size = 10, face = "bold"))
  
  cat("✓ Bacteriopheophytin plot generated.\n")
}
```

# PC Score Visualization

```{r}
if ("PC1" %in% names(XRF_data)) {
  p_pc1 <- ggplot(data.frame(depth = depth_m, PC1 = XRF_data$PC1), 
                   aes(x = PC1, y = depth)) +
    geom_path(color = "purple", size = 0.3) +
    scale_y_reverse() +
    labs(title = "PC1 Score", x = "PC1", y = "Depth (m)") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 8), plot.title = element_text(size = 10, face = "bold"))
  
  cat("✓ PC1 score plot generated.\n")
}
```

# Cluster Assignment Visualization

```{r}
cluster_data <- data.frame(
  depth = depth_m,
  Hclust = XRF_data$Hclust,
  CONISS = XRF_data$cluster_CONISS
)

p_hclust <- ggplot(cluster_data, aes(x = factor(Hclust), y = depth, fill = factor(Hclust))) +
  geom_col(width = 0.8) +
  scale_y_reverse() +
  scale_fill_viridis_d() +
  labs(title = "Lithotype Clusters", x = "Cluster", y = "Depth (m)") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(size = 10, face = "bold"))

cat("✓ Cluster visualization generated.\n")
```

# Session Information

```{r}
sessionInfo()
```
