---
title: "Plotting XRF data"
output: html_notebook
---

# Plotting PCA-selected data with clusters and HSI

Installing packages and loading libraries

```{r message=FALSE, warning=FALSE, include=FALSE}
#install.packages("pacman")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)

pacman::p_load(cluster, factoextra, FactoMineR, readxl, magrittr, readxl, dplyr, tidyr, openxlsx, vegan, ggplot2, rioja, psych, dendroextras, rcartocolor, latex2exp, ggsci, ggpubr, corrplot, viridis, cowplot, tidypaleo, grid, ggtext, bruceR, gapminer, forcats, ggthemes, ggmap, maps, ggimage, magick)
```

```{r}
# Create timestamped filename
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}
```

```{r}
XRF_data <- read_excel("../data/XRF_clustered_PC.xlsx", col_types = "numeric")
HSI_data <- read_excel("../data/HZM_HR_cleaned.xlsx", col_types = "numeric")
```

```{r}
#HSI_clean <- na.omit(HSI_data) # this doesn't work. I need to take the raw HSi, and do the downscale here. Also the calibration script needs to come.

#df_plot <- bind_cols(XRF_data[1:26], HSI_clean[3:11]) 

depth_m <- XRF_data$`Composite depth (mm)`/1000

HSI_data <- HSI_data %>% mutate(Zone = case_when(mean_VT22 > 14444 ~ "Pleniglacial",
                          mean_VT22  <= 14444 & mean_VT22  >= 13611  ~ "Bölling",
                          mean_VT22  <= 13611 & mean_VT22  >= 12727 & mean_VT22!=13011  ~ "Alleröd", 
                          mean_VT22 == 13011 ~ "LST",
                          mean_VT22  <= 12727 ~ "Younger Dryas")
  )

```

```{r}
Hclust1 <- as.factor(XRF_data$Hclust)
CONISS1 <- as.factor(XRF_data$cluster_CONISS)

```

```{r}
labels <- data.frame("depth" = seq(10989, 13822, by = 200))
depth_age <- LOOKUP(labels, c("depth"), HSI_data, c("depth_VT22"), "mean_VT22")
depth_age <- LOOKUP(labels, c("depth"), HSI_data, c("depth_VT22"), "mean_VT22")
names(depth_age)[names(depth_age)=="mean_VT22"] <- "age"
ages <- rev(as.vector(depth_age$age))
#breaks = c(13.792, 13.592, 13.392, 13.192, 12.992, 12.792, 12.592, 12.392, 12.192, 11.992, 11.792, 11.592, 11.392, 11.192, 10.992)
```

```{r}
PollenZ <- data.frame("age" = c(15342, 14444, 13611, 13011, 12727, 11625), "zone" = c("Pleniglacial", "Bølling", "Allerød", "LST", "Younger Dryas", ""))
PollenZ <- LOOKUP(PollenZ, c("age"), HSI_data, c("mean_VT22"), "depth_VT22")
PollenZ$depth <- PollenZ$depth_VT22/1000

PollenZ <- PollenZ %>% mutate(color= case_when(zone == "Pleniglacial" ~ "blue",
                           zone == "Bølling" ~ "red",
                           zone == "Allerød" ~ "red", 
                          zone == "LST" ~ "lightgray", 
                           zone == "Younger Dryas" ~ "blue")
   )

PollenZ <- PollenZ %>% mutate(line= case_when(zone == "Pleniglacial" ~ "solid",
                           zone == "Bølling" ~ "solid",
                           zone == "Allerød" ~ "solid", 
                          zone == "LST" ~ "solid", 
                           zone == "Younger Dryas" ~ "solid")
   )

PollenZ <- PollenZ %>% mutate(aplha= case_when(zone == "Pleniglacial" ~ "0.8",
                           zone == "Bølling" ~ "0.8",
                           zone == "Allerød" ~ "0.8", 
                          zone == "LST" ~ "0.2", 
                           zone == "Younger Dryas" ~ "0.8")
   )
```

```{r}
#AxisZ <- merge(depth_age, PollenZ, by = "age", all = T )

# age_depth <- LOOKUP(PollenZ, c("age"), HSI_data, c("mean_VT22"), "depth_VT22")
# zones <- rev(as.vector(age_depth$depth_VT22))
# age_depth <- age_depth %>% mutate(Zone = case_when(age > 14500 ~ "Pleniglacial",
#                           age == 14500   ~ "Bölling",
#                           age == 13750    ~ "Alleröd", 
#                           age == 13011 ~ "LST", 
#                           age <= 12750 ~ "Younger Dryas")
#   )


#LabZ <- LOOKUP(depth_age, c("depth"), HSI_data, c("depth_VT22"), "Zone")
#ages <- rev(as.vector(depth_age$mean_VT22))
```

# Plotting selected elements from the PCA of XRF

Ti, Cr_incCoh, Ca, Si/Ti for diatoms, PC1, PC2,

```{r}
# Plot numeric clusters (using factor levels)
#Ter <- textGrob("Detrital input", gp=gpar(fontsize=13, fontface="bold"))
Ti <- ggplot(XRF_data, aes(x=Ti, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "black", size= 0.5)+
 scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 0.2), digits = 2), expand=c(0,0)) +
 geom_path(alpha=.8, size=1.5, colour = "black")+
  scale_x_continuous(position = "bottom", expand=c(0,0))+ 
  labs(title = " ", x = "Ti (clr)\nDetrital input", y = " Depth (m) ")+
  theme_classic() +
  theme(legend.position="none", plot.caption=element_text(size=12, hjust=0), #margin=margin(15,0,0,0)), 
        axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,0.5), "cm"), panel.background = element_blank())


Fe <- ggplot(XRF_data, aes(x=Fe, y= depth_m))+
  geom_point(colour = "orange", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"),expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "orange") +
  labs(title = " ", x = " Fe (clr)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-1), "cm"), panel.background = element_blank())

Ca <- ggplot(XRF_data, aes(x=Ca, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "skyblue", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "skyblue") +
  labs(title = " ", x = "Carbonate precipitation \n Ca (clr)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-0.2), "cm"), panel.background = element_blank())

Si <- ggplot(XRF_data, aes(x=Si, y= depth_m))+
  geom_point(colour = "darkgrey", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "darkgrey") +
  labs(title = " ", x = " Si (clr)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-1), "cm"), panel.background = element_blank())

SiTi <- XRF_data$Si/XRF_data$Ti
SiTip<- ggplot(XRF_data, aes(x=SiTi, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "darkblue", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom" ,expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "darkblue") +
  labs(title = " ", x = " Si/Ti\n Diatoms", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-0.9), "cm"), panel.background = element_blank())


Mn <- ggplot(XRF_data, aes(x=Mn, y= depth_m))+
  geom_point(colour = "brown", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top" ,expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "brown") +
  labs(title = " ", x = " Mn (clr)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-1), "cm"), panel.background = element_blank())

S <- ggplot(XRF_data, aes(x=S, y= depth_m))+
  geom_point(colour = "goldenrod", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "goldenrod") +
  labs(title = " ", x = " S (clr)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-1), "cm"), panel.background = element_blank())

Crom <- ggplot(XRF_data, aes(x= Crinc_coh, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "brown", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0), labels = scales::number_format(accuracy = 0.1))+
 geom_path(alpha=.8, size=1.5, colour = "brown") +
  labs(title = " ", x = "Organic matter \n Cr[inc]/Cr[coh]", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-0.9), "cm"), panel.background = element_blank())

PC1ax <- ggplot(XRF_data, aes(x= XRF_data$PC1, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "lightblue", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "lightblue") +
  geom_vline(xintercept=0, color="lightblue", linetype = 2) +
  labs(title = " ", x = "PC1\nAutochtonous <-> Allochtonous", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,-0.2,0.5,0), "cm"), panel.background = element_blank()) 

PC2ax <- ggplot(XRF_data, aes(x= XRF_data$PC2, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "grey", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "grey") +
  geom_vline(xintercept=0, color="grey", linetype = 2) +
  labs(title = " ", x = "Organic/Redox <-> Minerals\nPC2" , y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0.1,0.5,-1.1), "cm"), panel.background = element_blank()) 


p4<- ggplot(XRF_data, aes(x = 0.05, y = XRF_data$order)) +  # Set transparency
  geom_tile(aes(fill = Hclust1)) + 
  scale_y_reverse(limits = c(2827,1), breaks = round(seq(1, 2827, by = 10), digits = 0), expand=c(0,0)) + 
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  labs(title = " ", x = "HClust", y = " ") +  # Adjust labels
  scale_fill_viridis_d(option = "H") +
  theme_classic() +
  theme(legend.position="none", axis.text = element_blank(), axis.line.x = element_line(colour = "white"), axis.ticks= element_blank(),  axis.line.y = element_blank(), plot.margin = unit(c(0,0,0.5,-1.1), "cm"), panel.background = element_blank()) #+ scale_x_continuous(expand = expansion(mult = c(0, 0.5)))


p5<- ggplot(XRF_data, aes(x = 0.05, y = XRF_data$order_CONISS)) +  # Set transparency
  geom_tile(aes(fill = CONISS1)) + 
 scale_y_reverse(limits = c(2827,1), breaks = round(seq(1, 2827, by = 10), digits = 0), expand=c(0,0)) + 
  scale_x_continuous(position = "top",expand=c(0,0))+
  labs(title = " ", x = "CONISS", y = " ") +  # Adjust labels
  #scale_fill_viridis_b(option = "D") +
  #scale_fill_fermenter (palette = "Paired")+
 # scale_fill_viridis_d(option = "C") +
  scale_fill_brewer(palette = "Dark2")+
  theme_classic() +
  theme(legend.position="none", axis.text = element_blank(), axis.line.x = element_line(colour = "white"), axis.ticks= element_blank(), axis.line.y = element_blank(), plot.margin = unit(c(0,0,0.5,-0.8), "cm"), panel.background = element_blank()) #+ #legend.text = element_blank(), legend.position = "top")
#scale_x_continuous(expand = expansion(mult = c(1, 0.5)))
```

## Plotting HSI data along with the XRF

```{r}
Chl <- ggplot(HSI_data, aes(x=RABD670calib, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "green3", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "green3") +
  labs(title = " ", x = " Total chlorophylls\n(µg/g wet sed)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-0.7), "cm"), panel.background = element_blank())

Bphe <- ggplot(HSI_data, aes(x=RABD845calib, y= depth_m))+
  geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  annotate("text", x= 5.2, y= 13.4, label = "Pleniglacial", color= "blue")+
  annotate("text", x= 5, y= 12.6, label = "Bølling", color= "darkred")+
  annotate("text", x= 7.7, y= 12.19, label = "Allerød", color= "darkred")+
  annotate("text", x= 6.2, y= 11.62, label = "Lachersee \ntephra", color= "black")+
  annotate("text", x= 5.5, y= 11.1, label = "Younger \nDryas", color= "blue")+
  geom_point(colour = "magenta4", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 0.2), digits = 2), labels = rev(ages), guide = guide_axis(position = "right"), expand=c(0,0)) +
  # scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "magenta4") +
  labs(title = " ", x = " Bacteriophyophytin-a\n(µg/g wet sed)", y = "Age (Cal yr BP)\nnon-linear axis") +
  theme_classic() +
  theme(legend.position="none", axis.text = element_text(size = 12), plot.margin = unit(c(0,0.3,0.5,-0.7), "cm"), panel.background = element_blank(), axis.title.y = element_text(angle = 270, vjust = 0.5))
  # theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-1), "cm"), panel.background = element_blank())


Cyano <- ggplot(HSI_data, aes(x=RABD620, y= depth_m))+
 # geom_text(aes(max(RABD620),PollenZ$depth,label = PollenZ$depth, vjust = -1))+
  geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "dodgerblue1", size= 0.5)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  # scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 0.2), digits = 2), labels = rev(ages), guide = guide_axis(position = "right"), expand=c(0,0), sec.axis = sec_axis(~., name = "Age (Cal yr BP) \n non-linear axis")) +
  #guides(y = "none"))+ #sec.axis = dup_axis(breaks = HSI_data$depth_VT22, labels = HSI_data$mean_VT22, name ="Age yr Cal BP"))) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=1.5, colour = "dodgerblue1") +
 geom_vline(xintercept=mean(HSI_data$RABD620), color="dodgerblue1", linetype = 1) +
  geom_vline(xintercept=mean(HSI_data$RABD620) + sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_vline(xintercept=mean(HSI_data$RABD620) - sd(HSI_data$RABD620), color="gray", linetype = 2) +
  labs(title = " ", x = " RABD620\n(Phycocyanin)", y = " ") +
  # annotate("text", x= 1.01, y= 13.4, label = "Pleniglacial", color= "blue")+
  # annotate("text", x= 1, y= 12.6, label = "Bølling", color= "darkred")+
  # annotate("text", x= 1, y= 12, label = "Allerød", color= "darkred")+
  # annotate("text", x= 1.017, y= 11.62, label = "Lachersee \n tephra", color= "black")+
  # annotate("text", x= 1.017, y= 11.1, label = "Younger \n Dryas", color= "blue")+
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,-0.7), "cm"), panel.background = element_blank())

#breaks = sort(c(ggplot_build(Cyano)$layout$panel_ranges[[1]]$y.major_source, PollenZ$depth))

```

```{r}
# img <- readPNG(system.file("img", "~/work/holzmaar/data/SourceData/all_concat_manu.png", package ="png"))
# 
# maps <- ggplot(XRF_data, aes(x = PC2, y = depth_m))+
#   geom_point(size = 2)+
#   theme_minimal(base_size = 12, base_family = 'Source Sans Pro') +
  
  # geom_image(data= tibble(PC2 = 0 , depth_m = 14), 
  #            aes(image = "~/work/holzmaar/data/SourceData/all_concat_manu.png")
  #            )
library(png)

img <- readPNG("../data/SourceData/all_concat_manuXRF-02.png")

maps <- ggplot(data= XRF_data, 
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  theme_cowplot()+
  annotation_custom(rasterGrob(img, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " Pigment \n intensities ", #Pigment intensity maps \n Cyanobacteria | Green Algae | Green Suplhur Bacteria | Purple Suphur Bacteria", 
       y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.7), "cm"), panel.background = element_blank(), axis.line.x = element_blank(), axis.ticks.x = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank())

PhyMap <- readPNG("../data/SourceData/Phyco.png")
TchlMap <- readPNG("../data/SourceData/TChl.png")
BpheMap <- readPNG("../data/SourceData/Bphe.png")

Phymaps <- ggplot(data= XRF_data, 
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  theme_cowplot()+
  annotation_custom(rasterGrob(PhyMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", #Pigment intensity maps \n Cyanobacteria | Green Algae | Green Suplhur Bacteria | Purple Suphur Bacteria", 
       y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.8), "cm"), panel.background = element_blank(), axis.line.x = element_blank(), axis.ticks.x = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank())

TChlmaps <- ggplot(data= XRF_data, 
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  theme_cowplot()+
  annotation_custom(rasterGrob(TchlMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", #Pigment intensity maps \n Cyanobacteria | Green Algae | Green Suplhur Bacteria | Purple Suphur Bacteria", 
       y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.8), "cm"), panel.background = element_blank(), axis.line.x = element_blank(), axis.ticks.x = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank())


Bphemaps <- ggplot(data= XRF_data, 
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  theme_cowplot()+
  annotation_custom(rasterGrob(BpheMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", #Pigment intensity maps \n Cyanobacteria | Green Algae | Green Suplhur Bacteria | Purple Suphur Bacteria", 
       y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.8), "cm"), panel.background = element_blank(), axis.line.x = element_blank(), axis.ticks.x = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank())




```

```{r}
# Si <- ggplot(XRF_data, aes(x=Si, y= depth_m))+
#   geom_point(colour = "darkgrey", size= 0.01)+
#   geom_path(alpha=.8, size=0.2, colour = "darkgrey") +
#   labs(title = " ", x = " Si (clr)", y = " ") +
#   scale_x_continuous(position = "bottom" ,expand=c(0,0))+
#  # labs(title = " ", x = " Si/Ti", y = " ")+
#   geom_point(aes(x=SiTi, y= depth_m), colour = "darkblue", size= 0.01)+
#   geom_path(aes(x=SiTi, y= depth_m), alpha=.8, size=0.2, colour = "darkblue")+
#   scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
#   scale_x_age_depth(SiTi, depth_name= "Si/Ti", expand=c(0,0))
#   
# 
#   # 
#   #  
#   # scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
#   # 
#  
#  
# Si
#   
#    theme_classic() +
#   theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
# 
# SiTi <- XRF_data$Si/XRF_data$Ti
# SiTip<- ggplot(XRF_data, aes(x=SiTi, y= depth_m))+
#   geom_point(colour = "darkblue", size= 0.01)+
#   scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
#   scale_x_continuous(position = "bottom" ,expand=c(0,0))+
#  geom_path(alpha=.8, size=0.2, colour = "darkblue") +
#   labs(title = " ", x = " Si/Ti", y = " ") +
#   theme_classic() +
#   theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

# Complete plot

```{r}
png(filename = timestamp_filename("../figs", "Figure1B"), width = 1754, height = 1240, units = "px", pointsize = 12, res=105)
linePlot <- plot_grid(Ti, Ca,  SiTip, Crom, p4, p5, PC1ax, PC2ax, Phymaps, Cyano, TChlmaps, Chl, Bphemaps, Bphe, #maps, 
nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1.4,0.8,0.6,0.6,
                        0.3,0.3,
                         1.3,1.2,
                         0.2,0.7,0.2,0.7,0.2,1.1), axis = "tb" )
linePlot
dev.off()
#linePlot
```

```{r}
cow_B<-plot_grid(NULL,linePlot,
                    rel_heights = c(0, 0.99),
                 rel_widths = c(0.01, 1),
                    ncol = 1,
                    #label_y = 1 + 0.07 /1.85,
                    labels = c(" ","B"),
                    #label_size = 18,
                    label_fontfamily = "sans")

cow_B
```

```{r}

Fig1AC <- readRDS("../figs/Fig1AC.rds")

png(filename = timestamp_filename("../figs", "Figure1"), width = 3000, height = 1500, units = "px", pointsize = 14, res=120)
#Fig1ABC <- 
plot_grid(cow_B, Fig1AC, rel_widths = c(3.4, 1.3), nrow=1)# axis ="tb", labels = c("B", " "))

dev.off()
#Fig1ABC

```

# Scientific Journal Version - Publication Ready Figure

This section creates a publication-ready version of Figure 1B optimized for scientific journals. The figure is sized at approximately 180mm width (suitable for double-column layout) at 300 DPI with appropriately sized fonts for print.

```{r}
# Font sizes for journal publication (adjusted for better readability)
journal_font_size <- 7
journal_axis_text_size <- 6
journal_title_size <- 7

# Create plots with appropriately sized fonts for journal publication
Ti_journal <- ggplot(XRF_data, aes(x=Ti, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "black", size= 0.3)+
 scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 0.2), digits = 2), expand=c(0,0)) +
 geom_path(alpha=.8, size=0.5, colour = "black")+
  scale_x_continuous(position = "bottom", expand=c(0,0))+ 
  labs(title = " ", x = "Ti (clr)\nDetrital input", y = " Depth (m) ")+
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", plot.caption=element_text(size=journal_font_size, hjust=0),
        axis.text = element_text(size = journal_axis_text_size), 
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,0.5), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

Ca_journal <- ggplot(XRF_data, aes(x=Ca, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "skyblue", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "skyblue") +
  labs(title = " ", x = "Carbonate precipitation \n Ca (clr)", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,0), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

SiTi_journal <- XRF_data$Si/XRF_data$Ti
SiTip_journal <- ggplot(XRF_data, aes(x=SiTi_journal, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "darkblue", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom" ,expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "darkblue") +
  labs(title = " ", x = " Si/Ti\n Diatoms", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

Crom_journal <- ggplot(XRF_data, aes(x= Crinc_coh, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "brown", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0), labels = scales::number_format(accuracy = 0.1))+
 geom_path(alpha=.8, size=0.5, colour = "brown") +
  labs(title = " ", x = "Organic matter \n Cr[inc]/Cr[coh]", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-1), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

p4_journal <- ggplot(XRF_data, aes(x = 0.05, y = XRF_data$order)) +
  geom_tile(aes(fill = Hclust1)) + 
  scale_y_reverse(limits = c(2827,1), breaks = round(seq(1, 2827, by = 10), digits = 0), expand=c(0,0)) + 
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  labs(title = " ", x = "HClust", y = " ") +
  scale_fill_viridis_d(option = "H") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.text = element_blank(), axis.line.x = element_line(colour = "white"), 
        axis.ticks= element_blank(), axis.line.y = element_blank(), 
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,-0.25,0.5,-0.7), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

p5_journal <- ggplot(XRF_data, aes(x = 0.1, y = XRF_data$order_CONISS)) +
  geom_tile(aes(fill = CONISS1)) + 
 scale_y_reverse(limits = c(2827,1), breaks = round(seq(1, 2827, by = 10), digits = 0), expand=c(0,0)) + 
  scale_x_continuous(position = "top",expand=c(0,0))+
  labs(title = " ", x = "CONISS", y = " ") +
  scale_fill_brewer(palette = "Dark2")+
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.text = element_blank(), axis.line.x = element_line(colour = "white"), 
        axis.ticks= element_blank(), axis.line.y = element_blank(), 
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0.2,0.5,-0.2), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

PC1ax_journal <- ggplot(XRF_data, aes(x= XRF_data$PC1, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "lightblue", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "lightblue") +
  geom_vline(xintercept=0, color="lightblue", linetype = 2) +
  labs(title = " ", x = "PC1\nAutochtonous <-> Allochtonous", y = " ", fontsize = 6) +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-0.2), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA)) 

PC2ax_journal <- ggplot(XRF_data, aes(x= XRF_data$PC2, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "grey", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "grey") +
  geom_vline(xintercept=0, color="grey", linetype = 2) +
  labs(title = " ", x = "Organic/Redox <-> Minerals\nPC2" , y = " ", fontsize = 6) +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-0.45), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

Phymaps_journal <- ggplot(data= XRF_data,
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  theme_cowplot(font_size = journal_font_size)+
  annotation_custom(rasterGrob(PhyMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(), 
        axis.line.y = element_blank(), axis.ticks.y = element_blank())

Cyano_journal <- ggplot(HSI_data, aes(x=RABD620, y= depth_m))+
  geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "dodgerblue1", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "dodgerblue1") +
 geom_vline(xintercept=mean(HSI_data$RABD620), color="dodgerblue1", linetype = 1) +
  geom_vline(xintercept=mean(HSI_data$RABD620) + sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_vline(xintercept=mean(HSI_data$RABD620) - sd(HSI_data$RABD620), color="gray", linetype = 2) +
  labs(title = " ", x = " RABD620\n(Phycocyanin)", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

TChlmaps_journal <- ggplot(data= XRF_data,
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  theme_cowplot(font_size = journal_font_size)+
  annotation_custom(rasterGrob(TchlMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(), 
        axis.line.y = element_blank(), axis.ticks.y = element_blank())

Chl_journal <- ggplot(HSI_data, aes(x=RABD670calib, y= depth_m))+
    geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(colour = "green3", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 1), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "green3") +
  labs(title = " ", x = " Total chlorophylls\n(µg/g wet sed)", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size),
        plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))

Bphemaps_journal <- ggplot(data= XRF_data,
         mapping = aes(x= PC2, y =depth_m)) + 
  geom_point(alpha = 0.01)  +
  scale_y_continuous(limits = c(10.989,13.822), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  theme_cowplot(font_size = journal_font_size)+
  annotation_custom(rasterGrob(BpheMap, width = 1, height = 1), 
                    xmin = -4.2, xmax = Inf,
                    ymin = 10.989 , ymax = 13.822 )+
  labs(title = " ", x = " ", y = " ") +
  theme(legend.position="none", axis.text = element_blank(), plot.margin = unit(c(0,0,0.5,-0.35), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.line.x = element_blank(), axis.ticks.x = element_blank(), 
        axis.line.y = element_blank(), axis.ticks.y = element_blank())

Bphe_journal <- ggplot(HSI_data, aes(x=RABD845calib, y= depth_m))+
  geom_hline(yintercept = PollenZ$depth, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  annotate("text", x= 5.9, y= 13.4, label = "Pleniglacial", color= "blue", size=2.5)+
  annotate("text", x= 5.1, y= 12.6, label = "Bølling", color= "darkred", size=2.5)+
  annotate("text", x= 7.72, y= 12.19, label = "Allerød", color= "darkred", size=2.5)+
  annotate("text", x= 6.4, y= 11.62, label = "Lachersee \ntephra", color= "black", size=1.5)+
  annotate("text", x= 5.8, y= 11.1, label = "Younger \nDryas", color= "blue", size=2.5)+
  geom_point(colour = "magenta4", size= 0.3)+
  scale_y_reverse(limits = c(13.822, 10.989), breaks = round(seq(10.989, 13.822, by = 0.2), digits = 2), labels = rev(ages), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
 geom_path(alpha=.8, size=0.5, colour = "magenta4") +
  labs(title = " ", x = " Bacteriophyophytin-a\n(µg/g wet sed)", y = "Age (Cal yr BP)\nnon-linear axis") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", 
        axis.text = element_text(size = journal_axis_text_size), 
        axis.title.x = element_text(size = journal_title_size),
        axis.title.y = element_text(angle = 270, vjust = 0.5, size = journal_title_size),
        plot.margin = unit(c(0,0.3,0.5,-0.3), "cm"), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

```{r}
# Create journal figure at 180mm width (double column) x ~120mm height
# Resolution: 300 DPI for journal quality
# 180mm = 7.09 inches, 120mm = 4.72 inches
# At 300 DPI: width = 2126 px, height = 1417 px

png(filename = timestamp_filename("../figs", "Figure1B_journal"), 
    width = 2126, height = 1417, units = "px", pointsize = journal_font_size, res=300, bg = "transparent")

linePlot_journal <- plot_grid(Ti_journal, Ca_journal, SiTip_journal, Crom_journal, 
                              p4_journal, p5_journal, 
                              PC1ax_journal, PC2ax_journal, 
                              Phymaps_journal, Cyano_journal, 
                              TChlmaps_journal, Chl_journal, 
                              Bphemaps_journal, Bphe_journal,
                              nrow = 1, align = "h",
                              rel_widths = c(1.1, 0.6, 0.6, 0.4,
                                           0.05, 0.35,
                                           0.8, 0.8,
                                           0.08, 0.55, 0.08, 0.55, 0.08, 1.1), 
                              axis = "tb")
linePlot_journal
dev.off()

# Display the plot
linePlot_journal
```
