---
title: "Clustering and PCA of Merged HSI and XRF Datasets"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Merge HSI (hyperspectral reflectance) and XRF (elemental) datasets via Additive Log Ratio (ALR) transformations. Conduct integrated clustering and PCA to explore combined geochemical and optical signals.

## Processing Steps
1. Load HSI and XRF datasets
2. Prepare HSI and XRF variables for ALR transformation
3. Apply ALR normalization to both datasets
4. Merge datasets to common depth scale
5. Perform CONISS constrained clustering on merged data
6. Perform unconstrained hierarchical clustering on merged data
7. Conduct PCA on merged ALR-transformed data (unscaled and z-scored)
8. Extract and visualize PC scores with cluster overlays
9. Export merged dataset with cluster and PC assignments

## Input Files
- ../data/HZM_HSI_calibrated.xlsx
- ../data/XRF_HZM_clr.xlsx

## Output Files
- ../data/merged_clustered.xlsx
- ../data/XRFHSI_clustered_PC.xlsx

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(cluster, factoextra, FactoMineR, readxl, magrittr, dplyr, tidyr, openxlsx, vegan, ggplot2, rioja, psych, dendroextras, rcartocolor, latex2exp, ggsci, ggpubr, corrplot, viridis, cowplot)

cat("Libraries loaded for merged dataset analysis.\n")
```

# Load Data

```{r}
HSI_raw <- read_excel("../data/HZM_HSI_calibrated.xlsx", col_types = "numeric")
XRF_raw <- read_excel("../data/XRF_HZM_clr.xlsx", col_types = "numeric")

cat("HSI data loaded: ", nrow(HSI_raw), " samples\n", sep="")
cat("XRF data loaded: ", nrow(XRF_raw), " samples\n", sep="")
```

# Data Preparation and ALR Transformation

ALR (Additive Log Ratio) transformation normalizes compositional data, converting counts or percentages to log-ratios that respect the closed nature of compositional datasets (where rows sum to constant values).

```{r}
HSI_selection <- select(HSI_raw, contains("calibrated"))
XRF_selection <- select(XRF_raw, 2:19)

HSI_depth <- as.numeric(HSI_raw[, 1]) + 10989
XRF_depth <- as.numeric(XRF_raw[, 1])

cat("Selected columns: HSI=", ncol(HSI_selection), ", XRF=", ncol(XRF_selection), "\n", sep="")
```

## ALR Transformation Function

```{r}
alr_transform <- function(x) {
  x_positive <- pmax(x, 1e-6)
  divisor <- 10^6 - rowSums(x_positive)
  divisor <- pmax(divisor, 1e-6)
  log_ratios <- log(x_positive / divisor)
  return(log_ratios)
}

HSI_alr <- alr_transform(HSI_selection)
XRF_alr <- alr_transform(XRF_selection)

cat("✓ ALR transformation applied to HSI (", ncol(HSI_alr), " variables)\n", sep="")
cat("✓ ALR transformation applied to XRF (", ncol(XRF_alr), " variables)\n", sep="")
```

## Merge Datasets to Common Depth Scale

```{r}
interpolate_to_xrf <- function(source_depth, source_data, target_depth) {
  result <- data.frame(matrix(NA, nrow = length(target_depth), ncol = ncol(source_data)))
  
  for (i in 1:ncol(source_data)) {
    valid_idx <- !is.na(source_data[, i])
    if (sum(valid_idx) > 1) {
      result[, i] <- approx(
        x = source_depth[valid_idx],
        y = as.numeric(source_data[valid_idx, i]),
        xout = target_depth,
        rule = 2
      )$y
    }
  }
  
  names(result) <- names(source_data)
  return(result)
}

HSI_alr_interp <- interpolate_to_xrf(HSI_depth, HSI_alr, XRF_depth)

merged_data <- cbind(XRF_alr, HSI_alr_interp)

cat("Merged dataset: ", nrow(merged_data), " samples × ", ncol(merged_data), " variables\n", sep="")
```

# CONISS Clustering on Merged Data

```{r}
clust_coniss <- chclust(dist(merged_data, method = "euclidean"), method = "coniss")

plot(clust_coniss, hang = -1, horiz = FALSE, x.rev = TRUE, cex = 0.8)
title("CONISS Clustering - Merged Dataset")

k_coniss <- 5
groups_coniss <- cutree(clust_coniss, k = k_coniss)
rect.hclust(clust_coniss, k = k_coniss, border = "red")
box(lty = "solid")

cat("✓ CONISS clustering complete. ", k_coniss, " zones identified.\n", sep="")
```

# Unconstrained Hierarchical Clustering

```{r}
clust_unconstr <- hclust(dist(merged_data, method = "euclidean"), method = "ward.D2")

plot(clust_unconstr, hang = -1, horiz = FALSE, x.rev = TRUE, cex = 0.8)
title("Unconstrained Clustering - Merged Dataset (Ward.D2)")

k_unconstr <- 3
groups_unconstr <- cutree(clust_unconstr, k = k_unconstr)
rect.hclust(clust_unconstr, k = k_unconstr, border = "blue")
box(lty = "solid")

cat("✓ Unconstrained clustering complete. ", k_unconstr, " lithotype clusters identified.\n", sep="")
```

# Save Clustered Merged Data

```{r}
merged_clustered <- data.frame(
  Depth_mm = XRF_depth,
  merged_data,
  CONISS_cluster = groups_coniss,
  Unconstrained_cluster = groups_unconstr
)

write.xlsx(merged_clustered, "../data/merged_clustered.xlsx", rowNames = FALSE)

cat("Merged clustered data exported to ../data/merged_clustered.xlsx\n")
```

# Principal Component Analysis on Merged Data

## PCA - Unscaled (Preserves Covariance)

```{r}
res_pca_unscaled <- PCA(merge_data, scale.unit = FALSE, graph = FALSE)
eig_unscaled <- get_eig(res_pca_unscaled)

cat("PCA (unscaled):\n")
print(eig_unscaled)

cat("PC1: ", round(eig_unscaled[1, 2], 1), "% variance\n", sep="")
cat("PC2: ", round(eig_unscaled[2, 2], 1), "% variance\n", sep="")
```

## PCA - Z-Scored

```{r}
res_pca_scaled <- PCA(merged_data, scale.unit = TRUE, graph = FALSE)
eig_scaled <- get_eig(res_pca_scaled)

cat("PCA (z-scored):\n")
print(eig_scaled)

cat("PC1: ", round(eig_scaled[1, 2], 1), "% variance\n", sep="")
cat("PC2: ", round(eig_scaled[2, 2], 1), "% variance\n", sep="")
```

# PC Score Extraction and Export

```{r}
pc_scores <- res_pca_scaled$ind$coord[, 1:4]

xrfhsi_clustered_pc <- data.frame(
  Depth_mm = XRF_depth,
  merged_clustered[, -1],
  PC1 = pc_scores[, 1],
  PC2 = pc_scores[, 2],
  PC3 = pc_scores[, 3],
  PC4 = pc_scores[, 4]
)

write.xlsx(xrfhsi_clustered_pc, "../data/XRFHSI_clustered_PC.xlsx", rowNames = FALSE)

cat("✓ Merged dataset PC scores exported to: ../data/XRFHSI_clustered_PC.xlsx\n")
cat("  Dimensions: ", nrow(xrfhsi_clustered_pc), " samples × ", ncol(xrfhsi_clustered_pc), " variables (including 4 PC axes)\n", sep="")
```

# Session Information

```{r}
sessionInfo()
```
