---
title: R Notebook
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Loading and installing packages

```{r include=FALSE}
#install.packages("pacman")
```

```{r include=FALSE}
library("pacman")

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("readxl", "dplyr", "tidyr", "openxlsx", "FactoMineR", "cluster", "factoextra", "magrittr", "vegan", "ggplot2", "rioja", "psych", "dendroextras", "rcartocolor", "latex2exp", "ggsci", "ggpubr", "corrplot", "viridis", "cowplot", "ggrepel", "janitor", "pheatmap","scatterplot3d", "tidypaleo", "grid", "ggtext", "bruceR", "gapminder", "forcats", "ggthemes", "ggmap", "maps", "reshape2")
```

```{r}
# Create timestamped filename
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}
```

# Loading data

Cleaning and preparing datasets of the low res data for the stats

First we load the low resolution dataset, which contains of: TOC, TN, TC, TS, sequential extraction fractions of Fe, Mn and P, diatom counts and pigment concentrations:

-   Sheet Redox_Dia_PigOM is pigments normalized by the organic matter content

-   Sheet Redox_Dia_PigBulk contains pigment concentrations normalized by bulk wet sediment weight and

-   Sheet Redox_Dia_PigOMug is the pigment concentrations normalized by OM content and converted from nmol/g to ug/g).

The conversion can be also done on the bulk normalized pigment concentration by using the molecular weights in the Sheet "Pig conversion".

```{r}
LR_all <- read_excel("../data/SourceData/LR_data.xlsx", sheet = "Redox_Dia_PigOMug")

depth_LR <- select(LR_all, 1)
age_LR <- select(LR_all, 3)
Pig <- select(LR_all, 20:106)
Pig_da <- select(LR_all, 1,3, 20:88, 94:95, 101:104, 108:109, 204)
Pig_totals <- select(LR_all, 107:111)
Redox <- select(LR_all, 8:19)
Dia <- select(LR_all, 112:202)
Dia_da <- select(LR_all, 1,3, 112:202, 204)
PolZ <- select(LR_all, 204)
Dia_totals <- select(LR_all, 203)
#LowRes_XRF_raw <- select(LowRes_all, 211:241)
#LowRes_HSI <- select(LowRes_all, 14:20)

```

```{r}
PollenZPig <- data.frame("age" = c(15342, 14444, 13611, 13011, 12727, 12607.4), "zone" = c("Pleniglacial", "Bølling", "Allerød", "LST", "Younger Dryas", "")) #14993.4 (bottom age for depth plots)
PollenZPig <- LOOKUP(PollenZPig, c("age"), LR_all, c("mean_VT22"), "depth_VT22")
PollenZPig$depth <- PollenZPig$depth_VT22/1000

PollenZPig <- PollenZPig %>% mutate(color= case_when(zone == "Pleniglacial" ~ "blue",
                           zone == "Bølling" ~ "red",
                           zone == "Allerød" ~ "red", 
                          zone == "LST" ~ "lightgray", 
                           zone == "Younger Dryas" ~ "blue")
   )

PollenZPig <- PollenZPig %>% mutate(line= case_when(zone == "Pleniglacial" ~ "solid",
                           zone == "Bølling" ~ "solid",
                           zone == "Allerød" ~ "solid", 
                          zone == "LST" ~ "solid", 
                           zone == "Younger Dryas" ~ "solid")
   )

PollenZPig <- PollenZPig %>% mutate(aplha= case_when(zone == "Pleniglacial" ~ "0.8",
                           zone == "Bølling" ~ "0.8",
                           zone == "Allerød" ~ "0.8", 
                          zone == "LST" ~ "0.2", 
                           zone == "Younger Dryas" ~ "0.8")
   )
```

# Pigment data exploration

First we clean up only the pigment data, as the resolution of the other data is even lower (e.g., diatoms), we explore datasets one by one.

```{r}
Pig_da.nan<-na.omit(Pig_da)
#Pig.clean <- Pig.nan[which(colMeans(=0(Pig.nan)) >= .95)]

remove_high_zero_columns <- function(data) {
  # Calculate the proportion of zeros in each column
  prop_zeros <- colMeans(data == 0)
  # Keep columns with less than or equal to 90% zeros
  data[, prop_zeros <= 0.5]
}
# Remove columns with high zero proportion
Pig_da.clean <- as.data.frame(remove_high_zero_columns(Pig_da.nan))

Pig.clean <- select(Pig_da.clean, 3:22)
```

## K-means clustering

deciding on the number of clusters, with the elbow method

```{r}
fviz_nbclust(Pig.clean, kmeans, method = "wss")
```

Plotting the selected number of clusters on top of the plot

```{r}
# To save the plot, you can uncomment the first and the last line of this section 
#png(filename = "No_clusters_K-means.png", width = 500, height = 500, units = "px", pointsize = 12) 

fviz_nbclust(Pig.clean, kmeans, method = "wss") + geom_vline(xintercept = 3, linetype = 2)+ 
  #here you can add the number of clusters to be shown in with a line 
  labs(subtitle = "Elbow method")
#dev.off()
```

### K-means clustering

```{r}
#make this example reproducible 
set.seed(100) 
#perform k-means clustering with k = 5 clusters 
km <- kmeans(Pig.clean, centers = 3, nstart = 25) 

#plot results of final k-means model 

#To save the plot, you can uncomment the first and the last line of this section 
#png(filename = "K-means_clusters.png", width = 500, height = 500, units = "px", pointsize = 12) 
fviz_cluster(km, data = Pig.clean) 
#dev.off()
```

```{r}
#LR 
cluster_assignments <- as.data.frame(km$cluster) 
colnames(cluster_assignments) <- 'K_means_cluster'
df_clustered <- cbind(Pig_da.clean, cluster_assignments) 
 #this may need to be changed depending on the number of pigments
```

## Hierarchical unconstrained clustering

Standardizing the data for clustering

```{r}
data <- as.matrix(Pig.clean) 
# Standardise data for CLUSTER analysis, z-scoring
data_standardized <- data 
for(i in 1:ncol(data_standardized) ) {
  data_standardized[ , i] <- data_standardized[,i]/sd(data_standardized[,i]) 
  }
```

```{r}
# Compute hierarchical clustering 
res.hc <- data_standardized %>% scale() %>% 
# Scale the data 
dist(method = "euclidean") %>% 
  # Compute dissimilarity matrix 
  hclust(method = "ward.D2")
# Compute hierachical clustering
```

Visualize the clusters using dendrogram

```{r}
# Visualize using factoextra 
# Cut in x groups and color by groups 

#To save the plot, you can uncomment the first and the last line of this section 
# png(filename = "~/work/holzmaar/figs/Pigments_Hclust_dendrogram.png", width = 800, height = 500, units = "px", pointsize = 12) 

fviz_dend(res.hc, k = 3, # Cut in four groups 
          cex = 0.5, # label size 
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07","#6e57d2" ), color_labels_by_k = TRUE, # color labels by groups 
          rect = TRUE # Add rectangle around groups 
          )
# dev.off()
```

Save the clusters to your dataset

```{r}
#LR 
clust <- hclust(dist(data_standardized, method="euclidean"), method="ward.D2") 
clustercut<-cbind(1:nrow(data_standardized),cutree(clust,3)) 
colnames(clustercut) <- c('order','Hcluster')
df_clustered <- bind_cols(df_clustered, clustercut) 
```

## Hierarchical constrained clustering CONISS

Standardize data before CONISS clustering

```{r}
data <- as.matrix(Pig.clean) 
# Standardise data for CLUSTER analysis 
data_standardized <- data 
for(i in 1:ncol(data_standardized) ) { 
  data_standardized[ , i] <- data_standardized[,i]/sd(data_standardized[,i]) 
  }
```

### Compute hierarchical clustering

```{r}
bstick(chclust(dist(data_standardized, method="euclidean"), method="coniss"), 15)
```

### Cluster

```{r}
#Dendrogram classic 
clust <- chclust(dist(data_standardized, method="euclidean"), method="coniss") 
#plot(clust, hang=-1, horiz=FALSE, x.rev=TRUE, cex= 0.8)

```

### Visualize dendrogram

```{r}
plot(clust, hang=-1, horiz=FALSE, x.rev=TRUE, cex= 0.8) 
# Number of clusters 
k <- 4
groups <- cutree(clust, k=k) 
rect.hclust(clust, k=k, border="red") 
box(lty="solid")
data.clust<-cbind(data,groups) 
t <- as.data.frame(groups)

```

```{r}
# Visualize using factoextra 
# Cut in 4 groups and color by groups 
#quartz() 

#To save the plot, you can uncomment the first and the last line of this section 
# png(filename = "~/work/holzmaar/figs/Pigments_CONISS_dendrogram.png", width = 800, height = 500, units = "px", pointsize = 12) 

fviz_dend(clust, k = 4, # Cut in four groups 
          cex = 0.5, # label size 
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07","#6e57d2", "yellow", "green", "black"), color_labels_by_k = TRUE, # color labels by groups 
          rect = TRUE # Add rectangle around groups 
          )

# dev.off()
```

### Visualise the CONISS clusters on the age-plotted pigment concetrations

```{r}
#Define colour scheme
p.col <- c(rep("forestgreen", times=4), rep("gold2", times=18), rep("blue", times=39))
# Define y axis 
y.scale <- Pig_da.clean$mean_VT22


# Cluster analysis

ma.dist <- vegdist(Pig.clean, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")
# png(filename = "~/work/holzmaar/figs/Pigment_CONISS_plot2.png", width = 1200, height = 800, units = "px", pointsize = 12)

polHZM.plot <- strat.plot(Pig.clean, yvar=y.scale, y.tks=y.scale,  ylim= c(11625, 15342), y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", #col.poly=p.col,
col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar pigments LG Samples", ylabel="Sample", clust=ma.chclust)

# Add cluster zones
addClustZone(polHZM.plot, ma.chclust, nZone=5, lwd=1.5, lty=2, col="grey25")
# # 
# dev.off()
```

Save the clusters to your dataset

```{r}
#pigment data 
cluster <- chclust(dist(data_standardized, method="euclidean"), method="coniss") 
clustercut<-cbind(1:nrow(data_standardized),cutree(cluster,4)) 
colnames(clustercut) <- c('orderCONISS','CONISScluster')
df_clustered <- bind_cols(df_clustered, clustercut) 
```

Get the file with assigned clusters for further plotting

```{r}
# Get a table with the content of each cluster 
#write.table(clustercut, "Pig_unconstrained_table_with_clusters_3.txt", sep="\t") # 
# write.xlsx(HR_cluster_all, "data/HR_all_clustered.xlsx", rowNames = FALSE) 
#low resolution
write.xlsx(df_clustered, "../data/cleaned_PigOMug_clustered.xlsx", rowNames = FALSE, col_type ="numeric")
```

# Principle component analysis

The next step in exploring our XRF data we perform principle component analysis to reduce the dimensionality of the dataset and to explore the correlation and variance of our variables. In the first step we aim to keep the covariance of the variables, this, we do not perform further data standardization (e.g., applying z-score normalization) as our data are already normalized via clr.

Load the XRF data

```{r}
PCA_pig <- read_excel("../data/cleaned_PigOMug_clustered.xlsx")

data <- select(PCA_pig, 3:22)
```

## ALR: Transforming pigment data - Additive log ratio

As compositional data are unidirectional, means that they have values between 0 and infinite, for statistical operations, we first need to tranform the data. For this we use alr.

In this way, we can then conduct PCA or clustering without worrying that our PCA axes will be affected by the unidirectional data character.

```{r}
# Function to replace zeros with half of minimum non-zero value
replace_zeros <- function(x) {
  min_value <- min(x[x != 0])  # Find minimum non-zero value
  if (is.na(min_value)) min_value <- 0  # Handle case with all zeros
  x[x == 0] <- min_value / 2  # Replace zeros with half of minimum
  return(x)
}

# Apply the function to each column of the data frame - replacing 0es
data <- as.data.frame(sapply(data, replace_zeros))
```

```{r}
prep <- function(x) {
  # Calculate the alr
  norm <- x / (10^6 - x) # we uses 10^6 because our units are ug/g
  # Normalize each value
}

prep_data <- as.data.frame(apply(data, 1, prep))
trans_data <- as.data.frame(t(prep_data))
alr <- log(trans_data)

alr_safe<- bind_cols("depth" = PCA_pig$comp_depth_mm, "age" = PCA_pig$mean_VT22, alr)

write.xlsx(alr_safe, "../data/pig_alr.xlsx", rowNames = FALSE, col_type ="numeric")
```

## Perform PCA - alr data

We perform the PCA on the clr normalized dataset and display the percentage of explained variance by each of the PC axis.

```{r}
detach("package:factoextra", unload = TRUE)
detach("package:FactoMineR", unload = TRUE)
library("FactoMineR")
library("factoextra")
res.pca <- PCA(alr, graph = F, scale.unit = F)
var <- get_pca_var(res.pca)
get_eig(res.pca)


png(filename = timestamp_filename("../figs", "LR_PCA-alr_explanatory_power"), width = 800, height = 500, units = "px", pointsize = 12)

fviz_screeplot(res.pca, addlabels = TRUE, ylim = c(0, 65))

dev.off()
```

We see that most of the data variance is explained by the first PC (55.3%) followed by the PC2 explaining 10.2% of the data variance. To be able to interpret our data and decrease the dimensionality by selection representative variable for a certain phenomenon or trend, we visualize the data in a bi-plot figure. We already added preliminary interpretations based on expert knowledge.

### Plot PCA bi-plot

-   Positively correlated variables are grouped together.

-   Negatively correlated variables are positioned on opposite sides of the plot origin (opposed quadrants).

-   The distance between variables and the origin measures the quality of the variables on the factor map. Variables that are away from the origin are well represented on the factor map.

```{r}
png(filename = timestamp_filename("../figs", "PCA_pig_PC1-2"), width = 800, height = 500, units = "px", pointsize = 12)

# autochtonous <- text_grob(Autochtonous, size=13, face=bold, color= lightblue)
# allochtonous <- text_grob(Allochtonous, size=13, face=bold, color= lightblue)
# Organic <- text_grob(Organic matter\nand redox, size=13, face=bold, color= grey,rot = 90)
# Mineral <- text_grob(Minerals, size=13, face=bold, color= grey, rot=90)

#PCA_clr <-
  fviz_pca_var(res.pca, col.var="contrib", axes = c(1,2), 
             ylim=c(-1,1), xlim = c(-0.7, 1.25), clip = ("off"),
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title = " ")#+
        #theme(plot.margin = unit(c(1,1,2,1), lines)) +
          # annotation_custom(allochtonous,xmin=-0.4,xmax=-0.4,ymin=-0.7,ymax=-0.7) + 
          # annotation_custom(autochtonous,xmin=0.7,xmax=0.7,ymin=-0.7,ymax=-0.7) +
          # annotation_custom(Organic,xmin=-0.7, xmax=-0.7,ymin=0.4,ymax=0.4) + 
          # annotation_custom(Mineral,xmin=-0.7,xmax=-0.7,ymin=-0.4,ymax=-0.4)
  #coord_cartesian(clip = off)

dev.off()
```

### Correlation of the variables with the principle components

We can further explore the correlation of the variable with the PC axis to get better understanding of what variables are exploring each axis.

```{r}
corrplot(var$cor, is.corr=FALSE)
```

### Plot the variables contribution to PCx

Further to statistically constrain/explore the contribution of the variable to the PC axis we plot the PC axis histograms.

#### Contributions of variables to PC1

```{r}
png(filename = timestamp_filename("../figs", "pig_PCA-alr_PC1_exp"), width = 800, height = 500, units = "px", pointsize = 12)
fviz_contrib(res.pca, choice = "var", axes = 1, top = 12)
dev.off()
```

#### Contributions of variables to PC2

```{r}
png(filename = timestamp_filename("../figs", "pig_PCA-alr_PC2_exp"), width = 800, height = 500, units = "px", pointsize = 12)
fviz_contrib(res.pca, choice = "var", axes = 2, top = 10)
dev.off()
```

#### Contributions of variables to PC3

```{r}
png(filename = timestamp_filename("../figs", "pig_PCA-alr_PC3_exp"), width = 800, height = 500, units = "px", pointsize = 12)
fviz_contrib(res.pca, choice ="var", axes = 3, top = 10)
dev.off()
```

## PCA on alr and z-scored

Even though our variables are unit less and were normalized in the way that they are ratios (clr), in the previous PCA we observe that certain variables are dominant compare to others. Thus, some of the relationships of the less represented variables are not clearly visible. To resolve this overpowering effect of the few dominant variables and the make the variables comparable, we scale the data using the z-score approach (standardizing by the standard deviation of the dataset).

```{r}
res.pca_scaled <- PCA(alr,  graph = FALSE, scale = 1) # in case one wants to scale the data using z-score
varS<- get_pca_var(res.pca_scaled)
get_eig(res.pca_scaled)
png(filename = "pig_PCA_explanatory_power_alr_scored.png", width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(res.pca_scaled, addlabels = TRUE, ylim = c(0, 55))
dev.off()
```

We can see that the PC1 axis has now lower percentage of the explained variances (45.3%), which brought more explained variance to the PC2 (11.5). In this way we get more explanatory power on PC2 which means for us that we can identify more than the one dominant process explaining the data variance (as it was in the first PCA of non-scaled data). Thus, we get more explanatory PCA when z-scoring the data which helps to our data interpretation.

### PCA on scaled data

-   Positively correlated variables are grouped together.

-   Negatively correlated variables are positioned on opposite sides of the plot origin (opposed quadrants).

-   The distance between variables and the origin measures the quality of the variables on the factor map. Variables that are away from the origin are well represented on the factor map.

```{r}
# autochtonous <- text_grob(Autochtonous, size=13, face=bold, color= lightblue)
# allochtonous <- text_grob(Allochtonous, size=13, face=bold, color= lightblue)
# Organic <- text_grob(Organic matter\nand redox, size=13, face=bold, color= grey,rot = 90)
# Mineral <- text_grob(Minerals, size=13, face=bold, color= grey, rot=90)

#PCA_clr_scaled <- 
#png(filename = timestamp_filename("../figs", "pig_PCA-alr_scaled_PC1-2"), width = 700, height = 500, units = "px", pointsize = 16)
  fviz_pca_var(res.pca_scaled, label = "var", axes = c(2,3), 
            gradient.cols = c("darkred", "orange", "forestgreen"), col.var="contrib",
            repel = TRUE, title = "" )+
   #theme(plot.margin = unit(c(0.1,8,0.1,2), "lines")) +
          # annotation_custom(autochtonous,xmin=-.5,xmax=-0.5,ymin=-1,ymax=-1) + 
          # annotation_custom(allochtonous,xmin=0.5,xmax=0.5,ymin=-1,ymax=-1) +
          # annotation_custom(Organic,xmin=-1, xmax=-1,ymin=0.7,ymax=0.7) + 
          # annotation_custom(Mineral,xmin=-1,xmax=-1,ymin=-0.5,ymax=-0.5)+
  coord_cartesian(clip = "off")
#dev.off()


```

### Correlation of the variables with the principle components - scaled

```{r}
corrplot(varS$cor, is.corr=FALSE)
```

### Plot the variables contribution to PCx

#### Contributions of variables to PC1

```{r}
fviz_contrib(res.pca_scaled, choice = "var", axes = 1, top = 15)
```

#### Contributions of variables to PC2

```{r}
fviz_contrib(res.pca_scaled, choice = "var", axes = 2, top = 10)
```

#### Contributions of variables to PC3

```{r}
fviz_contrib(res.pca_scaled, choice = "var", axes = 3, top = 10)
```

## PCA on alr data with Pollen Zones

For further data exploration and explanation we combine the results of clustering with the PCA bi-plots. In this way, we see what the important variables defining the specific lithotypes (the clusters).

```{r}
res.pca <- prcomp(alr, scale =F)
#res.pca <- PCA(df, scale = 0)
#var <- get_pca_var(res.pca)
#get_eig(res.pca)

# autochtonous <- text_grob(Autochtonous, size=13, face=bold, color= lightblue)
# allochtonous <- text_grob(Allochtonous, size=13, face=bold, color= lightblue) 
# Organic <- text_grob(Organic matter\nand redox, size=13, face=bold, color= grey,rot = 90) 
# Mineral <- text_grob(Minerals, size=13, face=bold, color= grey, rot=90) 


b_clr <- fviz_pca_biplot(res.pca, label="var", axes = c(1,3), repel = TRUE, 
                     title =" " , col.var = "black", #xlab = PC1 (55.3%), ylab=PC2 (10.2%),
                     alpha.var = 1, alpha.ind = 0.5, habillage=PCA_pig$PolZ,
               addEllipses=TRUE, ellipse.level=0.95, ellipse.alpha = 0.2) +
  labs(color = "Clusters", shape ="Clusters", fill = "Clusters")+
  # annotation_custom(autochtonous,xmin=-3,xmax=-3,ymin=-4,ymax=-4) + 
  # annotation_custom(allochtonous,xmin=3,xmax=3,ymin=-4,ymax=-4) + 
  # annotation_custom(Organic,xmin=-5, xmax=-5,ymin=-2,ymax=-2) + 
  # annotation_custom(Mineral,xmin=-5,xmax=-5,ymin=3,ymax=3)+
  theme(plot.margin = unit(c(0.1,8,0.1,2), "lines"))#+
   #coord_cartesian(xlim = c(-5, 5), ylim = c(-4, 4), clip = "off")



png(filename = timestamp_filename("../figs", "pig_PCA-alr_PolZ_PC1-3"), width = 700, height = 500, units = "px", pointsize = 16)
b_clr + scale_color_viridis_d(option ="H") + 
  scale_fill_viridis_d(option = "H")+
     theme_minimal()

dev.off()
```

## PCA on alr z-scored data with Pollen Zones

```{r}
res.pca_scaled <- prcomp(alr, scale =T)

# autochtonous <- text_grob(Autochtonous, size=13, face=bold, color= lightblue)
# allochtonous <- text_grob(Allochtonous, size=13, face=bold, color= lightblue) 
# Organic <- text_grob(Organic matter\nand redox, size=13, face=bold, color= grey,rot = 90) 
# Mineral <- text_grob(Minerals, size=13, face=bold, color= grey, rot=90) 


b_z <- fviz_pca_biplot(res.pca_scaled, label="var", axes = c(1,3), repel = TRUE, 
                     title =" " , col.var = "black", #xlab = PC1 (45.3%), ylab=PC2 (11.5%),
                     alpha.var = 1, alpha.ind = 0.5, habillage=PCA_pig$PolZ,
               #cluster_CONISS, #K_means_cluster, Hcluster
               addEllipses=TRUE, ellipse.level=0.95, ellipse.alpha = 0.2) +
  labs(color = "Clusters", shape ="Clusters", fill = "Clusters")+
  # annotation_custom(autochtonous,xmin=-5,xmax=-5,ymin=-6,ymax=-6) + 
  # annotation_custom(allochtonous,xmin=5,xmax=5,ymin=-6,ymax=-6) + 
  # annotation_custom(Organic,xmin=-8, xmax=-8,ymin=-4,ymax=-4) + 
  # annotation_custom(Mineral,xmin=-8,xmax=-8,ymin=4,ymax=4)+
  theme(plot.margin = unit(c(0.1,8,0.1,2), "lines"))#+
  #coord_cartesian(xlim = c(-8, 8), ylim = c(-6, 6), clip = "off")


png(filename = timestamp_filename("../figs", "pig_PCA-alr_scaled_PolZ_PC1-3"), width = 700, height = 500, units = "px", pointsize = 16)

b_z + scale_color_viridis_d(option ="H") + 
  scale_fill_viridis_d(option = "H")+
     theme_minimal()

dev.off()

```

Extracting the 4 PCs

One the PCA axis are well explored and the phenomena responsible to the data variation are explained, the PCA axis can be used as the reduced dataset replacing the original variables when representing the given/selected phenomena. Thus, we extract the PC axis scores and save them along with our data for further plotting or data analysis.

```{r}
#summary(res.pca)
res.pca <- prcomp(alr, scale. =T, rank. = 4) #center = T,
summary(res.pca)
```

Attaching the PC scores to the dataset

```{r}
# Extract principal component scores (first two components)
pc_scores <- res.pca$x  # Selects columns 1 and 2 for the first two PCs

# Extract loadings for all principal components
pc_loadings <- res.pca$rotation

df_PC <- bind_cols(PCA_pig, pc_scores)

write.xlsx(df_PC, "../data/Pig_clustered_PC.xlsx", rowNames = FALSE, col_type ="numeric")
```

## Trajectory plots

The sedimentary data can be preceived as time-series data as the process of settlement is happening though time. Thus, while using the CONISS clustering, which groups the sediment data with respect to the depth/time variable, the PC axis can be used to plot the data and be able to track the chronological development of the variables represented by the PC scores. These plots are here called trajectory and they represent the evolution of the sediment composition though depth/time.

```{r}
df_PC <- read.xlsx("../data/Pig_clustered_PC.xlsx")
depth_m <- as.data.frame(df_PC$comp_depth_mm)

df_PC <- df_PC%>% mutate(PZn = case_when(PolZ == "Pleniglacial" ~ "1",
                           PolZ == "Bølling" ~ "2",
                           PolZ == "Allerød" ~ "3",
                           PolZ == "YoungerDryas" ~ "4")
   )
```

```{r}
#plot(df_PC$PC1, df_PC$PC2, type= l, text = df_PC$comp_depth_mm)
End <- df_PC %>% # last trading day
  slice_max(df_PC$comp_depth_mm)
 
Start <- df_PC %>% # first trading day
  slice_min(df_PC$comp_depth_mm)
  #slice(which.min(depth_m))


s <- ggplot(df_PC, aes(PC1, PC3)) + 
  geom_point(aes(colour = factor(PZn),  shape = factor(PZn), alpha = 0.3))+
  labs(alpha = "Trajectory")+
 #geom_line()
  geom_path(alpha=0.3, col = "black", arrow = arrow(type = "closed", length = unit(0.25, "inches"), ends = "first")) +
  labs(colour = "PZn", shape = "PZn") +
# last trading day - round mark, annotation below
  geom_point(data = Start, aes(x = PC1, y = PC3), col = "green", shape = 21, fill = "green", size = 2, stroke = 1.7) +
  geom_text(data = Start, aes(x = PC1, y = PC3, label = sprintf("%0.0f", round(Start$comp_depth_mm, digits = 0))), size = 4, vjust = 2.5) +
  # first trading day - round mark, annotation above
  geom_point(data = End, aes(x = PC1, y = PC3), col = "red", shape = 21, fill = "red", size = 2, stroke = 1.7) +
  geom_text(data = End, aes(x = PC1, y = PC3, label = sprintf("%0.0f", round(End$comp_depth_mm, digits = 0))), size = 4, face= "bold", vjust = -1.5)



traj <- s + scale_color_brewer(palette="Dark2", labels = c("Pleniglacial", "Bølling-Allerød", "Allerød", "Younger Dryas"))+
  scale_shape_discrete(labels = c("Pleniglacial", "Bølling-Allerød", "Allerød", "Younger Dryas")) +#Paired
     theme_minimal()+
  # annotation_custom(autochtonous,xmin=-4,xmax=-4,ymin=-6,ymax=-6) + 
  # annotation_custom(allochtonous,xmin=5,xmax=5,ymin=-6,ymax=-6) + 
  # annotation_custom(Organic,xmin=-7, xmax=-7,ymin=-3,ymax=-3) + 
  # annotation_custom(Mineral,xmin=-7,xmax=-7,ymin=5,ymax=5)+
 # theme(plot.margin = unit(c(0.1,8,0.1,2), lines))+
  coord_cartesian(xlim = c(-10, 10), ylim = c(-8, 8), clip = "off")


png(filename = "../figs/pig_trajectory_PC1-3.png", width = 680, height = 600, units = "px", pointsize = 14, res=120)
traj
dev.off()
```

PCAs are not really explanatory, thus we try to apply multidimensional scaling (MDS) instead and correlation matrix or heat maps can be better approach.

# MDS

```{r}
mds_pig <- cmdscale(dist(alr), k=3)
mds_df <- as.data.frame(mds_pig)
df_PC <- bind_cols(df_PC, mds_df)
```

```{r}
plot(mds_pig[, 1], mds_pig[, 2])
points(mds_pig[, 1], mds_pig[, 2], pch =21, bg = df_PC$PZn)# here the color of the datapoints can be changed to show any of the made clusterings or the pollen zones
```

For the MSD visualization and interpretability, we use the pollen zones as the shape and color of our datapoints.

```{r}
png(filename = "../figs/pig_MDS_PolZ-1-3.png", width = 680, height = 600, units = "px", pointsize = 14, res=120)

ggscatter(df_PC, x = "V1", y = "V3",
          color = "PZn", shape = "PZn", fill = "PZn",
          palette = "jco",
          size = 3,
          ellipse = TRUE,
          ellipse.type = "convex",
          title = " ",
          xlab = "MDS Dimension 1",
          ylab = "MDS Dimension 3") +
    labs(color = "Pollen Zones", shape ="Pollen Zones", fill = "Pollen Zones") +
  scale_color_discrete(name = 'Pollen Zones', labels = c("Pleniglacial","Bølling","Allerød","Younder Dryas"))+
  scale_fill_discrete(name = 'Pollen Zones', labels = c("Pleniglacial","Bølling","Allerød","Younder Dryas"))+
  scale_shape_discrete(name = 'Pollen Zones', labels = c("Pleniglacial","Bølling","Allerød","Younder Dryas"))
  #geom_text_repel(aes(label = species), box.padding = 0.5)

dev.off()
```

The MDS showed us that the Pleniglacial data are different than the Bölling-Alleröd and Younger Dryas data. Likely we don't have sufficent number of data in the Younger Dryas to see whether the lake pigment composition returns to the same communities as in the Pleniglacial.

```{r}
# colors <- c("red", "blue", "pink")
# colors <- colors[as.numeric(df_PC$PZn)]
# scatterplot3d(mds_pig[,1:3], pch = 16,
#               xlab = "",
#               ylab = "",
#               zlab = "",
#               color=colors)
```

# Non-metric multidimensional scaling (NMDS)

Non-metric Multidimensional Scaling (NMDS) is a great tool for ecologists to understand community assemblages and identify their influencing factors. We use a data frame/matrix with relative abundances of different pigments in different time periods. Each pigment abundance in the time is an axis and each axis is a dimension. NMDS summarise all of that information into a 2-dimensional representation, showing the differences in pigment community composition. It tries to represent the original position of a community within a multidimensional space as accurately as possible while also minimising the number of dimensions to easily plot and visualise.

## Relative abundance of pigments (%)

Converting the absolute concentrations of pigments to relative %.

```{r}
community <- df_PC[,2:22]
commper <- adorn_percentages(community, denominator = "row", na.rm = TRUE, 2:21)
comPer <- commper[2:21]*100
```

Transforming the data frame into a matrix and performing the NMDS.

```{r}

ComMatrix <- as.matrix(comPer)

inv.NMDS <- metaMDS(ComMatrix, distance = "bray", k = 3, autotransform = TRUE, trymax=100)

# check stress value
inv.NMDS$stress # ideally stress value is < 0.2

plot(inv.NMDS)

#ordiplot(inv.NMDS, type = "n") # create blank ordination plot
orditorp(inv.NMDS, display = "species", col="red", air = 0.4) # add species (orders in this case) names in red
orditorp(inv.NMDS, display = "sites", cex = 1.25, air = 0.4) # add site numbers in black
```

The result plot shows 2 dimensions and each dot represents a different sample in time and each cross represents pigment. The closer 2 points are, the more similar those communities are in terms of composition.

```{r}
#png("~/work/holzmaar/figs/NMDS.png", width=6, height=5, units = "in", res = 300)
ordiplot(inv.NMDS) # plot shows communities (circles) and species (crosses)
orditorp(inv.NMDS, display = "species", col="red", air = 0.4)
orditorp(inv.NMDS, display = "sites", cex = 0.7, air = 0.4)
ordiellipse(inv.NMDS, df_PC$PZn, label = FALSE,
            col=c("dodgerblue", "tomato2", "tomato4", "deepskyblue"),
            draw = "polygon", alpha=120) # adding ellipses to the plot, grouping by distance (inverts$Distance)
legend("bottomleft", title="Pollen Zone", c("Pleniglacial","Bølling","Allerød","Younger Dryas"), 
       fill=c("dodgerblue", "tomato2", "tomato4",
                                   "deepskyblue"), horiz=FALSE, cex=.9) # adding a legend
#dev.off()
```

```{r}
nmds.scores <- as.data.frame(scores(inv.NMDS)$sites)

nmds.scores <- nmds.scores %>%
  mutate(Time = as.factor(df_PC$PZn))

png(filename = "../figs/pig_NMDS_1-3.png", width = 680, height = 600, units = "px", pointsize = 14, res=120)
# create ggplot
(inv_NMDS_plot <- ggplot(data = nmds.scores, aes(NMDS1, NMDS3)) +
# adding different colours and shapes for points at different distances
    geom_point(aes(color = Time, shape = Time)) + 
# adding covariance ellipses according to distance # use size argument if ggplot2 < v. 3.4.0
   # geom_path(data=nmds.scores, aes(x=NMDS1, y=NMDS2, colour=Time), linewidth=1) + 
# removes lines from legend
    guides(color = guide_legend(override.aes = list(linetype=c(NA,NA,NA,NA)))) +
    theme_bw() + 
    theme(panel.grid = element_blank()) +
# legend title
    scale_color_manual(name = "Pollen Zone", 
    # adjusting legend labels and colour
            labels = c("Pleniglacial","Bølling","Allerød","Younger Dryas"), 
            values = c("dodgerblue", "tomato2", "tomato4", "deepskyblue")) + 
    scale_shape_manual("Pollen Zone", 
            labels = c("Pleniglacial","Bølling","Allerød","Younger Dryas"), 
            values = c(17, 15, 3, 7))) # customising shapes
# save plot
#ggsave(filename = "~/work/holzmaar/figs/NMDSPaths.png", inv_NMDS_plot, device = "png")

dev.off()
```

Similarly to the MDS, we get information about the big difference of the Pleniglacial communities compared to the rest of the Late-Glacial periods.

# Correlation based heatmaps

To best present the pigment data, we need to decrees their dimensionality. We tried to use PCA, where we saw groups of pigments representing the community composition, but to visualize these communities on a time axis, we opt for correlation based heat maps.

```{r}
pig_prod <- read_excel("../data/SourceData/pig_prod.xlsx", sheet = "50%0")
#heet = "90%0T")
pig_prod_df <- as.data.frame(pig_prod)
rownames(pig_prod_df) <- pig_prod_df[,2]
pig_prod_df

col_names <- pig_prod_df[,2]
```

In case the scirpt starts here, we load all the data from the code above

```{r}
df_PC <- read.xlsx("../data/Pig_clustered_PC.xlsx")
depth_m <- as.data.frame(df_PC$comp_depth_mm)

df_PC <- df_PC%>% mutate(PZn = case_when(PolZ == "Pleniglacial" ~ "1",
                           PolZ == "Bølling" ~ "2",
                           PolZ == "Allerød" ~ "3",
                           PolZ == "YoungerDryas" ~ "4")
   )
```

```{r}
alr_safe <- read.xlsx("../data/pig_alr.xlsx")
```

```{r}
pig_prod_df <- pig_prod_df[,-1]
annotation_col = pig_prod_df[2]
```

```{r}
depth_mat <- as.data.frame(select(df_PC,1, 23, 2))

rownames(depth_mat) <- depth_mat[,1]
depth_mat <- depth_mat[,-1]
annotation_row = depth_mat[1]
```

```{r}
new_names <- alr_safe[,1]

alr_mat <- as.matrix(alr_safe[,3:22])
rownames(alr_mat) <- new_names
colnames(alr_mat) <- col_names

alr_scaled<- scale(alr_mat)

Pig.clean_ID <- as.matrix(Pig.clean)
rownames(Pig.clean_ID) <- new_names
colnames(Pig.clean_ID) <- col_names
```

```{r}
ann_colors = list(
   producer = c(Cryptophytes = "darkorange3", Cyanobacteria ="royalblue2", `Green algae` = "green4", `Green sulphur bacteria` = "greenyellow", `Purple non-sulphur bacteria`= "orchid2", `Purple sulphur bacteria` = "purple2", `Red blooming algae` = "red4", Silicifiers= "lightskyblue"),
    PolZ = c(Pleniglacial = "dodgerblue", Bølling = "tomato2", Allerød = "tomato4", `YoungerDryas` = "deepskyblue")) #`Lachersee tephra` = "gray29"
  #  GeneClass = c(Path1 = "#7570B3", Path2 = "#E7298A", Path3 = "#66A61E")
# , 

#Both dimension clustering
#png(filename = "~/work/holzmaar/figs/heatmap_corr_clustered_50_3percent.png", width = 1200, height = 1000, units = "px", pointsize = 18)
pheatmap(alr_scaled, show_rownames=T,show_colnames=T,
         annotation_col= annotation_col, annotation_row = annotation_row,
         annotation_colors = ann_colors,
         scale = "none",clustering_method="ward.D2", #cluster_rows = F,
         clustering_distance_cols="correlation", clustering_distance_rows = "euclidean",
         cutree_rows = 3,
         cutree_cols = 4, treeheight_row = 50,
         treeheight_col = 50, color = colorRampPalette(c("lightskyblue1","darkseagreen1","lightgreen","yellow", "orange", "red", "purple"))(15))
#dev.off()


# Only pigment clustering

obj <- pheatmap(alr_scaled, show_rownames=T,show_colnames=T,
         annotation_col=annotation_col, annotation_row = annotation_row,
         annotation_colors = ann_colors,
         fontsize = 24,
         fontsize_row = 10,
        #cellheight = 8, cellwidth = 20,
         scale = "none", #clustering_method="ward.D2",
         cluster_rows = F,
         clustering_distance_cols="euclidean", #clustering_distance_rows = "euclidean",
        # cutree_rows = 5,
         cutree_cols = 3, #treeheight_row = 50,
         treeheight_col = 50, color = colorRampPalette(c("lightskyblue1","darkseagreen1","lightgreen","yellow", "orange", "red", "purple"))(15))

png(filename = timestamp_filename("../figs", "SuppFigure3_heatmap_corr_50_3percent"), width = 1600, height = 1000, units = "px", pointsize = 24)
obj
dev.off()



```

Extract the clusters, make mean curve and use for a line plot

```{r}
obj$tree_row
obj$tree_col
PigClust <- cutree(obj$tree_col, k = 3) #fill in the number of observations (rows) and number of clusters acooriding to the dendrogram 
# df_clustered <- bind_cols(df_clustered, clustercut) 
# colnames(df_clustered)[29] <- 'order' # change the column number based on your dataset size 
# colnames(df_clustered)[30] <- 'Hcluster' # change the column number based on your dataset size

PigClust <- as.data.frame(t(PigClust))

alr_scaled <- as.data.frame(alr_scaled)
alr_scaledB<- rbind("PigClust" = PigClust , alr_scaled)
```

```{r}
df1 <- as.data.frame(t(alr_scaledB))

selection1 <- as.data.frame(t(df1[df1$PigClust == "1", ]))
selection2 <- as.data.frame(t(df1[df1$PigClust == "2", ]))
selection3 <- as.data.frame(t(df1[df1$PigClust == "3", ]))
```

```{r}
alr_scaled$C1 <- rowMeans(selection1[2:95,])
alr_scaled$C2 <- rowMeans(selection2[2:95,])
alr_scaled$C3 <- rowMeans(selection3[2:95,])


alr_scaled_clust<- bind_cols("depth" = PCA_pig$comp_depth_mm, "age" = PCA_pig$mean_VT22, alr_scaled)

write.xlsx(alr_scaled_clust, "../data/pig_alr_clust.xlsx", rowNames = FALSE, col_type ="numeric")
```

# Figure 2

```{r}
Pig_da.clean$TChl <- rowSums(Pig_da.clean[,20:22])
Pig_da.clean$Cyanos <- rowSums(Pig_da.clean[,11:13])

PollenZ <- select(Pig_da.clean, 1, 2, 23)

PollenZ <- PollenZ %>% mutate(color= case_when(PolZ == "Pleniglacial" ~ "blue",
                           PolZ == "Bølling" ~ "red",
                           PolZ == "Allerød" ~ "red", 
                          PolZ == "LST" ~ "lightgray", 
                           PolZ == "YoungerDryas" ~ "blue")
   )

PollenZ <- PollenZ %>% mutate(line= case_when(PolZ == "Pleniglacial" ~ "solid",
                           PolZ == "Bølling" ~ "solid",
                           PolZ == "Allerød" ~ "solid", 
                          PolZ == "LST" ~ "solid", 
                           PolZ == "YoungerDryas" ~ "solid")
   )

PollenZ <- PollenZ %>% mutate(aplha= case_when(PolZ == "Pleniglacial" ~ "0.8",
                           PolZ == "Bølling" ~ "0.8",
                           PolZ == "Allerød" ~ "0.8", 
                          PolZ == "LST" ~ "0.2", 
                           PolZ == "YoungerDryas" ~ "0.8")
   )
```

## High-resolution data

```{r}
HSI_data <- read_excel("../data/HZM_HR_cleaned.xlsx", col_types = "numeric")

PollenZ <- data.frame("age" = c(15342, 14444, 13611, 13011, 12727, 11625), "zone" = c("Pleniglacial", "Bølling", "Allerød", "LST", "Younger Dryas", ""))
PollenZ <- LOOKUP(PollenZ, c("age"), HSI_data, c("mean_VT22"), "depth_VT22")
PollenZ$depth <- PollenZ$depth_VT22/1000

PollenZ <- PollenZ %>% mutate(color= case_when(zone == "Pleniglacial" ~ "blue",
                           zone == "Bølling" ~ "red",
                           zone == "Allerød" ~ "red", 
                          zone == "LST" ~ "lightgray", 
                           zone == "Younger Dryas" ~ "blue")
   )

PollenZ <- PollenZ %>% mutate(line= case_when(zone == "Pleniglacial" ~ "solid",
                           zone == "Bølling" ~ "solid",
                           zone == "Allerød" ~ "solid", 
                          zone == "LST" ~ "solid", 
                           zone == "Younger Dryas" ~ "solid")
   )

PollenZ <- PollenZ %>% mutate(aplha= case_when(zone == "Pleniglacial" ~ "0.8",
                           zone == "Bølling" ~ "0.8",
                           zone == "Allerød" ~ "0.8", 
                          zone == "LST" ~ "0.2", 
                           zone == "Younger Dryas" ~ "0.8")
   )
```

```{r}
limYmin = min(HSI_data$mean_VT22) #HSI_data$depth_VT22
limYmax = max(HSI_data$mean_VT22) #HSI_data$depth_VT22

RABD620 <- ggplot()+
  geom_hline(data= PollenZPig, yintercept = PollenZPig$age,#depth_VT22, 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_hline(yintercept = PollenZ$age, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
   geom_vline(xintercept=mean(HSI_data$RABD620), color="dodgerblue1", linetype = 1) +
  geom_vline(xintercept=mean(HSI_data$RABD620) + sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_vline(xintercept=mean(HSI_data$RABD620) - sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD620), colour = "dodgerblue1", size= 0.5)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD620), alpha=.8, size=1.5, colour = "dodgerblue1") +
  #   annotate("text", x= 1.01, y= 15050 , label = "Pleniglacial", color= "blue")+ #x= 13700
  # annotate("text", x= 1.00, y= 14150, label = "Bølling", color= "darkred")+ #x= 12800
  # annotate("text", x= 1.00, y= 13550, label = "Allerød", color= "darkred")+ #x= 12.100
  # annotate("text", x= 1.01, y=12990, label = "Lachersee tephra", color= "black", size = 3)+ #x= 11600
  # annotate("text", x= 1.007, y= 12250, label = "Younger Dryas", color= "blue")+ #x= 11100
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  labs(title = " ", x = "RABD620\n(Phycocyanin)", y = "Age (Cal yr BP)" )+#y = " Depth (mm) ") +
  theme_classic() +
  theme(legend.position="none", axis.title.x.bottom = element_text(color = "dodgerblue1"), axis.text = element_text(size = 12), plot.margin = unit(c(0,0,0.5,1), "cm"), panel.background = element_blank())#axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), )

RABD660 <- ggplot()+
  geom_hline(data= PollenZPig, yintercept = PollenZPig$age,#depth_VT22, 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_hline(yintercept = PollenZ$age, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD670calib), colour = "limegreen", size= 0.5)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD670calib), alpha=.8, size=1.5, colour = "limegreen") +
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0))+
  labs(title = " ", x = "Total chlorophylls\n(µg/g wet sed)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.title.x.top = element_text(color = "limegreen"), axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.6), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

RABD845 <- ggplot()+
    geom_hline(data= PollenZPig, yintercept = PollenZPig$age,#depth_VT22, 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_hline(yintercept = PollenZ$age, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD845calib), colour = "violet", size= 0.5)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD845calib), alpha=.8, size=1.5, colour = "violet") +
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0))+
  labs(title = " ", x = "Bacteriopheophytin a\n(µg/g wet sed)", y = " ") +
  theme_classic() +
  theme(legend.position="none", axis.title.x.bottom = element_text(color = "violet"), axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-1), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())
```

```{r}
HIPlot <- plot_grid(RABD620,
                    RABD660, 
                    RABD845,
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1), axis = "tb" )
#dev.off()

png(filename = timestamp_filename("../figs", "HSIPlot"), width = 877, height = 620, units = "px", pointsize = 18, res=72)
HIPlot
dev.off()
```

### Combined, overlay plots

```{r}
Chlfs <- ggplot()+
  annotate("text", y= 580, x= 15050 , label = "Pleniglacial", color= "blue")+ #x= 13700
  annotate("text", y= 600, x= 14150, label = "Bølling", color= "darkred")+ #x= 12800
  annotate("text", y= 600, x= 13550, label = "Allerød", color= "darkred")+ #x= 12.100
  annotate("text", y= 520, x=12990, label = "Lachersee tephra", color= "black", size = 3)+ #x= 11600
  annotate("text", y= 500, x= 12550, label = "Younger Dryas", color= "blue")+ #x= 11100
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #Pig_da.clean$comp_depth_mm,
                                    y =Pig_da.clean$TChl), width = 20, stat = "identity", alpha = 0.6, fill = "palegreen4")+
  geom_point(data = HSI_data, aes(x=mean_VT22, #depth_VT22,
                                  y= RABD670calib*10), colour = "springgreen3", size= 0.01) +
  geom_path(data = HSI_data, aes(x= mean_VT22, #depth_VT22,
                                 y= RABD670calib*10), alpha=.5, size=0.2, colour = "springgreen3") +
  geom_vline(data= PollenZ, xintercept = PollenZ$age, #depth_VT,
             linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  coord_flip()+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0), name = "Age (Cal yr BP)") +
  scale_y_continuous(position = "right", expand=c(0,0), sec.axis = sec_axis(~ . /10, name = "Total chlorophylls\n(µg/g wet sed)"))+
  labs(title = " ", y = " Total chlorophylls\n(µg/g OM)", x = " ") +
  theme_classic() +
  theme(legend.position="none", #, axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) axis.text.y = element_blank(), 
        axis.title.x.bottom = element_text(color = "springgreen3"),  axis.title.x.top = element_text(color = "palegreen4"))


Bphe <- ggplot()+
  geom_vline(xintercept = PollenZ$age, #depth_VT,
             linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #Pig_da.clean$comp_depth_mm,
                                    y =Pig_da.clean$Okenone), width = 20, stat = "identity", alpha = 0.6, fill = "purple3")+
  geom_point(data = HSI_data, aes(x=mean_VT22, #depth_VT22
                                  y= RABD845calib*10), colour = "deeppink", size= 0.01) +
  geom_path(data = HSI_data, aes(x= mean_VT22, #depth_VT22
                                 y= RABD845calib*10), alpha=.5, size=0.2, colour = "deeppink") +
  coord_flip()+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0), sec.axis = sec_axis(~ . /10, name = "Bacteriophyophytin a\n(µg/g wet sed)"))+
  labs(title = " ", y = "Okenone\n(µg/g OM)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "deeppink"),  axis.title.x.top = element_text(color = "purple3"))
  
  # annotate("text", x= 1.015, y= 13.7, label = "Pleniglacial", color= "blue")+
  # annotate("text", x= 1.015, y= 12.8, label = "Bølling", color= "darkred")+
  # annotate("text", x= 1.015, y= 12.1, label = "Allerød", color= "darkred")+
  # annotate("text", x= 1.015, y= 11.6, label = "Lachersee tephra", color= "black")+
  # annotate("text", x= 1.015, y= 11.1, label = "Younger Dryas", color= "blue")+

Cyano <- ggplot()+
 # geom_text(aes(max(RABD620),PollenZ$depth,label = PollenZ$depth, vjust = -1))+
  geom_vline(xintercept = PollenZ$age, #depth_VT, 
             linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
  geom_hline(yintercept=mean(HSI_data$RABD620), color="dodgerblue1", linetype = 1) +
  geom_hline(yintercept=mean(HSI_data$RABD620) + sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_hline(yintercept=mean(HSI_data$RABD620) - sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_point(data= HSI_data, aes(x= mean_VT22, #depth_VT22
                                  y= RABD620), colour = "dodgerblue1", size= 0.01)+
  geom_path(data = HSI_data, aes(x= mean_VT22, #depth_VT22
                                 y= RABD620), alpha=.8, size=0.2, colour = "dodgerblue1") +
  coord_flip()+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
  labs(title = " ", y = "RABD620\n(Phycocyanin)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "dodgerblue1"))

```

```{r}
Combo <- plot_grid(Chlfs, Bphe, Cyano, 
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1), axis = "tb" )


png(filename = timestamp_filename("../figs", "ComboHSI"), width = 877, height = 620, units = "px", pointsize = 18, res=72)
Combo
dev.off()
```

## Pigment selection

```{r}

TChls <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22, 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #comp_depth_mm, 
                                    y =Pig_da.clean$TChl), width = 25, stat = "identity", alpha = 0.4, fill = "springgreen3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=TChl, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Total chlorophylls\n(µg/g OM) \n Green algae", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "springgreen3"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-0.65), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())
  # theme_classic() +
  # theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


# Oke <- ggplot()+
#   geom_hline(yintercept = PollenZ$depth_VT22, linetype = PollenZ$line, color= PollenZ$color, alpha = 0.8)+
#   geom_point(data=Pig_da.clean, aes(x=Pig_da.clean$Okenone, y=comp_depth_mm), colour = "purple", size= 0.01)+
#   scale_y_reverse(limits = c(13822, 10989), breaks = round(seq(10989, 13822, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
#   scale_x_continuous(position = "bottom", expand=c(0,0))+
#  geom_path(data=Pig_da.clean, aes(x=Okenone, y= comp_depth_mm), alpha=.8, size=0.2, colour = "purple") +
#   labs(title = " ", x = "Okenone \n(µg/g OM)", y = " ") +
#   annotate("text", x= 1.01, y= 13.7, label = "Pleniglacial", color= "blue")+
#   annotate("text", x= 1, y= 12.8, label = "Bølling", color= "darkred")+
#   annotate("text", x= 1, y= 12.1, label = "Allerød", color= "darkred")+
#   annotate("text", x= 1.005, y= 11.6, label = "Lachersee tephra", color= "black", size = 3)+
#   annotate("text", x= 1.01, y= 11.1, label = "Younger Dryas", color= "blue")+
#   theme_classic() +
#   theme(legend.position="none")
#   # theme_classic() +
#   # theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Oke <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 , 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #comp_depth_mm, 
                                      y =Pig_da.clean$Okenone), width = 25, stat = "identity", alpha = 0.4, fill = "maroon4")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Okenone\n(µg/g OM) \n Anoxic bacteria (PSB)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.top = element_text(color = "maroon4"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-1), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())


Fuco <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 ,
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #comp_depth_mm, 
                                    y =Pig_da.clean$Fuco), width = 25, stat = "identity", alpha = 0.4, fill = "darkorange3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Fucoxanthin\n(µg/g OM)\n Diatoms ", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange3"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-0.62), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())


Cyanos <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 ,
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, #comp_depth_mm, 
                                    y =Pig_da.clean$Cyanos), width = 25, stat = "identity", alpha = 0.4, fill = "lightseagreen")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Echi+Cantha+Zea\n(µg/g OM) \n Cynobacteria", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.top = element_text(color = "lightseagreen"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-.5), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

```

#### Pigment communities

```{r}
PigCom1 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 ,
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_col(data = alr_scaled_clust, aes(x=alr_scaled_clust$depth, y = alr_scaled_clust$C1), width = 25, stat = "identity", alpha = 0.4, fill = "royalblue1")+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C1), alpha=1, size=1.5, colour = "royalblue1") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Pigment community 1\n Pioneers \n relative abundance", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.top = element_text(color = "royalblue1"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-1), "cm"),axis.text = element_text(size = 12),panel.background = element_blank())

PigCom3 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 ,
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_col(data = alr_scaled_clust, aes(x=alr_scaled_clust$depth, y = alr_scaled_clust$C1), width = 25, stat = "identity", alpha = 0.4, fill = "royalblue1")+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C3), alpha=1, size=1.5, colour = "royalblue3") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Pigment community 2\n Secondary colonisers \n relative abundance", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "royalblue3"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-.9), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

PigCom2 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age,#depth_VT22 , 
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  # geom_col(data = alr_scaled_clust, aes(x=alr_scaled_clust$depth, y = alr_scaled_clust$C1), width = 25, stat = "identity", alpha = 0.4, fill = "royalblue1")+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C2), alpha=1, size=1.5, colour = "royalblue4") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Pigment community 3\n Diatoms+Bacteria \n relative abundance", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.top = element_text(color = "royalblue4"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-0.5), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())

```

```{r}
Pigments <- plot_grid(
  #RABD620, RABD660,  RABD845,  
  #Cyanos,TChls, Oke, 
  Chlfs, Bphe, Cyano, 
  PigCom1, PigCom3, PigCom2, 
  Oke, Fuco,
 # MnRedox,
  #Chlfs, Bphe, Cyano, 
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1,1,1,1,1,1,1), axis = "tb" )

png(filename = timestamp_filename("../figs", "Pigments"), width = 1754, height = 1240, units = "px", pointsize = 18, res=72)

Pigments

dev.off()

```

## Redox

```{r}
P_tot_pre <- as.data.frame(select(LR_all, 8:11))
P_23 <- as.data.frame(select(LR_all, 9:10))
#P_tot <- bind_cols(depth= LR_all$depth_VT22, P_tot_pre)
P_tot <- bind_cols(age= LR_all$mean_VT22, P_tot_pre)
#P_tot_23 <- bind_cols(depth= LR_all$depth_VT22, P_23)
P_tot_23 <- bind_cols(age= LR_all$mean_VT22, P_23)

P_stacked <- reshape2::melt(P_tot, var.names = names(P_tot)[2:ncol(P_tot)], measure.vars = names(P_tot)[2:ncol(P_tot)])

P_stacked23 <- reshape2::melt(P_tot_23, var.names = names(P_tot_23)[2:ncol(P_tot_23)], measure.vars = names(P_tot_23)[2:ncol(P_tot_23)])

Fe_tot_pre <- as.data.frame(select(LR_all, 16:19))
Fe_23 <- as.data.frame(select(LR_all, 17:18))
#Fe_tot <- bind_cols(depth= LR_all$depth_VT22, Fe_tot_pre)
Fe_tot <- bind_cols(age= LR_all$mean_VT22, Fe_tot_pre)
#Fe_tot_23 <- bind_cols(depth= LR_all$depth_VT22, Fe_23)
Fe_tot_23 <- bind_cols(age= LR_all$mean_VT22, Fe_23)

Fe_stacked <- reshape2::melt(Fe_tot, var.names = names(Fe_tot)[2:ncol(Fe_tot)], measure.vars = names(Fe_tot)[2:ncol(Fe_tot)])

Fe_stacked23 <- reshape2::melt(Fe_tot_23, var.names = names(Fe_tot_23)[2:ncol(Fe_tot_23)], measure.vars = names(Fe_tot_23)[2:ncol(Fe_tot_23)])

Mn_tot_pre <- as.data.frame(select(LR_all, 12:15))
Mn_23 <- as.data.frame(select(LR_all, 13:14))
Mn_23[is.na(Mn_23)] <- 0
#Mn_tot <- bind_cols(depth= LR_all$depth_VT22, Mn_tot_pre)
Mn_tot <- bind_cols(age= LR_all$mean_VT22, Mn_tot_pre)
#Mn_tot_23 <- bind_cols(depth= LR_all$depth_VT22, Mn_23)
Mn_tot_23 <- bind_cols(age= LR_all$mean_VT22, Mn_23)


Mn_stacked <- reshape2::melt(Mn_tot, var.names = names(Mn_tot)[2:ncol(Mn_tot)], measure.vars = names(Mn_tot)[2:ncol(Mn_tot)])

Mn_stacked23 <- reshape2::melt(Mn_tot_23, var.names = names(Mn_tot_23)[2:ncol(Mn_tot_23)], measure.vars = names(Mn_tot_23)[2:ncol(Mn_tot_23)])

#colorsAll <- c("red", "green", "blue", "purple")
```

### Percentage contribution

```{r}
P_tot$sum <- rowSums(P_tot[2:5])
P_tot$F1 <- (P_tot$`P-F1`/P_tot$sum)*100
P_tot$F2 <- (P_tot$`P-F2`/P_tot$sum)*100
P_tot$F3 <- (P_tot$`P-F3`/P_tot$sum)*100
P_tot$F4 <- (P_tot$`P-F4`/P_tot$sum)*100

P_stacked_per <- reshape2::melt(P_tot, var.names = names(P_tot)[7:ncol(P_tot)], measure.vars = names(P_tot)[7:ncol(P_tot)])


Mn_tot$sum <- rowSums(Mn_tot[2:5])
Mn_tot$F1 <- (Mn_tot$`Mn-F1`/Mn_tot$sum)*100
Mn_tot$F2 <- (Mn_tot$`Mn-F2`/Mn_tot$sum)*100
Mn_tot$F3 <- (Mn_tot$`Mn-F3`/Mn_tot$sum)*100
Mn_tot$F4 <- (Mn_tot$`Mn-F4`/Mn_tot$sum)*100

Mn_stacked_per <- reshape2::melt(Mn_tot, var.names = names(Mn_tot)[7:ncol(Mn_tot)], measure.vars = names(Mn_tot)[7:ncol(Mn_tot)])

Fe_tot$sum <- rowSums(Fe_tot[2:5])
Fe_tot$F1 <- (Fe_tot$`Fe-F1`/Fe_tot$sum)*100
Fe_tot$F2 <- (Fe_tot$`Fe-F2`/Fe_tot$sum)*100
Fe_tot$F3 <- (Fe_tot$`Fe-F3`/Fe_tot$sum)*100
Fe_tot$F4 <- (Fe_tot$`Fe-F4`/Fe_tot$sum)*100

Fe_stacked_per <- reshape2::melt(Fe_tot, var.names = names(Fe_tot)[7:ncol(Fe_tot)], measure.vars = names(Fe_tot)[7:ncol(Fe_tot)])
```

### Plot

#### Percentage plots

```{r}
# Create the plot
Pseq_per <- ggplot(P_stacked_per, aes(y = value, x = age, #depth 
                                      fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("seagreen4", "turquoise3", "seagreen2", "turquoise"))+
coord_flip()+ 
  labs(title = "", y = "P Fraction \n (%)", x = "") + 
  theme(legend.position="none") #axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

Mnseq_per <- ggplot(Mn_stacked_per, aes(y = value,  x = age, #depth 
                                        fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25,position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  #scale_fill_viridis_d(option = "H")+
  scale_fill_manual(values = c("dodgerblue1", "darkorchid2", "dodgerblue3", "darkorchid4"))+
coord_flip()+ 
  labs(title = "", y = "Mn Fraction \n (%)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())


Feseq_per <-  ggplot(Fe_stacked_per, aes(y = value, x = age, #depth 
                                         fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  #scale_fill_viridis_d(option = "H")+
  scale_fill_manual(values = c("coral3", "goldenrod1", "coral1", "goldenrod4"))+
coord_flip()+ 
  labs(title = "", y = "Fe Fraction \n(%)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())
```

#### Absolute values stacked

```{r}
# Create the plot
Pseq <- ggplot(P_stacked, aes(y = value,  x = age, #depth 
                              fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("seagreen4", "turquoise3", "seagreen2", "turquoise"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "P \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

# Create the plot
Mnseq <- ggplot(Mn_stacked, aes(y = value,  x = age, #depth 
                                fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("dodgerblue1", "darkorchid2", "dodgerblue3", "darkorchid4"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "Mn \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

# Create the plot
Feseq <- ggplot(Fe_stacked, aes(y = value,  x = age, #depth 
                                fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("coral3", "goldenrod1", "coral1", "goldenrod4"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "Fe \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())
```

#### Mn redox sensitive fractions

```{r}
Mn_totRed <- select(Mn_tot, 1, 3:4)
Mn_totRed[is.na(Mn_totRed)] <- 0
Mn_totRed$F2F3 <- rowSums(Mn_totRed[,2:3])

MnRedox2 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Mn_tot, aes( x = age, #depth  
                               y = Mn_tot$`Mn-F2`), width = 25, stat = "identity", alpha = 0.4, fill = "purple")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Mn redox F2 \n(µg/g)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "purple"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none")

MnRedox3 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Mn_tot, aes( x = age, #depth  
                               y = Mn_tot$`Mn-F3`), width = 25, stat = "identity", alpha = 0.4, fill = "purple3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Mn redox F3 \n(µg/g)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "purple3"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none")

MnRedox23 <- ggplot()+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
             linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_col(data = Mn_totRed, aes( x = age, #depth 
                                  y = F2F3), width = 25, stat = "identity", alpha = 0.4, fill = "purple3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0))+
 # geom_path(data=Pig_da.clean, aes(y=Okenone, x= comp_depth_mm), alpha=.8, size=0.2, colour = "green3") +
  labs(title = " ", y = "Mn redox F2+3 \n(µg/g)", x = " ") +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
  axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "purple3"))+
  # annotate("text", y= 600, x= 13700, label = "Pleniglacial", color= "blue")+
  # annotate("text", y= 600, x= 12800, label = "Bølling", color= "darkred")+
  # annotate("text", y= 600, x= 12.100, label = "Allerød", color= "darkred")+
  # annotate("text", y= 600, x= 11600, label = "Lachersee tephra", color= "black", size = 3)+
  # annotate("text", y= 600, x= 11100, label = "Younger Dryas", color= "blue")+
  coord_flip()+ 
  theme(legend.position="none")
```

#### Fraction 2 and 3 only

```{r}
Mnseq23 <- ggplot(Mn_stacked23, aes(y = value,  x = age, #depth 
                                    fill = factor(variable))) + 
    geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+ # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("darkorchid4","dodgerblue1", "darkorchid2", "dodgerblue3"))+
coord_flip()+ 
      scale_y_continuous(position = 'left',  expand=c(0,0)) +
  labs(title = "", y = "Mn redox fractions, \n F2 (purple),\n F3 (blue) (ug/g)  ", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(0,0,0.5,-0.63), "cm"),axis.text = element_text(size = 12), panel.background = element_blank()) #legend.position = "top", legend.title = element_blank())



Feseq23 <- ggplot(Fe_stacked23, aes(y = value,  x = age, #depth 
                                    fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
  labs(title = "",  # Adjust title (optional)
       x = "Depth",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("coral3", "goldenrod1", "coral1", "goldenrod4"))+
coord_flip()+ 
      scale_y_continuous(position = 'left',  expand=c(0,0)) +
  labs(title = "", y = "Fe redox fractions\n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())



Pseq23 <- ggplot(P_stacked23, aes(y = value,  x = age, #depth 
                                  fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
  labs(title = "",  # Adjust title (optional)
       x = "Depth",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("seagreen4", "turquoise3", "seagreen2", "turquoise"))+
coord_flip()+ 
      scale_y_continuous(position = 'left',  expand=c(0,0)) +
  labs(title = "", y = "P redox fractions\n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())
```

```{r}
Redox <- plot_grid(Pseq_per, Feseq_per, Mnseq_per, Pseq, Feseq, Mnseq, Pseq23,  Feseq23, Mnseq23,
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1,1,1,1), axis = "tb" )

png(filename = timestamp_filename("../figs", "Redox"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)

Redox

dev.off()
```

### Stacked plots

```{r}
PigmentsR <- plot_grid(
  #RABD620, RABD660,  RABD845,  
  #Cyanos,TChls, Oke, 
  Chlfs, Bphe, Cyano, 
  PigCom1, PigCom3, PigCom2, 
  Oke, Fuco,
  #MnRedox23,
  Mnseq23, #Feseq23, Pseq23,
#MnRedox3,  MnRedox2, 
  #Chlfs, Bphe, Cyano, 
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1,1,1,1,1,1,1, 1,1), axis = "tb" )
#dev.off()

png(filename = timestamp_filename("../figs", "PigmentsR"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)

PigmentsR

dev.off()
```

## Diatom selection

Load the diatom counts data and convert it to count%. Filter out all the columns/diatom species which have below 3% op counts.

```{r}
DiatL <- read_excel("../data/SourceData/diatoms_Lujan.xlsx") 

#Conversion of counts to %
diaL_perc <- adorn_percentages(DiatL, denominator = "row", na.rm = TRUE, 4:94) #na.rm = TRUE,
diaL_100 <- na.omit(cbind(diaL_perc[2:3], diaL_perc[4:94]*100)) 

# Threshold value of 3%
threshold <- 3 

# Function to filter columns based on threshold 
filter_columns <- function(df, threshold) {
  df %>% 
    select_if(~ any(. >= threshold)) 
  } 
# Apply the function to the data
filtered_diaL<- filter_columns(diaL_100, threshold)

write.xlsx(filtered_diaL, "../data/diatom_filtered.xlsx", rowNames = FALSE)

#df_plot <- filtered_data[3:18]
```

Load the diatom names

```{r}
DiaN <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "diatoms_name") #names <- as.list(DiaN[,1])
```

```{r}
df_names <- as.data.frame(DiaN) 
names <- as.vector(df_names[,1])
```

```{r}
rownames(df_names) <- names
```

```{r}
rownames(filtered_diaL) <- filtered_diaL[, 2] # 1 for depth
#rownames(DiaN) <- DiaN[ ,1] 
df_names<- df_names[,-1]
```

```{r}
#filtered_diaL <- filtered_diaL[,-1] 

#annotation_colD = df_names[1]
annotation_colA = df_names[1]
```

```{r}
PolZD <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "PolZ") 
#depth_mat <- as.data.frame(PolZD) 
age_mat <- as.data.frame(PolZD) 
#rownames(depth_mat) <- depth_mat[,1] 
rownames(age_mat) <- age_mat[,2] 
#depth_mat <- depth_mat[,-1] 
#annotation_rowD = depth_mat[3]
annotation_rowA = age_mat[3]
```

```{r}
#df_depth <- filtered_diaL[1] 
df_age <- filtered_diaL[2] 
df<- filtered_diaL[3:18] 
#df_cut<- filtered_diaL[1:20,2:17] 
#df_anox <- df_na_df[1:49, 3:23]
```

### Clustering

```{r}
png(filename = timestamp_filename("../figs", "diatoms"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)

y.scale <- filtered_diaL$mean_VT22
#y.scale <- filtered_diaL$`comp_depth_(mm)`

ma.dist <- vegdist(df, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")

dia.plot <- strat.plot(df, yvar=y.scale, y.tks=y.scale,  ylim= c(limYmin, limYmax), y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", #col.poly=p.col, 
col.poly.line="black", scale.percent=FALSE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar diatoms LG Samples", ylabel="Sample", clust=ma.chclust)

# Add cluster zones

addClustZone(dia.plot, ma.chclust, nZone=3, lwd=1.5, lty=2, col="grey25")

dev.off()
#dev.off()
```

```{r}
# remove_high_zero_columns <- function(data) { 
#   # # Calculate the proportion of zeros in each column 
#   prop_zeros <- colMeans(data == 0) 
#   # Keep columns with less than or equal to 95% zeros 
#   data[, prop_zeros <= 0.05] 
#   } #

# Remove columns with high zero proportion 
# df_clcl<- as.data.frame(remove_high_zero_columns(df_corr_cl)) 
#dfD_cl<- as.data.frame(remove_high_zero_columns(df_cut)) 
#dfD_scaled_cut<- scale(dfD_cl) #dfD_anox_scaled <- scale(df_anoxB)

dfD_scaled<- scale(df) 
```

### Heatmaps

```{r}
ann_colorsD = list( group = c(`benthic+epiphytic` = "darkorange3", planktonic ="royalblue2"), PolZD = c(`Pleniglacial` = "dodgerblue", `Bølling` = "tomato2", `Allerød` = "tomato4")) 

#Both dimension clustering 
png(filename = timestamp_filename("../figs", "diatoms_heat"), width = 1200, height = 1000, units = "px", pointsize = 14, res=120)

pheatmap(dfD_scaled, show_rownames=T,show_colnames=T, 
         annotation_col=annotation_colA, annotation_row = annotation_rowA, 
         annotation_colors = ann_colorsD, 
         scale = "none",clustering_method="ward.D2", #cluster_rows = F, 
         clustering_distance_cols="correlation", clustering_distance_rows = "euclidean",
         cutree_rows = 3, 
         cutree_cols = 3, treeheight_row = 50, 
         treeheight_col = 50, color = colorRampPalette(c("lightskyblue1","darkseagreen1","lightgreen","yellow", "orange", "red", "purple"))(15))

dev.off() 

# Only correlation clustering 
png(filename = timestamp_filename("../figs", "diatoms_heat_corr"), width = 1200, height = 1000, units = "px", pointsize = 14, res=120)

pheatmap(dfD_scaled, show_rownames=T,show_colnames=T, 
         annotation_col=annotation_colA, annotation_row = annotation_rowA, 
         annotation_colors = ann_colorsD, 
         scale = "none",clustering_method="ward.D2", cluster_rows = F, 
         clustering_distance_cols="correlation", #clustering_distance_rows = "euclidean", # cutree_rows = 5, 
         cutree_cols = 3, #treeheight_row = 50, 
         treeheight_col = 50, color = colorRampPalette(c("lightskyblue1","darkseagreen1","lightgreen","yellow", "orange", "red", "purple"))(15)) 

dev.off()
```

Based on the correlation clusters and the ecological meaning, we select further: Staurosira construens as a benthic oligotrophic species, Amphora indistinta as a oligo-to-mesotrophic species growing under low N conditions, Pantocekiella ocellata as a oligo-to-mesotrophic species growing in shallow/lithoral areas and finally Stephanodicus minutulus as a meso-to-eutrophic species growing in well-mixed, oxygenated water column.

### Plot

```{r}
Amph <- ggplot(filtered_diaL, aes(y= Amphora_indistinta, x= mean_VT22))+ #`comp_depth_(mm)`))+ 
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "blue")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
    scale_y_continuous(position = 'left', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Amphora\n indistinta (P) \n %counts ", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(0,0,0.5,-0.55), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

Pant <- ggplot(filtered_diaL , aes(y=Pantosekiella_ocellata,x= mean_VT22))+ #`comp_depth_(mm)`))+
   geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "brown")+
   scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
    scale_y_continuous(position = 'right', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
    labs(title = " ", y = "Pantosekiella\n ocellata (B) \n %counts " , x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

Stau <- ggplot(filtered_diaL , aes(y=Staurosira_construens, x= mean_VT22))+ #`comp_depth_(mm)`))+
   geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "brown")+
   scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
    scale_y_continuous(position = 'right', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
    scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
    theme_classic() +
  labs(title = " ", y = "Staurosira \n construens (B)\n %counts  ", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

Steph <-ggplot(filtered_diaL , aes(y=Stephanodiscus_minutulus, x= mean_VT22))+ #`comp_depth_(mm)`))+
   geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_bar(stat = "identity", width = 25, fill = "blue")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
    scale_y_continuous(position = 'left', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
    scale_fill_viridis_d(option = "H")+
  coord_flip()+ 
    labs(title = " ", y = "Stephanodiscus\n minutulus (P)\n %counts ", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())
```

```{r}
PigmentsRD <- plot_grid(
  #RABD620, RABD660,  RABD845,  
  #Cyanos,TChls, Oke, 
  Chlfs, Bphe, Cyano, 
  PigCom1, PigCom3, PigCom2, 
  Oke, Fuco,
  #MnRedox23,
  Mnseq23, #Feseq23, Pseq23,
#MnRedox3,  MnRedox2, 
  #Chlfs, Bphe, Cyano, 
  Stau, Amph, Pant, Steph,
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1,1,1,1,1,1,1,1,1, 1,1, 1,1 ,1), axis = "tb" )
#dev.off()

png(filename = timestamp_filename("../figs", "PigmentsRedDia"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)

PigmentsRD

dev.off()
```

## Pollen

```{r message=FALSE, warning=FALSE, include=FALSE}
PollenDat <- read_excel("../data/SourceData/Pollen_Litt_Usinger_stacked.xlsx", sheet = "LittLittStackedPer", col_types = "numeric")

PolDat<- PollenDat[3:155,]
PolDatStat <- PolDat[, 3:112]
```

```{r}
#PolDat[is.na(PolDat)] <- 0
# perc_pol <- adorn_percentages(PolDat, denominator = "row", na.rm = TRUE, 3:112)
# perc_pol_100 <- cbind(perc_pol[1:2], perc_pol[3:112]*100)


MeerPinus <- read_excel("../data/SourceData/Meerfeldemar_pollen.xlsx", sheet = "Pinus", col_types = "numeric")

MeerBetula <- read_excel("../data/SourceData/Meerfeldemar_pollen.xlsx", sheet = "Betula", col_types = "numeric")

MeerAP <- read_excel("../data/SourceData/Meerfeldemar_pollen.xlsx", sheet = "Herbs", col_types = "numeric")
```

### Clustering

```{r}
png(filename = timestamp_filename("../figs", "pollen"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)
y.scale <- PolDat$age
#y.scale <- filtered_diaL$`comp_depth_(mm)`

# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(PolDatStat) # Calculate column sums
pol_cl_per <- PolDatStat[, which(ma.sum > 10)] # Subset data

pol_df<- pol_cl_per[3:34]
pol_df[is.na(pol_df)] <-0

ma.dist <- vegdist(pol_df, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")

pol.plot <- strat.plot(pol_df, yvar=y.scale, y.tks=y.scale,  ylim= c(limYmin, limYmax),
                       y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", #col.poly=p.col, 
col.poly.line="black", scale.percent=FALSE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar pollen Litt Litt stack", ylabel="Sample", clust=ma.chclust)

# Add cluster zones
addClustZone(pol.plot, ma.chclust, nZone=10, lwd=1.5, lty=2, col="grey25")#dev.off()
```

```{r}
APNAP <- select(PolDat, 1:2, 116, 115)
Trees <- select(PolDat, 1,2,8,41,54,55,71)

pollen_stacked_per <- melt(APNAP, var.names = names(APNAP)[3:ncol(APNAP)], measure.vars = names(APNAP)[3:ncol(APNAP)])

Meer_pollen_stacked <- melt(MeerAP, var.names = names(MeerAP)[2:ncol(MeerAP)], measure.vars = names(MeerAP)[2:ncol(MeerAP)])
```

### Pollen plots

```{r}
poll<- ggplot(pollen_stacked_per, aes(y= value, x=pollen_stacked_per$age, fill=factor(variable)))+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25)+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
    scale_y_continuous(position = 'right', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_manual(values = c("goldenrod","springgreen2"))+
   coord_flip()+ 
  labs(title = " ", y = "AP and NAP \n(%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Pinus <- ggplot(Trees , aes(y= Pinus, x= age))+
  #geom_point(colour = "green", size= 0.02)+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  geom_bar(stat = "identity", width = 25, fill = "forestgreen")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2),expand=c(0,0)) +
    scale_y_continuous(position = 'left', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Pinus \n(%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Betula <- ggplot(Trees , aes(y= Betula, x= age))+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "burlywood3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2),expand=c(0,0)) +
    scale_y_continuous(position = 'right', expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Betula\n (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Juniperus <- ggplot(Trees , aes(y= Trees$Juniperus, x= age))+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "palegreen4")+
  annotate("text", y= 0.8, x= 14950, label = "Pleniglacial", color= "blue")+
  annotate("text", y= 0.8, x= 14050, label = "Bølling", color= "darkred")+
  annotate("text", y= 0.8, x= 13350, label = "Allerød", color= "darkred")+
  annotate("text", y= 0.8, x= 13000, label = "Lachersee \n tephra", color= "black")+
  annotate("text", y= 0.8, x= 12250, label = "Younger Dryas", color= "blue")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2),expand=c(0,0), position = "top") +
    scale_y_continuous(position = 'left',expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Juniperus\n (%)", x = " Age (Cal yr BP)" ) +
theme(legend.position="none", plot.margin = unit(c(0,0,0.5,-0.2), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())# axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


PollenFig <-plot_grid(poll, Pinus, Betula, Juniperus, nrow=1, rel_heights = c(1,1,1,1), hjust = -0.5, label_size = 18,
                    label_fontfamily = sans)

png(filename = timestamp_filename("../figs", "pollen_stack"), width = 1754, height = 1260, units = "px", pointsize = 14, res=120)

PollenFig

dev.off()
```

### Meerfeldemaar pollen underlay

```{r}
pollCombo<- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_col(data = Meer_pollen_stacked, aes(y= value, x=Meer_pollen_stacked$`Age varve BC`, fill=factor(variable)), stat = "identity", width = 20, alpha = 0.4)+
    geom_bar(data = pollen_stacked_per, aes(y= value, x=pollen_stacked_per$age, fill=factor(variable)), stat = "identity", width = 30)+
geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2), expand=c(0,0)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_manual(values = c("springgreen3","goldenrod","goldenrod3","springgreen"))+
   coord_flip()+ 
  scale_y_continuous(position = 'right', expand=c(0,0), sec.axis = sec_axis(~ . *1, name = "Holzmaar \n AP and NAP (%)")) +
  labs(title = " ", y = "Meerfeldmaar\n AP and NAP (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())

PinusCombo <- ggplot()+
  geom_col(data = MeerPinus, aes(y= MeerPinus$`Pinus%`, x=MeerPinus$`Age varve yr BP`), fill="forestgreen", stat = "identity", width = 20, alpha = 0.4 )+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data=Trees , aes(y= Pinus, x= age), stat = "identity", width = 30, fill = "darkseagreen4")+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2),expand=c(0,0)) +
    scale_y_continuous(position = 'left', expand=c(0,0), sec.axis = sec_axis(~ . *1, name = " Meerfeldemaar \n Pinus (%)")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Holzmaar \n Pinus (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"),axis.text = element_text(size = 12), panel.background = element_blank())

BetulaCombo <- ggplot()+
  geom_col(data = MeerBetula, aes(y= MeerBetula$`Betula%`, x=MeerBetula$`Age varve yr BP`), fill="burlywood", stat = "identity", width = 20, alpha = 0.4 )+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data=Trees , aes(y= Betula, x= age), stat = "identity", width = 30, fill = "burlywood4")+
  geom_vline(data= PollenZPig, xintercept = PollenZPig$age, #depth_VT22
               linetype = PollenZPig$line, color= PollenZPig$color, alpha = 0.8)+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 100), digits = 2),expand=c(0,0)) +
    scale_y_continuous(position = 'left', expand=c(0,0), sec.axis = sec_axis(~ . *1, name = "Meerfeldemaar \n Betula (%)")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Holzmaar \n Betula (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),plot.margin = unit(c(0,0,0.5,-0.55), "cm"), axis.text = element_text(size = 12), panel.background = element_blank())


```

# Figure 2 combined

```{r}
# png(filename = "~/work/holzmaar/data/fig/Figure2_alt.png", width = 3000, height = 1500, units = "px", pointsize = 13, res=110 )

Fig2 <- plot_grid(
  RABD620, RABD660,  RABD845,  
  Cyanos,TChls, Oke, Fuco,
  #Chlfs, Bphe, Cyano, 
  PigCom1, PigCom3, PigCom2, 
  #Oke, 
  #MnRedox23,
  Mnseq23, #Feseq23, Pseq23,
#MnRedox3,  MnRedox2, 
  #Chlfs, Bphe, Cyano, 
  Stau, Amph, Pant, Steph,
pollCombo, PinusCombo, BetulaCombo, Juniperus,
  nrow = 1, align = "h", #scale = c(0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 1, 1)) 
          rel_widths = c(1, 0.6,0.6,
                         0.65,0.65,0.65, 0.65,
                         0.8,0.8,0.8,
                         0.7,
                         0.6, 0.6, 0.6, 0.67,
                         0.6 ,0.6, 0.6, 1), 
          axis = "tb"
          #panel_spacing = unit(0.05, "lines")
)

png(filename = timestamp_filename("../figs", "Figure2_alt_res"), width = 1754, height = 1240, units = "px", pointsize = 14, res=72)

Fig2

dev.off()
```

# Figure 2 Journal Format - Recreate plots with smaller fonts

```{r}
# Journal format font sizes (following notebook 05 settings)
journal_font_size <- 5.5
journal_axis_text_size <- 5.5
journal_title_size <- 5.5

# Recreate HSI plots with journal formatting
RABD620_j <- ggplot()+
  geom_vline(xintercept=mean(HSI_data$RABD620), color="dodgerblue1", linetype = 1) +
  geom_vline(xintercept=mean(HSI_data$RABD620) + sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_vline(xintercept=mean(HSI_data$RABD620) - sd(HSI_data$RABD620), color="gray", linetype = 2) +
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD620), colour = "dodgerblue1", size= 0.3)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD620), alpha=.8, size=0.8, colour = "dodgerblue1") +
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0), breaks = c(1, 1.02))+
  labs(title = " ", x = "RABD620\n(Phycocyanin)", y = "Age (Cal yr BP)")+
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", 
        axis.title.x.bottom = element_text(color = "dodgerblue1", size = journal_title_size, face="bold"), 
        axis.title.y = element_text(size = journal_title_size, face="bold"),
        axis.text = element_text(size = journal_axis_text_size), 
        plot.margin = unit(c(0,0,0.5,0), "cm"), 
        panel.background = element_blank(),
        plot.background = element_blank())

RABD660_j <- ggplot()+
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD670calib), colour = "limegreen", size= 0.3)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD670calib), alpha=.8, size=0.8, colour = "limegreen") +
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "top", expand=c(0,0), breaks = c(0,20,40))+
  labs(title = " ", x = "Total Chls\n(µg/g wet sed)", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", 
        axis.title.x.top = element_text(color = "limegreen", size = journal_title_size, face="bold"), 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size), 
        panel.background = element_blank(),
        plot.background = element_blank())

RABD845_j <- ggplot()+
  geom_point(data= HSI_data, aes(y=mean_VT22, x= RABD845calib), colour = "violet", size= 0.3)+
  geom_path(data = HSI_data, aes(y=mean_VT22, x= RABD845calib), alpha=.8, size=0.8, colour = "violet") +
  scale_y_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_x_continuous(position = "bottom", expand=c(0,0), n.breaks = 3)+
  labs(title = " ", x = "Bacteriopheophytin a\n(µg/g wet sed)", y = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", 
        axis.title.x.bottom = element_text(color = "violet", size = journal_title_size, face="bold"), 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0.1,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size), 
        panel.background = element_blank(),
        plot.background = element_blank())

# Recreate pigment plots with journal formatting using Pig_da.clean
Cyanos_j <- ggplot()+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, y =Pig_da.clean$Cyanos), 
           width = 25, stat = "identity", alpha = 0.4, fill = "lightseagreen")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0), n.breaks = 3)+
  labs(title = " ", y = "Echi+Cantha+Zea\n(µg/g OM)", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(), 
        axis.title.x.top = element_text(color = "lightseagreen", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,0,0.5,-0.1), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

TChls_j <- ggplot()+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, y =Pig_da.clean$TChl), 
           width = 25, stat = "identity", alpha = 0.4, fill = "springgreen3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0), breaks = c(0,250, 500, 750))+
  labs(title = " ", y = "Total Chls\n(µg/g OM)", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.bottom = element_text(color = "springgreen3", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,-0.2,0.5,-0.55), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

Oke_j <- ggplot()+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, y =Pig_da.clean$Okenone), 
           width = 25, stat = "identity", alpha = 0.4, fill = "maroon4")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0), n.breaks = 3)+
  labs(title = " ", y = "Okenone\n(µg/g OM)", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.top = element_text(color = "maroon4", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,-0.2,0.5,0), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

Fuco_j <- ggplot()+
  geom_col(data = Pig_da.clean, aes(x=Pig_da.clean$mean_VT22, y =Pig_da.clean$Fuco), 
           width = 25, stat = "identity", alpha = 0.4, fill = "darkorange3")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0), breaks = c(0, 25, 50, 75))+
  labs(title = " ", y = "Fucoxanthin\n(µg/g OM)", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.bottom = element_text(color = "darkorange3", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,0,0.5,-0.35), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

# Recreate PCA component plots with journal formatting using alr_scaled_clust
PigCom1_j <- ggplot()+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C1), alpha=1, size=0.8, colour = "royalblue1") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0), n.breaks = 5)+
  labs(title = " ", y = "Pigment group 1\nPioneers", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.top = element_text(color = "royalblue1", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,0,0.5,-0.2), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

PigCom3_j <- ggplot()+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C3), alpha=1, size=0.8, colour = "royalblue3") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "left"), expand=c(0,0)) +
  scale_y_continuous(position = "left", expand=c(0,0), n.breaks = 3)+
  labs(title = " ", y = "Pigment group 2\nSecondary colonisers", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.bottom = element_text(color = "royalblue3", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,0.05,0.5,-0.55), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

PigCom2_j <- ggplot()+
  geom_path(data = alr_scaled_clust, aes(x=age, y= C2), alpha=1, size=0.8, colour = "royalblue4") +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), 
                  guide = guide_axis(position = "right"), expand=c(0,0)) +
  scale_y_continuous(position = "right", expand=c(0,0), n.breaks = 3)+
  labs(title = " ", y = "Pigment group 3\nDiatoms+Bacteria", x = " ") +
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x.top = element_text(color = "royalblue4", size = journal_title_size, face="bold"),
        plot.margin = unit(c(0,0,0.5,-0.05), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        panel.background = element_blank(),
        plot.background = element_blank())+
  coord_flip()

# Recreate Mn sequential extraction plot with journal formatting using Mn_stacked23
Mnseq23_j <- ggplot(Mn_stacked23, aes(y = value,  x = age, fill = factor(variable))) + 
  geom_bar(stat = "identity", width = 25, position = position_stack(reverse = TRUE)) +
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  scale_fill_manual(values = c("darkorchid4","dodgerblue1", "darkorchid2", "dodgerblue3"))+
  coord_flip()+ 
  scale_y_continuous(position = 'left', expand=c(0,0), n.breaks = 3) +
  labs(title = "", y = "Mn fractions \n F2+F3 (µg/g)", x = "") + 
  theme_classic(base_size = journal_font_size) +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        plot.margin = unit(c(0,0,0.5,-0.55), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

# Recreate diatom plots with journal formatting using filtered_diaL
Stau_j <- ggplot(filtered_diaL, aes(y=Staurosira_construens, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "brown")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  scale_y_continuous(position = 'right', expand=c(0,0), breaks = c(0, 25, 50, 75)) +
  coord_flip()+ 
  theme_classic(base_size = journal_font_size) +
  labs(title = " ", y = "Staurosira\nconstruens (%)", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

Amph_j <- ggplot(filtered_diaL, aes(y=Amphora_indistinta, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "brown")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  scale_y_continuous(position = 'left', expand=c(0,0), n.breaks = 3) +
  coord_flip()+ 
  theme_classic(base_size = journal_font_size) +
  labs(title = " ", y = "Amphora\nindistinta (%)", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), 
        plot.margin = unit(c(0,0,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

Pant_j <- ggplot(filtered_diaL, aes(y=Pantosekiella_ocellata, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "brown")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  scale_y_continuous(position = 'right', expand=c(0,0), n.breaks = 3) +
  coord_flip()+ 
  theme_classic(base_size = journal_font_size) +
  labs(title = " ", y = "Pantosekiella\nocellata (%)", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

Steph_j <- ggplot(filtered_diaL, aes(y=Stephanodiscus_minutulus, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "blue")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  scale_y_continuous(position = 'left', expand=c(0,0), breaks = c(0, 25, 50, 75)) +
  coord_flip()+ 
  theme_classic(base_size = journal_font_size) +
  labs(title = " ", y = "Stephanodiscus\nminutulus (%)", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.3), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

# Recreate pollen plots with journal formatting
pollCombo_j <- ggplot()+
  geom_col(data = Meer_pollen_stacked, aes(y= value, x=Meer_pollen_stacked$`Age varve BC`, fill=factor(variable)), stat = "identity", width = 20, alpha = 0.4)+
  geom_bar(data = pollen_stacked_per, aes(y= value, x=pollen_stacked_per$age, fill=factor(variable)), stat = "identity", width = 30)+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2), expand=c(0,0)) +
  theme_classic(base_size = journal_font_size) +
  scale_fill_manual(values = c("springgreen3","goldenrod","goldenrod3","springgreen"))+
  coord_flip()+ 
  scale_y_continuous(position = 'right', expand=c(0,0), breaks = c(25, 75), sec.axis = sec_axis(~ . *1, name = "\n    AP/NAP", breaks = c(25, 75))) +
  labs(title = " ", y = " \n AP/NAP", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.2), "cm"), 
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

PinusCombo_j <- ggplot()+
  geom_col(data = MeerPinus, aes(y= MeerPinus$`Pinus%`, x=MeerPinus$`Age varve yr BP`), fill="forestgreen", stat = "identity", width = 20, alpha = 0.4 )+
  geom_bar(data=Trees , aes(y= Pinus, x= age), stat = "identity", width = 30, fill = "darkseagreen4")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2),expand=c(0,0)) +
  scale_y_continuous(position = 'left', expand=c(0,0), n.breaks = 3, sec.axis = sec_axis(~ . *1, name = "Pinus")) +
  theme_classic(base_size = journal_font_size) +
  coord_flip()+
  labs(title = " ", y = "Pinus", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.25), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

BetulaCombo_j <- ggplot()+
  geom_col(data = MeerBetula, aes(y= MeerBetula$`Betula%`, x=MeerBetula$`Age varve yr BP`), fill="burlywood", stat = "identity", width = 20, alpha = 0.4 )+
  geom_bar(data=Trees , aes(y= Betula, x= age), stat = "identity", width = 30, fill = "burlywood4")+
  scale_x_reverse(limits = c(limYmax, limYmin), breaks = round(seq(limYmin, limYmax, by = 500), digits = 2),expand=c(0,0)) +
  scale_y_continuous(position = 'left', expand=c(0,0), breaks = c(0, 25, 50), sec.axis = sec_axis(~ . *1, name = "Betula", breaks = c(0, 25, 50))) +
  theme_classic(base_size = journal_font_size) +
  coord_flip()+
  labs(title = " ", y = "\n Betula", x = " ") +
  theme(legend.position="none", 
        axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.25), "cm"),
        axis.text = element_text(size = journal_axis_text_size),
        axis.title = element_text(size = journal_title_size, face="bold"),
        panel.background = element_blank(),
        plot.background = element_blank())

Juniperus_j <- ggplot(Trees , aes(y= Trees$Juniperus, x= age))+
  geom_bar(stat = "identity", width = 30, fill = "palegreen4")+
  scale_x_reverse(limits = c(limYmax, limYmin), expand=c(0,0), position = "top", breaks = NULL) +
  scale_y_continuous(position = 'left',expand=c(0,0), breaks = c(1)) +
  theme_classic(base_size = journal_font_size) +
  coord_flip()+
  labs(title = " ", y = "Juniperus", x = " ") +  # Remove x axis label
  theme(legend.position="none", 
        plot.margin = unit(c(0,-0.05,0.5,0), "cm"),  # Negative right margin
        axis.text.x.bottom = element_text(size = journal_axis_text_size),  # Juniperus values (after flip)
        axis.text.y = element_text(size = journal_axis_text_size),  # Keep age axis text (left)
        axis.text.x.top = element_blank(),   # Remove top axis text/numbers
        axis.ticks.x.top = element_blank(), 
        axis.title.x.bottom = element_text(size = journal_title_size, face="bold"),  # Juniperus label (after flip)
        axis.title.y = element_text(size = journal_title_size, face="bold"),  # Age label (left)
        axis.title.x.top = element_blank(),  # Remove top axis title
        axis.line.x.top = element_blank(),   # Remove top axis line
        panel.background = element_blank(),
        plot.background = element_blank())

# Create zone legend strip to replace top axis
zones_strip_j <- ggplot() +
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 15342, xmax = 14444, 
           fill = "lightblue", color = "black", size = 0.2) +  # Pleniglacial
  annotate("text", y = 0.4, x = (15342 + 14444)/2, label = "Pleniglacial", 
           size = 2, angle = 270, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 14444, xmax = 13611, 
           fill = "lightcoral", color = "black", size = 0.2) +  # Bølling
  annotate("text", y = 0.4, x = (14444 + 13611)/2, label = "Bølling", 
           size = 2, angle = 270, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 14070 , xmax = 14016, 
           fill = "white", color = "black", size = 0.2, alpha = 0.2) +  # Oldest Dryas
  annotate("text", y = 0.4, x = ((14070 + 14016)/2)-20, label = "Oldest\nDryas", 
           size = 1, angle = 0, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 13906, xmax = 13611, 
           fill = "white", color = "black", size = 0.2, alpha = 0.2) +  # Older Dryas
  annotate("text", y = 0.4, x = (13906 + 13611)/2, label = "Older\nDryas", 
           size = 1, angle = 0, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 13611, xmax = 12727, 
           fill = "lightsalmon", color = "black", size = 0.2) +  # Allerød
  annotate("text", y = 0.4, x = (13611 + 12727)/2, label = "Allerød", 
           size = 2, angle = 270, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 13011, xmax = 13011, 
           fill = "lightgray", color = "black", size = 0.2) +  # Lachersee tephra
  annotate("text", y = 0.4, x = 12970, label = "LST", 
           size = 2, angle = 0, hjust = 0.5) +
  
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 12727, xmax = 11625, 
           fill = "lightblue", color = "black", size = 0.2) +  # Younger Dryas
  annotate("rect", ymin = 0, ymax = 0.8, xmin = 12540, xmax = 11850, 
           fill = "white", alpha= 0.4, color = "black", linetype = "dashed", size = 0.3) +  # Hiatus
  annotate("text", y = 0.55, x = (12540 + 11850)/2, label = "Hiatus", 
           size = 2, angle = 270, hjust = 0.5) +
  annotate("text", y = 0.25, x = (12727 + 11625)/2, label = "Younger Dryas", 
           size = 2, angle = 270, hjust = 0.5) +
  scale_x_reverse(limits = c(limYmax, limYmin), expand=c(0,0)) +
  scale_y_continuous(limits = c(0, 1), expand=c(0,0)) +
  coord_flip() +
  theme_void() +
  theme(plot.background = element_blank(),
        panel.background = element_blank(),
        plot.margin = unit(c(0,0,0.5,-0.2), "cm"))  # Negative left margin to move closer
```

# Figure 2 combined - Journal Format (1/2 A4)

```{r}
# Journal format dimensions for 1/2 A4 page width (180mm = double column standard)
# Resolution: 300 DPI for publication quality
# 180mm = 7.09 inches = 2126 px at 300 DPI
# Height: 120mm = 4.72 inches = 1417 px at 300 DPI

# Total plots: 19 + 1 zone strip (3 HSI + 4 Pigments + 3 PCA + 1 Mn + 4 Diatoms + 4 Pollen + 1 zones)
# All plots have equal width, zone strip is narrower
base_grid <- plot_grid(
  RABD620_j, RABD660_j, RABD845_j,  
  Cyanos_j, TChls_j, Oke_j, Fuco_j,
  PigCom1_j, PigCom3_j, PigCom2_j, 
  Mnseq23_j,
  Stau_j, Amph_j, Pant_j, Steph_j,
  pollCombo_j, PinusCombo_j, BetulaCombo_j, Juniperus_j, zones_strip_j,
  nrow = 1, 
  align = "h",
  #rel_widths = c(rep(1, 19), 0.15),  # All plots equal width, zone strip narrow
  rel_widths = c(0.75, 0.4, 0.4,
                 0.4, 0.4, 0.4, 0.4,
                 0.4, 0.4, 0.4, 
                 0.4, 
                 0.4, 0.4, 0.4, 0.4,
                 0.4, 0.4, 0.4, 0.35, 0.15),  # zone strip at end with minimal width
  axis = "tb"
)

# Add pollen zone lines over the entire combined plot
# The plots use coord_flip() or scale_y_reverse(), with age range from limYmax to limYmin
# We need to calculate positions accounting for plot margins and the panel area
# Since expand=c(0,0) is used, the age scale spans exactly from limYmin to limYmax

# Age boundaries for climate zones
zone_ages <- c(15342, 14444, 14070, 14015, 13907, 13611, 12727, 11625)
lst_age <- 13011

# Function to convert age to normalized position (0 at limYmin/top, 1 at limYmax/bottom)
# Since plots are displayed with age decreasing upward (young at top, old at bottom)
age_to_y <- function(age) {
  1 - (age - limYmin) / (limYmax - limYmin)
}

zone_y <- sapply(zone_ages, age_to_y)
lst_y <- age_to_y(lst_age)

# Account for plot margins - the actual plot area doesn't span full 0-1
# Bottom margin is 0.5cm, top margin is 0
# Estimate: plot area spans from ~0.04 to ~0.98 of figure height
margin_bottom <- 0.1015
margin_top <- 0.086
plot_height <- 1 - margin_bottom - margin_top

# Adjust y positions to plot area
zone_y_adj <- margin_bottom + zone_y * plot_height
lst_y_adj <- margin_bottom + lst_y * plot_height

# Add zone lines as dashed lines across entire plot
Fig2_journal <- ggdraw(base_grid) +
  # Pleniglacial boundary
  # draw_line(x = c(0.03, 0.975), y = c(zone_y_adj[1], zone_y_adj[1]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # Meiendorf boundary
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[2], zone_y_adj[2]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # Oldest Dryas boundary
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[3], zone_y_adj[3]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # Bølling boundary  
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[4], zone_y_adj[4]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  #  # Older Drays boundary  
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[5], zone_y_adj[5]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # Allerød boundary
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[6], zone_y_adj[6]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # Younger Dryas boundary
  draw_line(x = c(0.05, 0.97), y = c(zone_y_adj[7], zone_y_adj[7]), color = "black", size = 0.3, linetype = "dashed", alpha = 0.8) +
  # LST marker (Lachersee tephra)
  draw_line(x = c(0.05, 0.97), y = c(lst_y_adj, lst_y_adj), color = "gray30", size = 0.3, linetype = "solid", alpha = 0.5) +
  # Add spanning label for HZM Pollen across pollCombo, Pinus, Betula, Juniperus
  # Pollen plots are at positions 16-19 out of 20 total plots
  # Calculate center: (0.75 + 11*0.4) + (4*0.4)/2 = 5.55/8.3 ≈ 0.668
  draw_label("HZM Pollen (%)", x = 0.9, y = 0.02, size = journal_title_size, fontface = "bold", hjust = 0.5, vjust = 0) +
  draw_line(x = c(0.79,0.97), y = 0.038, color = "gray30", size = 0.3, linetype = "solid", alpha = 1) +
  # Add spanning label for MER Pollen on top axis (secondary axis position)
  draw_label("MER Pollen (%)", x = 0.865, y = 0.98, size = journal_title_size, fontface = "bold", hjust = 0.5, vjust = 1) +
 draw_line(x = c(0.79,0.94), y = 0.963, color = "gray30", size = 0.3, linetype = "solid", alpha = 1) 

# Save as high-resolution PNG with transparent background
png(filename = timestamp_filename("../figs", "Figure2_journal_halfA4"),
    width = 2126, height = 1417, units = "px", pointsize = journal_font_size, res=300, bg = "transparent")
Fig2_journal
dev.off()

# Also save as PDF for vector graphics
# pdf(file = timestamp_filename("../figs", "Figure2_journal_halfA4", extension = ".pdf"),
#     width = 7.09, height = 4.72)  # inches at 180mm x 120mm
# Fig2_journal
# dev.off()

# Display the figure
Fig2_journal
```
