---
title: "Interpolation"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Installing and loading packages

```{r message=FALSE, warning=FALSE, include=FALSE}
#install.packages("pacman")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("tibble","readxl","dplyr","tidyr","openxlsx","car","rstatix", "DT", "data.table","reshape2")
```

## Calling functions

```{r message=FALSE, warning=FALSE, include=FALSE}
Scripts  <- file.path("../notebooks/Functions", fsep="/")
source(file.path(Scripts, "utility_functions.r", fsep="/" ))
```

## Preparing the data frames into a suitable format

As the XRF data are from a long core composite and we look only at the Late-Glacial section, the starting composite depth is not 0 but 10989 mm (10.9m). Our concatenated hyperspectral data are on a scale from 0mm. Therefore, we need to adjust the depth to the XRF depth.

```{r message=FALSE, warning=FALSE}
GDGT <- read_excel("../data/SourceData/External/Auel_ELSA_stack_GDGT.xlsx", col_types = "numeric")

Egelsee <- read_excel("../data/SourceData/External/Egelsee-LG-temp.xlsx", col_types = "numeric")

Ins <- read_excel("../data/SourceData/External/Insolation.xlsx", col_types = "numeric")

GerzenseeYD <- read_excel("../data/SourceData/External/Gerzensee2005.xlsx", sheet = "YD", col_types = "numeric")
GerzenseeOD <- read_excel("../data/SourceData/External/Gerzensee2005.xlsx", sheet = "OD", col_types = "numeric")

HerbsMer <- read_excel("../data/SourceData/External/Meerfeldemar_pollen.xlsx", sheet = "Herbs", col_types = "numeric")

PinusMer <- read_excel("../data/SourceData/External/Meerfeldemar_pollen.xlsx", col_types = "numeric", sheet = "Pinus")

BetulaMer <- read_excel("../data/SourceData/External/Meerfeldemar_pollen.xlsx", col_types = "numeric", sheet = "Betula")

PollenDat <- read_excel("../data/SourceData/Pollen_Litt_Usinger_stacked.xlsx", sheet = "LittLittStackedPerModel", col_types = "numeric")

PolDat<- PollenDat[3:230,]
APNAP <- select(PolDat, 1:2, 116, 115)
Trees <- select(PolDat, 1,2,8,41,54,55,71)

pollen_stacked_per <- melt(APNAP, var.names = names(APNAP)[3:ncol(APNAP)], measure.vars = names(APNAP)[3:ncol(APNAP)])

XRF_data <- read_excel("../data/XRF_clustered_PC.xlsx", col_types = "numeric")
HSI_data <- read_excel("../data/HZM_HR_cleaned.xlsx", col_types = "numeric")
LR_data <- read_excel("../data/SourceData/LR_data.xlsx", sheet = "Redox_Dia_PigOMug")
Redox <- select(LR_data, 3, 9:10, 13:14)

NGRIP <- read_excel("../data/SourceData/External/NGRIP.xlsx", col_types = "numeric") 

#NewScale <- read_excel("../data/SourceData/External/NGRIP.xlsx") 
base <- read_excel("../data/pig_alr.xlsx")
NewScale <- select(base, 2)

baseD <- read_excel("../data/diatom_filtered.xlsx") 
NewScaleD <- select(baseD, 2)

```

```{r}
# # Creating a new columns with converted HSI depths to the XRF depths (0 -> 10989)
# Source <- add_column(HSI_data, "Depth_LG_comp"= as.numeric(HSI_data$Sediment.Y..mm. + as.numeric(NewScale[1,"comp_depth_mm"])), .before = HSI_data$Sediment.Y..mm.) 

#Saving as a csv file
write.table(GDGT,file="../data/SourceData/External/csvs/GDGT.csv", sep=";",row.names = FALSE)

Egelsee_age <- select(Egelsee, 1, 3:6,8)
write.table(Egelsee_age,file="../data/SourceData/External/csvs/Egelsee.csv", sep=";", row.names = FALSE)

write.table(Ins,file="../data/SourceData/External/csvs/Insolation.csv", sep=";",row.names = FALSE)


Gerzensee <- merge(GerzenseeOD, GerzenseeYD, by = "age", all = T)
Gerzensee <- Gerzensee%>% replace(is.na(.), 0)
Gerzensee$JulyT <- Gerzensee$`July T.x`+ Gerzensee$`July T.y`
Gerzensee <- Gerzensee[,-2]
Gerzensee <- Gerzensee[,-2]
write.table(Gerzensee,file="../data/SourceData/External/csvs/Gerzensee.csv", sep=";",row.names = FALSE)

write.table(HerbsMer,file="../data/SourceData/External/csvs/MerHerbs.csv", sep=";",row.names = FALSE)
write.table(PinusMer,file="../data/SourceData/External/csvs/MerPinus.csv", sep=";",row.names = FALSE)
write.table(BetulaMer,file="../data/SourceData/External/csvs/MerBetula.csv", sep=";",row.names = FALSE)

HolzPol <- cbind(Trees, APNAP[,3:4])
write.table(HolzPol,file="../data/SourceData/External/csvs/HolzPol.csv", sep=";",row.names = FALSE)

XRF_data <- select(XRF_data, 26, 6, 7)
colnames(XRF_data)[1] <- 'age' 
write.table(XRF_data,file="../data/SourceData/External/csvs/XRF.csv", sep=";",row.names = FALSE)

write.table(HSI_data,file="../data/SourceData/External/csvs/HSI.csv", sep=";",row.names = FALSE)

write.table(Redox,file="../data/SourceData/External/csvs/Redox.csv", sep=";",row.names = FALSE)

write.table(NGRIP,file="../data/SourceData/External/csvs/NGRIP.csv", sep=";",row.names = FALSE)

#NewScale <- NewScale[,1]
#write.table(NewScale,file="../data/SourceData/External/csvs/NGRIPscale.csv", sep=";",row.names = FALSE)
write.table(NewScale,file="../data/SourceData/External/csvs/PigScale.csv", sep=";",row.names = FALSE)

write.table(NewScaleD,file="../data/SourceData/External/csvs/DiaScale.csv", sep=";",row.names = FALSE)
```

```{r}
# # Exctracting the XRF depth scale to a separate file (optional)
# Destin <- data.frame("depth"= as.integer(NewScale$comp_depth_mm))
# 
# #Saving as a csv file
# write.table(Destin,file="../data/XRF_scale.csv",sep=";", row.names = FALSE)
# 
# # Exctracting the XRF depth scale to a separate file (optional)
# Destin2 <- data.frame("depth"= as.integer(CleanedScale$`Composite depth (mm)`))
# 
# #Saving as a csv file
# write.table(Destin2,file="../data/XRF_scale_cleaned.csv",sep=";", row.names = FALSE)
```

# Fix interpolation

# Loading the prepared XRF and HSI data for interpolation

Here a base data needs to be selected as Destin. If we use diatoms, we need uncomment the DestinD line.

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/GDGT.csv", header=TRUE,sep=";")
# 
Destin <- read.csv(file="../data/SourceData/External/csvs/PigScale.csv", header=TRUE,sep=";")

#For other interplation, e.g., Diatoms, replace the Destin into the new name, e.g. DestinD in the following interpolation cells

# Destin <- read.csv(file="../data/SourceData/External/csvs/DiaScale.csv", header=TRUE,sep=";")

#DestinXRF <- read.csv(file="../data/XRF_scale.csv", header=TRUE,sep=";")
#Destin2 <- read.csv(file="../data/SourceData/External/csvs/NGRIP.csv", header=TRUE,sep=";") 
```

# Interpolating the data

```{r}
Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])
```

```{r}
GDGTinterpol <- Data
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/Gerzensee.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(GDGTinterpol, Data, by = "age_calBP", all = T)# here is some error as it changes the amount og observations
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/Egelsee.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/Insolation.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/NGRIP.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/XRF.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/Redox.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/HolzPol.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/MerBetula.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/MerPinus.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
```

```{r}
Source <- read.csv(file="../data/SourceData/External/csvs/MerHerbs.csv", header=TRUE, sep=";")

Names  <- colnames(Source)

#interpolate data for all columns
Data <- Destin
for (i in 2:ncol(Source)){
  temp     <- approx(Source[,1], Source[,i], xout=Destin[,1])
  Data[,i] <- temp$y
}

# Replace destination depth with interpolated depth (just to be sure it worked)
Data[,1] <- temp$x
# Replace column names with actual names 
colnames(Data) <- c("age_calBP",Names[2:length(Names)])

Interpol <- merge(Interpol, Data, by = "age_calBP", all = T)
#write.xlsx(Interpol, "../data/External_Interpol_pig.xlsx", rowNames = FALSE)
```

```{r}
forRDA <- select(Interpol, 1, 6, 7,9,15:17,19:26, 28:29, 31:38)
colnames(forRDA) <- c("age_calBP", "temp_GDGT", "JulyT_Ger", "JulyT_Egel", "insol_A", "insol_July", "insol_Jan", "18O_NGRIP", "Dust_NGRIP", "Ca", "Ti", "P_F2", "P_F3", "Mn_F2" , "Mn_F3","BetulaHZM", "JuniperusHZM", "PinusHZM", "SalixHZM", "NAP_HZM", "AP_HZM", "BetulaMer", "PinusMer", "NAP_Mer", "AP_Mer")


forRDA_pig <- select(Interpol, 1, 6, 7,9,15:17,19:22, 28:29, 31:38)
colnames(forRDA_pig) <- c("age_calBP", "temp_GDGT", "JulyT_Ger", "JulyT_Egel", "insol_A", "insol_July", "insol_Jan", "18O_NGRIP", "Dust_NGRIP", "Ca", "Ti","BetulaHZM", "JuniperusHZM", "PinusHZM", "SalixHZM", "NAP_HZM", "AP_HZM", "BetulaMer", "PinusMer", "NAP_Mer", "AP_Mer")

# forRDAD<- select(Interpol, 1, 6, 7,9,15:17,19:22, 28:29, 31:38)
# colnames(forRDAD) <- c("age_calBP", "temp_GDGT", "JulyT_Ger", "JulyT_Egel", "insol_A", "insol_July", "insol_Jan", "18O_NGRIP", "Dust_NGRIP", "Ca", "Ti","BetulaHZM", "JuniperusHZM", "PinusHZM", "SalixHZM", "NAP_HZM", "AP_HZM", "BetulaMer", "PinusMer", "NAP_Mer", "AP_Mer")
```

# Saving the interpolated HSI data

```{r}
 # write output file-
write.table(forRDA,file="../data/External_forRDA_pig_redox.csv",sep=";",row.names = FALSE)

# write.table(forRDAD,file="../data/External_forRDA_dia.csv",sep=";",row.names = FALSE)

write.xlsx(forRDA, "../data/External_forRDA_pig_redox.xlsx", rowNames = FALSE)

# write.xlsx(forRDAD, "../data/External_forRDA_dia.xlsx", rowNames = FALSE)

write.xlsx(forRDA_pig,file="../data/External_forRDA_pig.xlsx",rowNames = FALSE)
```

# Validation

Plots the old and the new data for comparison

```{r}
# for(j in 2:ncol(Source)){
# layout(matrix(c(1,2),1,2, byrow = TRUE))
# plot(Source[,j],-Source[,1], type="l", main="Original"    ,xlab=colnames(Source[j]),ylab=colnames(  Data[1]))
# plot(  Data[,j],  -Data[,1], type="l", main="interpolated",xlab=colnames(  Data[j]),ylab=colnames(  Data[1]))
# }
# 

#par(.def_par)

```
