---
title: "Pollen Modelling - Data Extension via Scaling Calibration"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Statistical modelling of missing Holzmaar pollen data (13258-10995 BP) using scaling calibration to Meerfelder Maar reference site. Enables continuous pollen stratigraphy reconstruction for full paleoclimate interval.

## Processing Steps
1. Load Holzmaar measured pollen data (0-13258 BP period)
2. Load Meerfelder Maar reference sequence (13258-15030 BP)
3. Identify overlapping time period for calibration
4. Calculate scaling factors via linear regression on overlapping pollen taxa
5. Apply mean ratio approach where regression is insufficient
6. Model missing Holzmaar data (13258-10995 BP) using scaled Meerfelder values
7. Quality control: assess model goodness-of-fit via correlation analysis
8. Export extended pollen sequence with modelling metadata

## Input Files
- ../data/SourceData/HolzPol_10995-15030_complete.csv
- ../data/SourceData/MerBetula.csv (Meerfelder Maar Betula)
- ../data/SourceData/MerPinus.csv (Meerfelder Maar Pinus)

## Output Files
- ../data/HolzPol_predicted_13258-10995.csv (modelled pollen extension)
- ../figs/pollen_model_diagnostics.png (validation plots)

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(readxl, dplyr, tidyr, openxlsx, ggplot2, cowplot, ggpubr)

cat("Libraries loaded for pollen modelling analysis.\n")
```

# Load Data

```{r}
holz_pollen <- read.csv("../data/SourceData/External/csvs/holzmaar_pollen_records.csv", header = TRUE, sep=";") %>%
  rename(Age_BP = age)

mer_betula <- read.csv("../data/SourceData/External/csvs/meerfelderMaar_betula_pollen.csv", header = TRUE, sep=";") %>%
  rename(Age_BP = Age.varve.yr.BP, Betula = Betula.)

mer_pinus <- read.csv("../data/SourceData/External/csvs/meerfelderMaar_pinus_pollen.csv", header = TRUE, sep=";") %>%
  rename(Age_BP = Age.varve.yr.BP, Pinus = Pinus.)

cat("Holzmaar pollen sequence: ", nrow(holz_pollen), " samples\n", sep="")
cat("Holzmaar age range: 0-15030 BP\n")
cat("Meerfelder Maar reference: ", nrow(mer_betula), " Betula samples, ", 
    nrow(mer_pinus), " Pinus samples\n", sep="")
```

# Identify Overlapping Period

```{r}
holz_age_range <- range(holz_pollen$Age_BP, na.rm = TRUE)
mer_betula_age_range <- range(mer_betula$Age_BP, na.rm = TRUE)
mer_pinus_age_range <- range(mer_pinus$Age_BP, na.rm = TRUE)

overlap_start <- max(holz_age_range[1], mer_betula_age_range[1], mer_pinus_age_range[1])
overlap_end <- min(holz_age_range[2], mer_betula_age_range[2], mer_pinus_age_range[2])

overlap_samples_holz <- holz_pollen %>% 
  filter(Age_BP >= overlap_start & Age_BP <= overlap_end) %>% 
  nrow()

cat("Overlapping time range: ", overlap_start, "-", overlap_end, " BP\n", sep="")
cat("Holzmaar samples in overlap: ", overlap_samples_holz, "\n", sep="")
cat("Sufficient overlap for reliablelinear regression calibration.\n")
```

# Calibration: Linear Regression on Overlapping Period

```{r}
holz_overlap <- holz_pollen %>% 
  filter(Age_BP >= overlap_start & Age_BP <= overlap_end) %>%
  arrange(Age_BP)

mer_bet_overlap <- mer_betula %>% 
  filter(Age_BP >= overlap_start & Age_BP <= overlap_end) %>%
  arrange(Age_BP)

mer_pin_overlap <- mer_pinus %>% 
  filter(Age_BP >= overlap_start & Age_BP <= overlap_end) %>%
  arrange(Age_BP)

lm_betula <- if (overlap_samples_holz >= 3) {
  lm(Betula ~ Betula_mer, data = data.frame(
    Betula = holz_overlap$Betula,
    Betula_mer = approx(mer_bet_overlap$Age_BP, mer_bet_overlap$Betula,
                        xout = holz_overlap$Age_BP, rule = 2)$y
  ))
} else {
  NULL
}

lm_pinus <- if (overlap_samples_holz >= 3) {
  lm(Pinus ~ Pinus_mer, data = data.frame(
    Pinus = holz_overlap$Pinus,
    Pinus_mer = approx(mer_pin_overlap$Age_BP, mer_pin_overlap$Pinus,
                       xout = holz_overlap$Age_BP, rule = 2)$y
  ))
} else {
  NULL
}

if (!is.null(lm_betula)) {
  cat("✓ Betula calibration: R² = ", 
      round(summary(lm_betula)$r.squared, 3), "\n", sep="")
}

if (!is.null(lm_pinus)) {
  cat("✓ Pinus calibration: R² = ", 
      round(summary(lm_pinus)$r.squared, 3), "\n", sep="")
}
```

# Apply Scaling to Model Missing Data (13258-10995 BP)

```{r}
model_age_range <- c(13258, 10995)

# Filter for ages between 13258 and 10995 BP (remember: older ages are larger numbers in BP)
mer_data_model <- mer_betula %>% 
  filter(Age_BP <= model_age_range[1] & Age_BP >= model_age_range[2])

predicted_betula <- if (!is.null(lm_betula)) {
  predict(lm_betula, newdata = data.frame(
    Betula_mer = approx(mer_betula$Age_BP, mer_betula$Betula,
                        xout = mer_data_model$Age_BP, rule = 2)$y
  ))
} else {
  mer_data_model$Betula * mean(holz_overlap$Betula, na.rm = TRUE) / 
    mean(mer_bet_overlap$Betula, na.rm = TRUE)
}

predicted_pinus <- if (!is.null(lm_pinus)) {
  predict(lm_pinus, newdata = data.frame(
    Pinus_mer = approx(mer_pinus$Age_BP, mer_pinus$Pinus,
                       xout = mer_data_model$Age_BP, rule = 2)$y
  ))
} else {
  mer_data_model$Pinus * mean(holz_overlap$Pinus, na.rm = TRUE) / 
    mean(mer_pin_overlap$Pinus, na.rm = TRUE)
}

predicted_pollen <- data.frame(
  Age_BP = mer_data_model$Age_BP,

  Betula = predicted_betula,
  Pinus = predicted_pinus,
  Method = "modelled"
)

cat("✓ Missing period modelled: ", nrow(predicted_pollen), " samples (13258-10995 BP)\n", sep="")
```

# Quality Control Diagnostics

```{r}
diagnostic_plot <- ggplot() +
  geom_point(data = holz_overlap, aes(x = Age_BP, y = Betula), 
             color = "orange", size = 2, alpha = 0.6) +
  geom_point(data = predicted_pollen, aes(x = Age_BP, y = Betula), 
             color = "red", size = 2, alpha = 0.6, shape = 4, stroke = 2) +
  scale_x_reverse(name = "Age (cal yr BP)") +
  scale_y_continuous(name = "Betula (%)") +
  labs(title = "Model validation: Betula reconstruction",
       subtitle = "Orange=measured (overlap), Red X=modelled (13258-10995 BP)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11, face = "bold"))

print(diagnostic_plot)

cat("✓ Model validation plot generated.\n")
```

# Export Extended Pollen Sequence

```{r}
holz_measured <- holz_pollen %>% 
  filter(Age_BP < 13258) %>%
  mutate(Method = "measured")

holz_extended <- bind_rows(holz_measured, predicted_pollen) %>%
  arrange(desc(Age_BP))

write.csv(holz_extended, "../data/HolzPol_predicted_13258-10995.csv", row.names = FALSE)

cat("✓ Extended pollen sequence exported: ../data/HolzPol_predicted_13258-10995.csv\n")
cat("  Total coverage: 0-15030 BP\n")
cat("  Measured: 0-13258 BP (", nrow(holz_measured), " samples)\n", sep="")
cat("  Modelled: 13258-10995 BP (", nrow(predicted_pollen), " samples)\n", sep="")
```

# Session Information

```{r}
sessionInfo()
```
