---
title: "Sequential Extraction Plots - Fe, Mn, P Fractionation"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Visualize sequential extraction results for redox-sensitive elements (Fe, Mn, P). Create stacked bar plots showing fraction distribution (F1: exchangeable, F2: Fe/Mn carbonate, F3: Fe/Mn oxides, F4: organic) and generate molar ratios to assess redox conditions.

## Processing Steps
1. Load sequential extraction dataset with element fractions
2. Prepare absolute concentration data for element speciation
3. Calculate percentage distributions across fractions
4. Create 100% stacked bar plots for each element
5. Calculate Fe/P molar ratios (indicator of redox state)
6. Overlay sediment pigment proxies (Okenone, bacteriopheophytin)
7. Generate combined publication figures

## Input Files
- ../data/SourceData/Sequential_Extraction.xlsx
- ../data/HZM_HR_cleaned.xlsx (HSI pigment data)

## Output Files
- ../figs/Fe_sequential_extraction_*.png
- ../figs/Mn_sequential_extraction_*.png
- ../figs/P_sequential_extraction_*.png
- ../figs/Fe_P_molar_ratio_*.png
- ../figs/Combined_redox_*.png

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(readxl, dplyr, tidyr, openxlsx, ggplot2, cowplot, viridis, ggpubr, reshape2)

cat("Libraries loaded for sequential extraction analysis.\n")
```

# Helper Functions

```{r}
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}

cat("Helper functions defined.\n")
```

# Load Data

```{r}
se_data <- read_excel("../data/SourceData/Sequential_Extraction.xlsx", col_types = "numeric")
hsi_data <- read_excel("../data/HZM_HR_cleaned.xlsx", col_types = "numeric")

cat("Sequential extraction data loaded: ", nrow(se_data), " samples\n", sep="")
cat("HSI pigment data loaded: ", nrow(hsi_data), " measurements\n", sep="")
```

# Prepare Element Fractionation Data

Sequential extraction fractions: F1 (exchangeable), F2 (carbonate), F3 (Fe/Mn oxides), F4 (organic/residual).

```{r}
extract_fractions <- function(se_data, element) {
  f1_col <- paste0(element, "_F1")
  f2_col <- paste0(element, "_F2")
  f3_col <- paste0(element, "_F3")
  f4_col <- paste0(element, "_F4")
  
  fractions <- data.frame(
    F1_exch = se_data[[f1_col]],
    F2_carb = se_data[[f2_col]],
    F3_oxide = se_data[[f3_col]],
    F4_org = se_data[[f4_col]]
  )
  
  fractions$Total <- rowSums(fractions, na.rm = TRUE)
  fractions$Pct_F1 <- 100 * fractions$F1_exch / fractions$Total
  fractions$Pct_F2 <- 100 * fractions$F2_carb / fractions$Total
  fractions$Pct_F3 <- 100 * fractions$F3_oxide / fractions$Total
  fractions$Pct_F4 <- 100 * fractions$F4_org / fractions$Total
  
  return(fractions)
}

Fe_frac <- extract_fractions(se_data, "Fe")
Mn_frac <- extract_fractions(se_data, "Mn")
P_frac <- extract_fractions(se_data, "P")

cat("✓ Sequential fractions extracted:\n")
cat("  Fe: total μmol/g = ", round(mean(Fe_frac$Total, na.rm = TRUE), 2), "\n", sep="")
cat("  Mn: total μmol/g = ", round(mean(Mn_frac$Total, na.rm = TRUE), 2), "\n", sep="")
cat("  P:  total μmol/g = ", round(mean(P_frac$Total, na.rm = TRUE), 2), "\n", sep="")
```

# Create 100% Stacked Bar Plots

```{r}
create_stacked_plot <- function(fractions, element_name, depth_m) {
  plot_data <- data.frame(
    depth = depth_m,
    F1_exch = fractions$Pct_F1,
    F2_carb = fractions$Pct_F2,
    F3_oxide = fractions$Pct_F3,
    F4_org = fractions$Pct_F4
  )
  
  plot_data_long <- plot_data %>%
    pivot_longer(cols = -depth, names_to = "Fraction", values_to = "Percentage") %>%
    mutate(Fraction = factor(Fraction, levels = c("F1_exch", "F2_carb", "F3_oxide", "F4_org"),
                              labels = c("F1: Exchangeable", "F2: Carbonate", 
                                        "F3: Fe/Mn Oxide", "F4: Organic")))
  
  p <- ggplot(plot_data_long, aes(x = depth, y = Percentage, fill = Fraction)) +
    geom_bar(stat = "identity", width = 0.8) +
    scale_y_continuous(name = "Percentage (%)", limits = c(0, 100)) +
    scale_x_reverse(name = "Depth (m)") +
    scale_fill_viridis_d(option = "H") +
    labs(title = paste0(element_name, " Sequential Extraction - 100% Stacked")) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold"),
      legend.position = "bottom"
    )
  
  return(p)
}

depth_m <- se_data$Depth_mm / 1000

p_Fe_stacked <- create_stacked_plot(Fe_frac, "Iron (Fe)", depth_m)
p_Mn_stacked <- create_stacked_plot(Mn_frac, "Manganese (Mn)", depth_m)
p_P_stacked <- create_stacked_plot(P_frac, "Phosphorus (P)", depth_m)

cat("✓ 100% stacked bar plots created for Fe, Mn, P\n")
```

# Calculate Fe/P Molar Ratio

Fe/P molar ratio indicates redox state: elevated ratios suggest oxidizing conditions.

```{r}
fe_p_ratio <- data.frame(
  depth = depth_m,
  Fe_P_molar = Fe_frac$Total / P_frac$Total,
  Fe_P_F2 = Fe_frac$F2_carb / P_frac$F2_carb,
  Fe_P_F3 = Fe_frac$F3_oxide / P_frac$F3_oxide
)

p_Fe_P <- ggplot(fe_p_ratio, aes(x = Fe_P_molar, y = depth)) +
  geom_path(color = "darkred", size = 0.5) +
  geom_point(size = 1.5, color = "darkred", alpha = 0.6) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Fe/P Molar Ratio") +
  labs(title = "Redox State Indicator - Fe/P Molar Ratio",
       subtitle = "Higher values indicate more oxidizing conditions") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 9),
    plot.title = element_text(size = 11, face = "bold")
  )

cat("✓ Fe/P molar ratio plot created\n")
```

# Combine with Pigment Proxy Data

```{r}
hsi_redox <- data.frame(
  depth = hsi_data$depth_VT22 / 1000,
  Okenone = if ("Okenone" %in% names(hsi_data)) hsi_data$Okenone else NA,
  Bphe = if ("Bphe_calibrated" %in% names(hsi_data)) hsi_data$Bphe_calibrated else NA
)

p_combined <- ggplot() +
  geom_path(data = hsi_redox, aes(x = Bphe, y = depth, color = "Bacteriopheophytin"), 
            size = 0.6) +
  geom_path(data = fe_p_ratio, aes(x = Fe_P_molar, y = depth, color = "Fe/P Ratio"),
            size = 0.6) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Normalized Value (scaled)") +
  scale_color_manual(values = c("Bacteriopheophytin" = "purple", "Fe/P Ratio" = "darkred")) +
  labs(title = "Combined Redox Indicators",
       subtitle = "Bacteriopheophytin (organic sedimentation) and Fe/P ratio (redox state)") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 9),
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "right"
  )

cat("✓ Combined redox indicator plot created\n")
```

# Export Sequential Extraction Plots

```{r}
png(filename = timestamp_filename("../figs", "Fe_sequential_extraction"),
    width = 800, height = 600, units = "px", res = 300, pointsize = 12)
print(p_Fe_stacked)
dev.off()

png(filename = timestamp_filename("../figs", "Mn_sequential_extraction"),
    width = 800, height = 600, units = "px", res = 300, pointsize = 12)
print(p_Mn_stacked)
dev.off()

png(filename = timestamp_filename("../figs", "P_sequential_extraction"),
    width = 800, height = 600, units = "px", res = 300, pointsize = 12)
print(p_P_stacked)
dev.off()

png(filename = timestamp_filename("../figs", "Fe_P_molar_ratio"),
    width = 800, height = 600, units = "px", res = 300, pointsize = 12)
print(p_Fe_P)
dev.off()

png(filename = timestamp_filename("../figs", "Combined_redox"),
    width = 1000, height = 600, units = "px", res = 300, pointsize = 12)
print(p_combined)
dev.off()

cat("✓ All sequential extraction plots exported: ../figs/\n")
cat("  - Fe_sequential_extraction_*.png\n")
cat("  - Mn_sequential_extraction_*.png\n")
cat("  - P_sequential_extraction_*.png\n")
cat("  - Fe_P_molar_ratio_*.png\n")
cat("  - Combined_redox_*.png\n")
```

# Session Information

```{r}
sessionInfo()
```
