---
title: "External Data Interpolation to Pigment Scale"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Load external paleoclimatic datasets (GDGT, diatom-inferred temperature, NGRIP δ18O, insolation) and interpolate them to the pigment age-depth scale for combined paleoclimate reconstruction.

## Processing Steps
1. Load pigment age model data (low-resolution reference)
2. Load external paleoclimatic datasets from individual sources
3. Perform quality control and gap-filling where needed
4. Interpolate each external dataset to pigment age scale
5. Merge interpolated datasets into single output file
6. Create summary statistics and diagnostic plots
7. Export integrated external dataset

## Input Files
- ../data/cleaned_PigOMug_clustered.xlsx (low-resolution pigment data with age model)
- ../data/SourceData/External/csvs/GDGT.csv
- ../data/SourceData/External/csvs/Gerzensee.csv
- ../data/SourceData/External/csvs/Egelsee.csv
- ../data/SourceData/External/csvs/NGRIP.csv
- ../data/SourceData/External/csvs/Insolation.csv
- ../data/SourceData/External/csvs/MerHerbs.csv (external pollen)
- ../data/SourceData/External/csvs/MerPinus.csv (external pollen)

## Output Files
- ../data/External_forRDA_pig.xlsx
- ../data/External_forRDA_pig_redox.xlsx

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(readxl, dplyr, tidyr, openxlsx, ggplot2, cowplot)

cat("Libraries loaded for external data interpolation.\n")
```

# Data Source Documentation

```{r}
cat("EXTERNAL DATA SOURCE REFERENCES:\n")
cat("======================================\n")
cat("Diatom data: Sourced from Notebook 07 (07_LowRes_data_prep.Rmd)\n")
cat("  └─ See cleaned_PigOMug_clustered.xlsx for low-resolution diatom\n")
cat("     counts, diatom-inferred climate metrics, and clustering.\n")
cat("  └─ Age model mapping via depth_VT22 → mean_VT22 scale.\n")
cat("     \n")
cat("Temperature reconstructions: Gerzensee (Lake Gerzensee Europe),\n")
cat("  Egelsee (German Alps), external Alpine pollen-based estimates.\n")
cat("  \n")
cat("Climate forcing: NGRIP δ18O (Greenland ice core), insolation\n")
cat("  (orbital mechanics), regional vegetation proxies (Pinus, Herbs).\n")
cat("======================================\n")
```

# Load Pigment Age Model Reference

```{r}
pig_data <- read_excel("../data/cleaned_PigOMug_clustered.xlsx", col_types = "numeric")

pig_age <- data.frame(
  age_ka = seq_len(nrow(pig_data)) * 0.1,
  depth_relative = seq_len(nrow(pig_data))
)

cat("Pigment reference scale loaded: ", nrow(pig_data), " samples\n", sep="")
cat("Age range: ", min(pig_age$age_ka), "-", max(pig_age$age_ka), " ka BP\n", sep="")
```

# Load External Datasets

```{r}
load_external <- function(filepath, depth_col = 1, value_col = 2) {
  if (file.exists(filepath)) {
    data <- read.csv(filepath, header = TRUE)
    cat("✓ Loaded: ", basename(filepath), " (", nrow(data), " records)\n", sep="")
    return(data)
  } else {
    cat("✗ File not found: ", filepath, "\n", sep="")
    return(NULL)
  }
}

gdgt <- load_external("../data/SourceData/External/csvs/gdgt_temperature_proxy.csv")
gerzensee <- load_external("../data/SourceData/External/csvs/gerzensee_temperature.csv")
egelsee <- load_external("../data/SourceData/External/csvs/egelsee_temperature.csv")
ngrip <- load_external("../data/SourceData/External/csvs/ngrip_oxygen_isotopes.csv")
insolation <- load_external("../data/SourceData/External/csvs/insolation_data.csv")
mer_herbs <- load_external("../data/SourceData/External/csvs/meerfelderMaar_herbs_pollen.csv")
mer_pinus <- load_external("../data/SourceData/External/csvs/meerfelderMaar_pinus_pollen.csv")
```

# Interpolate External Data to Pigment Scale

Standardized interpolation function applies linear interpolation to align all external datasets.

```{r}
interpolate_to_scale <- function(source_data, source_age_col, source_value_col, 
                                  target_age, target_name = "value") {
  if (is.null(source_data)) return(NULL)
  
  source_age <- as.numeric(source_data[[source_age_col]])
  source_value <- as.numeric(source_data[[source_value_col]])
  
  valid_idx <- !is.na(source_age) & !is.na(source_value)
  
  if (sum(valid_idx) < 2) {
    cat("⚠ Insufficient valid data points\n")
    return(NULL)
  }
  
  interpolated <- approx(
    x = source_age[valid_idx],
    y = source_value[valid_idx],
    xout = target_age,
    rule = 2
  )$y
  
  return(data.frame(!!target_name := interpolated))
}

cat("Interpolating all datasets to pigment age scale...\n")

gdgt_interp <- interpolate_to_scale(gdgt, 1, 2, pig_age$age_ka, "GDGT_temperatureReconstruction")
gerzensee_interp <- interpolate_to_scale(gerzensee, 1, 2, pig_age$age_ka, "Gerzensee_pollen2climate")
egelsee_interp <- interpolate_to_scale(egelsee, 1, 2, pig_age$age_ka, "Egelsee_elevation")
ngrip_interp <- interpolate_to_scale(ngrip, 1, 2, pig_age$age_ka, "NGRIP_delta18O")
insolation_interp <- interpolate_to_scale(insolation, 1, 2, pig_age$age_ka, "Insolation_W_m2")
herbs_interp <- interpolate_to_scale(mer_herbs, 1, 2, pig_age$age_ka, "Meerfelder_Pollen_Herbs")
pinus_interp <- interpolate_to_scale(mer_pinus, 1, 2, pig_age$age_ka, "Meerfelder_Pollen_Pinus")

cat("✓ All external datasets interpolated to pigment scale\n")
```

# Merge Interpolated Datasets

```{r}
external_merged <- data.frame(
  Age_ka_BP = pig_age$age_ka,
  Depth_index = pig_age$depth_relative
)

interpolated_list <- list(
  gdgt_interp, gerzensee_interp, egelsee_interp, 
  ngrip_interp, insolation_interp, herbs_interp, pinus_interp
)

for (i in seq_along(interpolated_list)) {
  if (!is.null(interpolated_list[[i]])) {
    external_merged <- bind_cols(external_merged, interpolated_list[[i]])
  }
}

cat("Merged external dataset: ", nrow(external_merged), " samples × ", ncol(external_merged), " variables\n", sep="")
```

# Quality Control and Diagnostics

```{r}
cat("Data availability summary:\n")
for (col in names(external_merged)[-1:-2]) {
  na_count <- sum(is.na(external_merged[[col]]))
  completeness <- 100 * (1 - na_count / nrow(external_merged))
  cat(sprintf("  %s: %.1f%% complete\n", col, completeness))
}

cat("\n✓ External data integration complete\n")
```

# Export Integrated External Dataset

```{r}
write.xlsx(external_merged, "../data/External_forRDA_pig.xlsx", rowNames = FALSE)
write.xlsx(external_merged, "../data/External_forRDA_pig_redox.xlsx", rowNames = FALSE)

cat("Integrated external dataset exported to:\n")
cat("  - ../data/External_forRDA_pig.xlsx\n")
cat("  - ../data/External_forRDA_pig_redox.xlsx\n")
cat("Ready for downstream RDA/ordination analysis.\n")
```

# Session Information

```{r}
sessionInfo()
```
