---
title: "Combined Figure: Discussion Plot and RDA"
output: html_notebook
---

# Combined Figure Creation

This notebook combines:
- Panel A: discussion_multiplot_v2_with_pigments_panel.png (2/3 A4 width = 14 cm)
- Panel B: RDA1-2_pigments_PolZ_communities_publication.png (1/3 A4 width = 7 cm)

Both panels at 1/3 A4 height (9.9 cm), arranged side by side.

```{r setup}
library(png)
library(grid)
library(gridExtra)
```

# Load the Images

```{r load-images}
# Load the two PNG files - using the panel-sized version of discussion plot
discussion_img <- readPNG("../figs/discussion_multiplot_v2_with_pigments_panel.png")
rda_img <- readPNG("../figs/RDA1-2_pigments_PolZ_communities_publication.png")
```

# Create Combined Figure

```{r combined-figure, fig.width=21/2.54, fig.height=7/2.54}
# A4 width = 21 cm, height for 1/3 = ~7 cm (A4 height is 29.7 cm, 1/3 ≈ 9.9 cm)
# But keeping proportional to your data: 1/3 A4 height ≈ 7-10 cm
# Using 7 cm for better aspect ratio

# Convert images to grobs
discussion_grob <- rasterGrob(discussion_img, interpolate = TRUE)
rda_grob <- rasterGrob(rda_img, interpolate = TRUE)

# Create layout: Panel A (2/3 width) + Panel B (1/3 width)
# Add panel labels A and B
grid.arrange(
  arrangeGrob(discussion_grob, 
              top = textGrob("A", x = 0.02, hjust = 0, 
                            gp = gpar(fontsize = 14, fontface = "bold"))),
  arrangeGrob(rda_grob, 
              top = textGrob("B", x = 0.02, hjust = 0, 
                            gp = gpar(fontsize = 14, fontface = "bold"))),
  ncol = 2,
  widths = c(2, 1)  # 2/3 and 1/3 proportions
)
```

# Save Combined Figure

```{r save-figure}
# A4 width = 21 cm
# Height: 1/3 A4 = 9.9 cm (29.7/3)

# Calculate pixels at 300 DPI
width_cm <- 21
height_cm <- 9.9  # Exact 1/3 of A4 height
dpi <- 300

width_px <- width_cm / 2.54 * dpi
height_px <- height_cm / 2.54 * dpi

png(filename = "../figs/combined_figure_discussion_RDA.png", 
    width = width_px, height = height_px, units = "px", res = dpi, bg = "transparent")

# Create layout with panel labels (without background boxes)
grid.newpage()

# Panel A
pushViewport(viewport(x = 0, y = 0, width = 2/3, height = 1, just = c("left", "bottom")))
grid.raster(discussion_img, interpolate = TRUE)
popViewport()

# Panel A label (without background)
pushViewport(viewport(x = 0.01, y = 0.98, width = 0.05, height = 0.05, just = c("left", "top")))
grid.text("A", gp = gpar(fontsize = 14, fontface = "bold", col = "black"))
popViewport()

# Panel B
pushViewport(viewport(x = 2/3, y = 0, width = 1/3, height = 1, just = c("left", "bottom")))
grid.raster(rda_img, interpolate = TRUE)
popViewport()

# Panel B label (without background)
pushViewport(viewport(x = 2/3 + 0.01, y = 0.98, width = 0.05, height = 0.05, just = c("left", "top")))
grid.text("B", gp = gpar(fontsize = 14, fontface = "bold", col = "black"))
popViewport()

dev.off()

cat("Combined figure saved to: ../figs/combined_figure_discussion_RDA.png\n")
cat(sprintf("Dimensions: %.0f × %.0f pixels (%.1f × %.1f cm at 300 DPI)\n", 
            width_px, height_px, width_cm, height_cm))
```

# Alternative Version with Tighter Layout

If you need tighter spacing between panels:

```{r save-figure-tight}
# Version with minimal spacing between panels
png(filename = "../figs/combined_figure_discussion_RDA_tight.png", 
    width = width_px, height = height_px, units = "px", res = dpi)

# Create a viewport layout for more control
grid.newpage()

# Panel A label
pushViewport(viewport(x = 0.01, y = 0.98, width = 0.05, height = 0.05, 
                      just = c("left", "top")))
grid.text("A", gp = gpar(fontsize = 14, fontface = "bold"))
popViewport()

# Panel A: Discussion plot (0 to 2/3 of width)
pushViewport(viewport(x = 0, y = 0, width = 2/3, height = 1, 
                      just = c("left", "bottom")))
grid.raster(discussion_img, interpolate = TRUE)
popViewport()

# Panel B label
pushViewport(viewport(x = 2/3 + 0.01, y = 0.98, width = 0.05, height = 0.05, 
                      just = c("left", "top")))
grid.text("B", gp = gpar(fontsize = 14, fontface = "bold"))
popViewport()

# Panel B: RDA plot (2/3 to end of width)
pushViewport(viewport(x = 2/3, y = 0, width = 1/3, height = 1, 
                      just = c("left", "bottom")))
grid.raster(rda_img, interpolate = TRUE)
popViewport()

dev.off()

cat("Tight version saved to: ../figs/combined_figure_discussion_RDA_tight.png\n")
```

# Custom Dimensions Version

If you need exact 1/3 A4 height (9.9 cm):

```{r save-figure-exact}
# Exact A4 proportions
width_cm_exact <- 21
height_cm_exact <- 29.7 / 3  # Exactly 1/3 of A4 height = 9.9 cm

width_px_exact <- width_cm_exact / 2.54 * dpi
height_px_exact <- height_cm_exact / 2.54 * dpi

png(filename = "../figs/combined_figure_discussion_RDA_exact.png", 
    width = width_px_exact, height = height_px_exact, units = "px", res = dpi)

grid.newpage()

# Panel A label
pushViewport(viewport(x = 0.01, y = 0.98, width = 0.05, height = 0.05, 
                      just = c("left", "top")))
grid.text("A", gp = gpar(fontsize = 14, fontface = "bold"))
popViewport()

# Panel A: Discussion plot
pushViewport(viewport(x = 0, y = 0, width = 2/3, height = 1, 
                      just = c("left", "bottom")))
grid.raster(discussion_img, interpolate = TRUE)
popViewport()

# Panel B label
pushViewport(viewport(x = 2/3 + 0.01, y = 0.98, width = 0.05, height = 0.05, 
                      just = c("left", "top")))
grid.text("B", gp = gpar(fontsize = 14, fontface = "bold"))
popViewport()

# Panel B: RDA plot
pushViewport(viewport(x = 2/3, y = 0, width = 1/3, height = 1, 
                      just = c("left", "bottom")))
grid.raster(rda_img, interpolate = TRUE)
popViewport()

dev.off()

cat("Exact dimensions version saved to: ../figs/combined_figure_discussion_RDA_exact.png\n")
cat(sprintf("Dimensions: %.0f × %.0f pixels (%.1f × %.1f cm at 300 DPI)\n", 
            width_px_exact, height_px_exact, width_cm_exact, height_cm_exact))
```
