---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Loading packages

```{r}
# install.packages("pacman")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library("pacman")

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load("vegan","factoextra", "FactoMineR", "ggsci", "readxl", "dplyr"," ggplot2", "grid", "ggvegan", "BiodiversityR", "ggrepel","ggforce", "tibble","tidyr","openxlsx","car","rstatix", "DT", "data.table","reshape2", "psych", "cluster", "dendroextras", "rioja", "mclust", "corto", "corrplot","ggpubr", "vegan","bruceR")# "magrittr",  , "rcartocolor", "latex2exp", "ggpubr", "corrplot", "viridis", "cowplot", "janitor", "ggtext",  "gapminer", "forcats", "ggthemes", "ggmap", "maps")
# # data(varespec)
# # data(varechem)

```

# Loading data

Pigments or Diatoms as the base from LR_data

```{r}
base <- read_excel("../data/pig_alr.xlsx")

# Using diatoms as a base requires also new interpolation of the external data (notebook 08)
baseD <- read_excel("../data/diatom_filtered.xlsx") 
baseD <- as.data.frame(cbind("age"= baseD$mean_VT22, scale(baseD[,3:18])))

#baseAll <- merge(base, baseD, by = "age", all = T )
```

Loading the external data from already downsampled file (see notebook 08):

Pollen from Pollen_Litt_Usinger, GDGT temp from Auel_ELSA_stack_GDGT, 18O and dust from NGRIP, Insulation and summer T from Chironimids from <https://vo.imcce.fr/insola/earth/online/earth/online/index.php>

All the external data are then downscale/interpolated to the resolution of the pigment dataset. It can be rerun on the diatom base (use the 08 notebook).

```{r}
ext_all <- read_excel("../data/External_forRDA_pig.xlsx")
Names  <- colnames(ext_all)
colnames(ext_all) <- c("age",Names[2:length(Names)])

ext_dia <- read_excel("../data/External_forRDA_dia.xlsx")
Names  <- colnames(ext_dia)
colnames(ext_dia) <- c("age",Names[2:length(Names)])

ext <- ext_all
ext_all <- select(ext_all, 1:17) #removing Meerfeldemaar data, as they have Nans and shorten our dataset during cleaning

ext_dia <- select(ext_dia, 1:17)


ext_rex <- read_excel("../data/External_forRDA_pig_redox.xlsx")
Names  <- colnames(ext_rex)
colnames(ext_rex) <- c("age",Names[2:length(Names)])
ext_red <- select(ext_rex, 1:20)
```

Loading the primary producers communities from the correlation based heatmap clustering (notebook 07) and conducting CONISS clustering on the 3 representative communities.

```{r}
ProducerG <-  read_excel("../data/pig_alr_clust.xlsx")
ProdG <- select(ProducerG, 23, 24, 25)

clust <- chclust(dist(ProdG, method="euclidean"), method="coniss") 
fviz_dend(clust, k = 3, # Cut in four groups 
          cex = 0.5, # label size 
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07","#6e57d2", "yellow", "green", "black"), color_labels_by_k = TRUE, # color labels by groups 
          rect = TRUE # Add rectangle around groups 
          )

clustercut<-cbind(1:94,cutree(clust,3)) 
ProdG <- bind_cols(ProducerG[,2] , ProdG, clustercut) 
colnames(ProdG)[5] <- 'orderCONISS' 
colnames(ProdG)[6] <- 'CONISScluster'

ProdG_cl <- select(ProdG, 1, 6)
```

## Preparing all the datasets for RDA - merging, cleaning, splitting

```{r}
groups <- read_excel("../data/Pig_clustered_PC.xlsx")
groups <- groups%>% mutate(PZn = case_when(PolZ == "Pleniglacial" ~ "1",
                           PolZ == "Bølling" ~ "2",
                           PolZ == "Allerød" ~ "3",
                           PolZ == "YoungerDryas" ~ "4")
   )
# Column 2 is mean_VT22 (age), column 23 is PolZ
groups <- select(groups, 2, 23)
colnames(groups) <- c("age","PolZ")

groups <- merge(groups, ProdG_cl)

cleaning_full <-  merge(base, ext, by = "age", all = T)
cleaning_full <- merge(groups, cleaning_full, by = "age", all = T)
cleaning <- merge(base, ext_all, by = "age", all = T)
cleaning <- merge(groups, cleaning, by = "age", all = T)

cleaningRed <- merge(base, ext_red, by = "age", all = T)
cleaningRed <- merge(groups, cleaningRed, by = "age", all = T)
```

```{r}
dia_base <- baseD %>% mutate(Zone = case_when(age > 14444 ~ "Pleniglacial",
                          age  <= 14444 & age  >= 13611  ~ "Bølling",
                          age  <= 13611 & age  >= 12727 & age!=13011  ~ "Allerød", 
                          age  == 13011 ~ "LST", 
                          age <= 12727 ~ "Younger Dryas"))


groupD <- as.data.frame(select(dia_base, 1, 18))
groupD <- groupD%>% mutate(PolZ = case_when(Zone == "Pleniglacial" ~ "1",
                           Zone == "Bølling" ~ "2",
                           Zone == "Allerød" ~ "3",
                           Zone == "YoungerDryas" ~ "4"))
                           
colnames(groupD) <- c("age","PolZ")

cleaning_D <- merge(baseD, ext_dia, by = "age", all =T)

#cleaningAll <- merge(baseAll, ext_all, by ="age", all =T)
```

### Removing Nans

```{r}
clean <- na.omit(cleaning)
cleanD <- na.omit(cleaning_D)
#cleanAll <- na.omit(cleaningAll) - only 8 observation left as the diatoms are samples at different ages(depths)

cleanRed <- na.omit(cleaningRed)

groupsRed <- select(cleanRed, 1, 2)
```

```{r}
#Base layer of pigments
base_cl <- select(clean, 5:24)
#external variables for the pigments
ext_cl <- select(clean, 25:40)
time <- select(clean, 2) # age/Pollen zone
pigG <- select(clean, 3) # pigment producerd groups

# base layer of daitoms
baseD_cl <- select(cleanD, 2:17) 
#external variables scaled to diatoms resolution
extD_cl <- select(cleanD, 18:33)

# base layer of pigments on the Redox data resolution
base_cl_rex <- select(cleanRed, 5:24)
ext_cl_rex <- select(cleanRed, 25:43) #48
time_rex <- select(cleanRed, 2) #pollen
pigG_rex <- select(cleanRed, 3)

base_cl_rexPBA <- cleanRed[31:86,5:24]
ext_cl_rexPBA <- cleanRed[31:86,25:43]
time_PBA <- as.data.frame(cleanRed[31:86,2])
colnames(time_PBA)[1] <- "PolZ" 
pigG_PBA <- as.data.frame(cleanRed[31:86, 3])
colnames(pigG_PBA)[1] <- "CONISScluster"
```

Selection of Bölling-Alleröd and Younger Dryas data only - to focus on anoxia

```{r}
cleanBAY <- filter(cleaning_full, between(age, 12607, 14200 ))
cleanBAY 
base_cl_BAY <- select(cleanBAY, 5:24)
ext_cl_BAY <- select(cleanBAY, 25:44)
time_BAY <- select(cleanBAY, 2)
pigG_BAY <- select(cleanBAY, 3)
```

# Testing the relationships of the external variables

The external variables should not have significant correlations, otherwise the dataset should be reduced only to the non-correlated varibales.

## Conducting PCA to identify relationships between the external variables

```{r}
# All extarnal data
ext.PCA <- prcomp(ext_cl, scale. =T)
summary(ext.PCA)

var<- get_pca_var(ext.PCA)
get_eig(ext.PCA)

png(filename = "../figs/ext_scaled_PCA_explanatory_power.png", width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(ext.PCA, addlabels = TRUE, ylim = c(0, 55), title = "Pigments_external")
dev.off()

#------------External data in the diatoms extend
ext.PCAD <- prcomp(extD_cl, scale. =T)
summary(ext.PCAD)

varD<- get_pca_var(ext.PCAD)
get_eig(ext.PCAD)
png(filename = "../figs/ext_dia_PCA_explanatory_power.png", width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(ext.PCAD, addlabels = TRUE, ylim = c(0, 55), title = "Diatoms_external")
dev.off()

#---------------Externals only Bölling/Alleröd/Younger Dryas
ext_cl_BAY <- na.omit(ext_cl_BAY)
ext.PCA_BAY <- prcomp(ext_cl_BAY, scale. =T)
summary(ext.PCA_BAY)

varBAY<- get_pca_var(ext.PCA_BAY)
get_eig(ext.PCA_BAY)

png(filename = "../figs/ext_BAY_explanatory_power.png", width = 800, height = 500, units = "px", pointsize = 12)
fviz_screeplot(ext.PCA_BAY, addlabels = TRUE, ylim = c(0, 55), title = "Bo-Al-YD-pigments_external")
dev.off()


#------------------External with redox included, in redox eztend
ext.PCA_Rex <- prcomp(ext_cl_rex, scale. =T)
summary(ext.PCA_Rex)

varRed<- get_pca_var(ext.PCA_Rex)
get_eig(ext.PCA_Rex)

png(filename = "../figs/ext_red_PCA_explanatory_power.png", width = 800, height = 500, units = "px", pointsize = 12)

fviz_screeplot(ext.PCA_Rex, addlabels = TRUE, ylim = c(0, 55), title = "PCA with Redox")

dev.off()
```

```{r}
biplot <- fviz_pca_var(ext.PCA, col.var="contrib", axes = c(1,2), 
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title = "Pigments externals")

biplotD <- fviz_pca_var(ext.PCAD, col.var="contrib", axes = c(1,2),
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title = "Diatoms externals")

biplotBAY <- fviz_pca_var(ext.PCA_BAY, col.var="contrib", axes = c(1,2),
            gradient.cols = c("darkred", "orange", "forestgreen"),
            repel = TRUE, title = "Bo-Al-YD-pigments_external")

biplotRex <- fviz_pca_var(ext.PCA_Rex, col.var="contrib", axes = c(1,2),
             gradient.cols = c("darkred", "orange", "forestgreen"),
             repel = TRUE, title = "Redox pigments external")

# continue here
png(filename = "../figs/biplot_ext_pig.png", width = 800, height = 500, units = "px", pointsize = 12)
biplot
dev.off()

png(filename = "../figs/biplot_ext_dia.png", width = 800, height = 500, units = "px", pointsize = 12)
biplotD
dev.off()

png(filename = "../figs/biplot_ext_BAY.png", width = 800, height = 500, units = "px", pointsize = 12)
biplotBAY
dev.off()

png(filename = "../figs/biplot_ext_Redox.png", width = 800, height = 500, units = "px", pointsize = 12)
biplotRex
dev.off()

png(filename = "../figs/ext_PCAs.png", width = 1600, height = 1000, units = "px", pointsize = 12)
ggarrange(biplot, biplotD, biplotBAY, biplotRex, ncol = 2, nrow = 2)
dev.off()
```

```{r}
#contributors to PCA ext pigment scaled
fviz_contrib(ext.PCA, choice = "var", axes = 1, top = 12)
fviz_contrib(ext.PCA, choice = "var", axes = 2, top = 12)

#contributors to PCA ext diatom scaled
fviz_contrib(ext.PCAD, choice = "var", axes = 1, top = 12)
fviz_contrib(ext.PCAD, choice = "var", axes = 2, top = 12)
```

## Testing correlation of the external variables

```{r}
res <- cor.mtest(ext_cl, conf.level = .95)
corrplot(cor(ext_cl), type = "upper", p.mat = res$p, insig = "label_sig", sig.level = c(.001, .01, .05), pch.cex = 0.8, pch.col = "yellow", tl.col="black")
corrplot(var$cor, is.corr=FALSE)

resD <- cor.mtest(extD_cl, conf.level = .95)
corrplot(cor(extD_cl), type = "upper", p.mat = resD$p, insig = "label_sig", sig.level = c(.001, .01, .05), pch.cex = 0.8, pch.col = "yellow", tl.col="black")
corrplot(varD$cor, is.corr=FALSE)

resBAY <- cor.mtest(ext_cl_BAY, conf.level = .95)
corrplot(cor(ext_cl_BAY), type = "upper", p.mat = resBAY$p, insig = "label_sig", sig.level = c(.001, .01, .05), pch.cex = 0.8, pch.col = "yellow", tl.col="black")
corrplot(varBAY$cor, is.corr=FALSE)

resRex <- cor.mtest(ext_cl_rex, conf.level = .95)
corrplot(cor(ext_cl_rex), type = "upper", p.mat = resRex$p, insig = "label_sig", sig.level = c(.001, .01, .05), pch.cex = 0.8, pch.col = "yellow", tl.col="black")
corrplot(varRed$cor, is.corr=FALSE)
```

```{r}
#pairs.panels(ext_cl)
```

Based on the PCA and correlations, we can remove variables which strongly correlate or anticorrelate.

Removing (ext_cl_red & extD_cl_red)

-   annual insolation, as it anticorrelates with insolation in July, but also insolation July correlates with Betula_HZM -\> keeping only Betula_HZM in the pigment dataset

-   annual insolation removed also from diatoms dataset as it correlates with the NAP

-   Areboreal pollen (AP_HZM), as it is a % of NAP - keeping only NAP - in both datasets

Updated (ext_cl_red2)

-   removing all the HZM pollen as it is extrapolated from 13258 cal yr BP, keeping only the insolation July

Focusing on anoxia only (ext_cl_BAY)

-   Reduced dataset spanning only from 14200 until 12607 cal yr BP, using only Meerfeldemar Data, as HZM pollen ends up at 13258 cal yr BP

-   Removed AP_Mer

```{r}
ext_cl_red <- select(ext_cl, 1:3, 6:15)
extD_cl_red <- select(extD_cl, 1:3, 5:15)

ext_cl_red2 <- select(ext_cl, 1:10)
ext_cl_red_BAY <- select(ext_cl_BAY, 1:10, 17:19)

# ext_cl_red_rx <- select(ext_cl_red, 1:10, 17:19)
```

# RDA

```{r}
# Regular RDA without grouping constraint - this will give you more axes
rda_model_pig <- rda(base_cl ~ ., data= ext_cl_red, scale=T)
rda_model_pig

rda_model_pig2 <- rda(base_cl ~ ., data= ext_cl_red2, scale=T)
rda_model_pig2

rda_model_dia <- rda(baseD_cl ~ ., data= extD_cl_red, scale=T)
rda_model_dia

rda_model_BAY <- rda(base_cl_BAY ~ ., data= ext_cl_red_BAY, scale=T)
rda_model_BAY

rda_model_redox <- rda(base_cl_rex ~ ., data= ext_cl_rex, scale=T)
rda_model_redox

rda_model_redoxPBA <- rda(base_cl_rexPBA ~ ., data= ext_cl_rexPBA, scale=T)
rda_model_redoxPBA

# Note: Removed 'grouping = groups$PolZ, means = T' to get full RDA axes
# The grouping will still be used for coloring points in the plots
```

The 13 RDAs explain 64% of the variance (based on the Constrained proportion data, or also can be calculated as the sum of the RDAs contrained eigenvalues/total inertia), for more details see: <https://www.davidzeleny.net/anadat-r/doku.php/en:rda_cca_examples>

# Plotting RDA - ggplot version

```{r}
library(ggplot2)
library(ggrepel)  # For better label positioning

#colors = c("dodgerblue", "tomato2", "tomato4", "deepskyblue")

# Function to create RDA biplot
plot_rda_biplot <- function(rda_model, groups_data, title = " ") {
  # 1. Extract Site Scores (Points)
  site_scores <- data.frame(scores(rda_model, display = "sites", scaling = 3))
  
  # Match the grouping data to site scores by rownames/indices
  if (nrow(site_scores) == nrow(groups_data)) {
    site_scores$Category <- as.character(groups_data$PolZ)
    
    # Clean up the category labels - convert to proper display names
    # Only include levels that are actually present in the data
    present_levels <- unique(site_scores$Category)
    all_levels <- c("Pleniglacial", "Bølling", "Allerød", "YoungerDryas")
    all_labels <- c("Pleniglacial", "Bølling", "Allerød", "Younger Dryas")
    
    # Filter to only present levels
    level_mask <- all_levels %in% present_levels
    used_levels <- all_levels[level_mask]
    used_labels <- all_labels[level_mask]
    
    site_scores$Category <- factor(site_scores$Category, 
                                   levels = used_levels,
                                   labels = used_labels)
    
    # If any NAs, print warning with actual values found
    if (any(is.na(site_scores$Category))) {
      warning(paste("Some categories not matched. Found:", 
                   paste(unique(as.character(groups_data$PolZ)), collapse=", ")))
    }
  } else {
    warning(paste0("Site scores and groups length mismatch. Sites: ", 
                   nrow(site_scores), ", Groups: ", nrow(groups_data)))
    site_scores$Category <- rep(NA, nrow(site_scores))
  }
  
  # 2. Extract Constraint Scores (Arrows for environmental variables)
  env_scores <- data.frame(scores(rda_model, display = "bp", scaling = 3))
  env_scores$Variable <- rownames(env_scores)
  
  # Get explained variance (for axis labels)
  rda_summary <- summary(rda_model)
  eig <- rda_summary$cont$importance
  if (ncol(eig) >= 2) {
    perc_explained <- round(100 * eig[2, 1:2], 2)
  } else {
    perc_explained <- c(round(100 * eig[2, 1], 2), 0)
  }
  
  # 3. Calculate symmetrical axis limits
  # Get the maximum absolute value from both axes to make them equal
  max_x <- max(abs(c(site_scores$RDA1, env_scores$RDA1)), na.rm = TRUE)
  max_y <- max(abs(c(site_scores$RDA2, env_scores$RDA2)), na.rm = TRUE)
  axis_limit <- max(max_x, max_y) * 1.1  # Add 10% margin
  
  # 3.5. Set up colors based on present categories
  all_colors <- c("Pleniglacial" = "dodgerblue", 
                  "Bølling" = "tomato2", 
                  "Allerød" = "tomato4", 
                  "Younger Dryas" = "deepskyblue")
  used_colors <- all_colors[used_labels]
  
  # 4. Create the plot
  p <- ggplot(site_scores, aes(x = RDA1, y = RDA2)) +
    # Plot Points (Sites), colored by category
    geom_point(aes(color = Category), size = 4, alpha = 0.7) +  # Increased point size
    # Add Category Ellipses
    stat_ellipse(aes(group = Category, color = Category), level = 0.95, linewidth = 1.5) +  # 1.5pt lines
    scale_color_manual(name = "Pollen Zone", 
                       values = used_colors) + 
    
    # Plot Arrows (Environmental Variables)
    geom_segment(data = env_scores, 
                 aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
                 arrow = arrow(length = unit(0.3, "cm"), type = "closed"), 
                 color = "black", linewidth = 1.5) +  # 1.5pt lines
    
    # Add Arrow Labels with better positioning - 12pt font
    geom_text_repel(data = env_scores, 
                    aes(x = RDA1, y = RDA2, label = Variable),
                    color = "black", size = 4.5, fontface = "bold",  # size ~12pt
                    box.padding = 0.5, point.padding = 0.3) +
    
    # Reference lines
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", alpha = 0.5, linewidth = 0.75) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", alpha = 0.5, linewidth = 0.75) +
    
    # Labels and theme
    labs(x = paste0("RDA1 (", perc_explained[1], "%)"),
         y = paste0("RDA2 (", perc_explained[2], "%)"),
         title = title,
         color = "Pollen Zone") +
    # Set symmetrical axis limits for square plot
    xlim(-axis_limit, axis_limit) +
    ylim(-axis_limit, axis_limit) +
    theme_classic(base_size = 14) +  # Base font size 14pt for all text
    theme(legend.position = "right",
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  # 16pt title
          axis.title = element_text(size = 14, face = "bold"),  # 14pt axis titles
          axis.text = element_text(size = 12),  # 12pt axis text
          legend.title = element_text(size = 14, face = "bold"),  # 14pt legend title
          legend.text = element_text(size = 12),  # 12pt legend text
          axis.line = element_line(linewidth = 1.5),  # 1.5pt axis lines
          axis.ticks = element_line(linewidth = 1.5),  # 1.5pt tick marks
          aspect.ratio = 1) +  # Force 1:1 aspect ratio
    coord_fixed(ratio = 1)
  
  return(p)
}
```

```{r fig.width=10, fig.height=10}
# Plot RDA for pigment model
p_pig <- plot_rda_biplot(rda_model_pig, groups, title = "RDA Biplot - Pigments")

# Save with square dimensions
png(filename = "../figs/RDA_pigments_1-2.png", width = 800, height = 800, units = "px", pointsize = 12)
print(p_pig)
dev.off()
```

```{r fig.width=10, fig.height=10}
# Plot RDA for pigment model 2
p_pig2 <- plot_rda_biplot(rda_model_pig2, groups, title = "RDA Biplot - Pigments (no pollen)")

png(filename = "../figs/RDA_pigmentsNoPollen_1-2.png", width = 800, height = 800, units = "px", pointsize = 12)
print(p_pig2)
dev.off()
```

```{r fig.width=10, fig.height=8}
# Plot RDA for BAY model
# Note: Need to subset groups to match base_cl_BAY dimensions
groups_BAY <- groups[groups$age >= 12607 & groups$age <= 14200, ]
p_BAY <- plot_rda_biplot(rda_model_BAY, groups_BAY, title = "RDA Biplot - Bølling, Allerød, Younger Dryas")

png(filename = "../figs/RDA_pig_BAY_1-2.png", width = 800, height = 800, units = "px", pointsize = 12)
print(p_BAY)
dev.off()
```

```{r fig.width=10, fig.height=8}

p_pigD<- plot_rda_biplot(rda_model_dia, groupD, title = "RDA Biplot - Diatoms")

png(filename = "../figs/RDA_diatoms_1-2.png", width = 800, height = 800, units = "px", pointsize = 12)
print(p_pigD)
dev.off()
```

```{r fig.width=10, fig.height=8}
# Plot RDA for redox
p_red <- plot_rda_biplot(rda_model_redox, groupsRed, title = "RDA Biplot -  redox")

png(filename = "../figs/RDA_pigmentsRedox_1-2.png", width = 800, height = 800, units = "px", pointsize = 12)
print(p_red)
dev.off()
```

```{r}
plot(rda_model_pig, type ="text")
plot(rda_model_dia)
plot(rda_model_pig2)
plot(rda_model_BAY)
plot(rda_model_redox)
```

```{r}
screeplot(rda_model_pig)
screeplot(rda_model_dia)
screeplot(rda_model_pig2)
screeplot(rda_model_BAY)
screeplot(rda_model_redox)
```

```{r}
signif.full <- anova.cca(rda_model_pig, parallel=getOption("mc.cores"))
signif.full # test for model significance

signif.fullD <- anova.cca(rda_model_dia, parallel=getOption("mc.cores"))
signif.fullD # test for model significance
```

Get variance inflation factors for each predictor

```{r}
vif.cca(rda_model_pig) 

vif.cca(rda_model_dia)
```

# Plotting RDA

## Basic

Plotting RDA with the pollen zone and primary producers clustering color-coding

Note: The ggplot2 versions using plot_rda_biplot() function are already created above (lines 368-517). The base R plot() versions below are kept for reference or additional customization if needed.

```{r}
# Prepare factor labels for base R plotting (if needed)
pigG$CONISScluster <- factor(pigG$CONISScluster, labels = c("Diatoms + Bacteria", "Secondary colonisers", "Pioneers"))
predictions <- cbind(ext_cl, time, pigG)

pigG_BAY$CONISScluster <- factor(pigG_BAY$CONISScluster, labels = c("Diatoms + Bacteria"))
predictionsBAY <- cbind(ext_cl_red_BAY, time_BAY, pigG_BAY)

pigG_rex$CONISScluster - factor(pigG_rex$CONISScluster, labels = c("Diatoms + Bacteria", "Secondary colonisers", "Pioneers"))
predictionsRed <- cbind(ext_cl_rex, time_rex, pigG_rex)
```

```{r}
# Setup colors and factor levels for base R plotting
eco <- as.factor(predictions$PolZ)
levels(eco) <- c("Allerød", "Bølling", "Pleniglacial", "Younger Dryas")
bgT <- c("tomato4", "tomato2", "dodgerblue3", "deepskyblue")

ecoRed <- as.factor(predictionsRed$PolZ)
levels(ecoRed) <- c("Allerød", "Bølling", "Pleniglacial")
bgTRed <- c("tomato4", "tomato2", "dodgerblue3")

ecoBAY <- as.factor(predictionsBAY$PolZ)
levels(ecoBAY) <- c("Allerød", "Bølling", "YD")
bgTBAY <- c("tomato4", "tomato2","deepskyblue")
```

# Base R Plots (Original Style)

## Basic RDA plots with pollen zones

```{r fig.width=10, fig.height=10}

png(filename = "../figs/RDA1-2_pigments_PolZ.png", width = 800, height = 800, units = "px", pointsize = 12)

# Set graphics parameters for publication standards
par(cex.axis = 1.2,   # Axis tick labels = 12pt
    cex.lab = 1.2,    # Axis labels = 12pt
    lwd = 2,          # Line width = 1.5pt (R uses different scale, 2 ≈ 1.5pt)
    mar = c(5, 5, 4, 2) + 0.1)  # Margins

plot(rda_model_pig, scaling =2, type="n")  # type="n" prevents automatic drawing
points(rda_model_pig, display="sites", pch=c(21), cex=2, bg=bgT[eco], scaling=2, lwd=2)
# text(rda_model_pig, display="sites", cex=0.65, pos=1, col="blue", labels = base$age)  # Removed age labels to reduce overlap
text(rda_model_pig, display="species", cex=1.0, pos=4, col="black", scaling = 2, offset=0.8)  # Species labels, slightly smaller
text(rda_model_pig, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)  # Arrow labels bold, no pos to avoid overlap
legend("bottomleft", legend=c(levels(eco)), bty="n", col="gray5", pch=21, cex=1.2, pt.bg=bgT, ncol = 1, pt.lwd=2)

dev.off()
```

```{r fig.width=10, fig.height=10}
png(filename = "../figs/RDA1-2_pigments_BAY_PolZ.png", width = 800, height = 800, units = "px", pointsize = 12)

# Set graphics parameters for publication standards
par(cex.axis = 1.2,   # Axis tick labels = 12pt
    cex.lab = 1.2,    # Axis labels = 12pt
    lwd = 2,          # Line width = 1.5pt
    mar = c(5, 5, 4, 2) + 0.1)  # Margins

plot(rda_model_BAY, scaling =2, type ="n")
points(rda_model_BAY, display="sites", pch=c(21), cex=2, bg=bgTBAY[ecoBAY], scaling=2, lwd=2)
# text(rda_model_BAY, display="sites", cex=0.65, pos=1, col="blue", labels = base$age)  # Removed age labels to reduce overlap
text(rda_model_BAY, display="species", cex=1.0, pos=4, col="black", scaling = 2, offset=0.8)  # Species labels, slightly smaller
text(rda_model_BAY, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)  # Arrow labels bold, no pos to avoid overlap
legend("bottomleft", legend=c(levels(ecoBAY)), bty="n", col="gray5", pch=21, cex=1.2, pt.bg=bgT, ncol = 1, pt.lwd=2) 

dev.off()
```

## CONISS RDA

Adding CONISS clusters performed on the 3 averaged pigment communities originating from the correlation heat maps. This shows what is influencing the pigment clusters through time.

```{r}
Pig <- as.factor(predictions$CONISScluster)
levels(Pig) <- c("Pigment cluster 3", "Pigment cluster 2", "Pigment cluster 1")
bgP <- c("royalblue4", "royalblue2", "lightblue")
pchP <- c(25,23,24)
```

```{r fig.width=10, fig.height=10}

png(filename = "../figs/RDA1-2_pigments_PolZ_clustered.png", width = 800, height = 800, units = "px", pointsize = 12)
# Set graphics parameters for publication standards
par(cex.axis = 1.2, cex.lab = 1.2, lwd = 2, mar = c(5, 5, 4, 2) + 0.1)

plot(rda_model_pig, scaling =2, type="n")
# Single layer: pollen zone colors + pigment cluster shapes
points(rda_model_pig, display="sites", pch=pchP[Pig], cex=2.5, bg=bgT[eco], col=bgT[eco], scaling=2, lwd=2)
text(rda_model_pig, display="species", cex=1.0, pos=4, col="black", scaling = 2, offset=0.8)
text(rda_model_pig, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)
# Two-part legend: pollen zones (colors) and pigment clusters (shapes)
legend("bottomleft", 
       legend=c(levels(eco), "", levels(Pig)), 
       bty="n", 
       col=c(rep("gray32", 4), "", rep("gray32", 3)), 
       pch=c(rep(22, 4), NA, pchP),  # Square for color legend, actual shapes for clusters
       pt.bg=c(bgT, "", rep("gray80", 3)),  # Colors for pollen zones, gray for shape examples
       cex=1.0, 
       pt.lwd=2,
       pt.cex=c(rep(2, 4), NA, rep(2, 3)),
       ncol = 1)

dev.off()
```

```{r fig.width=10, fig.height=10}
png(filename = "../figs/RDA1-2_pigments_red_PolZ_clustered.png", width = 800, height = 800, units = "px", pointsize = 12)

# Set graphics parameters for publication standards
par(cex.axis = 1.2, cex.lab = 1.2, lwd = 2, mar = c(5, 5, 4, 2) + 0.1)

plot(rda_model_redox, scaling =2, type="n")
# Single layer: pollen zone colors + pigment cluster shapes
points(rda_model_redox, display="sites", pch=pchP[Pig], cex=2.5, bg=bgTRed[ecoRed], col=bgTRed[ecoRed], scaling=2, lwd=2)
text(rda_model_redox, display="species", cex=1.0, pos=4, col="black", scaling = 2, offset=0.8)
text(rda_model_redox, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)
# Two-part legend: pollen zones (colors) and pigment clusters (shapes)
legend("bottomleft", 
       legend=c(levels(ecoRed), "", levels(Pig)), 
       bty="n", 
       col=c(rep("gray32", 3), NA, rep("gray32", 3)), 
       pch=c(rep(22, 3), NA, pchP),  # Square for color legend, actual shapes for clusters
       pt.bg=c(bgTRed, NA, rep("gray80", 3)),  # Colors for pollen zones, gray for shape examples
       cex=1.0, 
       pt.lwd=2,
       pt.cex=c(rep(2, 3), NA, rep(2, 3)),
       ncol = 1)

dev.off()
```

## Pigment producers groups

Using the correlation clusters of pigments from the heat map (3 groups), to display through time, to see which community (Pioneers, Secondary colonizers and Diatoms + BPSB) is influenced by what external factors.

```{r}
pig_prod <- read_excel("../data/SourceData/HZM_pigment_production.xlsx", sheet = "50%0")
pig_prod_df <- as.data.frame(pig_prod)
rownames(pig_prod_df) <- pig_prod_df[,2]
pig_prod_df$alternative <- c("P", "P", "P", "D", "P", "P", "P", "S", "P", "S", "P", "P", "D", "P", "S", "P", "S", "S","S", "S")
```

```{r}
PigClass <- as.factor(pig_prod_df$alternative)
levels(PigClass) <- c("Diatoms + Bacteria", "Pioneers", "Secondary colonisers")
bgClass <- c("mediumpurple3", "cyan", "green3")
pchClass <- c(25,23,24)
```

```{r fig.width=10, fig.height=10}
# RDA with HZM pollen - full time period
png(filename = "../figs/RDA1-2_pigments_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

# Set graphics parameters for publication standards
par(cex.axis = 1.2, cex.lab = 1.2, lwd = 2, mar = c(5, 5, 4, 2) + 0.1)

plot(rda_model_pig, scaling =2, type= "n")
points(rda_model_pig, display="sites", pch=c(21), cex=2.5, bg=bgT[eco], scaling=2, lwd=2)
points(rda_model_pig, display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=2, scaling=2, lwd=2)
# text(rda_model_pig, display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)  # Removed to avoid overlap
text(rda_model_pig, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)
legend("bottomleft", 
       legend=c(levels(eco), "", levels(PigClass)), 
       bty="n", 
       col=c(rep("gray32", 4), "", bgClass), 
       pch=c(rep(21, 4), NA, pchClass), 
       pt.bg=c(bgT, "", bgClass), 
       cex=1.0, 
       pt.lwd=2,
       pt.cex=c(rep(2, 4), NA, rep(2, 3)),
       ncol = 1)

dev.off()
```

```{r fig.width=10, fig.height=10}
# RDA for Bølling-Allerød-Younger Dryas period only
png(filename = "../figs/RDA1-2_pigments_BAY_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

# Set graphics parameters for publication standards
par(cex.axis = 1.2, cex.lab = 1.2, lwd = 2, mar = c(5, 5, 4, 2) + 0.1)

plot(rda_model_BAY, scaling =2, type ="n")
points(rda_model_BAY, display="sites", pch=c(21), cex=2.5, bg=bgTBAY[ecoBAY], scaling=2, lwd=2)
points(rda_model_BAY, display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=2, scaling=2, lwd=2)
# text(rda_model_BAY, display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)  # Removed to avoid overlap
text(rda_model_BAY, scaling=2, display="bp", col="#0868ac", cex=1.0, font=2, offset=0.8)
legend("bottomleft", 
       legend=c(levels(ecoBAY), "", levels(PigClass)), 
       bty="n", 
       col=c(rep("gray32", 3), "", bgClass), 
       pch=c(rep(21, 3), NA, pchClass), 
       pt.bg=c(bgTBAY, "", bgClass), 
       cex=1.0, 
       pt.lwd=2,
       pt.cex=c(rep(2, 3), NA, rep(2, 3)),
       ncol = 1)
dev.off()
```

# RDA plot with everything

```{r fig.width=10, fig.height=10}
# RDA including redox data - PUBLICATION VERSION
# Size: 1 column (8.8 cm width = 346 px at 300 dpi)
# Font: 12pt main, 9pt minor
# Line width: 1.5pt
png(filename = "../figs/RDA1-2_pigments_redox_PolZ_communities.png", 
    width = 8.8, height = 8.8, units = "cm", res = 300, pointsize = 9)

# Publication-standard graphics parameters
par(cex.axis = 1.0,      # Axis numbers = 9pt (base)
    cex.lab = 1.33,      # Axis labels = 12pt (9 * 1.33)
    lwd = 1.5,           # Line width = 1.5pt
    mgp = c(2.5, 0.7, 0), # Axis label spacing
    mar = c(4.5, 4.5, 1, 1),  # Margins: bottom, left, top, right
    tcl = -0.3)          # Tick length

plot(rda_model_redox, scaling = 2, type = "n")
# Site points (pollen zones)
points(rda_model_redox, display = "sites", pch = 21, cex = 0.8, 
       bg = bgTRed[ecoRed], scaling = 2, lwd = 1.5)
# Species points (pigment communities)
points(rda_model_redox, display = "species", pch = pchClass[PigClass], 
       bg = bgClass[PigClass], col = bgClass[PigClass], cex = 0.8, 
       scaling = 2, lwd = 1.5)

# Environmental arrows with labels - smaller font to avoid overlap
arrow_coords <- scores(rda_model_redox, display = "bp", scaling = 2)
# Plot arrows
arrows(0, 0, arrow_coords[,1], arrow_coords[,2], 
       length = 0.1, col = "#0868ac", lwd = 1.5)
# Add labels with smaller font (7pt) to reduce overlap
text(arrow_coords[,1] * 1.1, arrow_coords[,2] * 1.1, 
     labels = rownames(arrow_coords),
     col = "#0868ac", cex = 0.78, font = 2)  # 7pt bold for arrow labels (9 * 0.78)

# Legend with proper sizing
legend("bottomleft", 
       legend = c(levels(ecoRed), "", levels(PigClass)), 
       bty = "n", 
       col = c(rep("gray32", 3), NA, bgClass), 
       pch = c(rep(21, 3), NA, pchClass), 
       pt.bg = c(bgTRed, NA, bgClass), 
       cex = 0.8,           # 9pt legend text
       pt.lwd = 1.5,        # 1.5pt point borders
       pt.cex = 1.0,        # Point size in legend
       seg.len = 1.5,       # Length of legend lines
       x.intersp = 0.8,     # Horizontal spacing
       y.intersp = 1.0)     # Vertical spacing to avoid overlap

dev.off()

```

# Exploring RDA 1 and 3

```{r}
png(filename = "../figs/RDA1-3_pigments_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

plot(rda_model_pig, scaling =2, choices=c(1,3))
points(rda_model_pig, choices=c(1,3), display="sites", pch=c(21), cex=2, bg=bgT[eco], scaling=2)# col=bgT[eco])
#text(rda_model_pig, display="sites", cex=0.65, pos=1, col="blue") # the toads
points(rda_model_pig, choices=c(1,3), display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=1, scaling=2)#, bg=bgP[Pig])       bg=bgP, col=bgP,    # the phenotypes
text(rda_model_pig, choices=c(1,3), display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)
# the environmental predictors
text(rda_model_pig, choices=c(1,3), scaling=2, display="bp", col="#0868ac", cex=1)  
legend("bottomleft", legend=c(levels(eco), levels(PigClass)), bty="n", col=c("gray5", "gray5", "gray5", "gray5",  bgClass), pch=c(21,21,21, 21, pchClass), cex=0.8, pt.bg=c(bgT, bgClass), ncol = 1)

dev.off()


png(filename = "../figs/RDA1-3_pigments_BAY_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

plot(rda_model_BAY, scaling =2, type ="n", choices=c(1,3))
points(rda_model_BAY, choices=c(1,3), display="sites", pch=c(21), cex=2, bg=bgTBAY[ecoBAY], scaling=2)
points(rda_model_BAY,choices=c(1,3), display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=1, scaling=2)
text(rda_model_BAY, choices=c(1,3), display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)
# the environmental predictors
text(rda_model_BAY, choices=c(1,3), scaling=2, display="bp", col="#0868ac", cex=1)  
legend("bottomleft", legend=c(levels(ecoBAY),levels(PigClass)), bty="n", col=c("gray5", "gray5", "gray5",  bgClass), pch=c(21,21,21, pchClass), cex=0.8, pt.bg=c(bgTBAY, bgClass), ncol = 1) 

dev.off()
```

Exploring RDA 2 and 3

```{r}
png(filename = "../figs/RDA2-3_pigments_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

plot(rda_model_pig, scaling =2, choices=c(2,3))
points(rda_model_pig, choices=c(2,3), display="sites", pch=c(21), cex=2, bg=bgT[eco], scaling=2)# col=bgT[eco])
#text(rda_model_pig, display="sites", cex=0.65, pos=1, col="blue") # the toads
points(rda_model_pig, choices=c(2,3), display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=1, scaling=2)#, bg=bgP[Pig])       bg=bgP, col=bgP,    # the phenotypes
text(rda_model_pig, choices=c(2,3), display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)
# the environmental predictors
text(rda_model_pig, choices=c(2,3), scaling=2, display="bp", col="#0868ac", cex=1)  
legend("bottomleft", legend=c(levels(eco), levels(PigClass)), bty="n", col=c("gray5", "gray5", "gray5", "gray5",  bgClass), pch=c(21,21,21, 21, pchClass), cex=0.8, pt.bg=c(bgT, bgClass), ncol = 1)

dev.off()


png(filename = "../figs/RDA2-3_pigments_BAY_PolZ_communities.png", width = 800, height = 800, units = "px", pointsize = 12)

plot(rda_model_BAY, scaling =2, type ="n", choices=c(2,3))
points(rda_model_BAY, choices=c(2,3), display="sites", pch=c(21), cex=2, bg=bgTBAY[ecoBAY], scaling=2)
points(rda_model_BAY,choices=c(2,3), display="species", pch= pchClass[PigClass], bg=bgClass[PigClass], col =bgClass[PigClass], cex=1, scaling=2)
text(rda_model_BAY, choices=c(2,3), display="species", cex=1, pos=1, col="black", scaling = 2, adj = 0.5)
# the environmental predictors
text(rda_model_BAY, choices=c(2,3), scaling=2, display="bp", col="#0868ac", cex=1)  
legend("bottomleft", legend=c(levels(ecoBAY),levels(PigClass)), bty="n", col=c("gray5", "gray5", "gray5",  bgClass), pch=c(21,21,21, pchClass), cex=0.8, pt.bg=c(bgTBAY, bgClass), ncol = 1) 

dev.off()
```

# Test ggplot RDA function

Quick diagnostic to check if the updated plotting function works correctly

```{r}
# Check what values are actually in groups$PolZ
cat("Unique values in groups$PolZ:\n")
print(unique(groups$PolZ))
cat("\nClass of groups$PolZ:", class(groups$PolZ), "\n")
cat("Number of rows in groups:", nrow(groups), "\n")
```

# Publication Figure - RDA with Reduced External Variables

Removing highly correlated external variables (P_F3, P_F2, JulyT_Egel, Insol_July, Insol_A) for clearer visualization.

```{r}
# Create reduced external dataset by removing highly correlated variables and interpolated data
ext_cl_rex_reduced <- ext_cl_rex %>% 
  select(-P_F3, -P_F2, -JulyT_Egel, -insol_July, -insol_A, -SalixHZM, -JuniperusHZM)

# Merge Mn_F2 and Mn_F3 into a single Mn_redox variable (average)
if("Mn_F2" %in% colnames(ext_cl_rex_reduced) & "Mn_F3" %in% colnames(ext_cl_rex_reduced)) {
  ext_cl_rex_reduced$Mn_redox <- (ext_cl_rex_reduced$Mn_F2 + ext_cl_rex_reduced$Mn_F3) / 2
  ext_cl_rex_reduced <- ext_cl_rex_reduced %>% select(-Mn_F2, -Mn_F3)
}

# Check remaining columns
cat("Remaining external variables:\n")
print(colnames(ext_cl_rex_reduced))
```

```{r}
# Run RDA with reduced external variables
rda_model_redox_reduced <- rda(base_cl_rex ~ ., data = ext_cl_rex_reduced, scale = TRUE)
rda_model_redox_reduced

# Check model significance
signif_reduced <- anova.cca(rda_model_redox_reduced, parallel = getOption("mc.cores"))
signif_reduced

# Check VIF
vif.cca(rda_model_redox_reduced)
```

```{r fig.width=10, fig.height=10}
# Publication-ready RDA plot with reduced variables and nice labels
# Size: 1 column (8.8 cm width for publication)
# Font: 12pt main, 9pt minor
# Line width: 1.5pt

png(filename = "../figs/RDA1-2_pigments_redox_PolZ_communities_reduced.png", 
    width = 8.8, height = 8.8, units = "cm", res = 300, pointsize = 9)

# Publication-standard graphics parameters - reduced margins and spacing
par(cex.axis = 0.9,       # Axis numbers
    cex.lab = 1.1,        # Axis labels
    lwd = 1.5,            # Line width = 1.5pt
    mgp = c(1.8, 0.5, 0), # Axis label spacing (tighter)
    mar = c(3.2, 3.2, 0.5, 0.5),  # Margins: bottom, left, top, right (reduced)
    tcl = -0.3)           # Tick length

# Calculate symmetrical axis limits
rda_scores <- scores(rda_model_redox_reduced, scaling = 2)
max_limit <- max(abs(c(rda_scores$sites, rda_scores$species))) * 1.1
xlim <- c(-max_limit, max_limit)
ylim <- c(-max_limit, max_limit)

# Get variance explained for axis labels
eig <- eigenvals(rda_model_redox_reduced)
var_exp <- eig / sum(eig) * 100

# Create blank plot without using plot.rda to avoid black cross
plot.new()
plot.window(xlim = xlim, ylim = ylim, asp = 1)

# Add gray dashed reference lines FIRST (so they appear behind everything)
abline(h = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)
abline(v = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)

# Add axes and labels
axis(1)
axis(2)
box()
mtext(sprintf("RDA1 (%.1f%%)", var_exp[1]), side = 1, line = 2, cex = par("cex.lab"))
mtext(sprintf("RDA2 (%.1f%%)", var_exp[2]), side = 2, line = 2, cex = par("cex.lab"))

# Site points (pollen zones) - semi-transparent, no border
site_scores <- scores(rda_model_redox_reduced, display = "sites", scaling = 2)
points(site_scores[,1], site_scores[,2], pch = 21, cex = 1.0, 
       bg = adjustcolor(bgTRed[ecoRed], alpha.f = 0.6), 
       col = adjustcolor(bgTRed[ecoRed], alpha.f = 0.6), lwd = 0)

# Species points (pigment communities)
species_scores <- scores(rda_model_redox_reduced, display = "species", scaling = 3.5)
points(species_scores[,1], species_scores[,2], pch = pchClass[PigClass], 
       bg = bgClass[PigClass], col = bgClass[PigClass], cex = 0.9, 
       lwd = 1.5)

# Environmental arrows with formatted labels - make arrows much longer
arrow_coords <- scores(rda_model_redox_reduced, display = "bp", scaling = 2)
arrow_names <- rownames(arrow_coords)

# Scale factor to make arrows longer (2x longer for better visibility)
arrow_scale <- 2.0
arrow_coords_scaled <- arrow_coords * arrow_scale

# Create mapping from original names to nice labels
label_map <- list(
  "`18O_NGRIP`" = expression(delta^18*O[NGRIP]),
  "temp_GDGT" = expression(T[GDGT]),
  "JulyT_Ger" = expression(T[Ger]~"(July)"),
  "BetulaHZM" = expression(italic(Betula)),
  "PinusHZM" = expression(italic(Pinus)),
  "Dust_NGRIP" = expression(Dust[NGRIP]),
  "NAP_HZM" = expression(Herbs),
  "insol_Jan" = expression("Winter insolation"),
  "Mn_redox" = expression(Mn[redox]),
  "SalixHZM" = expression(italic(Salix)),
  "JunineprusHZM" = expression(italic(Juniperus)),
  "Ca" = expression(Ca[XRF]),
  "Ti" = expression(Ti[XRF])
)

# Plot longer arrows in black
arrows(0, 0, arrow_coords_scaled[,1], arrow_coords_scaled[,2], 
       length = 0.1, col = "black", lwd = 1.5)

# Add labels with nice formatting - positioned further from arrows to avoid overlap
for (i in 1:nrow(arrow_coords_scaled)) {
  var_name <- arrow_names[i]
  label_expr <- label_map[[var_name]]
  
  # If no mapping exists, use original name
  if (is.null(label_expr)) {
    label_expr <- var_name
  }
  
  # Position labels further out to avoid overlap with arrows
  x_coord <- arrow_coords_scaled[i,1]
  y_coord <- arrow_coords_scaled[i,2]
  
  # Smart positioning: use pos parameter based on arrow direction
  if (abs(x_coord) > abs(y_coord)) {
    # Horizontal arrows: place left or right
    text_pos <- ifelse(x_coord > 0, 4, 2)  # 4=right, 2=left
  } else {
    # Vertical arrows: place above or below
    text_pos <- ifelse(y_coord > 0, 3, 1)  # 3=above, 1=below
  }
  
  # Special adjustment for Mn_redox to avoid overlap with JulyT_Ger
  if (var_name == "Mn_redox") {
    text_pos <- 1  # Force below position
  }
  
  text(x_coord, y_coord, 
       labels = label_expr,
       col = "black", cex = 0.7, font = 2, pos = text_pos, offset = 0.2)
}

# Legend with smaller font and no borders on legend symbols
# Reversed order: oldest (Pleniglacial) at bottom, youngest (Allerød) at top
legend_labels_red <- c("Allerød", "Bølling", "Pleniglacial", "", 
                       "Diatoms + Bacteria", "Secondary colonisers", "Pioneers")
legend_colors_red <- c(bgTRed[1], bgTRed[2], bgTRed[3], NA, bgClass[1], bgClass[3], bgClass[2])
legend_pch_red <- c(21, 21, 21, NA, pchClass[1], pchClass[3], pchClass[2])

legend("bottomleft", 
       legend = legend_labels_red, 
       bty = "n", 
       col = c(rep(NA, 3), NA, bgClass[1], bgClass[3], bgClass[2]),  # No border color for pollen zones
       pch = legend_pch_red, 
       pt.bg = legend_colors_red, 
       cex = 0.75,           # Smaller legend text
       pt.lwd = 0,           # No border on legend symbols
       pt.cex = 0.9,         # Point size in legend
       seg.len = 1.2,        # Length of legend lines
       x.intersp = 0.7,      # Horizontal spacing
       y.intersp = 0.9)      # Vertical spacing

dev.off()
```

# Publication Figure - RDA with Full Dataset (All Pollen Zones)

Similar formatting to the reduced RDA plot above, but including all external variables and all four pollen zones.

```{r}
# Create reduced external dataset by removing only the interpolated data
ext_cl_red_pub <- ext_cl_red %>% 
  select(-SalixHZM, -JuniperusHZM, JulyT_Egel) %>%
  select(-contains("Salix"), -contains("Junineprus"), -contains("JulyT_Egel"))

# Check remaining columns
cat("External variables for full RDA:\n")
print(colnames(ext_cl_red_pub))
```

```{r}
# Run RDA with publication-ready external variables
rda_model_pig_pub <- rda(base_cl ~ ., data = ext_cl_red_pub, scale = TRUE)
rda_model_pig_pub

# Check model significance
signif_pig_pub <- anova.cca(rda_model_pig_pub, parallel = getOption("mc.cores"))
signif_pig_pub

# Check VIF
vif.cca(rda_model_pig_pub)
```

```{r fig.width=10, fig.height=10}
# Publication-ready RDA plot with all pollen zones
# Size: 1 column (8.8 cm width for publication)
# Font: 12pt main, 9pt minor
# Line width: 1.5pt

png(filename = "../figs/RDA1-2_pigments_PolZ_communities_publication.png", 
    width = 8.8, height = 8.8, units = "cm", res = 300, pointsize = 9, bg = "transparent")

# Publication-standard graphics parameters - reduced margins and spacing
par(cex.axis = 0.9,       # Axis numbers
    cex.lab = 1.1,        # Axis labels
    lwd = 1.5,            # Line width = 1.5pt
    mgp = c(1.8, 0.5, 0), # Axis label spacing (tighter)
    mar = c(3, 3, 0.3, 0.3),  # Margins: bottom, left, top, right (minimized)
    tcl = -0.3)           # Tick length

# Calculate symmetrical axis limits
rda_scores <- scores(rda_model_pig_pub, scaling = 2)
max_limit <- max(abs(c(rda_scores$sites, rda_scores$species))) * 1.1
xlim <- c(-max_limit, max_limit)
ylim <- c(-max_limit, max_limit)

# Get variance explained for axis labels
eig <- eigenvals(rda_model_pig_pub)
var_exp <- eig / sum(eig) * 100

# Create blank plot without using plot.rda to avoid black cross
plot.new()
plot.window(xlim = xlim, ylim = ylim, asp = 1)

# Add gray dashed reference lines FIRST (so they appear behind everything)
abline(h = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)
abline(v = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)

# Add axes and labels
axis(1)
axis(2)
box()
mtext(sprintf("RDA1 (%.1f%%)", var_exp[1]), side = 1, line = 2, cex = par("cex.lab"))
mtext(sprintf("RDA2 (%.1f%%)", var_exp[2]), side = 2, line = 2, cex = par("cex.lab"))

# Site points (pollen zones) - semi-transparent, no border
site_scores <- scores(rda_model_pig_pub, display = "sites", scaling = 2)
points(site_scores[,1], site_scores[,2], pch = 21, cex = 1.0, 
       bg = adjustcolor(bgT[eco], alpha.f = 0.6), 
       col = adjustcolor(bgT[eco], alpha.f = 0.6), lwd = 0)

# Species points (pigment communities)
species_scores <- scores(rda_model_pig_pub, display = "species", scaling = 3.5)
points(species_scores[,1], species_scores[,2], pch = pchClass[PigClass], 
       bg = bgClass[PigClass], col = bgClass[PigClass], cex = 0.9, 
       lwd = 1.5)

# Environmental arrows with formatted labels - make arrows much longer
arrow_coords <- scores(rda_model_pig_pub, display = "bp", scaling = 2)
arrow_names <- rownames(arrow_coords)

# Scale factor to make arrows longer (2x longer for better visibility)
arrow_scale <-2.8
arrow_coords_scaled <- arrow_coords * arrow_scale

# Create mapping from original names to nice labels (same as reduced version)
label_map <- list(
  "`18O_NGRIP`" = expression(delta^18*O[NGRIP]),
  "temp_GDGT" = expression(T[GDGT]),
  "JulyT_Ger" = expression(T[Ger]~"(July)"),
  "BetulaHZM" = expression(italic(Betula)),
  "PinusHZM" = expression(italic(Pinus)),
  "Dust_NGRIP" = expression(Dust[NGRIP]),
  "NAP_HZM" = expression(Herbs),
  "insol_Jan" = expression("Winter \ninsolation"),
  "Mn_redox" = expression(Mn[redox]),
  "SalixHZM" = expression(italic(Salix)),
  "JunineprusHZM" = expression(italic(Juniperus)),
  "Ca" = expression(Ca[XRF]),
  "Ti" = expression(Ti[XRF])
)

# Plot longer arrows in black
arrows(0, 0, arrow_coords_scaled[,1], arrow_coords_scaled[,2], 
       length = 0.1, col = "black", lwd = 1.5)

# Add labels with nice formatting - positioned to avoid overlap
for (i in 1:nrow(arrow_coords_scaled)) {
  var_name <- arrow_names[i]
  label_expr <- label_map[[var_name]]
  
  # If no mapping exists, use original name
  if (is.null(label_expr)) {
    label_expr <- var_name
  }
  
  # Position labels to avoid overlap with arrows
  x_coord <- arrow_coords_scaled[i,1]
  y_coord <- arrow_coords_scaled[i,2]
  
  # Smart positioning: use pos parameter based on arrow direction
  if (abs(x_coord) > abs(y_coord)) {
    # Horizontal arrows: place left or right
    text_pos <- ifelse(x_coord > 0, 4, 2)  # 4=right, 2=left
  } else {
    # Vertical arrows: place above or below
    text_pos <- ifelse(y_coord > 0, 3, 1)  # 3=above, 1=below
  }
  
  # Special adjustments to avoid overlap
  if (var_name == "BetulaHZM") {
    text_pos <- 1.2  # Force below position for Betula
  }
  
  text(x_coord, y_coord, 
       labels = label_expr,
       col = "black", cex = 0.7, font = 2, pos = text_pos, offset = 0.2)
}

# Legend with smaller font and no borders on legend symbols
# Reversed order: oldest (Pleniglacial) at bottom, youngest (Younger Dryas) at top
legend_labels <- c("Younger Dryas", "Allerød", "Bølling", "Pleniglacial", "", 
                   "Diatoms + Bacteria", "Secondary colonisers", "Pioneers")
legend_colors <- c(bgT[4], bgT[1], bgT[2], bgT[3], NA, bgClass[1], bgClass[3], bgClass[2])
legend_pch <- c(21, 21, 21, 21, NA, pchClass[1], pchClass[3], pchClass[2])

legend("bottomleft", 
       legend = legend_labels, 
       bty = "n", 
       col = c(rep(NA, 4), NA, bgClass[1], bgClass[3], bgClass[2]),  # No border color
       pch = legend_pch, 
       pt.bg = legend_colors, 
       cex = 0.75,           # Smaller legend text
       pt.lwd = 0,           # No border on legend symbols
       pt.cex = 0.9,         # Point size in legend
       seg.len = 1.2,        # Length of legend lines
       x.intersp = 0.7,      # Horizontal spacing
       y.intersp = 0.9)      # Vertical spacing

dev.off()
```
