---
title: "Pollen Discussion Plot - Pinus and Betula with Modelled Extension"
output:
  html_notebook:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Overview

## Purpose
Generate pollen-specific discussion visualization focusing on Pinus and Betula abundance. Extends coverage to earlier periods (13258-10995 BP) using modelled data from statistical calibration to Meerfelder Maar reference sequence.

## Processing Steps
1. Load Holzmaar pollen data (measured and modelled portions)
2. Load modelled data extension (13258-10995 BP) from Notebook 11
3. Combine measured and modelled records
4. Create stratigraphic plot with Pinus/Betula as primary taxa
5. Highlight modelled section with transparency or annotation
6. Overlay climate zones and chronological boundaries
7. Export publication-quality figure

## Input Files
- ../data/SourceData/HolzPol_10995-15030_complete.csv (main pollen sequence)
- ../data/XRF_clustered_PC.xlsx (age model reference)

## Output Files
- ../figs/pollen_discussion_PinusBetula_*.png

# Setup and Library Loading

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pacman)
pacman::p_load(readxl, read.csv, dplyr, tidyr, openxlsx, ggplot2, cowplot, viridis, ggrepel)

cat("Libraries loaded for pollen visualization.\n")
```

# Load Pollen Data

```{r}
pollen_full <- read.csv("../data/SourceData/HolzPol_10995-15030_complete.csv", header = TRUE)

cat("Complete pollen sequence loaded: ", nrow(pollen_full), " samples\n", sep="")
cat("Depth range: Holzmaar 10995-15030 BP\n")
cat("Coverage: Full measured period (0-13258 BP) + modelled extension (13258-10995 BP)\n")
```

# Extract Pinus and Betula

```{r}
pollen_taxa <- data.frame(
  depth_cm = pollen_full$Depth,
  age_bp = pollen_full$Age_BP,
  Pinus_pct = pollen_full$Pinus,
  Betula_pct = pollen_full$Betula,
  is_modelled = grepl("modelled|predicted", tolower(pollen_full$Method), ignore.case = TRUE)
)

cat("Pollen taxa extracted: ", nrow(pollen_taxa), " samples\n", sep="")
cat("Measured vs modelled: ")
cat(sum(!pollen_taxa$is_modelled), " measured, ")
cat(sum(pollen_taxa$is_modelled), " modelled\n", sep="")
```

# Create Pollen Discussion Plot

```{r}
p_pollen <- ggplot(pollen_taxa, aes(x = age_bp)) +
  geom_path(aes(y = Pinus_pct), color = "darkgreen", size = 0.5) +
  geom_path(aes(y = Betula_pct), color = "orange", size = 0.5) +
  geom_point(aes(y = Pinus_pct, fill = is_modelled), 
             size = 1.5, shape = 21, color = "darkgreen", alpha = 0.6) +
  geom_point(aes(y = Betula_pct, fill = is_modelled), 
             size = 1.5, shape = 21, color = "orange", alpha = 0.6) +
  scale_x_reverse(name = "Age (cal yr BP)") +
  scale_y_continuous(name = "Pollen % (Pinus & Betula)", limits = c(0, 100)) +
  scale_fill_manual(values = c("TRUE" = "lightyellow", "FALSE" = "gray20"),
                    labels = c("FALSE" = "Measured", "TRUE" = "Modelled")) +
  labs(title = "Pollen Reconstruction: Pinus and Betula",
       subtitle = "Measured (0-13258 BP) and Modelled Extension (13258-10995 BP)",
       fill = "Data Type") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "italic"),
    legend.position = "bottom"
  )

cat("✓ Pollen discussion plot created with measured/modelled distinction.\n")
```

# Climate Zone Overlays

```{r}
zones <- data.frame(
  age_start = c(15342, 14444, 13611, 13011, 12727),
  age_end = c(14444, 13611, 13011, 12727, 11625),
  zone = c("Pleniglacial", "Bølling", "Allerød", "LST", "Younger Dryas")
)

for (i in 1:nrow(zones)) {
  p_pollen <- p_pollen +
    annotate("rect",
             xmin = zones$age_start[i], xmax = zones$age_end[i],
             ymin = -Inf, ymax = Inf,
             fill = c("blue", "red", "red", "lightgray", "blue")[i],
             alpha = 0.08)
}

cat("Climate zones overlay added.\n")
```

# Export Pollen Plot

```{r}
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}

png(filename = timestamp_filename("../figs", "pollen_discussion_PinusBetula"),
    width = 900, height = 600, units = "px", res = 300, pointsize = 12)

print(p_pollen)

dev.off()

cat("✓ Pollen discussion plot exported.\n")
```

# Session Information

```{r}
sessionInfo()
```
