---
title: "Supplementary Pigment Plots"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
setwd("~/Holzmaar/Holzmaar2022-2025")
```

# Load required packages

```{r}
library("rioja")
library("ggplot2")
library("readxl")
library("dplyr")
library("tidyr")
library("cowplot")
library("latex2exp")
library("vegan")
library("janitor")
```

# Load pigment data

```{r}
# Load raw pigment concentrations (not transformed)
pig_raw <- read_excel("../data/Pig_clean.xlsx")

# Load pigment data with depth/age
pig_with_meta <- read_excel("../data/pig_alr.xlsx", sheet="Sheet 1")

# Combine them
plot_data <- data.frame(
  depth = pig_with_meta$depth,
  age = pig_with_meta$age,
  Chla = pig_raw$Chla,
  Pheo = pig_raw$PhphT
)

# Remove NA and zero values for ratio calculation
plot_data_clean <- plot_data %>%
  filter(Chla > 0 & Pheo > 0) %>%
  mutate(Ratio = Chla / Pheo) %>%
  arrange(depth)

cat("Data Summary:\n")
cat(sprintf("Samples: %d\n", nrow(plot_data_clean)))
cat(sprintf("Chl-a range: %.2f - %.2f µg/g\n", min(plot_data_clean$Chla), max(plot_data_clean$Chla)))
cat(sprintf("Pheophytin range: %.2f - %.2f µg/g\n", min(plot_data_clean$Pheo), max(plot_data_clean$Pheo)))
cat(sprintf("Ratio range: %.3f - %.3f\n", min(plot_data_clean$Ratio), max(plot_data_clean$Ratio)))
cat(sprintf("OM Ratio MIN: %.4f\n", min(plot_data_clean$Ratio)))
cat(sprintf("OM Ratio MAX: %.4f\n\n", max(plot_data_clean$Ratio)))
```

# Prepare data for stratigraphic plot

```{r}
# Load full pigment dataset for stratigraphic plot
LowRes_all <- read_excel("../data/SourceData/LR_data.xlsx")

# Extract pigment columns
LowRes_pig_d <- select(LowRes_all, 1, 25:111)

# Remove NA values
LR_pig <- na.omit(LowRes_pig_d)
depth_pig <- LR_pig[, 1]

# Function to remove columns with high proportion of zeros
remove_high_zero_columns <- function(data) {
  prop_zeros <- colMeans(data == 0)
  data[, prop_zeros <= 0.95]
}

# Remove columns with >95% zeros
df_pig <- as.data.frame(remove_high_zero_columns(LR_pig))

cat("Pigment dataset:\n")
cat(sprintf("Samples: %d\n", nrow(df_pig)))
cat(sprintf("Pigment variables: %d\n", ncol(df_pig) - 1))

# Debug: Check for near-zero variance and mostly-NA columns
cat("\nDetailed pigment column check:\n")
for(i in 2:min(15, ncol(df_pig))) {
  col_data <- df_pig[[i]]
  n_zeros <- sum(col_data == 0, na.rm = TRUE)
  n_nas <- sum(is.na(col_data))
  pct_zero <- 100 * n_zeros / length(col_data)
  col_max <- max(col_data, na.rm = TRUE)
  col_mean <- mean(col_data, na.rm = TRUE)
  cat(sprintf("  %s: zeros=%.1f%%, NAs=%d, max=%.2f, mean=%.4f\n", 
              names(df_pig)[i], pct_zero, as.integer(n_nas), col_max, col_mean))
}
```

# Stratigraphic plot: All pigments

```{r}
# Convert to percentages
perc_pig <- adorn_percentages(df_pig[, 2:ncol(df_pig)], denominator = "row", na.rm = TRUE)
perc_100 <- perc_pig * 100

# Remove pigments where total abundance is less than 5%
ma.sum <- colSums(perc_100)
df_pig_perc <- perc_100[, which(ma.sum > 5)]

cat(sprintf("Pigments > 5%% total abundance: %d\n", ncol(df_pig_perc)))
cat("\nAll pigments passing 5% threshold:\n")
pigments_above_threshold <- names(df_pig_perc)
for(i in seq_along(pigments_above_threshold)) {
  cat(sprintf("  %2d. %s (total: %.2f%%)\n", i, pigments_above_threshold[i], ma.sum[pigments_above_threshold[i]]))
}

# List of pigments being plotted
plotted_pigments <- c("Fuco", "Alfa_car", "Echi1", "Diadi", "Canta", "Zeax", "Lute", "OH_Sph.ene", "Isore", "Chloxant", "Okenone", "PhphT", "PhideT")
cat("\n\nPigments currently being plotted:\n")
for(i in seq_along(plotted_pigments)) {
  cat(sprintf("  %2d. %s\n", i, plotted_pigments[i]))
}

# Check for missing pigments
missing_pigments <- setdiff(pigments_above_threshold, plotted_pigments)
if(length(missing_pigments) > 0) {
  cat("\n⚠ WARNING: Pigments above 5% threshold but NOT plotted:\n")
  for(pig in missing_pigments) {
    cat(sprintf("  - %s (total: %.2f%%)\n", pig, ma.sum[pig]))
  }
} else {
  cat("\n✓ All pigments above 5% threshold are being plotted!\n")
}
```

```{r}
# Standardize data for cluster analysis
data <- as.matrix(df_pig_perc)
data_standardized <- data
for(i in 1:ncol(data_standardized)) {
  data_standardized[, i] <- data_standardized[, i] / sd(data_standardized[, i])
}

# Perform hierarchical cluster analysis
ma.dist <- vegdist(data_standardized, method = "euclidean", binary = FALSE, diag = FALSE, upper = FALSE, na.rm = FALSE)
ma.chclust <- chclust(ma.dist, method = "coniss")
```

```{r}
# Define color scheme for pigments
# Adjust colors based on pigment types
p.col <- c(rep("darkgreen", times = 5), rep("green", times = 5), 
           rep("orange", times = 5), rep("red", times = 5), 
           rep("purple", times = 5))

# Set depth axis
y.scale <- as.numeric(depth_pig[[1]])

# Create stratigraphic plot with all pigments
pig.plot <- strat.plot(df_pig_perc, 
                        yvar = y.scale, 
                        y.tks = y.scale, 
                        y.rev = TRUE, 
                        plot.line = FALSE, 
                        plot.poly = TRUE, 
                        plot.bar = TRUE, 
                        col.bar = "black", 
                        col.poly = p.col, 
                        col.poly.line = "black", 
                        scale.percent = TRUE, 
                        xSpace = 0.01, 
                        x.pc.inc = 5, 
                        x.pc.lab = TRUE, 
                        x.pc.omit0 = TRUE, 
                        srt.xlabel = 45, 
                        las = 2, 
                        col.exag = "auto", 
                        exag.mult = 3, 
                        exag.alpha = 0.50, 
                        title = "Holzmaar Pigments - All Types", 
                        ylabel = "Depth (cm)", 
                        clust = ma.chclust)

# Add cluster zones
addClustZone(pig.plot, ma.chclust, nZone = 5, lwd = 1.5, lty = 2, col = "grey25")
```

# Data preparation: Convert µmol to µg using molecular weights

```{r}
# First, let's identify all sheets in LR_data.xlsx
library(readxl)

all_sheets <- excel_sheets("../data/SourceData/LR_data.xlsx")
cat("Available sheets in LR_data.xlsx:\n")
for(i in seq_along(all_sheets)) {
  cat(sprintf("  %d. %s\n", i, all_sheets[i]))
}
```

```{r}
# Load molecular weights table
mw_table_raw <- read_excel("../data/SourceData/LR_data.xlsx", sheet = "Pig conversion")
cat("Molecular weights sheet loaded:\n")
cat(sprintf("  Total entries: %d\n", nrow(mw_table_raw)))
cat(sprintf("  Column names: %s\n", paste(names(mw_table_raw), collapse = ", ")))

cat("\n=== First 10 rows of raw data ===\n")
print(head(mw_table_raw, 10))

# Select columns 2 and 3 (Name and molecular weight g/mol), then rename
mw_table <- mw_table_raw %>%
  select(2, 3) %>%
  setNames(c("pigment_name", "molecular_weight")) %>%
  filter(!is.na(molecular_weight)) %>%
  mutate(molecular_weight = as.numeric(molecular_weight))

cat("\n=== After proper renaming and conversion ===\n")
cat(sprintf("  Pigments with valid weights: %d\n", sum(!is.na(mw_table$molecular_weight))))
cat("  First 10 entries:\n")
print(head(mw_table, 10))

# Create named vector for molecular weight lookup 
mw_vector <- setNames(mw_table$molecular_weight, mw_table$pigment_name)

cat("\nSample of molecular weight vector:\n")
print(head(mw_vector, 15))
cat("\nChecking for key pigments in MW vector:\n")
cat(sprintf("  PhphT present: %s\n", "PhphT" %in% names(mw_vector)))
cat(sprintf("  PhideT present: %s\n", "PhideT" %in% names(mw_vector)))

# If PhphT and PhideT don't have MWs, add them (these are total pheophytines/pheophorbides)
if (!("PhphT" %in% names(mw_vector))) {
  mw_vector["PhphT"] <- 893.5  # Approximate MW for Pheophytin a
  cat("  Added PhphT with MW = 893.5\n")
}
if (!("PhideT" %in% names(mw_vector))) {
  mw_vector["PhideT"] <- 879.5  # Approximate MW for Pheophorbide a
  cat("  Added PhideT with MW = 879.5\n")
}

# Define metadata columns
metadata_cols <- c("comp_depth_mm", "mean_VT22")

# Load OM-normalized pigment data
LR_all_raw <- read_excel("../data/SourceData/LR_data.xlsx", sheet = "Redox_Dia_PigOM")
cat("\n\nOM dataset (Redox_Dia_PigOM) columns:\n")
print(names(LR_all_raw))
cat("Total columns:", length(names(LR_all_raw)), "\n")

# Check if PhphT and PhideT exist
cat("\nChecking for pheophytin/pheophorbide columns:\n")
cat(sprintf("  PhphT in OM: %s\n", "PhphT" %in% names(LR_all_raw)))
cat(sprintf("  PhideT in OM: %s\n", "PhideT" %in% names(LR_all_raw)))

# Print all columns that might be pheophytins or pheophorbides
pheo_cols <- grep("pheo|Pheo|Phph|Phid", names(LR_all_raw), ignore.case = TRUE, value = TRUE)
cat("Columns matching 'pheo' pattern:", paste(pheo_cols, collapse = ", "), "\n")

# Filter to keep metadata + pigments with MWs + PhphT and PhideT if they exist
cols_to_keep <- c(metadata_cols, names(mw_vector))
cols_to_keep_adjusted <- cols_to_keep[cols_to_keep %in% names(LR_all_raw)]

# Add PhphT and PhideT even if not in mw_vector columns originally
if("PhphT" %in% names(LR_all_raw) && !("PhphT" %in% cols_to_keep_adjusted)) {
  cols_to_keep_adjusted <- c(cols_to_keep_adjusted, "PhphT")
  cat("\n✓ Adding PhphT to OM dataset\n")
}
if("PhideT" %in% names(LR_all_raw) && !("PhideT" %in% cols_to_keep_adjusted)) {
  cols_to_keep_adjusted <- c(cols_to_keep_adjusted, "PhideT")
  cat("✓ Adding PhideT to OM dataset\n")
}

cat("\nFinal columns to select from OM dataset:\n")
print(cols_to_keep_adjusted)

LR_all <- LR_all_raw %>% 
  select(all_of(cols_to_keep_adjusted))

cat("\nOM dataset after selection has columns:", paste(names(LR_all), collapse = ", "), "\n")

depth_LRall <- (LR_all$comp_depth_mm)/1000

# Load bulk pigment data
LR_all_bulk_raw <- read_excel("../data/SourceData/LR_all_bulk.xlsx")

cat("\n\nBulk dataset (LR_all_bulk) columns:\n")
print(names(LR_all_bulk_raw))
cat("Total columns:", length(names(LR_all_bulk_raw)), "\n")

pheo_cols_bulk <- grep("pheo|Pheo|Phph|Phid", names(LR_all_bulk_raw), ignore.case = TRUE, value = TRUE)
cat("Columns matching 'pheo' pattern:", paste(pheo_cols_bulk, collapse = ", "), "\n")

# Filter to keep metadata + pigments with MWs + PhphT and PhideT if they exist  
cols_to_keep_bulk <- cols_to_keep[cols_to_keep %in% names(LR_all_bulk_raw)]

if("PhphT" %in% names(LR_all_bulk_raw) && !("PhphT" %in% cols_to_keep_bulk)) {
  cols_to_keep_bulk <- c(cols_to_keep_bulk, "PhphT")
  cat("\n✓ Adding PhphT to bulk dataset\n")
}
if("PhideT" %in% names(LR_all_bulk_raw) && !("PhideT" %in% cols_to_keep_bulk)) {
  cols_to_keep_bulk <- c(cols_to_keep_bulk, "PhideT")
  cat("✓ Adding PhideT to bulk dataset\n")
}

cat("\nFinal columns to select from bulk dataset:\n")
print(cols_to_keep_bulk)

LR_all_bulk <- LR_all_bulk_raw %>% 
  select(all_of(cols_to_keep_bulk))

cat("\nBulk dataset after selection has columns:", paste(names(LR_all_bulk), collapse = ", "), "\n")

depth_LRall_bulk <- (LR_all_bulk$comp_depth_mm)/1000

cat("\nData filtered to pigments with molecular weights:\n")
cat(sprintf("  LR_all (OM): %d rows × %d columns\n", nrow(LR_all), ncol(LR_all)))
cat(sprintf("  LR_all_bulk (bulk): %d rows × %d columns\n", nrow(LR_all_bulk), ncol(LR_all_bulk)))

# Function to convert nmol/g to µg/g using molecular weights
# Conversion: µg/g = (nmol/g × MW(g/mol)) / 1000
convert_umol_to_ug <- function(data, mw_vector, metadata_cols = c("comp_depth_mm", "mean_VT22")) {
  # Identify pigment columns (all columns except metadata)
  pigment_cols <- setdiff(names(data), metadata_cols)
  
  # Create a copy for conversion
  data_converted <- data
  
  cat("  Conversion details:\n")
  # Convert each pigment column
  for(pigment in pigment_cols) {
    if(pigment %in% names(mw_vector)) {
      # Get molecular weight (g/mol) and ensure it's numeric
      mw <- as.numeric(mw_vector[pigment])
      
      # Get original data
      data_orig <- data[[pigment]]
      cat(sprintf("    %s:\n", pigment))
      cat(sprintf("      Original class: %s\n", class(data_orig)))
      cat(sprintf("      Original NA count: %d (of %d values)\n", sum(is.na(data_orig)), length(data_orig)))
      
      # Show first few original values
      cat(sprintf("      First 5 original values: %s\n", paste(head(data_orig, 5), collapse = ", ")))
      
      # Try to convert to numeric
      data_numeric <- as.numeric(data_orig)
      cat(sprintf("      After as.numeric - class: %s, NA count: %d\n", class(data_numeric), sum(is.na(data_numeric))))
      cat(sprintf("      First 5 converted values: %s\n", paste(head(data_numeric, 5), collapse = ", ")))
      
      # Check if mw is valid
      cat(sprintf("      MW value: %s (class: %s)\n", mw, class(mw)))
      
      # Do conversion with explicit checks
      if(!is.na(mw) && !all(is.na(data_numeric))) {
        result <- (data_numeric * mw) / 1000  # nmol/g × MW / 1000 = µg/g
        cat(sprintf("      Multiplication result - class: %s, NA count: %d\n", class(result), sum(is.na(result))))
        cat(sprintf("      First 5 result values: %s\n", paste(head(result, 5), collapse = ", ")))
        data_converted[[pigment]] <- result
      } else {
        cat(sprintf("      ⚠ Skipping multiplication: mw is NA or all data is NA\n"))
      }
    } else {
      cat(sprintf("    ⚠ %s not found in molecular weights\n", pigment))
    }
  }
  
  return(data_converted)
}

# Convert OM data
cat("\nConverting OM data (nmol/g → µg/g):\n")
cat(sprintf("Input data - rows: %d, cols: %d\n", nrow(LR_all), ncol(LR_all)))
cat("Column names in LR_all:\n")
print(names(LR_all))

LR_all_ug <- convert_umol_to_ug(LR_all, mw_vector)

cat("\nOutput OM data:\n")
cat(sprintf("  Rows: %d, Cols: %d\n", nrow(LR_all_ug), ncol(LR_all_ug)))
print(head(LR_all_ug, 3))

# Convert bulk data
cat("\n\nConverting bulk data (nmol/g → µg/g):\n")
cat(sprintf("Input data - rows: %d, cols: %d\n", nrow(LR_all_bulk), ncol(LR_all_bulk)))
cat("Column names in LR_all_bulk:\n")
print(names(LR_all_bulk))

LR_all_bulk_ug <- convert_umol_to_ug(LR_all_bulk, mw_vector)

cat("\nOutput Bulk data:\n")
cat(sprintf("  Rows: %d, Cols: %d\n", nrow(LR_all_bulk_ug), ncol(LR_all_bulk_ug)))
print(head(LR_all_bulk_ug, 3))

cat("\n✓ Data conversion complete\n")
```

# Diagnostic: Check data loading and conversion

```{r}
cat("=== Data Verification After Conversion ===\n\n")

cat("OM data (LR_all_ug) - converted to µg:\n")
cat(sprintf("  Dimensions: %d rows × %d columns\n", nrow(LR_all_ug), ncol(LR_all_ug)))
cat(sprintf("  Sample pigment values (first row):\n"))
sample_cols <- intersect(c("Fuco", "Alfa_car", "Zeax", "PhphT", "PhideT"), names(LR_all_ug))
for(col in sample_cols) {
  cat(sprintf("    %s: %.4f µg\n", col, LR_all_ug[[col]][1]))
}

cat("\nBulk data (LR_all_bulk_ug) - converted to µg:\n")
cat(sprintf("  Dimensions: %d rows × %d columns\n", nrow(LR_all_bulk_ug), ncol(LR_all_bulk_ug)))
cat(sprintf("  Sample pigment values (first row):\n"))
for(col in sample_cols) {
  if(col %in% names(LR_all_bulk_ug)) {
    cat(sprintf("    %s: %.4f µg\n", col, LR_all_bulk_ug[[col]][1]))
  }
}

cat("\nKey metadata columns:\n")
cat(sprintf("  LR_all_ug - mean_VT22: %s\n", ifelse("mean_VT22" %in% names(LR_all_ug), "✓ present", "✗ missing")))
cat(sprintf("  LR_all_bulk_ug - mean_VT22: %s\n", ifelse("mean_VT22" %in% names(LR_all_bulk_ug), "✓ present", "✗ missing")))

# Create pigment subset for plotting - select only pigments with valid molecular weights
# Remove pigments that are all zero (no molecular weight available)
metadata_cols <- c("comp_depth_mm", "mean_VT22")
pigment_cols <- setdiff(names(LR_all_bulk_ug), metadata_cols)

# Keep only pigments with non-zero values (i.e., they had molecular weights)
non_zero_pigments <- pigment_cols[colSums(LR_all_bulk_ug[, pigment_cols] != 0, na.rm = TRUE) > 0]

# Create subset with metadata + valid pigments
pig_all_bulk_ug <- as.data.frame(select(LR_all_bulk_ug, all_of(metadata_cols), all_of(non_zero_pigments)))

cat("\nPigment subset (pig_all_bulk_ug):\n")
cat(sprintf("  Dimensions: %d rows × %d columns\n", nrow(pig_all_bulk_ug), ncol(pig_all_bulk_ug)))
cat(sprintf("  Metadata: %s\n", paste(metadata_cols, collapse = ", ")))
cat(sprintf("  Pigments with valid data: %d\n", length(non_zero_pigments)))
cat(sprintf("  First 10 column names: %s\n", paste(names(pig_all_bulk_ug)[1:min(10, ncol(pig_all_bulk_ug))], collapse = ", ")))

cat("\n✓ All data prepared and ready for plotting\n")
```

```         

# Create pigment plots using consistent style from quick_diatom_plotting
Fuco <- ggplot(LR_all_ug , aes(y= Fuco, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "orange")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Fucoxanthine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Alpha <- ggplot(LR_all_ug , aes(y= Alfa_car, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "α-Carotene (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Echi <- ggplot(LR_all_ug , aes(y= Echi1, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Echinenone (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Diato <- ggplot(LR_all_ug , aes(y= Diadi, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Diadinoxanthine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Canta <- ggplot(LR_all_ug , aes(y= Canta, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Canthaxanthine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Zea <- ggplot(LR_all_ug , aes(y= Zeax, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Zeaxanthine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Lut <- ggplot(LR_all_ug , aes(y= Lute, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Lutein (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

OHSp <- ggplot(LR_all_ug , aes(y= OH_Sph.ene, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "OH-Spheroidene (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Iso <- ggplot(LR_all_ug , aes(y= Isore, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Isorenoretine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Chlx <- ggplot(LR_all_ug , aes(y= Chloxant, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Chloraxanthine (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Oke <- ggplot(LR_all_ug , aes(y= Okenone, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "purple")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Okenone (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Phph <- ggplot(LR_all_ug , aes(y= PhphT, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Total Pheophytines (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Phide <- ggplot(LR_all_ug , aes(y= PhideT, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Total Pheophorbides (µg)", x = " ") +
  theme(legend.position="none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

# Ratio plot (transparent bars, same scale as pigment plots)
Ratio_plot <- ggplot(plot_data_clean, aes(y = Ratio, x = depth)) +
  geom_bar(stat = "identity", width = 25, fill = alpha("steelblue", 0.5)) +
  scale_x_reverse(limits = c(max(plot_data_clean$depth, na.rm=TRUE), min(plot_data_clean$depth, na.rm=TRUE)), breaks = round(seq(min(plot_data_clean$depth, na.rm=TRUE), max(plot_data_clean$depth, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right') +
  theme_classic() +
  coord_flip() +
  labs(title = " ", y = "Chl/Pheo Ratio", x = " ") +
  theme(legend.position = "none", 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank(), 
        axis.text.y = element_blank())
```

# Pigments overlay (Bulk vs. OM-normalized)

```{r}
# Overlay plots comparing bulk sediment (LR_all_bulk_ug) with OM-normalized data (LR_all_ug)
# Both datasets converted from µmol to µg using molecular weights
# Bulk data shown in saturated color, OM-normalized shown in semi-transparent overlay

# First identify which pigments are available in both datasets (non-zero values)
metadata_cols <- c("comp_depth_mm", "mean_VT22")
pigment_cols_om <- setdiff(names(LR_all_ug), metadata_cols)
pigment_cols_bulk <- setdiff(names(LR_all_bulk_ug), metadata_cols)

# Find pigments with non-zero values in each dataset
om_pigments_nonzero <- pigment_cols_om[colSums(LR_all_ug[, pigment_cols_om] != 0, na.rm = TRUE) > 0]
bulk_pigments_nonzero <- pigment_cols_bulk[colSums(LR_all_bulk_ug[, pigment_cols_bulk] != 0, na.rm = TRUE) > 0]

# Find common pigments (those in both datasets with non-zero values)
common_pigments <- intersect(om_pigments_nonzero, bulk_pigments_nonzero)

# Force include PhphT and PhideT if they exist in either dataset
if("PhphT" %in% pigment_cols_om | "PhphT" %in% pigment_cols_bulk) {
  common_pigments <- c(common_pigments, "PhphT")
}
if("PhideT" %in% pigment_cols_om | "PhideT" %in% pigment_cols_bulk) {
  common_pigments <- c(common_pigments, "PhideT")
}
common_pigments <- unique(common_pigments)

cat("\nPigments available for overlay plots:\n")
cat(sprintf("  OM dataset non-zero pigments: %d\n", length(om_pigments_nonzero)))
cat(sprintf("  Bulk dataset non-zero pigments: %d\n", length(bulk_pigments_nonzero)))
cat(sprintf("  Common pigments (in both): %d\n", length(common_pigments)))
cat(sprintf("  List: %s\n", paste(sort(common_pigments), collapse = ", ")))

# Check which columns actually exist in datasets
cols_in_om <- intersect(names(LR_all_ug), common_pigments)
cols_in_bulk <- intersect(names(LR_all_bulk_ug), common_pigments)

cat("\nColumns found in OM dataset:", paste(cols_in_om, collapse = ", "), "\n")
cat("Columns found in bulk dataset:", paste(cols_in_bulk, collapse = ", "), "\n")

# Add missing columns with 0 values if needed
all_cols_needed <- unique(c(cols_in_om, cols_in_bulk))
for (col in all_cols_needed) {
  if (!col %in% names(LR_all_ug)) {
    LR_all_ug[[col]] <- 0
  }
  if (!col %in% names(LR_all_bulk_ug)) {
    LR_all_bulk_ug[[col]] <- 0
  }
}

# Filter datasets to common pigments
LR_all_ug_overlay <- select(LR_all_ug, all_of(metadata_cols), all_of(all_cols_needed))
LR_all_bulk_ug_overlay <- select(LR_all_bulk_ug, all_of(metadata_cols), all_of(all_cols_needed))

cat("✓ Overlay datasets have columns:", paste(setdiff(names(LR_all_ug_overlay), metadata_cols), collapse = ", "), "\n")

Fuco2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay, aes(y= Fuco*15, x= mean_VT22), stat = "identity", width = 25, fill = alpha("darkorange",1))+
  geom_bar(data= LR_all_ug_overlay , aes(y= Fuco, x= mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 50), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Fucoxanthine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Fucoxanthine\n (µg/g OM)", x = "Age (cal yr BP)") +
  theme(legend.position="none", axis.ticks.y= element_line(), axis.line.y = element_blank(), axis.text.y = element_text(size = 8), axis.title.x.bottom = element_text(color = "darkorange1"), axis.title.x.top = element_text(color = "gold3"))

Alpha2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Alfa_car*100, x= mean_VT22), stat = "identity", width = 25, fill = "firebrick4")+
  geom_bar(data= LR_all_ug_overlay , aes(y= Alfa_car, x= mean_VT22), stat = "identity", width = 25, fill = alpha("coral1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /100, name = "α-Carotene\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "α-Carotene\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "firebrick4"), axis.title.x.top = element_text(color = "coral2"))

Echi2 <- ggplot()+
  geom_bar(data=LR_all_bulk_ug_overlay , aes(y= Echi1*15, x= mean_VT22), stat = "identity", width = 25, fill = "firebrick4") +
  geom_bar(data= LR_all_ug_overlay, aes(y= Echi1, x= mean_VT22), stat = "identity", width = 25, fill = alpha("coral1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Echinenone\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Echinenone\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "firebrick4"), axis.title.x.top = element_text(color = "coral2"))

Diato2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Diadi*50, x= mean_VT22), stat = "identity", width = 25, fill = alpha("orangered",1))+
  geom_bar(data= LR_all_ug_overlay, aes(y= Diadi, x= mean_VT22), stat = "identity", width = 25, fill = alpha("yellow2", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /50, name = "Diadinoxanthine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Diadinoxanthine\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "orangered3"), axis.title.x.top = element_text(color = "yellow3"))

Canta2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Canta*50, x= mean_VT22), stat = "identity", width = 25, fill = "orangered")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Canta, x= mean_VT22), stat = "identity", width = 25, fill = alpha("yellow", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /50, name = "Canthaxanthine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Canthaxanthine\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "orangered3"), axis.title.x.top = element_text(color = "yellow3"))

Zea2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Zeax*50, x= mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Zeax, x= mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /50, name = "Zeaxanthine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Zeaxanthine\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))

Lut2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Lute*50, x= mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Lute, x= mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /50, name = "Lutein\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Lutein\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))

OHSp2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= OH_Sph.ene*50, x= mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
  geom_bar(data= LR_all_ug_overlay, aes(y= OH_Sph.ene, x= mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /50, name = "OH-Spheroidene\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "OH-Spheroidene\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))

Iso2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Isore*15, x= mean_VT22), stat = "identity", width = 25, fill = alpha("springgreen3", 1))+
  geom_bar(data= LR_all_ug_overlay, aes(y= Isore, x= mean_VT22), stat = "identity", width = 25, fill = alpha("olivedrab1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Isorenoretine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Isorenoretine\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "Springgreen3"), axis.title.x.top = element_text(color = "olivedrab3"))

Chlx2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Chloxant*15, x= mean_VT22), stat = "identity", width = 25, fill = "springgreen3")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Chloxant, x= mean_VT22), stat = "identity", width = 25, fill = alpha("olivedrab1", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Chloraxanthine\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Chloraxanthine\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "springgreen3"), axis.title.x.top = element_text(color = "olivedrab3"))

Oke2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Okenone*10, x= mean_VT22), stat = "identity", width = 25, fill = "purple3")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Okenone, x= mean_VT22), stat = "identity", width = 25, fill = alpha("violet", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /10, name = "Okenone\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Okenone\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "purple3"), axis.title.x.top = element_text(color = "violet"))


Chla2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= Chla*15, x= mean_VT22), stat = "identity", width = 25, fill = "darkgreen")+
  geom_bar(data= LR_all_ug_overlay, aes(y= Chla, x= mean_VT22), stat = "identity", width = 25, fill = alpha("green3", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Chlorophyll a\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Chlorophyll a\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkgreen"), axis.title.x.top = element_text(color = "green3"))

BChl2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= BChl*15, x= mean_VT22), stat = "identity", width = 25, fill = "darkgreen")+
  geom_bar(data= LR_all_ug_overlay, aes(y= BChl, x= mean_VT22), stat = "identity", width = 25, fill = alpha("green3", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Bacteriochlorophyll\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Bacteriochlorophyll\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkgreen"), axis.title.x.top = element_text(color = "green3"))


Phph2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= PhphT*15, x= mean_VT22), stat = "identity", width = 25, fill = "darkgreen")+
  geom_bar(data= LR_all_ug_overlay, aes(y= PhphT, x= mean_VT22), stat = "identity", width = 25, fill = alpha("green3", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Total Pheophytines\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Total Pheophytines\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkgreen"), axis.title.x.top = element_text(color = "green3"))

Phide2 <- ggplot()+
  geom_bar(data= LR_all_bulk_ug_overlay , aes(y= PhideT*15, x= mean_VT22), stat = "identity", width = 25, fill = "forestgreen")+
  geom_bar(data= LR_all_ug_overlay, aes(y= PhideT, x= mean_VT22), stat = "identity", width = 25, fill = alpha("green", 0.5))+
  scale_x_reverse(limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 500), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /15, name = "Total Pheophorbides\n (µg/g bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Total Pheophorbides\n (µg/g OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "forestgreen"), axis.title.x.top = element_text(color = "green"))

# Ratio plot - Overlay (Bulk vs. OM-normalized)
# Calculate ratios for both datasets using LR_all files for consistency
ratio_om <- LR_all_ug %>%
  filter(Chla > 0 & PhphT > 0) %>%
  mutate(Ratio = Chla / PhphT) %>%
  select(mean_VT22, Ratio)

ratio_bulk <- LR_all_bulk_ug %>%
  filter(Chla > 0 & PhphT > 0) %>%
  mutate(Ratio = Chla / PhphT) %>%
  select(mean_VT22, Ratio)

# Calculate ratio statistics for axis scaling
ratio_om_clean <- ratio_om %>% filter(!is.na(Ratio), is.finite(Ratio))
ratio_bulk_clean <- ratio_bulk %>% filter(!is.na(Ratio), is.finite(Ratio))

cat("\n=== RATIO STATISTICS FOR AXIS SCALING ===\n")
cat("OM-normalized Ratio (Chla / PhphT):\n")
cat(sprintf("  Min: %.4f\n", min(ratio_om_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Max: %.4f\n", max(ratio_om_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Mean: %.4f\n", mean(ratio_om_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Range: %.4f\n\n", max(ratio_om_clean$Ratio, na.rm = TRUE) - min(ratio_om_clean$Ratio, na.rm = TRUE)))

cat("Bulk Ratio (Chla / PhphT):\n")
cat(sprintf("  Min: %.4f\n", min(ratio_bulk_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Max: %.4f\n", max(ratio_bulk_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Mean: %.4f\n", mean(ratio_bulk_clean$Ratio, na.rm = TRUE)))
cat(sprintf("  Range: %.4f\n\n", max(ratio_bulk_clean$Ratio, na.rm = TRUE) - min(ratio_bulk_clean$Ratio, na.rm = TRUE)))


# Ratio plot - Overlay (Bulk vs. OM-normalized)
# Calculate ratios for both datasets using LR_all files for consistency
ratio_zoo <- LR_all_ug %>%
  filter(PhideT > 0 & PhphT > 0) %>%
  mutate(Ratio_z = PhideT / PhphT) %>%
  select(mean_VT22, Ratio_z)

ratio_zoo_bulk <- LR_all_bulk_ug %>%
  filter(PhideT > 0 & PhphT > 0) %>%
  mutate(Ratio_z = PhideT / PhphT) %>%
  select(mean_VT22, Ratio_z)

Ratio_3 <- ggplot()+
  geom_bar(data= ratio_zoo_bulk, aes(y= Ratio_z*2, x= mean_VT22), stat = "identity", width = 25, fill = "darkslateblue")+
  geom_bar(data= ratio_zoo, aes(y= Ratio_z, x= mean_VT22), stat = "identity", width = 25, fill = alpha("steelblue2", 0.5))+
  scale_x_reverse(position = 'top', limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 50), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /2, name = "Phide/Pheo Ratio\n (bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Phide/Pheo Ratio\n (OM)", x = "Age (cal yr BP)") +
  theme(legend.position="none", axis.ticks.y= element_line(), axis.line.y = element_blank(), axis.text.y = element_text(size = 8), axis.title.x.bottom = element_text(color = "darkslateblue"), axis.title.x.top = element_text(color = "steelblue2"))


Ratio2 <- ggplot()+
  geom_bar(data= ratio_bulk, aes(y= Ratio*2, x= mean_VT22), stat = "identity", width = 25, fill = "darkslateblue")+
  geom_bar(data= ratio_om, aes(y= Ratio, x= mean_VT22), stat = "identity", width = 25, fill = alpha("steelblue2", 0.5))+
  scale_x_reverse(position = 'top', limits = c(max(LR_all_ug$mean_VT22, na.rm=TRUE), min(LR_all_ug$mean_VT22, na.rm=TRUE)), breaks = round(seq(min(LR_all_ug$mean_VT22, na.rm=TRUE), max(LR_all_ug$mean_VT22, na.rm=TRUE), by = 50), digits = 0)) +
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /2, name = "Chl/Pheo Ratio\n (bulk)")) +
  theme_classic() +
  coord_flip()+ 
  labs(title = " ", y = "Chl/Pheo Ratio\n (OM)", x = " ") +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkslateblue"), axis.title.x.top = element_text(color = "steelblue2"))

```

# Wide plot: Overlay plots (Bulk vs. OM-normalized) + Ratio

```{r}

# Check which pigment plots exist (only include plots for pigments in common_pigments)
all_pigment_plots <- list(
  Fuco = if("Fuco" %in% common_pigments) Fuco2 else NULL,
  Alpha = if("Alfa_car" %in% common_pigments) Alpha2 else NULL,
  Echi = if("Echi1" %in% common_pigments) Echi2 else NULL,
  Diato = if("Diadi" %in% common_pigments) Diato2 else NULL,
  Canta = if("Canta" %in% common_pigments) Canta2 else NULL,
  Zea = if("Zeax" %in% common_pigments) Zea2 else NULL,
  Lut = if("Lute" %in% common_pigments) Lut2 else NULL,
  OHSp = if("OH_Sph.ene" %in% common_pigments) OHSp2 else NULL,
  Iso = if("Isore" %in% common_pigments) Iso2 else NULL,
  Chlx = if("Chloxant" %in% common_pigments) Chlx2 else NULL,
  Chla = if("Chla" %in% common_pigments) Chla2 else NULL,
  BChl = if("BChl" %in% common_pigments) BChl2 else NULL,
  Oke = if("Okenone" %in% common_pigments) Oke2 else NULL,
  Phph = if("PhphT" %in% common_pigments) Phph2 else NULL,
  Phide = if("PhideT" %in% common_pigments) Phide2 else NULL,
  Ratio2 = if("Ratio2" %in% common_pigments) Ratio2 else NULL,
  Ratio_3=if("Ratio_3" %in% common_pigments) Ratio_3 else NULL
)

# Remove NULL entries (pigments not in common_pigments)
plots_to_use <- Filter(Negate(is.null), all_pigment_plots)

cat("\nPlots to include in overlay grid:\n")
cat(sprintf("  Pigment plots: %d (out of 13 attempted)\n", length(plots_to_use)))
cat(sprintf("  List: %s\n", paste(names(plots_to_use), collapse = ", ")))

# # Add ratio plot (overlay: bulk vs OM-normalized)
plots_to_use$Ratio <- Ratio2
plots_to_use$Ratio_z <- Ratio_3

# Modify y-axis styling: show axis line only on first (left) and last (right) plots
# First plot: show axis.line.y on left side
if(length(plots_to_use) > 0) {
  plots_to_use[[1]] <- plots_to_use[[1]] + theme(axis.line.y = element_line())
  
  # Last plot: show axis.line.y on right side  
  if(length(plots_to_use) > 1) {
    plots_to_use[[length(plots_to_use)]] <- plots_to_use[[length(plots_to_use)]] + theme(axis.line.y = element_line())
  }
}

# Combine all plots in plot_grid
wide_overlay_plot <- do.call(plot_grid, c(plots_to_use, list(nrow = 1, align = "h", rel_widths = rep(1, length(plots_to_use)))))

print(wide_overlay_plot)

# Save the comprehensive overlay plot
ggsave("../figs/Pigments_Overlay_WithRatio.png", wide_overlay_plot, 
       width = (28 * length(plots_to_use) / 14), height = 8, dpi = 300, bg = "white")
cat("✓ Comprehensive overlay pigment plot saved: figs/Pigments_Overlay_WithRatio.png\n")
```

# Optional: Individual Chlorophyll-a and Pheophytin plots (if needed separately)

```{r}
# These are alternative individual plots if you want them separately
# Uncomment to use

# Bar plot 1: Chlorophyll-a
# p_chla <- ggplot(plot_data_clean, aes(y = Chla, x = plot_data_clean$depth)) +
#   geom_bar(stat = "identity", width = 25, fill = "darkgreen") +
#   scale_x_reverse(breaks = round(seq(min(plot_data_clean$depth), max(plot_data_clean$depth), by = 100), digits = 0)) +
#   scale_y_continuous(position = 'right') +
#   theme_classic() +
#   coord_flip() +
#   labs(title = " ", y = "Chlorophyll-a (µg/g)", x = " ") +
#   theme(legend.position = "none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

# Bar plot 2: Pheophytin-a
# p_pheo <- ggplot(plot_data_clean, aes(y = Pheo, x = plot_data_clean$depth)) +
#   geom_bar(stat = "identity", width = 25, fill = "orange") +
#   scale_x_reverse(breaks = round(seq(min(plot_data_clean$depth), max(plot_data_clean$depth), by = 100), digits = 0)) +
#   scale_y_continuous(position = 'right') +
#   theme_classic() +
#   coord_flip() +
#   labs(title = " ", y = "Pheophytin-a (µg/g)", x = " ") +
#   theme(legend.position = "none", axis.ticks.y = element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

# Plot information

```{r}
cat("\n=== PIGMENT PLOT DETAILS ===\n\n")
cat(sprintf("Total pigments displayed: %d\n", length(pig_individual_plots) - 1))
cat(sprintf("Pigments included:\n"))
for(i in 1:(length(pigment_names))) {
  cat(sprintf("  %d. %s (color: %s)\n", i, pigment_names[i], pig_colors[i]))
}
cat(sprintf("\nFinal plot: Chl-a / Pheophytin Ratio (color: steelblue)\n\n"))
cat("All plots are aligned vertically at the same depth scale.\n")
```

# Summary Statistics

```{r}
cat("\n=== PIGMENT SUMMARY STATISTICS ===\n\n")

# Chlorophyll statistics
cat("Chlorophyll-a:\n")
cat(sprintf("  Mean: %.2f µg/g\n", mean(plot_data_clean$Chla)))
cat(sprintf("  Median: %.2f µg/g\n", median(plot_data_clean$Chla)))
cat(sprintf("  Std Dev: %.2f µg/g\n", sd(plot_data_clean$Chla)))
cat(sprintf("  Range: %.2f - %.2f µg/g\n\n", min(plot_data_clean$Chla), max(plot_data_clean$Chla)))

# Pheophytin statistics
cat("Pheophytin-a:\n")
cat(sprintf("  Mean: %.2f µg/g\n", mean(plot_data_clean$Pheo)))
cat(sprintf("  Median: %.2f µg/g\n", median(plot_data_clean$Pheo)))
cat(sprintf("  Std Dev: %.2f µg/g\n", sd(plot_data_clean$Pheo)))
cat(sprintf("  Range: %.2f - %.2f µg/g\n\n", min(plot_data_clean$Pheo), max(plot_data_clean$Pheo)))

# Ratio statistics
cat("Chl-a / Pheophytin Ratio:\n")
cat(sprintf("  Mean: %.3f\n", mean(plot_data_clean$Ratio)))
cat(sprintf("  Median: %.3f\n", median(plot_data_clean$Ratio)))
cat(sprintf("  Std Dev: %.3f\n", sd(plot_data_clean$Ratio)))
cat(sprintf("  Range: %.3f - %.3f\n", min(plot_data_clean$Ratio), max(plot_data_clean$Ratio)))
```
