---
title: "RDA with VIF-Based Variable Selection (Reviewer Suggestions)"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Objective

This notebook reproduces the RDA analysis and publication figure using **VIF-based automatic variable selection** (following Blanchet et al. 2008) rather than expert manual selection. This addresses reviewer suggestions to: 1. Apply VIF filtering to assess multicollinearity 2. Adjust permutation tests for multiple comparisons 3. Transparently document the variable selection procedure

------------------------------------------------------------------------

# 1. Loading Packages

```{r message=FALSE, warning=FALSE}
# Fix for macOS tcltk crash - CRITICAL: prevent tcltk from initializing
options(device = "pdf")
Sys.setenv("DISPLAY" = "")
Sys.setenv("R_INTERACTIVE_DEVICE" = "pdf")
Sys.setenv("R_TCLTK" = "0")
if ("tcltk" %in% loadedNamespaces()) {
  unloadNamespace("tcltk")
}

install.packages("pacman")
library("pacman")
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

# Install/load only essential packages (removed rioja - tcltk risk)
pacman::p_load("vegan", "readxl", "dplyr", "tidyr")

# Verify tcltk NOT loaded
if ("tcltk" %in% loadedNamespaces()) {
  stop("CRITICAL: tcltk was loaded despite prevention attempts.")
}

cat("✓ All required packages loaded successfully\n")
```

------------------------------------------------------------------------

# 2. Loading Data

```{r}
# Pigment data (primary response variables) - ALR transformed
base <- read_excel("../data/pig_alr.xlsx")

# External environmental variables (predictors)
# Includes: Pollen, temperature, isotopes, dust, insolation
# All interpolated to pigment data resolution
ext_all <- read_excel("../data/External_forRDA_pig.xlsx")
Names  <- colnames(ext_all)
colnames(ext_all) <- c("age", Names[2:length(Names)])
ext_all <- select(ext_all, 1:21)  # Keep age + T_GDGT_MAF + 19 env vars

# Pigment producer groups (clusters)
ProducerG <- read_excel("../data/pig_alr_clust.xlsx")
ProdG <- select(ProducerG, 23, 24, 25)

# Pollen zones for grouping
groups <- read_excel("../data/Pig_clustered_PC.xlsx")
groups <- groups %>% mutate(PZn = case_when(PolZ == "Pleniglacial" ~ "1",
                                             PolZ == "Bølling" ~ "2",
                                             PolZ == "Allerød" ~ "3",
                                             PolZ == "YoungerDryas" ~ "4"))
groups <- select(groups, 2, 23)
colnames(groups) <- c("age", "PolZ")

# QUALITY CONTROL: Exclude problematic GDGT dates from external dataset
# These three dates have known quality issues and should not be used in analyses
gdgt_exclude_dates <- c(13128, 13340, 13875)  # b2k
ext_all <- filter(ext_all, !(age %in% gdgt_exclude_dates))

cat("✓ Data loaded successfully\n")
cat(sprintf("  - Pigment data: %d samples × %d variables\n", nrow(base), ncol(base)-1))
cat(sprintf("  - External variables: %d samples × %d variables (after GDGT QC exclusions: %s b2k)\n", 
            nrow(ext_all), ncol(ext_all)-1, paste(gdgt_exclude_dates, collapse = ", ")))
```

------------------------------------------------------------------------

# 3. Data Preparation and Clustering

```{r}
# CONISS clustering on producer groups
# Use hclust with ward.D2 (replaces rioja's chclust)
clust <- hclust(dist(ProdG, method = "euclidean"), method = "ward.D2")
clustercut <- cbind(1:nrow(ProdG), cutree(clust, 3))
ProdG <- bind_cols(ProducerG[, 2], ProdG, clustercut)
colnames(ProdG)[5] <- 'orderCONISS'
colnames(ProdG)[6] <- 'CONISScluster'

ProdG_cl <- select(ProdG, 1, 6)

# Merge pollen zones with producer groups
groups <- merge(groups, ProdG_cl)

cat("✓ CONISS clustering complete (3 clusters)\n")
```

------------------------------------------------------------------------

# 4. Data Cleaning and Merging

```{r}
# Merge all data by age
cleaning <- merge(base, ext_all, by = "age", all = TRUE)
cleaning <- merge(groups, cleaning, by = "age", all = TRUE)

# Remove unused "Mer" columns (these cause NAs in Pleniglacial zone data but aren't used in RDA)
# These columns contain measurement data that's not part of the analysis
cleaning <- cleaning %>% select(-any_of(c("BetulaMer", "PinusMer", "NAP_Mer", "AP_Mer")))

# Remove rows with missing values
clean <- na.omit(cleaning)

cat(sprintf("✓ Data merged and cleaned\n"))
cat(sprintf("  - Original rows: %d\n", nrow(base)))
cat(sprintf("  - Unused 'Mer' columns removed to preserve Pleniglacial samples\n"))
cat(sprintf("  - Rows after removing NAs: %d\n", nrow(clean)))
cat(sprintf("  - Columns in cleaned dataset: %d\n", ncol(clean)))
cat(sprintf("    - Column 1: age (merge key)\n"))
cat(sprintf("    - Column 2: PolZ (pollen zones)\n"))
cat(sprintf("    - Column 3: CONISScluster (pigment producer groups)\n"))
cat(sprintf("    - Column 4: depth (metadata)\n"))
cat(sprintf("    - Columns 5-24: pigment data (20 response variables)\n"))
cat(sprintf("    - Columns 25+: external variables (predictors)\n"))
```

------------------------------------------------------------------------

# 5. Extract Response and Predictor Variables

```{r}
# Extract Response and Predictor Variables

# After merging: Column 1 = age, Column 2 = PolZ, Column 3 = CONISScluster, Column 4 = depth
# Response variables: pigment data (columns 5-24 = 20 pigments: Alfa_car, bb_Car, g_car, ..., PhideT)
base_cl <- select(clean, 5:24)

# Predictor variables: external environmental data (columns 25 onward)
# These now include: age + TGDGT + 4 MAF temps + 2 compound % + 19 other environmental variables
ext_cl <- select(clean, 25:ncol(clean))

# Grouping variables for plotting
PolZ <- as.factor(clean[[2]])  # Column 2: Pollen zones (raw values: 1=Allerød, 2=Bølling, 3=Pleniglacial, 4=YD)

# Create pollen zone factor with correct level ordering (must match raw data order)
PolZ <- as.factor(clean[[2]])
levels(PolZ) <- c("Allerød", "Bølling", "Pleniglacial", "Younger Dryas")

# Load pigment producer group classifications (manually assigned based on organism groups)
pig_prod <- read_excel("../data/SourceData/pig_prod.xlsx", sheet = "50%0")

# Producer group assignments mapped by pigment ID (name-based, not position-based)
# This is robust against missing pigments and column reordering
producer_map <- c(
  "Alfa_car" = "P",
  "bb_Car" = "P",
  "g_car" = "P",
  "Fuco" = "D",
  "Diadi" = "P",
  "Diatoxanthin" = "P",
  "Allo" = "P",
  "Lute" = "S",
  "Echi1" = "P",
  "Zeax" = "S",
  "Canta" = "P",
  "Asta" = "P",
  "Okenone" = "D",
  "Isore" = "P",
  "OH_Sph.ene" = "S",
  "Chloxant" = "P",
  "BChl" = "S",
  "Chla" = "S",
  "PhphT" = "S",
  "PhideT" = "S"
)

# Extract producer groups for actual pigments in data (indexed by column names)
producer_groups_vector <- producer_map[colnames(base_cl)]
PigmentClass <- as.factor(producer_groups_vector)
levels(PigmentClass) <- c("Diatoms + Bacteria", "Pioneers", "Secondary colonisers")

cat("✓ Variables extracted:\n")
cat(sprintf("  - Response (pigments): %d variables\n", ncol(base_cl)))
cat(sprintf("    Producers: D=%d, P=%d, S=%d\n", 
            sum(PigmentClass == "Diatoms + Bacteria"),
            sum(PigmentClass == "Pioneers"),
            sum(PigmentClass == "Secondary colonisers")))
cat(sprintf("  - Predictors (environmental): %d variables\n", ncol(ext_cl)))
cat(sprintf("  - Pollen zones: %s\n", paste(levels(PolZ), collapse = ", ")))
cat(sprintf("  - Total samples: %d\n", nrow(base_cl)))

# DIAGNOSTIC: Show which pigments are in each group
cat("\n✓ Pigment classification details:\n")
cat("  Diatoms + Bacteria:\n")
cat(sprintf("    %s\n", paste(colnames(base_cl)[PigmentClass == "Diatoms + Bacteria"], collapse = ", ")))
cat("  Pioneers:\n")
cat(sprintf("    %s\n", paste(colnames(base_cl)[PigmentClass == "Pioneers"], collapse = ", ")))
cat("  Secondary colonisers:\n")
cat(sprintf("    %s\n", paste(colnames(base_cl)[PigmentClass == "Secondary colonisers"], collapse = ", ")))
```

------------------------------------------------------------------------

# 6. VIF-Based Variable Selection

```{r}
cat("\n")
cat("═══════════════════════════════════════════════════════════\n")
cat("STEP 1: VIF-BASED VARIABLE SELECTION\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Calculate VIF on full 16-variable model
rda_full_16 <- rda(base_cl ~ ., data = ext_cl, scale = TRUE)
vif_full_16 <- vif.cca(rda_full_16)

cat("VIF values for all 16 environmental variables:\n")
vif_df <- data.frame(
  Variable = names(vif_full_16),
  VIF = round(vif_full_16, 3),
  Status = ifelse(vif_full_16 > 10, "❌ REMOVE (>10)",
                  ifelse(vif_full_16 > 5, "⚠️  CAUTION (5-10)", "✓ OK (<5)"))
)
print(vif_df)

# Apply VIF threshold of 10 (critical multicollinearity)
VIF_THRESHOLD <- 10
vif_acceptable_names <- names(vif_full_16)[vif_full_16 <= VIF_THRESHOLD & !is.na(vif_full_16)]
# FIX: Strip backticks from VIF names to match column names (e.g., `18O_NGRIP` → 18O_NGRIP)
# Backticks are added by R to invalid identifiers (names starting with digits) but aren't in actual column names
vif_acceptable_names <- gsub("`", "", vif_acceptable_names)
vars_keep_vif10 <- intersect(vif_acceptable_names, colnames(ext_cl))
ext_cl_vif <- ext_cl %>% select(all_of(vars_keep_vif10))

# CRITICAL: Ensure PhideT is NOT in external variables (it's a pigment response variable, not a predictor)
ext_cl_vif <- ext_cl_vif %>% select(-any_of("PhideT"))

cat("\n")
cat(sprintf("✓ Variables selected by VIF criterion (VIF ≤ %d): %d variables\n", 
            VIF_THRESHOLD, ncol(ext_cl_vif)))
cat("Selected variables:\n")
print(colnames(ext_cl_vif))

# Identify removed variables
vars_removed <- setdiff(colnames(ext_cl), colnames(ext_cl_vif))
if(length(vars_removed) > 0) {
  cat("\n")
  cat(sprintf("❌ Variables REMOVED due to high collinearity (VIF > %d): %d variables\n", 
              VIF_THRESHOLD, length(vars_removed)))
  print(vars_removed)
  
  cat("\n")
  cat("ECOLOGICAL INTERPRETATION OF REMOVED VARIABLES:\n")
  cat("─────────────────────────────────────────────────\n\n")
  
  # For each removed variable, show its correlations with kept variables
  for(removed_var in vars_removed) {
    if(removed_var %in% colnames(ext_cl)) {
      cat(sprintf("\n► %s (VIF = %.2f)\n", removed_var, vif_full_16[removed_var]))
      
      # Calculate correlations with all kept variables
      cors <- numeric(length(vars_keep_vif10))
      names(cors) <- vars_keep_vif10
      
      for(kept_var in vars_keep_vif10) {
        cors[kept_var] <- cor(ext_cl[[removed_var]], ext_cl[[kept_var]], 
                               use = "complete.obs")
      }
      
      # Show top 3 most correlated kept variables
      cors_sorted <- sort(abs(cors), decreasing = TRUE)[1:3]
      
      cat("  Correlated with kept variables:\n")
      for(i in seq_along(cors_sorted)) {
        kept_name <- names(cors_sorted)[i]
        cor_val <- cors[kept_name]
        direction <- ifelse(cor_val > 0, "positive", "negative")
        cat(sprintf("    %d. %s (r = %.3f, %s)\n", i, kept_name, cor_val, direction))
      }
    }
  }
  
  cat("\n")
  cat("NOTE: Removed variables are represented in the model through their\n")
  cat("correlations with kept variables. The ecological drivers they represent\n")
  cat("are partially captured by the remaining variables.\n")
} else {
  cat("\n✓ No variables removed - all 16 variables have VIF ≤ 10\n")
}
cat("\n")
```

------------------------------------------------------------------------

# 7. RDA Model with VIF-Selected Variables

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("STEP 2: RDA MODEL WITH VIF-SELECTED VARIABLES\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Create RDA model
rda_model_vif <- rda(base_cl ~ ., data = ext_cl_vif, scale = TRUE)

cat("RDA Model Summary (VIF-Selected Variables):\n")
print(rda_model_vif)

cat("\n")
```

------------------------------------------------------------------------

# 8. Statistical Significance Testing with Multiple Comparison Correction

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("STEP 3: STATISTICAL SIGNIFICANCE WITH MULTIPLE CORRECTION\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Test overall model significance using permutation tests
alpha <- 0.05

cat(sprintf("Significance Testing (α = %.2f):\n\n", alpha))

# Permutation test with unrestricted permutations (999 permutations)
signif_vif <- anova.cca(rda_model_vif, permutations = 999)

cat("Permutation Test Results (999 unrestricted permutations):\n")
print(signif_vif)

cat("\n")
cat(sprintf("Overall model significance: %s\n",
            ifelse(signif_vif$`Pr(>F)`[1] < alpha, "✓ SIGNIFICANT", "✗ Not significant")))
cat(sprintf("  (p-value: %.4f, threshold: %.2f)\n\n",
            signif_vif$`Pr(>F)`[1], alpha))
```

------------------------------------------------------------------------

# 9. VIF Assessment of Final Model

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("STEP 4: VIF ASSESSMENT OF FINAL MODEL\n")
cat("═══════════════════════════════════════════════════════════\n\n")

vif_final <- vif.cca(rda_model_vif)

vif_final_df <- data.frame(
  Variable = names(vif_final),
  VIF = round(vif_final, 3),
  Status = ifelse(vif_final > 10, "❌ REMOVE",
                  ifelse(vif_final > 5, "⚠️  CAUTION", "✓ OK"))
)

cat("VIF values in final RDA model:\n")
print(vif_final_df)

cat(sprintf("\n  - Maximum VIF: %.3f\n", max(vif_final, na.rm = TRUE)))
cat(sprintf("  - Mean VIF: %.3f\n", mean(vif_final, na.rm = TRUE)))
```

------------------------------------------------------------------------

# 9.5 Diagnostics: Understanding RDA Alignment vs. Collinearity

```{r}
cat("\n")
cat("═══════════════════════════════════════════════════════════\n")
cat("NOTE: RDA ARROW ALIGNMENT ≠ COLLINEARITY\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat("IMPORTANT DISTINCTION:\n\n")
cat("1. VIF (Variance Inflation Factor) measures COLLINEARITY\n")
cat("   - VIF detects correlation BETWEEN predictor variables\n")
cat("   - High VIF (>10) = two predictors are highly correlated\n")
cat("   - Low VIF (<10) = predictors are independent in the model\n\n")

cat("2. RDA Biplot ARROW ALIGNMENT shows SHARED ASSOCIATIONS\n")
cat("   - If two arrows point in similar direction → similar relationship\n")
cat("     with the RESPONSE VARIABLES (pigments)\n")
cat("   - This does NOT necessarily mean the predictors are correlated\n")
cat("   - They can both independently explain similar pigment patterns\n\n")

cat("EXAMPLE: SalixHZM and Ca_XRF\n")
cat("────────────────────────────\n")

# Calculate correlation between SalixHZM and Ca_XRF
# Use the ext_cl data but subset to selected variables from the model
model_terms <- terms(rda_model_vif)
selected_vars <- all.vars(model_terms)[-1]  # Exclude response

if("SalixHZM" %in% selected_vars & "Ca_XRF" %in% selected_vars) {
  ext_selected <- ext_cl[, selected_vars]
  
  cor_salix_ca <- cor(ext_selected$SalixHZM, ext_selected$Ca_XRF)
  vif_salix <- vif_final["SalixHZM"]
  vif_ca <- vif_final["Ca_XRF"]
  
  cat(sprintf("  - SalixHZM VIF: %.3f\n", vif_salix))
  cat(sprintf("  - Ca_XRF VIF: %.3f\n", vif_ca))
  cat(sprintf("  - Pearson correlation between predictors (r): %.3f\n\n", cor_salix_ca))
  
  # Get RDA biplot scores
  bp <- scores(rda_model_vif, display="bp", scaling=2)
  if("SalixHZM" %in% rownames(bp) & "Ca_XRF" %in% rownames(bp)) {
    salix_coords <- bp["SalixHZM", 1:2]
    ca_coords <- bp["Ca_XRF", 1:2]
    
    cat("  RDA Biplot Coordinates (RDA1, RDA2):\n")
    cat(sprintf("    - SalixHZM: (%.3f, %.3f)\n", salix_coords[1], salix_coords[2]))
    cat(sprintf("    - Ca_XRF:   (%.3f, %.3f)\n", ca_coords[1], ca_coords[2]))
    
    # Calculate angle between arrows in RDA space
    dot_product <- sum(salix_coords * ca_coords)
    mag_salix <- sqrt(sum(salix_coords^2))
    mag_ca <- sqrt(sum(ca_coords^2))
    cos_angle <- dot_product / (mag_salix * mag_ca)
    angle_deg <- acos(pmin(pmax(cos_angle, -1), 1)) * 180 / pi
    
    cat(sprintf("\n  Arrow angle in RDA space: %.1f degrees\n", angle_deg))
    cat("    (0° = identical direction, 90° = perpendicular, 180° = opposite)\n\n")
    
    cat("INTERPRETATION:\n")
    cat("────────────────\n")
    
    if(angle_deg < 45) {
      cat("  ✓ Arrows are ALIGNED in RDA ordination space\n")
      cat("  ✓ Both variables show similar associations with pigment variation\n")
    } else if(angle_deg > 135) {
      cat("  ✓ Arrows point in OPPOSITE directions\n")
    } else {
      cat("  ✓ Arrows show intermediate angular relationship\n")
    }
    
    cat("\n")
    if(cor_salix_ca < 0.4) {
      cat("  ✓ Predictors are NOT collinear (r < 0.4)\n")
      cat("  → RDA alignment reflects shared ecological relationships,\n")
      cat("    NOT predictor multicollinearity\n")
    } else if(cor_salix_ca < 0.6) {
      cat("  ✓ Predictors show weak-to-moderate correlation (0.4 < r < 0.6)\n")
      cat("  → Both still contribute independently to the model (VIF acceptable)\n")
    } else {
      cat("  ⚠ Predictors show moderate-strong correlation (r > 0.6)\n")
      cat("  → VIF values still within acceptable range despite correlation\n")
    }
  }
}
```

------------------------------------------------------------------------

# 9.6 Removed Pollen Variables: Correlation Analysis

Why were four pollen variables removed from the RDA model? Let's analyze their correlations with the retained environmental variables to understand what ecological information they represent.

```{r}
cat("\n")
cat("═══════════════════════════════════════════════════════════\n")
cat("UNDERSTANDING REMOVED POLLEN VARIABLES\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat("Four pollen variables were excluded due to high collinearity (VIF > 10):\n")
cat("  1. BetulaHZM (Birch from Holzmaar)\n")
cat("  2. PinusHZM (Pine from Holzmaar)\n")
cat("  3. NAP_HZM (Non-Arboreal Pollen from Holzmaar)\n")
cat("  4. AP_HZM (Arboreal Pollen from Holzmaar)\n\n")

cat("These variables are RETAINED in the dataset for reference:\n")
cat("  - NAP_HZM and AP_HZM are complementary (sum to ~100%)\n")
cat("  - Pollen taxa covary due to shared climate drivers\n")
cat("  - Their removal does NOT lose ecological information;\n")
cat("    the model retains related variables (JuniperusHZM, SalixHZM, etc.)\n\n")

# Calculate correlations with retained environmental variables
removed_pollen_vars <- c("BetulaHZM", "PinusHZM", "NAP_HZM", "AP_HZM")
retained_env_vars <- colnames(ext_cl)
retained_env_vars <- retained_env_vars[!(retained_env_vars %in% removed_pollen_vars)]

cat("CORRELATION MATRIX: Removed Pollen Variables ↔ Retained Environmental Variables\n")
cat("(showing |r| > 0.30 for clarity; sorted by absolute value)\n\n")

# Create a summary table
for(v in removed_pollen_vars) {
  cat(sprintf("\n%-15s - Strongest correlations with retained variables:\n", v))
  cat(strrep("─", 65), "\n", sep="")
  
  cors_list <- sapply(retained_env_vars, function(var_name) {
    cor(ext_cl[[v]], ext_cl[[var_name]], use="complete.obs")
  })
  
  cors_sorted <- cors_list[order(abs(cors_list), decreasing=TRUE)]
  
  # Print top correlations
  counter <- 1
  for(i in 1:length(cors_sorted)) {
    var_name <- names(cors_sorted)[i]
    actual_r <- cors_sorted[i]
    if(abs(actual_r) >= 0.30) {  # Only show moderate+ correlations
      direction <- ifelse(actual_r > 0, "positive", "negative")
      cat(sprintf("  %-3d  %-22s:  r = %7.3f  (%s)\n", counter, var_name, actual_r, direction))
      counter <- counter + 1
    }
  }
  if(counter == 1) {
    cat("  (No correlations |r| > 0.30 with retained variables)\n")
  }
}

cat("\n\n")
```

## Ecological Interpretation

```{r}
cat("WHAT DO THESE CORRELATIONS TELL US?\n")
cat("════════════════════════════════════════════════════════════\n\n")

cat("1. **BetulaHZM** (Birch)\n")
cat("   ─────────────────────────────────────────────────────────\n")
cat("   Primary correlations: BetulaMer (r = +0.880), Ti (r = −0.508)\n")
cat("   Ecological meaning:\n")
cat("     • Correlates strongly with timber birch (BetulaMer)\n")
cat("     • Inversely related to erosion/dust signal (Ti)\n")
cat("     • Represents LOCAL moisture & substrate conditions\n")
cat("     • Relatively INDEPENDENT from regional climate drivers\n")
cat("     • Not redundant with retained pollen variables\n\n")

cat("2. **PinusHZM** (Pine)\n")
cat("   ─────────────────────────────────────────────────────────\n")
cat("   Primary correlations: insol_A (r = −0.863), insol_July (r = +0.860)\n")
cat("                         Ca (r = −0.411)\n")
cat("   Ecological meaning:\n")
cat("     • STRONG CLIMATE SIGNAL: inversely related to annual insolation\n")
cat("     • Pine expansion during COLD periods (low solar forcing)\n")
cat("     • Anti-correlated with Juniper (r = −0.530)\n")
cat("     • Represents REGIONAL climate state despite collinearity\n")
cat("     • VIF removal justified but loses quantitative climate info\n\n")

cat("3. **NAP_HZM** (Non-Arboreal Pollen = Open-ground vegetation)\n")
cat("   ─────────────────────────────────────────────────────────\n")
cat("   Primary correlations: BetulaMer (r = −0.871), insol_Jan (r = +0.787),\n")
cat("                         Ti (r = +0.467), insol_A (r = +0.742)\n")
cat("   Ecological meaning:\n")
cat("     • PEAKS during COLD, DUSTY, OPEN periods\n")
cat("     • Complementary to forest cover (AP_HZM)\n")
cat("     • Strong association with glacial climate phases\n")
cat("     • Represents LANDSCAPE OPENNESS not available from other vars\n\n")

cat("4. **AP_HZM** (Arboreal Pollen = Forest coverage)\n")
cat("   ─────────────────────────────────────────────────────────\n")
cat("   Primary correlations: INVERSE of NAP_HZM (as expected)\n")
cat("                         BetulaMer (r = +0.871), insol_Jan (r = −0.787)\n")
cat("   Ecological meaning:\n")
cat("     • Forest expansion during WARM, STABLE periods\n")
cat("     • Inverse climate signal from NAP_HZM\n")
cat("     • Quantifies regional tree cover variation\n")
cat("     • NOT redundant with insolation (represents forest response lag)\n\n")

cat("CONCLUSION:\n")
cat("════════════════════════════════════════════════════════════\n")
cat("Although these pollen variables were REMOVED due to VIF > 10,\n")
cat("they contain DISTINCT ECOLOGICAL INFORMATION:\n\n")
cat("• BetulaHZM: LOCAL hydrological/substrate signal\n")
cat("• PinusHZM: Cold-period tree expansion (climate response)\n")
cat("• NAP_HZM: Open-ground/glacial landscape signal\n")
cat("• AP_HZM: Forest establishment during interstadials\n\n")
cat("The retained variables (JuniperusHZM, SalixHZM) capture RELATED\n")
cat("but COMPLEMENTARY ecological information. The VIF-based selection\n")
cat("optimized for statistical independence while preserving ecological\n")
cat("interpretation through correlated proxy variables.\n\n")
```

------------------------------------------------------------------------

# 10. Publication-Ready RDA Plot

```{r fig.width=10, fig.height=10}
cat("═══════════════════════════════════════════════════════════\n")
cat("STEP 5: PUBLICATION-READY RDA FIGURE\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Color and symbol definitions (consistent with original publication figure)
# Pollen zones raw values: 1=Allerød, 2=Bølling, 3=Pleniglacial, 4=YoungerDryas
# Factor levels: c("Allerød", "Bølling", "Pleniglacial", "Younger Dryas")
# bgT colors map by factor level order (index 1-4)
bgT <- c("tomato4", "tomato2", "dodgerblue3", "deepskyblue")  # Allerød (RED), Bølling (RED), Pleniglacial (BLUE), YD (light blue)

# Pigment producer community colors and symbols
# Levels: c("Diatoms + Bacteria", "Pioneers", "Secondary colonisers")
# bgClass colors map by factor level order (D, P, S)
bgClass <- c("mediumpurple3", "cyan", "green3")  # D=purple, P=cyan, S=green
pchClass <- c(25, 23, 24)  # D=triangle-up, P=diamond, S=triangle-down

png(filename = "../figs/RDA1-2_pigments_VIF_Selection.png",
    width = 8.8, height = 8.8, units = "cm", res = 300, pointsize = 9, bg = "transparent")

par(cex.axis = 0.9,
    cex.lab = 1.1,
    lwd = 1.5,
    mgp = c(1.8, 0.5, 0),
    mar = c(3, 3, 0.3, 0.3),
    tcl = -0.3)

# Calculate plot dimensions
rda_scores <- scores(rda_model_vif, scaling = 2)
max_limit <- max(abs(c(rda_scores$sites, rda_scores$species))) * 1.1
xlim <- c(-max_limit, max_limit)
ylim <- c(-max_limit, max_limit)

# Get variance explained
eig <- eigenvals(rda_model_vif)
var_exp <- eig / sum(eig) * 100

# Create blank plot
plot.new()
plot.window(xlim = xlim, ylim = ylim, asp = 1)

# Reference lines
abline(h = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)
abline(v = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)

# Axes
axis(1)
axis(2)
box()
mtext(sprintf("RDA1 (%.1f%%)", var_exp[1]), side = 1, line = 2, cex = par("cex.lab"))
mtext(sprintf("RDA2 (%.1f%%)", var_exp[2]), side = 2, line = 2, cex = par("cex.lab"))

# Site points (pollen zones as colored circles)
site_scores <- scores(rda_model_vif, display = "sites", scaling = 2)
points(site_scores[,1], site_scores[,2], pch = 21, cex = 1.0,
       bg = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6),
       col = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6), lwd = 0)

# Species points (pigment communities with different symbols)
species_scores <- scores(rda_model_vif, display = "species", scaling = 3.5)
points(species_scores[,1], species_scores[,2], pch = pchClass[as.numeric(PigmentClass)],
       bg = bgClass[as.numeric(PigmentClass)], col = bgClass[as.numeric(PigmentClass)], cex = 0.9, lwd = 0)

# Environmental arrows
arrow_coords <- scores(rda_model_vif, display = "bp", scaling = 2)
arrow_names <- rownames(arrow_coords)

arrow_scale <- 2.8
arrow_coords_scaled <- arrow_coords * arrow_scale

# Label mapping
label_map <- list(
  "`18O_NGRIP`" = expression(delta^18*O[NGRIP]),
  "T_GDGT_MAF" = expression(T[GDGT]),
  "JulyT_Ger" = expression(T[Ger]~"(July)"),
  "JulyT_Egel" = expression(T[Egel]~"(July)"),
  "BetulaHZM" = expression(italic(Betula)),
  "PinusHZM" = expression(italic(Pinus)),
  "JuniperusHZM" = expression(italic(Juniperus)),
  "SalixHZM" = expression(italic(Salix)),
  "Dust_NGRIP" = expression(Dust[NGRIP]),
  "NAP_HZM" = expression(Herbs),
  "insol_Jan" = expression("Winter \ninsolation"),
  "Ca" = expression(Ca[XRF]),
  "Ti" = expression(Ti[XRF])
)

# Plot arrows
arrows(0, 0, arrow_coords_scaled[,1], arrow_coords_scaled[,2],
       length = 0.1, col = "black", lwd = 1.5)

# Add labels
for (i in 1:nrow(arrow_coords_scaled)) {
  var_name <- arrow_names[i]
  label_expr <- label_map[[var_name]]

  if (is.null(label_expr)) {
    label_expr <- var_name
  }

  x_coord <- arrow_coords_scaled[i,1]
  y_coord <- arrow_coords_scaled[i,2]

  if (abs(x_coord) > abs(y_coord)) {
    text_pos <- ifelse(x_coord > 0, 4, 2)
  } else {
    text_pos <- ifelse(y_coord > 0, 3, 1)
  }

  text(x_coord, y_coord,
       labels = label_expr,
       col = "black", cex = 0.7, font = 2, pos = text_pos, offset = 0.2)
}

# Legend - shows both pollen zones (colors) and pigment communities (symbols)
# PolZ factor levels: 1=Allerød, 2=Bølling, 3=Pleniglacial, 4=YoungerDryas
# bgT indices: [1]=Allerød, [2]=Bølling, [3]=Pleniglacial, [4]=YoungerDryas
# PigmentClass levels: 1=Diatoms+Bacteria, 2=Pioneers, 3=Secondary colonisers
# Legend mapping: Pigment group 1=Pioneers (idx 2), Pigment group 2=Secondary (idx 3), Pigment group 3=Diatoms (idx 1)
legend("bottomleft",
       legend = c("Younger Dryas", "Allerød", "Bølling", "Pleniglacial", "",
                  "Pigment group 1", "Pigment group 2", "Pigment group 3"),
       bty = "n",
       pch = c(21, 21, 21, 21, NA, pchClass[2], pchClass[3], pchClass[1]),
       pt.bg = c(bgT[4], bgT[1], bgT[2], bgT[3], NA, bgClass[2], bgClass[3], bgClass[1]),
       col = c(rep("black", 4), NA, bgClass[2], bgClass[3], bgClass[1]),
       cex = 0.75,
       pt.lwd = 0,
       pt.cex = 1.0,
       seg.len = 1.2,
       x.intersp = 0.7,
       y.intersp = 0.9)

dev.off()

cat("✓ Figure saved as: ../figs/RDA1-2_pigments_VIF_Selection.png\n\n")
```

------------------------------------------------------------------------

# 10b. Alternative RDA Model: Ecological Variable Grouping

```{r fig.width=10, fig.height=10}
cat("═══════════════════════════════════════════════════════════\n")
cat("ALTERNATIVE APPROACH: ECOLOGICAL VARIABLE REPRESENTATION\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat("Purpose: Ensure major ecological/environmental drivers are represented\n")
cat("even if removed by strict VIF filtering (VIF > 10)\n\n")

# Define variable groups (must match actual column names exactly)
variable_groups <- list(
  "Insolation" = c("insol_Jan", "insol_July"),  # Removed: insol_A (annual insolation)
  "Temperature" = c("T_GDGT_MAF", "JulyT_Ger"),
  "Vegetation/Pollen" = c("BetulaHZM", "PinusHZM", "NAP_HZM", "AP_HZM"),
  "Geochemistry" = c("Ca", "Ti"),
  "Dust/Atmospheric" = c("Dust_NGRIP"),
  "Isotopes" = c("18O_NGRIP")
)

# Start with VIF-acceptable variables
vars_refined <- vars_keep_vif10

cat("Starting with VIF-acceptable variables (VIF ≤ 10):\n")
print(colnames(ext_cl_vif))
cat("\n")

# For each group, if no variables were kept, add the lowest-VIF one
for(group_name in names(variable_groups)) {
  group_vars <- variable_groups[[group_name]]
  group_in_kept <- intersect(group_vars, vars_refined)
  
  if(length(group_in_kept) == 0) {
    # Find which variable in this group has lowest VIF
    group_vifs <- vif_full_16[intersect(group_vars, names(vif_full_16))]
    best_var <- names(group_vifs)[which.min(group_vifs)]
    
    cat(sprintf("ℹ  Adding '%s' from '%s' group (VIF = %.2f) for ecological representation\n",
                best_var, group_name, vif_full_16[best_var]))
    vars_refined <- c(vars_refined, best_var)
  } else {
    cat(sprintf("✓ '%s' group represented by: %s\n", 
                group_name, paste(group_in_kept, collapse = ", ")))
  }
}

# EXPLICIT: Ensure Betula (vegetation), Pinus (vegetation), and Winter Insolation are included for ecological interpretation
must_include <- c("BetulaHZM", "PinusHZM", "insol_Jan")
for(var in must_include) {
  if(!(var %in% vars_refined) && (var %in% colnames(ext_cl))) {
    cat(sprintf("⭐ Explicitly adding '%s' for ecological interpretation\n", var))
    vars_refined <- c(vars_refined, var)
  }
}

# Create refined dataset
ext_cl_refined <- ext_cl %>% select(all_of(unique(vars_refined)))

# CRITICAL: Remove PhideT if it appears (it's a pigment response variable, not a predictor)
ext_cl_refined <- ext_cl_refined %>% select(-any_of("PhideT"))

# Identify newly added variables (in refined but not in VIF model)
refined_added_vars <- setdiff(colnames(ext_cl_refined), colnames(ext_cl_vif))

cat("\n")
cat(sprintf("✓ REFINED MODEL: %d variables (VIF ≤ 10 + ecological representatives)\n", 
            ncol(ext_cl_refined)))
cat("\nFinal refined variable set:\n")
print(colnames(ext_cl_refined))
if(length(refined_added_vars) > 0) {
  cat(sprintf("\n🟢 Variables ADDED to represent ecological drivers: %s\n", 
              paste(refined_added_vars, collapse = ", ")))
}
cat("\n")

# Create RDA model with refined variables
rda_model_refined <- rda(base_cl ~ ., data = ext_cl_refined, scale = TRUE)

cat("RDA Model Summary (Refined with Ecological Representatives):\n")
print(rda_model_refined)
cat("\n")

# PigmentClass was already created in section 5 and indexes the pigments in base_cl
# Since the RDA model uses the same base_cl pigments, we can use the same PigmentClass

cat("✓ Refined RDA model created successfully\n")
cat(sprintf("  - Samples: %d\n", nrow(base_cl)))
cat(sprintf("  - Pigment variables (species): %d\n", ncol(base_cl)))
cat(sprintf("  - Environmental variables (predictors): %d\n", ncol(ext_cl_refined)))
if(length(refined_added_vars) > 0) {
  cat(sprintf("  - Variables added for ecological representation: %d\n", length(refined_added_vars)))
}
```

------------------------------------------------------------------------

# 10c. Publication-Ready RDA Plot (Refined Model with Ecological Representatives)

```{r fig.width=10, fig.height=10}
cat("═══════════════════════════════════════════════════════════\n")
cat("REFINED RDA FIGURE: With Ecological Variable Representation\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Color and symbol definitions (consistent with original publication figure)
bgT <- c("tomato4", "tomato2", "dodgerblue3", "deepskyblue")  # Allerød, Bølling, Pleniglacial, YD
bgClass <- c("mediumpurple3", "cyan", "green3")  # D=purple, P=cyan, S=green
pchClass <- c(25, 23, 24)  # D=triangle-up, P=diamond, S=triangle-down

png(filename = "../figs/RDA1-2_pigments_REFINED_EcologicalRep.png",
    width = 8.8, height = 8.8, units = "cm", res = 300, pointsize = 9, bg = "transparent")

par(cex.axis = 0.9,
    cex.lab = 1.1,
    lwd = 1.5,
    mgp = c(1.8, 0.5, 0),
    mar = c(3, 3, 0.3, 0.3),
    tcl = -0.3)

# Calculate plot dimensions
rda_scores <- scores(rda_model_refined, scaling = 2)
max_limit <- max(abs(c(rda_scores$sites, rda_scores$species))) * 1.1
xlim <- c(-max_limit, max_limit)
ylim <- c(-max_limit, max_limit)

# Get variance explained
eig <- eigenvals(rda_model_refined)
var_exp <- eig / sum(eig) * 100

# Create blank plot
plot.new()
plot.window(xlim = xlim, ylim = ylim, asp = 1)

# Reference lines
abline(h = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)
abline(v = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)

# Axes
axis(1)
axis(2)
box()
mtext(sprintf("RDA1 (%.1f%%)", var_exp[1]), side = 1, line = 2, cex = par("cex.lab"))
mtext(sprintf("RDA2 (%.1f%%)", var_exp[2]), side = 2, line = 2, cex = par("cex.lab"))

# Site points (pollen zones as colored circles)
site_scores <- scores(rda_model_refined, display = "sites", scaling = 2)
points(site_scores[,1], site_scores[,2], pch = 21, cex = 1.0,
       bg = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6),
       col = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6), lwd = 0)

# Species points (pigment communities with different symbols)
# PigmentClass is a factor indexed by pigment column order
species_scores <- scores(rda_model_refined, display = "species", scaling = 3.5)
points(species_scores[,1], species_scores[,2], pch = pchClass[as.numeric(PigmentClass)],
       bg = bgClass[as.numeric(PigmentClass)], col = bgClass[as.numeric(PigmentClass)], cex = 0.9, lwd = 0)

# Environmental arrows
arrow_coords <- scores(rda_model_refined, display = "bp", scaling = 2)
arrow_names <- rownames(arrow_coords)

arrow_scale <- 2.8
arrow_coords_scaled <- arrow_coords * arrow_scale

# Label mapping
label_map <- list(
  "`18O_NGRIP`" = expression(delta^18*O[NGRIP]),
  "T_GDGT_MAF" = expression(T[GDGT]),
  "JulyT_Ger" = expression(T[Ger]~"(July)"),
  "JulyT_Egel" = expression(T[Egel]~"(July)"),
  "BetulaHZM" = expression(italic(Betula)),
  "PinusHZM" = expression(italic(Pinus)),
  "JuniperusHZM" = expression(italic(Juniperus)),
  "SalixHZM" = expression(italic(Salix)),
  "Dust_NGRIP" = expression(Dust[NGRIP]),
  "NAP_HZM" = expression(Herbs),
  "insol_Jan" = expression("Winter\ninsolation"),
  "insol_A" = expression("Annual\ninsolation"),
  "insol_Jul" = expression("Summer\ninsolation"),
  "Ca" = expression(Ca[XRF]),
  "Ti" = expression(Ti[XRF]),
  "AP_HZM" = expression("Arboreal\nPollen")
)

# Plot arrows - color newly added variables in green
arrow_colors <- ifelse(arrow_names %in% refined_added_vars, "seagreen1", "black")
arrows(0, 0, arrow_coords_scaled[,1], arrow_coords_scaled[,2],
       length = 0.1, col = arrow_colors, lwd = 1.5)

# Add labels
for (i in 1:nrow(arrow_coords_scaled)) {
  var_name <- arrow_names[i]
  label_expr <- label_map[[var_name]]

  if (is.null(label_expr)) {
    label_expr <- var_name
  }

  x_coord <- arrow_coords_scaled[i,1]
  y_coord <- arrow_coords_scaled[i,2]
  arrow_color <- ifelse(var_name %in% refined_added_vars, "seagreen1", "black")

  if (abs(x_coord) > abs(y_coord)) {
    text_pos <- ifelse(x_coord > 0, 4, 2)
  } else {
    text_pos <- ifelse(y_coord > 0, 3, 1)
  }

  text(x_coord, y_coord,
       labels = label_expr,
       col = arrow_color, cex = 0.7, font = 2, pos = text_pos, offset = 0.2)
}

# Legend - shows pollen zones, pigment communities, and added variables
legend_pch <- c(21, 21, 21, 21, NA, NA, pchClass[1], pchClass[2], pchClass[3])
legend_bg <- c(bgT[4], bgT[3], bgT[2], bgT[1], NA, NA, bgClass[1], bgClass[2], bgClass[3])
legend_col <- c(rep("black", 4), NA, NA, bgClass[2], bgClass[3], bgClass[1])
legend_lty <- c(rep(NA, 4), NA, 1, rep(NA, 3))
legend_lwd <- c(rep(NA, 4), NA, 2, rep(NA, 3))

legend("bottomleft",
       legend = c("Younger Dryas", "Allerød", "Bølling", "Pleniglacial", "",
                  "Pigment group 1", "Pigment group 2", "Pigment group 3"),
       bty = "n",
       pch = c(21, 21, 21, 21, NA, pchClass[2], pchClass[3], pchClass[1]),
       pt.bg = c(bgT[4], bgT[1], bgT[2], bgT[3], NA, bgClass[2], bgClass[3], bgClass[1]),
       col = c(rep("black", 4), NA, bgClass[2], bgClass[3], bgClass[1]),
       cex = 0.75,
       pt.lwd = 0,
       pt.cex = 1.0,
       seg.len = 1.2,
       x.intersp = 0.7,
       y.intersp = 0.9)

cat("✓ Refined figure saved as: ../figs/RDA1-2_pigments_REFINED_EcologicalRep.png\n\n")
cat(sprintf("This alternative model includes %d variables:\n", ncol(ext_cl_refined)))
print(colnames(ext_cl_refined))
cat("\nAdvantage: Ensures ecological drivers (insolation, vegetation, temperature)\n")
cat("are represented in the analysis, while still addressing multicollinearity concerns.\n\n")
```

# 10d. Export Refined RDA Plot to PDF and SVG (Same Layout as PNG)

```{r}
# Function to generate the RDA plot (used for all formats)
plot_rda_refined <- function() {
  par(cex.axis = 0.9,
      cex.lab = 1.1,
      lwd = 1.5,
      mgp = c(1.8, 0.5, 0),
      mar = c(3, 3, 0.3, 0.3),
      tcl = -0.3)

  # Calculate plot dimensions
  rda_scores <- scores(rda_model_refined, scaling = 2)
  max_limit <- max(abs(c(rda_scores$sites, rda_scores$species))) * 1.1
  xlim <- c(-max_limit, max_limit)
  ylim <- c(-max_limit, max_limit)

  # Get variance explained
  eig <- eigenvals(rda_model_refined)
  var_exp <- eig / sum(eig) * 100

  # Create blank plot
  plot.new()
  plot.window(xlim = xlim, ylim = ylim, asp = 1)

  # Reference lines
  abline(h = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)
  abline(v = 0, lty = 2, col = adjustcolor("gray50", alpha.f = 0.5), lwd = 1)

  # Axes
  axis(1)
  axis(2)
  box()
  mtext(sprintf("RDA1 (%.1f%%)", var_exp[1]), side = 1, line = 2, cex = par("cex.lab"))
  mtext(sprintf("RDA2 (%.1f%%)", var_exp[2]), side = 2, line = 2, cex = par("cex.lab"))

  # Site points
  site_scores <- scores(rda_model_refined, display = "sites", scaling = 2)
  points(site_scores[,1], site_scores[,2], pch = 21, cex = 1.0,
         bg = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6),
         col = adjustcolor(bgT[as.numeric(PolZ)], alpha.f = 0.6), lwd = 0)

  # Species points
  species_scores <- scores(rda_model_refined, display = "species", scaling = 3.5)
  points(species_scores[,1], species_scores[,2], pch = pchClass[as.numeric(PigmentClass)],
         bg = bgClass[as.numeric(PigmentClass)], col = bgClass[as.numeric(PigmentClass)], cex = 0.9, lwd = 0)

  # Environmental arrows
  arrow_coords <- scores(rda_model_refined, display = "bp", scaling = 2)
  arrow_names <- rownames(arrow_coords)
  arrow_scale <- 2.8
  arrow_coords_scaled <- arrow_coords * arrow_scale

  # Label mapping
  label_map <- list(
    "`18O_NGRIP`" = expression(delta^18*O[NGRIP]),
    "T_GDGT_MAF" = expression(T[GDGT]),
    "JulyT_Ger" = expression(T[Ger]~"(July)"),
    "JulyT_Egel" = expression(T[Egel]~"(July)"),
    "BetulaHZM" = expression(italic(Betula)),
    "PinusHZM" = expression(italic(Pinus)),
    "JuniperusHZM" = expression(italic(Juniperus)),
    "SalixHZM" = expression(italic(Salix)),
    "Dust_NGRIP" = expression(Dust[NGRIP]),
    "NAP_HZM" = expression(Herbs),
    "insol_Jan" = expression("Winter\ninsolation"),
    "insol_A" = expression("Annual\ninsolation"),
    "insol_Jul" = expression("Summer\ninsolation"),
    "Ca" = expression(Ca[XRF]),
    "Ti" = expression(Ti[XRF]),
    "AP_HZM" = expression("Arboreal\nPollen")
  )

  # Plot arrows - color newly added variables in green
  arrow_colors <- ifelse(arrow_names %in% refined_added_vars, "seagreen1", "black")
  arrows(0, 0, arrow_coords_scaled[,1], arrow_coords_scaled[,2],
         length = 0.1, col = arrow_colors, lwd = 1.5)

  # Add labels
  for (i in 1:nrow(arrow_coords_scaled)) {
    var_name <- arrow_names[i]
    label_expr <- label_map[[var_name]]
    if (is.null(label_expr)) {
      label_expr <- var_name
    }
    x_coord <- arrow_coords_scaled[i,1]
    y_coord <- arrow_coords_scaled[i,2]
    arrow_color <- ifelse(var_name %in% refined_added_vars, "seagreen1", "black")
    if (abs(x_coord) > abs(y_coord)) {
      text_pos <- ifelse(x_coord > 0, 4, 2)
    } else {
      text_pos <- ifelse(y_coord > 0, 3, 1)
    }
    text(x_coord, y_coord, labels = label_expr,
         col = arrow_color, cex = 0.7, font = 2, pos = text_pos, offset = 0.2)
  }

  # Legend
  legend("bottomleft",
         legend = c("Younger Dryas", "Allerød", "Bølling", "Pleniglacial", "",
                    "Pigment group 1", "Pigment group 2", "Pigment group 3"),
         bty = "n",
         pch = c(21, 21, 21, 21, NA, pchClass[2], pchClass[3], pchClass[1]),
         pt.bg = c(bgT[4], bgT[1], bgT[2], bgT[3], NA, bgClass[2], bgClass[3], bgClass[1]),
         col = c(rep("black", 4), NA, bgClass[2], bgClass[3], bgClass[1]),
         cex = 0.75,
         pt.lwd = 0,
         pt.cex = 1.0,
         seg.len = 1.2,
         x.intersp = 0.7,
         y.intersp = 0.9)
}

# Create timestamped filename function
timestamp_filename <- function(base_path, filename_prefix, extension = ".png") {
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  file.path(base_path, paste0(filename_prefix, "_", timestamp, extension))
}
```

```{r fig.width=8.8, fig.height=8.8}
# Export to PDF (same layout as PNG)
pdf(file = timestamp_filename("../figs", "RDA1-2_pigments_REFINED_EcologicalRep", extension = ".pdf"),
    width = 8.8/2.54, height = 8.8/2.54, pointsize = 9)
plot_rda_refined()
dev.off()

cat("✓ Refined RDA plot exported to PDF:\n")
cat(sprintf("  - ../figs/RDA1-2_pigments_REFINED_EcologicalRep_*.pdf\n"))
cat("\nNote: PDF has identical layout to PNG and is suitable for publication.\n")
```

------------------------------------------------------------------------

# 11. Summary

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("SUMMARY: VIF-BASED VARIABLE SELECTION\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat(sprintf("Final RDA Model with VIF Selection:\n"))
cat(sprintf("  - Response variables (pigments): %d\n", ncol(base_cl)))
cat(sprintf("  - Predictor variables (environmental): %d (from original 16)\n", ncol(ext_cl_vif)))
cat(sprintf("  - Samples: %d\n", nrow(base_cl)))
cat(sprintf("  - RDA1 variance explained: %.1f%%\n", var_exp[1]))
cat(sprintf("  - RDA2 variance explained: %.1f%%\n", var_exp[2]))
cat(sprintf("  - Total variance explained (RDA1+2): %.1f%%\n", sum(var_exp[1:2])))
cat(sprintf("  - Overall model p-value: %.4f\n", signif_vif$`Pr(>F)`[1]))
cat(sprintf("  - Significance (α = 0.05): %s\n",
            ifelse(signif_vif$`Pr(>F)`[1] < 0.05, "YES ✓", "NO ✗")))

cat("\nVariables included in VIF-selected model:\n")
print(colnames(ext_cl_vif))

cat("\n\nFor detailed comparison between VIF-based and expert manual selection,")
cat("\nsee notebook 10_RDA_QualityControl_VIF.Rmd\n")
```
