---
title: "Chlorophyll a/Pheophytin Ratio and Diatom Community Analysis"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Objective

Investigate the relationship between the Chlorophyll a/Pheophytin (Chl a/Pheo) ratio and diatom community composition. The Chl a/Pheo ratio is a proxy for lake level/water depth: - **Lower ratio** (degraded Chl a) → shallow water, benthic diatoms dominate - **Higher ratio** (fresh Chl a) → deep water, planktonic diatoms dominate

This analysis tests this hypothesis using three complementary approaches.

------------------------------------------------------------------------

# 1. Loading Packages and Data

```{r message=FALSE, warning=FALSE}
install.packages("pacman")
library("pacman")
pacman::p_load("readxl", "dplyr", "tidyr", "vegan", "corrplot")

cat("✓ All required packages loaded successfully\n")
```

```{r}
# Load diatom data
diatoms_raw <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "diatoms")
diatom_names <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "diatoms_name")
diatom_total <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "total_abundance")

cat("✓ Diatom data loaded\n")
cat(sprintf("  - Diatom species data: %d samples × %d species\n", nrow(diatoms_raw), ncol(diatoms_raw)-3))
cat(sprintf("  - Diatom classification: %d species\n", nrow(diatom_names)))
cat(sprintf("  - Diatom groups: Planktonic (%d) + Benthic (%d)\n",
            sum(diatom_names$cat == "P"), sum(diatom_names$cat == "B")))

# Create mapping of species to category
species_cat_map <- setNames(diatom_names$cat, diatom_names$names)
```

```{r}
# Load Chl a/Pheo ratio - calculated same way as notebook 13
# Load raw pigment concentrations
pig_raw <- read_excel("../data/Pig_clean.xlsx")
pig_with_meta <- read_excel("../data/pig_alr.xlsx")

# Combine and calculate ratio
chl_pheo_data <- data.frame(
  depth = pig_with_meta$depth,
  age = pig_with_meta$age,
  Chla = pig_raw$Chla,
  Pheo = pig_raw$PhphT
)

# Remove NA and zero values
chl_pheo_clean <- chl_pheo_data %>%
  filter(Chla > 0 & Pheo > 0) %>%
  mutate(ratio = Chla / Pheo) %>%
  arrange(depth)

cat("Chl a/Pheophytin Ratio Data:\n")
cat(sprintf("  - Samples: %d\n", nrow(chl_pheo_clean)))
cat(sprintf("  - Chl a range: %.2f - %.2f µg/g\n", min(chl_pheo_clean$Chla), max(chl_pheo_clean$Chla)))
cat(sprintf("  - Pheophytin range: %.2f - %.2f µg/g\n", min(chl_pheo_clean$Pheo), max(chl_pheo_clean$Pheo)))
cat(sprintf("  - Ratio range: %.3f - %.3f\n", min(chl_pheo_clean$ratio), max(chl_pheo_clean$ratio)))
cat(sprintf("  - Mean ratio: %.3f\n\n", mean(chl_pheo_clean$ratio)))
```

------------------------------------------------------------------------

# 2. Data Preparation

```{r}
# Standardize diatom data to proportions (relative abundance)
# Remove the first 3 metadata columns
diatom_species <- diatoms_raw[, 4:ncol(diatoms_raw)]
diatom_ages <- diatoms_raw$mean_VT22

# Convert to relative abundance (proportions)
diatom_prop <- sweep(diatom_species, 1, rowSums(diatom_species), "/")

# Create planktonic/benthic index
# Get column names that are diatom species
species_names <- colnames(diatom_species)

# Map species to categories
species_cats <- sapply(species_names, function(sp) {
  # Match species name to diatom_names
  idx <- which(diatom_names$names == sp)
  if(length(idx) > 0) {
    return(diatom_names$cat[idx[1]])
  } else {
    return(NA)
  }
})

cat("Species categorization:\n")
cat(sprintf("  - Planktonic species in data: %d\n", sum(species_cats == "P", na.rm=TRUE)))
cat(sprintf("  - Benthic species in data: %d\n", sum(species_cats == "B", na.rm=TRUE)))
cat(sprintf("  - Unknown: %d\n", sum(is.na(species_cats))))

# Calculate benthic/planktonic proportions for each sample
benthic_prop <- rowSums(diatom_prop[, which(species_cats == "B")], na.rm=TRUE)
planktonic_prop <- rowSums(diatom_prop[, which(species_cats == "P")], na.rm=TRUE)

# Benthic/Planktonic index (higher = more benthic)
benthic_index <- benthic_prop / (benthic_prop + planktonic_prop)

cat("\nBenthic/Planktonic Index:\n")
cat(sprintf("  - Mean benthic proportion: %.3f\n", mean(benthic_index, na.rm=TRUE)))
cat(sprintf("  - Range: %.3f - %.3f\n", min(benthic_index, na.rm=TRUE), max(benthic_index, na.rm=TRUE)))
```

```{r}
# Merge diatom data with Chl a/Pheo ratio by DEPTH (not age)
# Depth should be more reliable for alignment

diatom_merge_data <- data.frame(
  depth = diatoms_raw$`comp_depth (mm)`,
  age = diatoms_raw$mean_VT22,
  benthic_prop = benthic_prop,
  planktonic_prop = planktonic_prop,
  benthic_index = benthic_index
)

# Merge with Chl/Pheo ratio data by DEPTH with tolerance
# Using nearest neighbor matching with max 10 mm tolerance
combined_data <- data.frame()
diatom_row_indices <- c()

for(i in 1:nrow(diatom_merge_data)) {
  diatom_depth <- diatom_merge_data$depth[i]
  
  # Find closest pigment sample (within 10 mm tolerance)
  pigment_idx <- which.min(abs(chl_pheo_clean$depth - diatom_depth))
  depth_diff <- abs(chl_pheo_clean$depth[pigment_idx] - diatom_depth)
  
  if(depth_diff <= 10) {  # 10 mm tolerance
    combined_data <- rbind(combined_data, data.frame(
      depth_diatom = diatom_depth,
      age_diatom = diatom_merge_data$age[i],
      benthic_prop = diatom_merge_data$benthic_prop[i],
      planktonic_prop = diatom_merge_data$planktonic_prop[i],
      benthic_index = diatom_merge_data$benthic_index[i],
      depth_chlpheo = chl_pheo_clean$depth[pigment_idx],
      Chla = chl_pheo_clean$Chla[pigment_idx],
      Pheo = chl_pheo_clean$Pheo[pigment_idx],
      ratio = chl_pheo_clean$ratio[pigment_idx],
      depth_diff_mm = depth_diff
    ))
    diatom_row_indices <- c(diatom_row_indices, i)
  }
}

cat("Combined Dataset (Diatoms + Chl a/Pheo Ratio - Depth-Based Merge):\n")
cat(sprintf("  - Merge tolerance: ± 10 mm depth difference\n"))
cat(sprintf("  - Samples with both data: %d\n", nrow(combined_data)))
cat(sprintf("  - Samples in diatom data only: %d\n", nrow(diatom_merge_data) - nrow(combined_data)))
cat(sprintf("  - Samples in Chl/Pheo data only: %d\n", nrow(chl_pheo_clean) - nrow(combined_data)))
cat(sprintf("  - Depth alignment success rate: %.1f%%\n", 100 * nrow(combined_data) / nrow(diatom_merge_data)))

if(nrow(combined_data) > 0) {
  cat(sprintf("\n  Average depth difference between matched samples: %.2f mm\n", mean(combined_data$depth_diff_mm)))
}

cat("\n✓ Data ready for correlation analyses\n")
```

------------------------------------------------------------------------

# 3. OPTION 1: Direct Correlation with Diatom Species Composition

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("OPTION 1: Direct Correlation (Ratio vs Diatom Species)\n")
cat("═══════════════════════════════════════════════════════════\n\n")

if(nrow(combined_data) == 0) {
  cat("WARNING: No overlapping samples. Cannot perform correlation.\n")
} else {
  # Extract diatom species proportions for overlapping samples (using stored indices)
  diatom_prop_overlap <- diatom_prop[diatom_row_indices, ]
  
  # Calculate correlations between Chl a/Pheo ratio and each species
  cor_list <- list()
  for(sp in colnames(diatom_prop_overlap)) {
    cor_test <- cor.test(combined_data$ratio, diatom_prop_overlap[, sp], 
                         method = "pearson", use = "complete.obs")
    cor_list[[sp]] <- data.frame(
      species = sp,
      correlation = as.numeric(cor_test$estimate),
      p_value = cor_test$p.value,
      category = species_cats[sp],
      stringsAsFactors = FALSE
    )
  }
  
  # Combine into data frame
  cor_results <- do.call(rbind, cor_list)
  rownames(cor_results) <- NULL
  
  # Sort by correlation strength
  cor_results <- cor_results[order(abs(cor_results$correlation), decreasing = TRUE), ]
  
  # Identify significant correlations (p < 0.05)
  cor_results$significant <- cor_results$p_value < 0.05
  
  cat("Correlation Results (Pearson r):\n")
  cat(sprintf("  - Planktonic species correlations:\n"))
  plank_cors <- cor_results[cor_results$category == "P", ]
  if(nrow(plank_cors) > 0) {
    for(i in 1:nrow(plank_cors)) {
      sig_marker <- ifelse(plank_cors$significant[i], " *", "")
      cat(sprintf("    %s: r = %.3f, p = %.4f%s\n", 
                  plank_cors$species[i], plank_cors$correlation[i], 
                  plank_cors$p_value[i], sig_marker))
    }
  }
  
  cat(sprintf("  - Benthic species correlations (top 10):\n"))
  benthic_cors <- cor_results[cor_results$category == "B", ]
  benthic_cors <- benthic_cors[order(abs(benthic_cors$correlation), decreasing = TRUE), ]
  if(nrow(benthic_cors) > 10) benthic_cors <- benthic_cors[1:10, ]
  for(i in 1:nrow(benthic_cors)) {
    sig_marker <- ifelse(benthic_cors$significant[i], " *", "")
    cat(sprintf("    %s: r = %.3f, p = %.4f%s\n", 
                benthic_cors$species[i], benthic_cors$correlation[i], 
                benthic_cors$p_value[i], sig_marker))
  }
  
  cat("\n* = p < 0.05\n\n")
  
  # Summary statistics
  cat("Summary:\n")
  cat(sprintf("  - Planktonic species with positive correlation: %d/%d\n",
              sum(plank_cors$correlation > 0), nrow(plank_cors)))
  cat(sprintf("  - Benthic species with negative correlation: %d/%d\n",
              sum(benthic_cors$correlation < 0), nrow(benthic_cors)))
  cat(sprintf("  - Significant correlations (p < 0.05): %d\n",
              sum(cor_results$significant)))
}
```

------------------------------------------------------------------------

# 4. OPTION 2: Comparing Planktonic vs Benthic-Dominated Communities

```{r fig.width=8, fig.height=6}
cat("═══════════════════════════════════════════════════════════\n")
cat("OPTION 2: Community Type Comparison (Planktonic vs Benthic)\n")
cat("═══════════════════════════════════════════════════════════\n\n")

if(nrow(combined_data) == 0) {
  cat("WARNING: No overlapping samples. Cannot perform analysis.\n")
} else {
  # Classify combined samples as planktonic or benthic-dominated
  threshold <- 0.5
  combined_data$community_type <- ifelse(combined_data$benthic_index > threshold, 
                                         "Benthic-dominated", "Planktonic-dominated")
  
  # Summary
  cat(sprintf("Sample classification (threshold: %d%% benthic):\n", threshold*100))
  cat(sprintf("  - Benthic-dominated: %d samples\n", sum(combined_data$community_type == "Benthic-dominated")))
  cat(sprintf("  - Planktonic-dominated: %d samples\n", sum(combined_data$community_type == "Planktonic-dominated")))
  
  # Compare Chl a/Pheo ratio between groups
  benthic_ratio <- combined_data$ratio[combined_data$community_type == "Benthic-dominated"]
  planktonic_ratio <- combined_data$ratio[combined_data$community_type == "Planktonic-dominated"]
  
  cat("\nChl a/Pheo Ratio by Community Type:\n")
  cat(sprintf("  - Benthic-dominated: Mean = %.3f, SD = %.3f, n = %d\n", 
              mean(benthic_ratio, na.rm=TRUE), sd(benthic_ratio, na.rm=TRUE), length(benthic_ratio)))
  cat(sprintf("  - Planktonic-dominated: Mean = %.3f, SD = %.3f, n = %d\n",
              mean(planktonic_ratio, na.rm=TRUE), sd(planktonic_ratio, na.rm=TRUE), length(planktonic_ratio)))
  
  # Statistical test (check if we have enough observations)
  if(length(na.omit(benthic_ratio)) >= 2 && length(na.omit(planktonic_ratio)) >= 2) {
    # Mann-Whitney U test - more robust for small samples
    mw_test <- wilcox.test(benthic_ratio, planktonic_ratio)
    cat(sprintf("\nMann-Whitney U Test:\n"))
    cat(sprintf("  - p-value: %.4f\n", mw_test$p.value))
    cat(sprintf("  - Significant difference (p < 0.05): %s\n",
                ifelse(mw_test$p.value < 0.05, "YES ✓", "NO")))
    
    # Also run t-test for comparison
    t_test <- t.test(benthic_ratio, planktonic_ratio)
    cat(sprintf("\nIndependent t-test (parametric):\n"))
    cat(sprintf("  - p-value: %.4f\n", t_test$p.value))
    cat(sprintf("  - Significant difference (p < 0.05): %s\n",
                ifelse(t_test$p.value < 0.05, "YES ✓", "NO")))
  } else {
    cat("\nWARNING: Insufficient observations in one or both groups for statistical testing.\n")
  }
  
  # Create boxplot only if we have data
  if(length(na.omit(benthic_ratio)) > 0 || length(na.omit(planktonic_ratio)) > 0) {
    par(mfrow=c(1,1), mar=c(5, 4, 2, 2))
    boxplot(list(benthic_ratio, planktonic_ratio),
            names = c("Benthic-dominated", "Planktonic-dominated"),
            ylab = "Chl a/Pheophytin Ratio",
            col = c("lightblue", "lightcyan"),
            main = "Chl a/Pheo Ratio by Diatom Community Type")
    
    # Add individual points
    stripchart(list(benthic_ratio, planktonic_ratio),
               vertical = TRUE,
               at = c(1, 2),
               pch = 19,
               col = rgb(0, 0, 0, 0.5),
               add = TRUE,
               method = "jitter")
  }
}
```

------------------------------------------------------------------------

# 5. OPTION 3: Benthic/Planktonic Index Correlation

```{r fig.width=8, fig.height=6}
cat("═══════════════════════════════════════════════════════════\n")
cat("OPTION 3: Benthic/Planktonic Index Correlation\n")
cat("═══════════════════════════════════════════════════════════\n\n")

if(nrow(combined_data) == 0) {
  cat("WARNING: No overlapping samples. Cannot perform analysis.\n")
} else {
  # Calculate correlation between benthic index and Chl a/Pheo ratio
  cor_pearson <- cor.test(combined_data$benthic_index, combined_data$ratio, 
                          method = "pearson")
  cor_spearman <- cor.test(combined_data$benthic_index, combined_data$ratio, 
                           method = "spearman")
  
  cat("Index Definition:\n")
  cat("  Benthic Index = Benthic proportion / (Benthic + Planktonic)\n")
  cat("  Range: 0 (100% planktonic) to 1 (100% benthic)\n\n")
  
  cat("Correlation Results:\n")
  cat(sprintf("  - Pearson r = %.3f, p-value = %.4f\n", 
              cor_pearson$estimate, cor_pearson$p.value))
  cat(sprintf("  - Spearman rho = %.3f, p-value = %.4f\n",
              cor_spearman$estimate, cor_spearman$p.value))
  
  # Interpret correlation
  expected_direction <- "negative"
  actual_direction <- ifelse(cor_pearson$estimate < 0, "negative", "positive")
  direction_match <- ifelse(expected_direction == actual_direction, "✓ EXPECTED", "✗ UNEXPECTED")
  
  cat(sprintf("\nInterpretation:\n"))
  cat(sprintf("  - Expected: NEGATIVE correlation\n"))
  cat(sprintf("  - Observed: %s correlation (%s)\n", actual_direction, direction_match))
  
  # Calculate R² to show explained variance
  r_squared <- cor_pearson$estimate^2 * 100
  cat(sprintf("  - Explained variance (R²): %.1f%%\n", r_squared))
  
  # Fit linear regression
  lm_model <- lm(combined_data$ratio ~ combined_data$benthic_index)
  
  # Create scatter plot with regression line
  plot(combined_data$benthic_index, combined_data$ratio,
       xlab = "Benthic/Planktonic Index",
       ylab = "Chl a/Pheophytin Ratio",
       main = "Relationship between Diatom Community Type and Pigment Ratio",
       pch = 19,
       col = rgb(0, 0, 0, 0.6),
       cex = 1.2)
  
  # Add regression line
  abline(lm_model, col = "red", lwd = 2)
  
  # Add correlation info to plot
  text(0.05, 0.95 * max(combined_data$ratio, na.rm=TRUE),
       sprintf("r = %.3f, p = %.4f\nR² = %.1f%%", 
               cor_pearson$estimate, cor_pearson$p.value, r_squared),
       pos = 4,
       cex = 1,
       bg = "white")
  
  # Print regression summary
  cat("\nLinear Regression Model:\n")
  print(summary(lm_model))
}
```

------------------------------------------------------------------------

# 6. Data Structure for Next Steps

```{r}
cat("═══════════════════════════════════════════════════════════\n")
cat("DATA READY FOR ANALYSIS\n")
cat("═══════════════════════════════════════════════════════════\n\n")

if(nrow(combined_data) == 0) {
  cat("⚠ WARNING: No samples matched between datasets!\n")
  cat("Samples with diatom data: %d\n", nrow(diatom_merge_data))
  cat("Samples with Chl/Pheo data: %d\n", nrow(chl_pheo_clean))
  cat("\nTroubleshooting:\n")
  cat("- Check age/depth alignment between datasets\n")
  cat("- May need to merge by depth instead of age\n")
  cat("- Consider interpolation if sampling resolutions differ\n")
} else {
  cat("✓ ANALYSIS COMPLETE\n\n")
  
  cat("Summary of Results:\n")
  cat(sprintf("- Samples analyzed: %d\n", nrow(combined_data)))
  cat(sprintf("- Benthic-dominated: %d (%.1f%%)\n", 
              sum(combined_data$benthic_index > 0.5),
              100 * sum(combined_data$benthic_index > 0.5) / nrow(combined_data)))
  cat(sprintf("- Planktonic-dominated: %d (%.1f%%)\n",
              sum(combined_data$benthic_index <= 0.5),
              100 * sum(combined_data$benthic_index <= 0.5) / nrow(combined_data)))
  
  cat("\nKey Data Structures:\n")
  cat("  1. combined_data: Merged diatom + Chl a/Pheo ratio data\n")
  cat("     - Columns: age, depth_diatom, benthic_prop, planktonic_prop,\n")
  cat("               benthic_index, depth_chlpheo, Chla, Pheo, ratio, community_type\n\n")
  
  cat("  2. diatom_prop_overlap: Species proportions for overlapping samples\n")
  cat(sprintf("     - Dimensions: %d samples × %d species\n", 
              nrow(diatom_prop_overlap), ncol(diatom_prop_overlap))
  )
  
  cat("\nAnalyses Performed:\n")
  cat("  ✓ OPTION 1: Direct correlation with diatom species\n")
  cat("  ✓ OPTION 2: Community type comparison (planktonic vs benthic)\n")
  cat("  ✓ OPTION 3: Benthic index regression (continuous)\n\n")
  
  cat("Next Steps:\n")
  cat("  1. Examine species-level correlations (Option 1)\n")
  cat("  2. Compare community types (Option 2)\n")
  cat("  3. Test benthic index relationship (Option 3)\n")
  cat("  4. Integrate results with RDA findings (lake level paleoclimate)\n")
  cat("  5. Create publication-quality figures combining all approaches\n")
}
```
