---
title: "Diatom Community Dendrogram with Barplots"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Objective

Create a hierarchical clustering visualization of diatom communities with:
- Dendrogram showing sample clustering
- Color bars indicating pollen zone (PolZ) and diatom group (planktonic vs benthic)
- Stacked barplots showing diatom species composition for each sample
- Similar structure to heatmap but with composition bars

------------------------------------------------------------------------

# 1. Load Libraries and Data

```{r message=FALSE, warning=FALSE}
install.packages("pacman")
library("pacman")
pacman::p_load("readxl", "dplyr", "tidyr", "vegan", "dendextend", "colorspace")

cat("✓ All required packages loaded successfully\n")
```

```{r}
# Load diatom data from diatoms_Lujan file
diatoms_raw <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "diatoms")
diatom_names <- read_excel("../data/SourceData/diatoms_Lujan.xlsx", sheet = "diatoms_name")

cat("✓ Diatom data loaded\n")
cat(sprintf("  - Samples: %d\n", nrow(diatoms_raw)))
cat(sprintf("  - Species: %d\n", ncol(diatoms_raw)-3))
cat(sprintf("  - Classification: %d species\n", nrow(diatom_names)))
```

```{r}
# Extract metadata and species data
sample_meta <- diatoms_raw %>% select(1:3) %>% 
  rename(
    sample_id = sample,
    depth_mm = `comp_depth (mm)`,
    age = mean_VT22
  )

# Extract diatom species counts (columns 4 onwards)
diatom_species <- diatoms_raw[, 4:ncol(diatoms_raw)]

# Convert to proportions
diatom_prop <- sweep(diatom_species, 1, rowSums(diatom_species), "/")

# Remove rows with NAs
na_rows <- which(rowSums(is.na(diatom_prop)) > 0)
if(length(na_rows) > 0) {
  cat(sprintf("Removing %d samples with NA values in diatom composition\n", length(na_rows)))
  diatom_prop <- diatom_prop[-na_rows, ]
  sample_meta <- sample_meta[-na_rows, ]
}

cat("Data preparation complete\n")
cat(sprintf("  - Sample metadata: %d rows\n", nrow(sample_meta)))
cat(sprintf("  - Species proportion matrix: %d × %d\n", nrow(diatom_prop), ncol(diatom_prop)))
```

------------------------------------------------------------------------

# 2. Assign Ecological Groups and Pollen Zones

```{r}
# Create diatom group classification (planktonic vs benthic)
species_cats <- sapply(colnames(diatom_species), function(sp) {
  idx <- which(diatom_names$names == sp)
  if(length(idx) > 0) {
    return(diatom_names$cat[idx[1]])
  } else {
    return(NA)
  }
})

# Create group factor for visualization
group_colors <- ifelse(species_cats == "P", "#1f77b4", "#ff7f0e")  # Blue for planktonic, Orange for benthic

cat("Species classification:\n")
cat(sprintf("  - Planktonic: %d\n", sum(species_cats == "P", na.rm=TRUE)))
cat(sprintf("  - Benthic+Epiphytic: %d\n", sum(species_cats == "B", na.rm=TRUE)))
```

```{r}
# Assign pollen zones based on age (simplified classification)
assign_pollenzone <- function(age) {
  if(is.na(age)) return(NA)
  if(age > 14000) return("Pleniglacial")
  if(age > 13000) return("Bølling")
  if(age > 12700) return("Allerød")
  if(age > 11700) return("Younger Dryas")
  return("Holocene")
}

sample_meta$PolZ <- sapply(sample_meta$age, assign_pollenzone)

# Define PolZ colors (matching your heatmap)
polz_color_map <- c(
  "Pleniglacial" = "#4575b4",
  "Bølling" = "#d73027",
  "Allerød" = "#fee090",
  "Younger Dryas" = "#91bfdb",
  "Holocene" = "#1a9850"
)

sample_meta$PolZ_color <- polz_color_map[sample_meta$PolZ]

cat("Pollen zone assignment:\n")
print(table(sample_meta$PolZ))
```

```{r}
# Create group labels (planktonic-dominated vs benthic-dominated)
# Based on proportions
benthic_prop <- rowSums(diatom_prop[, which(species_cats == "B")], na.rm=TRUE)
sample_meta$benthic_index <- benthic_prop / (benthic_prop + rowSums(diatom_prop[, which(species_cats == "P")], na.rm=TRUE))
sample_meta$group <- ifelse(sample_meta$benthic_index > 0.5, "Benthic-dominated", "Planktonic-dominated")

# Group colors
group_color_map <- c(
  "Planktonic-dominated" = "#1f77b4",
  "Benthic-dominated" = "#ff7f0e"
)
sample_meta$group_color <- group_color_map[sample_meta$group]

cat("Sample grouping:\n")
print(table(sample_meta$group))
```

------------------------------------------------------------------------

# 3. Hierarchical Clustering

```{r}
# Perform hierarchical clustering on species composition
# Using Bray-Curtis distance
dist_matrix <- vegdist(diatom_prop, method = "bray")

# Hierarchical clustering using complete linkage
hc <- hclust(dist_matrix, method = "complete")

# Create dendextend dendrogram
dend <- as.dendrogram(hc)

# Order samples by dendrogram
sample_order <- order.dendrogram(dend)

# Reorder data
diatom_prop_ordered <- diatom_prop[sample_order, ]
sample_meta_ordered <- sample_meta[sample_order, ]

cat("Hierarchical clustering complete\n")
cat(sprintf("  - Distance metric: Bray-Curtis\n"))
cat(sprintf("  - Linkage method: Complete\n"))
```

------------------------------------------------------------------------

# 4. Prepare Top Abundant Species

```{r}
# Get top species by mean abundance
mean_abundance <- colMeans(diatom_prop)
top_n <- 15
top_species <- names(sort(mean_abundance, decreasing = TRUE)[1:top_n])

cat(sprintf("Top %d most abundant species:\n", top_n))
for(i in 1:length(top_species)) {
  cat(sprintf("  %d. %s: %.2f%%\n", i, top_species[i], mean_abundance[top_species[i]]*100))
}

# Get group classification for top species
top_species_group <- species_cats[top_species]
```

------------------------------------------------------------------------

# 5. Create Stratigraphic-Style Dendrogram Figure

```{r fig.width=20, fig.height=14}
# Create stratigraphic plot with dendrogram on top
# Layout: [Dendrogram on top]
#         [Depth axis] [Species bars]

png(filename = sprintf("../figs/Diatom_Dendrogram_Barplot_%s.png", 
                       format(Sys.time(), "%Y%m%d_%H%M%S")),
    width = 2000, height = 1400, res = 150)

# Create layout: top=dendrogram, bottom left=depth axis, bottom right=species bars
layout_matrix <- matrix(c(
  1, 1, 1, 1, 1, 1, 1, 1,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3
), nrow=6, byrow=TRUE)

layout(layout_matrix)

# Panel 1: Dendrogram on top
par(mar=c(2,4,3,2))

# Create dendrogram with leaves colored by group
polz_bars <- sample_meta_ordered$PolZ_color
group_bars <- sample_meta_ordered$group_color

dend_colored <- dendextend::set(dend, "leaves_col", group_bars)

# Plot dendrogram (vertical orientation shown horizontally at top)
plot(dend_colored, 
     main="Diatom Community Clustering (Bray-Curtis)",
     xlab="",
     ylab="Distance",
     horiz=FALSE, leaflab="none")

# Panel 2: Depth/Age axis on the left (narrower)
par(mar=c(5,1,0,0.5))
plot(NA, xlim=c(0, 1), ylim=c(max(sample_meta_ordered$depth_mm), min(sample_meta_ordered$depth_mm)),
     xlab="", ylab="Depth (mm)", main="",
     xaxt="n", cex.lab=0.9)
axis(2, at=seq(min(sample_meta_ordered$depth_mm), max(sample_meta_ordered$depth_mm), by=500),
     las=1, cex.axis=0.7)

# Add PolZ colored background bands (narrower zones)
polz_unique <- unique(sample_meta_ordered[, c("PolZ", "PolZ_color")])
for(i in 1:nrow(polz_unique)) {
  zones <- which(sample_meta_ordered$PolZ == polz_unique$PolZ[i])
  if(length(zones) > 0) {
    min_depth <- min(sample_meta_ordered$depth_mm[zones])
    max_depth <- max(sample_meta_ordered$depth_mm[zones])
    rect(0, min_depth, 1, max_depth, 
         col=adjustcolor(polz_unique$PolZ_color[i], alpha.f=0.5),
         border=NA)
  }
}

# Panel 3: Species individual bars (like stratigraphic pollen diagram)
par(mar=c(5,0.5,0,2))

# Create space for all species
n_species <- length(top_species)
plot(NA, xlim=c(0, n_species), ylim=c(max(sample_meta_ordered$depth_mm), min(sample_meta_ordered$depth_mm)),
     xlab="", ylab="", xaxt="n", yaxt="n", main="")

# Draw species bars
for(sp_idx in 1:n_species) {
  species_col <- top_species[sp_idx]
  species_abund <- diatom_prop_ordered[, species_col]
  
  # Scale abundance to width (normalized by max)
  scaled_abund <- (species_abund / max(species_abund)) * 0.8
  
  # Draw polygon for each species showing abundance profile
  for(i in 1:nrow(diatom_prop_ordered)) {
    depth <- sample_meta_ordered$depth_mm[i]
    left <- (sp_idx - 1) + 0.1
    right <- left + scaled_abund[i]
    
    # Color by group type
    species_group <- species_cats[species_col]
    sp_color <- ifelse(species_group == "P", "#1f77b4", "#ff7f0e")
    
    polygon(c(left, right, right, left),
            c(depth-10, depth-10, depth+10, depth+10),
            col=sp_color, border=NA, alpha=0.7)
  }
  
  # Add species name at top
  text(sp_idx - 0.5, min(sample_meta_ordered$depth_mm) - 150, 
       species_col, srt=45, adj=c(0, 0.5), cex=0.7, xpd=TRUE)
  
  # Add vertical line between species
  lines(c(sp_idx, sp_idx), 
        c(min(sample_meta_ordered$depth_mm), max(sample_meta_ordered$depth_mm)),
        col="lightgray", lwd=0.5)
}

dev.off()

cat("✓ Stratigraphic figure saved\n")
```

------------------------------------------------------------------------

# 6. Create Alternative PDF Version

```{r fig.width=20, fig.height=14}
pdf(file = sprintf("../figs/Diatom_Dendrogram_Barplot_%s.pdf", 
                   format(Sys.time(), "%Y%m%d_%H%M%S")),
    width = 20, height = 14)

# Create layout: top=dendrogram, bottom left=depth axis, bottom right=species bars
layout_matrix <- matrix(c(
  1, 1, 1, 1, 1, 1, 1, 1,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3,
  2, 3, 3, 3, 3, 3, 3, 3
), nrow=6, byrow=TRUE)

layout(layout_matrix)

# Panel 1: Dendrogram on top
par(mar=c(2,4,3,2))

# Create dendrogram with leaves colored by group
polz_bars <- sample_meta_ordered$PolZ_color
group_bars <- sample_meta_ordered$group_color

dend_colored <- dendextend::set(dend, "leaves_col", group_bars)

# Plot dendrogram (vertical orientation shown horizontally at top)
plot(dend_colored, 
     main="Diatom Community Clustering (Bray-Curtis)",
     xlab="",
     ylab="Distance",
     horiz=FALSE, leaflab="none")

# Panel 2: Depth/Age axis on the left (narrower)
par(mar=c(5,1,0,0.5))
plot(NA, xlim=c(0, 1), ylim=c(max(sample_meta_ordered$depth_mm), min(sample_meta_ordered$depth_mm)),
     xlab="", ylab="Depth (mm)", main="",
     xaxt="n", cex.lab=0.9)
axis(2, at=seq(min(sample_meta_ordered$depth_mm), max(sample_meta_ordered$depth_mm), by=500),
     las=1, cex.axis=0.7)

# Add PolZ colored background bands (narrower zones)
polz_unique <- unique(sample_meta_ordered[, c("PolZ", "PolZ_color")])
for(i in 1:nrow(polz_unique)) {
  zones <- which(sample_meta_ordered$PolZ == polz_unique$PolZ[i])
  if(length(zones) > 0) {
    min_depth <- min(sample_meta_ordered$depth_mm[zones])
    max_depth <- max(sample_meta_ordered$depth_mm[zones])
    rect(0, min_depth, 1, max_depth, 
         col=adjustcolor(polz_unique$PolZ_color[i], alpha.f=0.5),
         border=NA)
  }
}

# Panel 3: Species individual bars (like stratigraphic pollen diagram)
par(mar=c(5,0.5,0,2))

# Create space for all species
n_species <- length(top_species)
plot(NA, xlim=c(0, n_species), ylim=c(max(sample_meta_ordered$depth_mm), min(sample_meta_ordered$depth_mm)),
     xlab="", ylab="", xaxt="n", yaxt="n", main="")

# Draw species bars
for(sp_idx in 1:n_species) {
  species_col <- top_species[sp_idx]
  species_abund <- diatom_prop_ordered[, species_col]
  
  # Scale abundance to width (normalized by max)
  scaled_abund <- (species_abund / max(species_abund)) * 0.8
  
  # Draw polygon for each species showing abundance profile
  for(i in 1:nrow(diatom_prop_ordered)) {
    depth <- sample_meta_ordered$depth_mm[i]
    left <- (sp_idx - 1) + 0.1
    right <- left + scaled_abund[i]
    
    # Color by group type
    species_group <- species_cats[species_col]
    sp_color <- ifelse(species_group == "P", "#1f77b4", "#ff7f0e")
    
    polygon(c(left, right, right, left),
            c(depth-10, depth-10, depth+10, depth+10),
            col=sp_color, border=NA, alpha=0.7)
  }
  
  # Add species name at top
  text(sp_idx - 0.5, min(sample_meta_ordered$depth_mm) - 150, 
       species_col, srt=45, adj=c(0, 0.5), cex=0.7, xpd=TRUE)
  
  # Add vertical line between species
  lines(c(sp_idx, sp_idx), 
        c(min(sample_meta_ordered$depth_mm), max(sample_meta_ordered$depth_mm)),
        col="lightgray", lwd=0.5)
}

dev.off()

cat("✓ PDF figure saved\n")
```

------------------------------------------------------------------------

# 7. Summary Statistics

```{r}
cat("════════════════════════════════════════════\n")
cat("ANALYSIS SUMMARY\n")
cat("════════════════════════════════════════════\n\n")

cat("Dataset:\n")
cat(sprintf("  - Total samples: %d\n", nrow(sample_meta)))
cat(sprintf("  - Diatom species: %d\n", ncol(diatom_species)))
cat(sprintf("  - Depth range: %.2f - %.2f mm\n", min(sample_meta$depth_mm), max(sample_meta$depth_mm)))

cat("\nSample Classification:\n")
for(polz in unique(sample_meta$PolZ)) {
  n <- sum(sample_meta$PolZ == polz)
  cat(sprintf("  - %s: %d samples\n", polz, n))
}

cat("\nCommunity Type:\n")
for(group in unique(sample_meta$group)) {
  n <- sum(sample_meta$group == group)
  cat(sprintf("  - %s: %d samples\n", group, n))
}

cat("\nTop 15 Species (by mean abundance):\n")
for(i in 1:length(top_species)) {
  cat(sprintf("  %2d. %-40s (%.2f%%, %s)\n", 
              i, top_species[i], 
              mean_abundance[top_species[i]]*100,
              top_species_group[i]))
}

cat("\n✓ Analysis complete!\n")
cat("  - Figure saved: ../figs/Diatom_Dendrogram_Barplot_[timestamp].png\n")
cat("  - PDF saved: ../figs/Diatom_Dendrogram_Barplot_[timestamp].pdf\n")
```

