---
title: "Pollen Modelling: Predicting Holzmaar from Meerfelder Maar Using Scaling Factors"
output: html_notebook
---

# Objective

Model missing Holzmaar pollen data (13258 to 10995 BP) using Meerfelder Maar pollen data with a scaling factor approach.

**Strategy:** 1. Compare Meerfelder Maar and Holzmaar data in the overlapping period (where both exist) 2. Calculate scaling factors and offsets for each pollen type 3. Check if trends are similar between the two sites 4. Apply the scaling to Meerfelder Maar data in the missing period (13258-10995 BP)

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(zoo)

# Set working directory
knitr::opts_knit$set(root.dir = here::here())
```

## Load Data

```{r load_data}
# Load the main Holzmaar pollen dataset from Excel
PollenDat <- read_excel("data/SourceData/Pollen_Litt_Usinger_stacked.xlsx", 
                        sheet = "LittLittStackedPer", col_types = "numeric")

cat("=== Excel Pollen Data ===\n")
cat("Dimensions:", nrow(PollenDat), "rows x", ncol(PollenDat), "columns\n")
cat("Column names:", paste(head(names(PollenDat), 20), collapse = ", "), "...\n")
cat("Age range:", range(PollenDat[,1], na.rm = TRUE), "\n\n")

# Check if this dataset covers the target period (10995-13258 BP)
target_data <- PollenDat %>%
  filter(!is.na(.[,1]) & .[,1] >= 10995 & .[,1] <= 13258)

cat("Data points in target period (10995-13258 BP):", nrow(target_data), "\n")
if(nrow(target_data) > 0) {
  cat("  Age range in target period:", range(target_data[,1], na.rm = TRUE), "\n")
  cat("  This Excel file HAS data in the target period!\n\n")
} else {
  cat("  No data in target period from Excel file\n\n")
}

# Load Holzmaar pollen data from CSV (simplified version)
HolzPol <- read.csv("data/SourceData/External/csvs/HolzPol.csv", sep = ";")
cat("=== CSV Holzmaar Data ===\n")
cat("Age range:", range(HolzPol$age), "\n")
cat("Columns:", paste(names(HolzPol), collapse = ", "), "\n\n")

# Load Meerfelder Maar pollen data (predictors - complete dataset)
MerPinus <- read.csv("data/SourceData/External/csvs/MerPinus.csv", sep = ";")
MerBetula <- read.csv("data/SourceData/External/csvs/MerBetula.csv", sep = ";")
MerHerbs <- read.csv("data/SourceData/External/csvs/MerHerbs.csv", sep = ";")

# Rename columns for consistency
colnames(MerPinus) <- c("age", "Pinus_Mer")
colnames(MerBetula) <- c("age", "Betula_Mer")
colnames(MerHerbs) <- c("age", "Herbs_Mer", "TreesShrubs_Mer")

cat("=== Meerfelder Maar Data ===\n")
cat("Pinus age range:", range(MerPinus$age), "\n")
cat("Betula age range:", range(MerBetula$age), "\n")
cat("Herbs age range:", range(MerHerbs$age), "\n")
```

## Check for Existing Holzmaar Data in Target Period

```{r check_excel_data}
# Extract relevant columns from Excel data
# Assuming first column is age, look for Betula, Pinus, AP/NAP columns
if(nrow(target_data) > 0) {
  cat("=== Exploring Excel Data in Target Period ===\n\n")
  
  # Show first few rows to understand structure
  cat("First 10 rows of Excel data in target period:\n")
  print(head(target_data, 10))
  
  # Try to identify relevant columns
  # Common pollen column patterns: Betula, Pinus, AP, NAP, etc.
  pollen_cols <- grep("Betula|Pinus|AP|NAP|Tree|Herb", names(PollenDat), 
                      ignore.case = TRUE, value = TRUE)
  
  if(length(pollen_cols) > 0) {
    cat("\n\nPotential pollen columns found:\n")
    cat(paste(pollen_cols, collapse = ", "), "\n\n")
    
    # Extract these columns for the target period
    Excel_target <- target_data %>%
      select(1, any_of(pollen_cols))
    
    cat("Sample of Excel pollen data in target period:\n")
    print(head(Excel_target, 10))
  }
}
```

## Merge Meerfelder Maar Data

```{r merge_meerfelder}
# Merge all Meerfelder Maar datasets
Mer_combined <- MerPinus %>%
  full_join(MerBetula, by = "age") %>%
  full_join(MerHerbs, by = "age") %>%
  arrange(age)

# Check for the target range (10995-13258 BP - where we need to predict Holzmaar)
Mer_target <- Mer_combined %>%
  filter(age >= 10995 & age <= 13258)

cat("Meerfelder Maar data points in target range (10995-13258):", nrow(Mer_target), "\n")
cat("Age range of Meerfelder Maar target data:", range(Mer_target$age, na.rm = TRUE), "\n")
head(Mer_target)
tail(Mer_target)
```

## Analyze Overlapping Period

```{r analyze_overlap}
# Find the overlapping period where both datasets exist
# Holzmaar: 0-13258 BP
# Meerfelder Maar: check its range
mer_min <- min(Mer_combined$age, na.rm = TRUE)
mer_max <- max(Mer_combined$age, na.rm = TRUE)

cat("Holzmaar age range: 0 -", max(HolzPol$age), "BP\n")
cat("Meerfelder Maar age range:", mer_min, "-", mer_max, "BP\n\n")

# Define overlapping period for comparison
# The overlap is where BOTH datasets have data
overlap_start <- max(0, mer_min)
overlap_end <- min(max(HolzPol$age), mer_max)

cat("Overlapping period for analysis:", overlap_start, "-", overlap_end, "BP\n\n")

if(overlap_start >= overlap_end) {
  stop("No overlap between Holzmaar and Meerfelder Maar datasets!\n",
       "Holzmaar: 0-", max(HolzPol$age), " BP\n",
       "Meerfelder Maar: ", mer_min, "-", mer_max, " BP")
}

# Filter Holzmaar data for overlapping period
Holz_overlap <- HolzPol %>%
  filter(age >= overlap_start & age <= overlap_end) %>%
  select(age, Betula, Pinus, NAP, AP)

cat("Holzmaar data points in overlap:", nrow(Holz_overlap), "\n")

# Check if we have Meerfelder Maar data in this range
Mer_in_range <- Mer_combined %>%
  filter(age >= overlap_start & age <= overlap_end) %>%
  summarise(
    n = n(),
    has_pinus = sum(!is.na(Pinus_Mer)),
    has_betula = sum(!is.na(Betula_Mer)),
    has_trees = sum(!is.na(TreesShrubs_Mer))
  )

cat("Meerfelder Maar data in overlap:\n")
print(Mer_in_range)
cat("\n")

# Interpolate Meerfelder Maar to Holzmaar ages in the overlapping period
# Only if we have enough Meerfelder Maar data
if(Mer_in_range$n < 2) {
  stop("Not enough Meerfelder Maar data points in overlapping period.\n",
       "Need at least 2 points for interpolation, have ", Mer_in_range$n)
}

overlap_comparison <- Holz_overlap %>%
  mutate(
    Pinus_Mer = approx(Mer_combined$age, Mer_combined$Pinus_Mer, xout = age, rule = 1)$y,
    Betula_Mer = approx(Mer_combined$age, Mer_combined$Betula_Mer, xout = age, rule = 1)$y,
    Herbs_Mer = approx(Mer_combined$age, Mer_combined$Herbs_Mer, xout = age, rule = 1)$y,
    TreesShrubs_Mer = approx(Mer_combined$age, Mer_combined$TreesShrubs_Mer, xout = age, rule = 1)$y
  )

# Count how many complete cases we have
complete_cases <- overlap_comparison %>%
  filter(!is.na(Pinus_Mer) & !is.na(Betula_Mer) & !is.na(TreesShrubs_Mer))

cat("Matched data points with complete data:", nrow(complete_cases), "\n\n")

if(nrow(complete_cases) < 10) {
  cat("WARNING: Only", nrow(complete_cases), "complete data points available.\n")
  cat("This may not be enough for reliable scaling factors.\n\n")
}

head(overlap_comparison, 20)
```

## Compare Trends and Calculate Scaling Factors

```{r compare_trends}
# Calculate summary statistics for both sites
cat("=== Summary Statistics in Overlapping Period ===\n\n")

cat("BETULA:\n")
cat("  Holzmaar - Mean:", round(mean(overlap_comparison$Betula, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$Betula, na.rm = TRUE), 2), "\n")
cat("  Meerfelder Maar - Mean:", round(mean(overlap_comparison$Betula_Mer, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$Betula_Mer, na.rm = TRUE), 2), "\n")
cat("  Correlation:", round(cor(overlap_comparison$Betula, overlap_comparison$Betula_Mer, use = "complete.obs"), 3), "\n\n")

cat("PINUS:\n")
cat("  Holzmaar - Mean:", round(mean(overlap_comparison$Pinus, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$Pinus, na.rm = TRUE), 2), "\n")
cat("  Meerfelder Maar - Mean:", round(mean(overlap_comparison$Pinus_Mer, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$Pinus_Mer, na.rm = TRUE), 2), "\n")
cat("  Correlation:", round(cor(overlap_comparison$Pinus, overlap_comparison$Pinus_Mer, use = "complete.obs"), 3), "\n\n")

cat("AP (Arboreal Pollen):\n")
cat("  Holzmaar - Mean:", round(mean(overlap_comparison$AP, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$AP, na.rm = TRUE), 2), "\n")
cat("  Meerfelder Maar - Mean:", round(mean(overlap_comparison$TreesShrubs_Mer, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$TreesShrubs_Mer, na.rm = TRUE), 2), "\n")
cat("  Correlation:", round(cor(overlap_comparison$AP, overlap_comparison$TreesShrubs_Mer, use = "complete.obs"), 3), "\n\n")

cat("NAP (Non-Arboreal Pollen):\n")
cat("  Holzmaar - Mean:", round(mean(overlap_comparison$NAP, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$NAP, na.rm = TRUE), 2), "\n")
cat("  Meerfelder Maar - Mean:", round(mean(overlap_comparison$Herbs_Mer, na.rm = TRUE), 2), 
    "%, SD:", round(sd(overlap_comparison$Herbs_Mer, na.rm = TRUE), 2), "\n")
cat("  Correlation:", round(cor(overlap_comparison$NAP, overlap_comparison$Herbs_Mer, use = "complete.obs"), 3), "\n\n")

# Calculate scaling factors using linear regression
# Model: Holzmaar = slope * Meerfelder + intercept

# Remove NAs for modeling
model_data <- overlap_comparison %>%
  select(Betula, Betula_Mer, Pinus, Pinus_Mer, AP, TreesShrubs_Mer) %>%
  filter(complete.cases(.))

cat("Data points available for modeling (after removing NAs):", nrow(model_data), "\n\n")

if(nrow(model_data) < 5) {
  cat("\n!!! WARNING !!!\n")
  cat("Very few data points available (", nrow(model_data), ").\n")
  cat("Scaling factors may not be reliable.\n")
  cat("Consider using a simpler approach or checking data quality.\n\n")
  
  # If we have some data, use simple mean ratios instead of regression
  if(nrow(model_data) > 0) {
    cat("Using mean ratio approach instead of linear regression:\n")
    scaling_factors <- list(
      betula = list(
        slope = mean(model_data$Betula / model_data$Betula_Mer, na.rm = TRUE),
        intercept = 0
      ),
      pinus = list(
        slope = mean(model_data$Pinus / model_data$Pinus_Mer, na.rm = TRUE),
        intercept = 0
      ),
      ap = list(
        slope = mean(model_data$AP / model_data$TreesShrubs_Mer, na.rm = TRUE),
        intercept = 0
      )
    )
    
    cat("Betula ratio:", round(scaling_factors$betula$slope, 3), "\n")
    cat("Pinus ratio:", round(scaling_factors$pinus$slope, 3), "\n")
    cat("AP ratio:", round(scaling_factors$ap$slope, 3), "\n\n")
    
    # Create dummy models for compatibility
    lm_betula <- list(coefficients = c(0, scaling_factors$betula$slope))
    lm_pinus <- list(coefficients = c(0, scaling_factors$pinus$slope))
    lm_ap <- list(coefficients = c(0, scaling_factors$ap$slope))
    
  } else {
    stop("No overlapping data available for scaling!")
  }
} else {
  # Sufficient data - use linear regression
  lm_betula <- lm(Betula ~ Betula_Mer, data = model_data)
  lm_pinus <- lm(Pinus ~ Pinus_Mer, data = model_data)
  lm_ap <- lm(AP ~ TreesShrubs_Mer, data = model_data)

  cat("=== Linear Regression Scaling Factors ===\n\n")
  cat("BETULA: Holzmaar =", round(coef(lm_betula)[2], 3), "* Meerfelder +", 
      round(coef(lm_betula)[1], 3), "\n")
  cat("  R² =", round(summary(lm_betula)$r.squared, 3), "\n\n")

  cat("PINUS: Holzmaar =", round(coef(lm_pinus)[2], 3), "* Meerfelder +", 
      round(coef(lm_pinus)[1], 3), "\n")
  cat("  R² =", round(summary(lm_pinus)$r.squared, 3), "\n\n")

  cat("AP: Holzmaar =", round(coef(lm_ap)[2], 3), "* TreesShrubs_Mer +", 
      round(coef(lm_ap)[1], 3), "\n")
  cat("  R² =", round(summary(lm_ap)$r.squared, 3), "\n\n")

  # Store scaling factors
  scaling_factors <- list(
    betula = list(slope = coef(lm_betula)[2], intercept = coef(lm_betula)[1]),
    pinus = list(slope = coef(lm_pinus)[2], intercept = coef(lm_pinus)[1]),
    ap = list(slope = coef(lm_ap)[2], intercept = coef(lm_ap)[1])
  )
}
```

## Visualize Relationships in Overlapping Period

```{r visualize_overlap, fig.width=12, fig.height=10}
par(mfrow = c(2, 3))

# Scatterplots showing relationships
# Betula
plot(overlap_comparison$Betula_Mer, overlap_comparison$Betula,
     xlab = "Meerfelder Maar Betula (%)", ylab = "Holzmaar Betula (%)",
     main = paste("Betula (r =", round(cor(overlap_comparison$Betula, overlap_comparison$Betula_Mer, use = "complete.obs"), 2), ")"),
     pch = 16, col = rgb(0, 0, 1, 0.3))
if(all(is.finite(coef(lm_betula)))) {
  abline(lm_betula, col = "red", lwd = 2)
}
abline(0, 1, col = "gray", lty = 2)
legend("topleft", c("Linear fit", "1:1 line"), col = c("red", "gray"), lty = c(1, 2), lwd = 2, cex = 0.8)

# Pinus
plot(overlap_comparison$Pinus_Mer, overlap_comparison$Pinus,
     xlab = "Meerfelder Maar Pinus (%)", ylab = "Holzmaar Pinus (%)",
     main = paste("Pinus (r =", round(cor(overlap_comparison$Pinus, overlap_comparison$Pinus_Mer, use = "complete.obs"), 2), ")"),
     pch = 16, col = rgb(0, 0, 1, 0.3))
if(all(is.finite(coef(lm_pinus)))) {
  abline(lm_pinus, col = "red", lwd = 2)
}
abline(0, 1, col = "gray", lty = 2)
legend("topleft", c("Linear fit", "1:1 line"), col = c("red", "gray"), lty = c(1, 2), lwd = 2, cex = 0.8)

# AP
plot(overlap_comparison$TreesShrubs_Mer, overlap_comparison$AP,
     xlab = "Meerfelder Maar Trees/Shrubs (%)", ylab = "Holzmaar AP (%)",
     main = paste("AP (r =", round(cor(overlap_comparison$AP, overlap_comparison$TreesShrubs_Mer, use = "complete.obs"), 2), ")"),
     pch = 16, col = rgb(0, 0, 1, 0.3))
if(all(is.finite(coef(lm_ap)))) {
  abline(lm_ap, col = "red", lwd = 2)
}
abline(0, 1, col = "gray", lty = 2)
legend("topleft", c("Linear fit", "1:1 line"), col = c("red", "gray"), lty = c(1, 2), lwd = 2, cex = 0.8)

# Time series showing trends
plot(overlap_comparison$age, overlap_comparison$Betula, type = "l", col = "blue", lwd = 2,
     xlab = "Age (cal yr BP)", ylab = "Betula (%)", main = "Betula Trends Over Time")
lines(overlap_comparison$age, overlap_comparison$Betula_Mer, col = "orange", lwd = 2)
legend("topright", c("Holzmaar", "Meerfelder Maar"), col = c("blue", "orange"), lwd = 2)

plot(overlap_comparison$age, overlap_comparison$Pinus, type = "l", col = "blue", lwd = 2,
     xlab = "Age (cal yr BP)", ylab = "Pinus (%)", main = "Pinus Trends Over Time")
lines(overlap_comparison$age, overlap_comparison$Pinus_Mer, col = "orange", lwd = 2)
legend("topright", c("Holzmaar", "Meerfelder Maar"), col = c("blue", "orange"), lwd = 2)

plot(overlap_comparison$age, overlap_comparison$AP, type = "l", col = "blue", lwd = 2,
     xlab = "Age (cal yr BP)", ylab = "AP (%)", main = "AP Trends Over Time")
lines(overlap_comparison$age, overlap_comparison$TreesShrubs_Mer, col = "orange", lwd = 2)
legend("topright", c("Holzmaar", "Meerfelder Maar"), col = c("blue", "orange"), lwd = 2)
```

## Check Meerfelder Maar Data in Target Period (13258-10995 BP)

```{r apply_scaling}
# Create prediction ages at ~30 year intervals
prediction_ages <- seq(from = 13258, to = 10995, by = -30)

cat("Creating predictions for", length(prediction_ages), "time points\n")
cat("Age range:", range(prediction_ages), "BP\n\n")

# Interpolate Meerfelder Maar data to prediction ages
prediction_data <- data.frame(age = prediction_ages) %>%
  mutate(
    Pinus_Mer = approx(Mer_combined$age, Mer_combined$Pinus_Mer, xout = age, rule = 2)$y,
    Betula_Mer = approx(Mer_combined$age, Mer_combined$Betula_Mer, xout = age, rule = 2)$y,
    TreesShrubs_Mer = approx(Mer_combined$age, Mer_combined$TreesShrubs_Mer, xout = age, rule = 2)$y
  )

# Apply scaling factors to predict Holzmaar values
prediction_data <- prediction_data %>%
  mutate(
    Betula = scaling_factors$betula$slope * Betula_Mer + scaling_factors$betula$intercept,
    Pinus = scaling_factors$pinus$slope * Pinus_Mer + scaling_factors$pinus$intercept,
    AP = scaling_factors$ap$slope * TreesShrubs_Mer + scaling_factors$ap$intercept
  )

# Constrain to valid percentage range [0, 100]
prediction_data <- prediction_data %>%
  mutate(
    Betula = pmax(0, pmin(100, Betula)),
    Pinus = pmax(0, pmin(100, Pinus)),
    AP = pmax(0, pmin(100, AP)),
    NAP = 100 - AP  # Maintain NAP = 100 - AP relationship
  )

cat("Predictions generated:\n")
cat("Betula range:", round(range(prediction_data$Betula), 2), "%\n")
cat("Pinus range:", round(range(prediction_data$Pinus), 2), "%\n")
cat("AP range:", round(range(prediction_data$AP), 2), "%\n")
cat("NAP range:", round(range(prediction_data$NAP), 2), "%\n\n")

cat("Mean values in predicted period:\n")
cat("Betula:", round(mean(prediction_data$Betula), 2), "%\n")
cat("Pinus:", round(mean(prediction_data$Pinus), 2), "%\n")
cat("AP:", round(mean(prediction_data$AP), 2), "%\n")
cat("NAP:", round(mean(prediction_data$NAP), 2), "%\n")
```

## Create Final Predicted Dataset

```{r create_predicted_dataset}
# Create final predicted dataset matching Holzmaar format
Holz_predicted <- data.frame(
  age = prediction_data$age,
  depth = NA,  # Depth would need to be interpolated separately
  Betula = prediction_data$Betula,
  Juniperus = NA,  # Not modeled
  Picea = NA,      # Not modeled
  Pinus = prediction_data$Pinus,
  Salix = NA,      # Not modeled
  NAP = prediction_data$NAP,
  AP = prediction_data$AP
)

cat("Final predicted dataset:\n")
cat("Data points:", nrow(Holz_predicted), "\n")
cat("Age range:", range(Holz_predicted$age), "BP\n")
cat("Time resolution: ~30 years\n\n")

head(Holz_predicted, 10)
tail(Holz_predicted, 10)
```

## Combine with Original Data

```{r combine_data}
# Original Holzmaar data covers 0-13258 BP
# We're adding predictions for 13258-10995 BP
Holz_existing <- HolzPol %>%
  filter(age <= 13258)

# For the combined dataset, we need to match the structure
# Add the missing columns to predicted data (as NA)
Holz_predicted_full <- Holz_predicted %>%
  mutate(
    Juniperus = NA,
    Picea = NA,
    Salix = NA
  ) %>%
  select(names(HolzPol))

# Combine datasets: original data (0-13258) + predicted data (13258-10995)
Holz_extended <- bind_rows(
  Holz_existing,
  Holz_predicted_full
) %>%
  arrange(age)

cat("Original Holzmaar dataset size:", nrow(Holz_existing), "\n")
cat("Predicted dataset size:", nrow(Holz_predicted_full), "\n")
cat("Extended dataset size:", nrow(Holz_extended), "\n")
cat("Extended age range:", range(Holz_extended$age), "\n\n")

# Show the transition area around 13258 BP
cat("Data around the transition (13000-11000 BP):\n")
Holz_extended %>%
  filter(age >= 11000 & age <= 13300) %>%
  select(age, Betula, Pinus, NAP, AP) %>%
  arrange(desc(age)) %>%
  print(n = 30)
```

## Visualize Results

```{r visualize_results, fig.width=10, fig.height=8}
# Create long format for plotting - focus on the modeled period
plot_data <- Holz_extended %>%
  filter(age >= 10500 & age <= 14000) %>%
  mutate(source = ifelse(age <= 13258, "Original", "Predicted")) %>%
  select(age, Betula, Pinus, NAP, AP, source) %>%
  pivot_longer(cols = c(Betula, Pinus, NAP, AP), 
               names_to = "Variable", 
               values_to = "Percentage")

# Plot
ggplot(plot_data, aes(x = age, y = Percentage, color = source)) +
  geom_line(size = 1) +
  geom_point(alpha = 0.5, size = 2) +
  facet_wrap(~Variable, scales = "free_y", ncol = 1) +
  geom_vline(xintercept = 13258, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 13258, y = Inf, label = "Prediction starts (13258 BP)", 
           vjust = 1.5, hjust = 1.1, color = "red", size = 3.5) +
  annotate("rect", xmin = 10995, xmax = 13258, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "orange") +
  scale_color_manual(values = c("Original" = "blue", "Predicted" = "orange")) +
  labs(title = "Holzmaar Pollen Data: Original vs Predicted",
       subtitle = "Orange shaded area (13258-10995 BP): Predicted using Meerfelder Maar data at ~30 year intervals",
       x = "Age (cal yr BP)",
       y = "Percentage",
       color = "Data Source") +
  theme_bw() +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold")) +
  scale_x_reverse()  # Reverse x-axis so older dates are on the left
```

## Save Results

```{r save_results}
# Create the final dataset from 10995 to 15030 BP (full range)
Holz_final <- Holz_extended %>%
  filter(age >= 10995) %>%
  arrange(age)

# Save the complete dataset with predictions
write.csv(Holz_final, "data/HolzPol_10995-15030_complete.csv", row.names = FALSE)

# Also save just the extended part (with predictions)
write.csv(Holz_extended, "data/HolzPol_extended_with_predictions.csv", row.names = FALSE)

# Save just the predicted portion (13258-10995 BP)
write.csv(Holz_predicted, "data/HolzPol_predicted_13258-10995.csv", row.names = FALSE)

cat("Results saved to:\n")
cat("- data/HolzPol_10995-15030_complete.csv (complete dataset:", 
    nrow(Holz_final), "rows from", min(Holz_final$age), "to", max(Holz_final$age), "BP)\n")
cat("- data/HolzPol_extended_with_predictions.csv (full extended dataset from 0 BP)\n")
cat("- data/HolzPol_predicted_13258-10995.csv (predicted portion only,", 
    nrow(Holz_predicted), "rows at ~30 year intervals)\n\n")

# Summary of the predicted portion
cat("=== Predicted Data Summary (13258-10995 BP) ===\n")
cat("Number of predicted points:", nrow(Holz_predicted), "\n")
cat("Age range:", range(Holz_predicted$age), "\n")
cat("Age interval: ~30 years\n")
cat("Mean Betula:", round(mean(Holz_predicted$Betula, na.rm = TRUE), 2), "%\n")
cat("Mean Pinus:", round(mean(Holz_predicted$Pinus, na.rm = TRUE), 2), "%\n")
cat("Mean NAP:", round(mean(Holz_predicted$NAP, na.rm = TRUE), 2), "%\n")
cat("Mean AP:", round(mean(Holz_predicted$AP, na.rm = TRUE), 2), "%\n")
cat("\nConstraint verification:\n")
cat("All Betula values in [0,100]:", all(Holz_predicted$Betula >= 0 & Holz_predicted$Betula <= 100), "\n")
cat("All Pinus values in [0,100]:", all(Holz_predicted$Pinus >= 0 & Holz_predicted$Pinus <= 100), "\n")
cat("All AP values in [0,100]:", all(Holz_predicted$AP >= 0 & Holz_predicted$AP <= 100), "\n")
cat("All NAP values in [0,100]:", all(Holz_predicted$NAP >= 0 & Holz_predicted$NAP <= 100), "\n")
cat("AP + NAP = 100 for all rows:", all(abs((Holz_predicted$AP + Holz_predicted$NAP) - 100) < 0.001), "\n")
```

## Model Diagnostics

```{r diagnostics, fig.width=10, fig.height=10}
par(mfrow = c(2, 2))
gam.check(model_betula)
gam.check(model_pinus)
gam.check(model_nap)
gam.check(model_ap)
```
