---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
setwd("~/Library/CloudStorage/ProtonDrive-zahajska@pm.me-folder/Germany-ELSA/Eifel/data_in_process/HZM_stats_2024")
```

# packages

```{r}
library("rioja")
library("ggplot2")
library("cluster")
library("readxl")
library("dplyr")
library("tidyr")
library("openxlsx")
library("viridis")
library("rcartocolor")
library("cowplot")
library("latex2exp")
library("vegan")
library(corrplot)
library("PerformanceAnalytics")
library("reshape2")
library(janitor)
```

# Diatoms

```{r}
#depth_D <- select(LowRes_all, 1)
#LowRes_dia_D <- cbind(depth_dia_cl, LowRes_dia)
```

```{r}
#write.xlsx(LowRes_dia_D, "data/LR_dia_depth.xlsx", rowNames = FALSE, col_types = "numeric")
```

```{r}
LowRes_dia_D <- read_excel("data/LR_dia_depth.xlsx")
```

```{r}
LowRes_dia_NaN <- as.data.frame(na.omit(LowRes_dia_D))
#depth_dia_cl <- select(LowRes_dia_NaN, 1)
LowRes_dia_cl <- select(LowRes_dia_NaN, 2:92)
depth_diatom <- LowRes_dia_NaN[,1]
```

```{r}
library(janitor)

perc <- adorn_percentages(LowRes_dia_cl, denominator = "row", na.rm = TRUE, 1:91)
perc_100 <- perc*100

```

```{r}
# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_100) # Calculate column sums
df_cl <- perc_100[, which(ma.sum > 3)] # Subset data
```

```{r}

#png(filename = "fig/dia_3per_corr_values.png", width = 1754, height = 1754, units = "px", pointsize = 10, res = 200)
chart.Correlation(df_cl, histogram=TRUE, pch=19)
#dev.off()
```

```{r}

png(filename = "fig/dia_3perc_corr.png", width = 1754, height = 1754, units = "px", pointsize = 12)
#visualize correlation matrix
corrplot(cor(df_cl))
dev.off()
```

```{r}
#png(filename = "fig/quick.png", width = 1754, height = 1240, units = "px", pointsize = 12)
strat.plot(df_cl)
#dev.off()
```

```{r}
# Define colour scheme
p.col <- c(rep("forestgreen", times=4), rep("gold2", times=18), rep("blue", times=39))
# Define y axis 
y.scale <- depth_diatom
#y.override <- row.names(depth_dia_cl)
# Plot bar plot
pol.plot <- strat.plot(df_cl, yvar=y.scale, y.tks=y.scale, ylim= c(10989, 13369), y.rev=TRUE, plot.line=FALSE, plot.poly=FALSE, plot.bar=TRUE, col.bar=p.col, lwd.bar=10, scale.percent=TRUE, xSpace=0.01, x.pc.lab=TRUE, x.pc.omit0=TRUE, las=2)
```

```{r}
pol.plot <- strat.plot(df_cl, yvar=y.scale, y.tks=y.scale, y.rev=TRUE, ylim= c(10989, 13369), plot.line=FALSE, plot.poly=TRUE, plot.bar=FALSE, col.poly=p.col, col.poly.line="black", lwd.poly=1, scale.percent=TRUE, xSpace=0.01, x.pc.lab=TRUE, x.pc.omit0=TRUE, las=2)
```

```{r}
# Cluster analysis

ma.dist <- vegdist(df_cl, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")
```

```{r}
# Filled plot with cluster analysis
png(filename = "fig/dia_3perc_limits.png", width = 1754, height = 1240, units = "px", pointsize = 12)
pol.plot <- strat.plot(df_cl, yvar=y.scale, y.tks=y.scale,  ylim= c(10989, 13369), y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar diatoms LG Samples", ylabel="Sample", clust=ma.chclust)

# Add cluster zones
addClustZone(pol.plot, ma.chclust, nZone=6, lwd=1.5, lty=2, col="grey25")
dev.off()
```

```{r}
perc <- adorn_percentages(LowRes_dia_cl, denominator = "row", na.rm = TRUE, 1:91)
perc_100 <- perc*100

```

```{r}
# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_100) # Calculate column sums
df_cl_perc <- perc_100[, which(ma.sum > 10)] # Subset data
```

```{r}
# Cluster analysis
ma.dist <- vegdist(df_cl_perc, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")
```

```{r}
# Filled plot with cluster analysis
png(filename = "fig/dia_10perc_all.png", width = 1754, height = 1240, units = "px", pointsize = 12)
pol.plot <- strat.plot(df_cl_perc, yvar=y.scale, y.tks=y.scale, y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar diatoms LG Samples", ylabel="Depth", clust=ma.chclust)

# Add cluster zones
addClustZone(pol.plot, ma.chclust, nZone=6, lwd=1.5, lty=2, col="grey25")
dev.off()
```

```{r}
df_cl_perc_dia<- cbind(depth_diatom, df_cl_perc)
write.xlsx(df_cl_perc_dia, "data/LR_dia_10perc.xlsx", rowNames = FALSE, col_types = "numeric")
```

# Pigments

```{r}
#LowRes_pig <- read_excel("data/LR_clustered_pig.xlsx")
LowRes_all <- read_excel("LowRes_all.xlsx")
# 
LowRes_pig1 <- LowRes_all[,25:111]
depth <- LowRes_all[,1]
LowRes_pig <- cbind(depth, LowRes_pig1)

LowRes_pig_cl <- na.omit(LowRes_pig)
depth_cl <- LowRes_pig_cl[,1]
LowRes_pig_NaN <- LowRes_pig_cl[,2:88]


```

```{r}
 remove_high_zero_columns <- function(data) {
#   # Calculate the proportion of zeros in each column
   prop_zeros <- colMeans(data == 0)
  # Keep columns with less than or equal to 95% zeros
   data[, prop_zeros <= 0.95]
 }
 # Remove columns with high zero proportion
df_filtered<- as.data.frame(remove_high_zero_columns(LowRes_pig_NaN))
```

```{r}
library(corrplot)
#png(filename = "fig/pig_95perc_nonzero_corr.png", width = 1754, height = 1754, units = "px", pointsize = 12)
#visualize correlation matrix
corrplot(cor(df_filtered))
#dev.off()
```

```{r}
library("PerformanceAnalytics")
#png(filename = "fig/pig_95perc_nonzero_corr_values.png", width = 1754, height = 1754, units = "px", pointsize = 10, res = 200)
chart.Correlation(df_filtered, histogram=TRUE, pch=19)
#dev.off()
```

```{r}
library(janitor)

perc_pig <- adorn_percentages(LowRes_pig_NaN, denominator = "row", na.rm = TRUE, 1:87)
perc_100 <- perc_pig*100

```

```{r}
# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_100) # Calculate column sums
df_cl_perc <- perc_100[, which(ma.sum > 5)] # Subset data
```

```{r}
# Cluster analysis
library("vegan")

data <- as.matrix(df_cl_perc)
# Standardise data for CLUSTER analysis
data_standardized <- data
for(i in 1:ncol(data_standardized) ) {
  data_standardized[ , i] <- data_standardized[,i]/sd(data_standardized[,i])
}

ma.dist <- vegdist(data_standardized, method="euclidean", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")
```

```{r}
# Define colour scheme
p.col <- c(rep("forestgreen", times=4), rep("gold2", times=18), rep("blue", times=39))
# Define y axis 
y.scale <- depth_cl

# Filled plot with cluster analysis
#png(filename = "fig/pig_5perc_std.png", width = 1754, height = 1240, units = "px", pointsize = 12) df_cl_perc
pig.plot <- strat.plot(df_cl_perc, yvar=y.scale, y.tks=y.scale, ylim=(10989, 13369),  y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar pigments LG ", ylabel="Depth", clust=ma.chclust)

# Add cluster zones
addClustZone(pig.plot, ma.chclust, nZone=4, lwd=1.5, lty=2, col="grey25")
#dev.off()
```

# All selected proxies

```{r}
{r}
LowRes_all <- read_excel("data/LR_all.xlsx")

LowRes_pig_d <- select(LowRes_all,1,25:111)
LowRes_redox_d <- select(LowRes_all, 1:13)
LowRes_dia_d <- select(LowRes_all, 1, 117:207)
LowRes_XRF_raw_d <- select(LowRes_all, 1, 211:241)
LowRes_HSI_d <- select(LowRes_all, 1, 14:20)
#depth_pig <- select(LowRes_all, 1, 25:111)
#dep <- select(LowRes_all, 1)
```

```{r}
{r}
LR_pig <- na.omit(LowRes_pig_d)
LR_red <- na.omit(LowRes_redox_d)
LR_dia <- na.omit(LowRes_dia_d)
LR_XRF <- na.omit(LowRes_XRF_raw_d)
LR_HSI <- na.omit(LowRes_HSI_d)


 remove_high_zero_columns <- function(data) {
#   # Calculate the proportion of zeros in each column
   prop_zeros <- colMeans(data == 0)
  # Keep columns with less than or equal to 95% zeros
   data[, prop_zeros <= 0.95]
 }
 # Remove columns with high zero proportion
df_pig<- as.data.frame(remove_high_zero_columns(LR_pig))
df_red<- as.data.frame(remove_high_zero_columns(LR_red))
df_dia<- as.data.frame(remove_high_zero_columns(LR_dia))
df_XRF<- as.data.frame(remove_high_zero_columns(LR_XRF))
df_HSI<- as.data.frame(remove_high_zero_columns(LR_HSI))


```

```{r}
library(janitor)
perc <- adorn_percentages(df_dia, denominator = "row", na.rm = TRUE, 2:69)
perc_100 <- perc*100

# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_100) # Calculate column sums
df_dia_per <- perc_100[, which(ma.sum > 10)] # Subset data

```

```{r}
# Define colour scheme
p.col <- c(rep("forestgreen", times=4), rep("gold2", times=18), rep("blue", times=39))
# Define y axis 
y.scale <- df_pig[,1]

# Filled plot with cluster analysis
#png(filename = "fig/pig_5perc_std.png", width = 1754, height = 1240, units = "px", pointsize = 12) df_cl_perc
pig.plot <- strat.plot(df_pig[2:38], yvar=y.scale, y.tks=y.scale, y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar pigments LG ", ylabel="Depth")#, #clust=ma.chclust)

# Add cluster zones
#addClustZone(pig.plot, ma.chclust, nZone=4, lwd=1.5, lty=2, col="grey25")
#dev.off()
```

```{r}
dia.plot <- strat.plot(df_dia_per[2:25], yvar=df_dia_per[,1], y.tks=df_dia_per[,1], y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar diatoms LG Samples", ylabel="Depth" )#,clust=ma.chclust)

```

```{r}
png(filename = "fig/big_plot.png", width = 1750, height = 1750, units = "px", pointsize = 12)
plot_grid(pig.plot,dia.plot)# nrow = 1, align = "h",  rel_widths = c(0.5, 0.5))#labels = c('A', 'B'), , label_size = 12
#multiplot(pig.plot,dia.plot)#p3, Pseq, Mnseq, Feseq, p4, p5, p6, Chl, Ti, Fe, cols=13)
dev.off()

```

# Combining Diatoms, Pigments, HSI and XRF

```{r}
HR <- read_excel("data/HR_raw_forPCA.xlsx")
depth_HR <- (HR$Depth_mm_HSI_composite + 10989)/1000
age_HR <- HR$mean_VT22
pig <- read_excel("data/pig_clustered_PCA.xlsx")
depth_pig <- pig$comp_depth_mm/1000
dia <- read_excel("data/LR_dia_10perc.xlsx")
depth_dia <- dia$depth_diatom/1000
redox <- read_excel("data/LR_redox_PCA_clustered.xlsx")
depth_redox <- redox$comp_depth_mm/1000
pollen <- read_excel("data/Meerfeldemar_pollen.xlsx")
age_pol <- pollen$`Age varve BC`
LR_all <- read_excel("LowRes_all.xlsx")
depth_LRall <- (LR_all$comp_depth_mm)/1000
LR_all_bulk <- read_excel("data/LR_all_bulk.xlsx")
depth_LRall_bulk <- (LR_all_bulk$comp_depth_mm)/1000
Pin <- read_excel("data/Meerfeldemar_pollen.xlsx", sheet = "Pinus")
age_pin <- Pin$`Age varve yr BP`
Bet <- read_excel("data/Meerfeldemar_pollen.xlsx", sheet = "Betula")
age_bet <- Bet$`Age varve yr BP`
```

```{r}
pig_all <- as.data.frame(select(LR_all, 1,3, 31:120))
pig_all_bulk <- as.data.frame(select(LR_all_bulk, 1:2, 26:115))
dia_all <- as.data.frame(select(LR_all, 1,3, 123:214))
redox_all <- as.data.frame(select(LR_all, 1,3, 8:19))
age_all <- LR_all$mean_VT22
depth_all <- LR_all$depth_VT22/1000
```

## High_res

```{r}
Chl <- ggplot(HR, aes(x=RABD671cal, y= age_HR))+
  geom_point(colour = "darkgreen", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) + # 13.822, 10.898, seq(10.898, 13.822, by = 0.1)
 geom_path(alpha=.8, size=0.2, colour = "darkgreen") +
  labs(title = " ", x = "Chloropigments \n(ug/g)", y = " ") +
    scale_x_continuous(position = 'top') +
  theme_classic() +
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Cya <- ggplot(HR, aes(x=RABD620, y= age_HR))+
  geom_point(colour = "darkblue", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "darkblue") +
  labs(title = " ", x = "RABD620 \nCyanobacteria", y = " ") +
  theme_classic() +
theme(legend.position="none") #axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

RABD715norm= HR$RABD715/HR$Rmean

GSB <- ggplot(HR, aes(x=RABD715norm, y= age_HR))+
  geom_point(colour = "limegreen", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "limegreen") +
  labs(title = " ", x = "RABD715 \nGreen Suplhur Bacteria", y = " ") +
  theme_classic() +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

PSB <- ggplot(HR, aes(x=RABD845cal, y= age_HR))+
  geom_point(colour = "purple", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "purple") +
  labs(title = " ", x = "Bacteriopheophytine-a \n(ug/g)", y = " ") +
  theme_classic() +
    scale_x_continuous(position = 'top') +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Ti <- ggplot(HR, aes(x=Ti, y= age_HR))+
  geom_point(colour = "black", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "black") +
  labs(title = " ", x = "Ti \n(%)", y = " ") +
  theme_classic() +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Si <- ggplot(HR, aes(x=Si, y= age_HR))+
  geom_point(colour = "blue", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "blue") +
  labs(title = " ", x = "Si \n(%)", y = " ") +
  theme_classic() +
    scale_x_continuous(position = 'top') +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

HR$Si_Ti <- HR$Si / HR$Ti

SiTi <- ggplot(HR, aes(x=Si_Ti, y= age_HR))+
  geom_point(colour = "lightblue", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "lightblue") +
  labs(title = " ", x = "Si/Ti \n (ratio)", y = " ") +
  theme_classic() +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

S <- ggplot(HR, aes(x=S, y= age_HR))+
  geom_point(colour = "darkrod", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "darkrod") +
  labs(title = " ", x = "S \n(%)", y = " ") +
  theme_classic() +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

## Low_res

```{r}

TOC <- ggplot(LR_all, aes(x=LR_all$`TOC [%]`, y= age_all))+
  geom_point(colour = "black", size= 0.01)+
  scale_y_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 geom_path(alpha=.8, size=0.2, colour = "black") +
  labs(title = " ", x = "TOC \n(%)", y = " ") +
  theme_classic() +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

### Diatoms to percentages

```{r}
dia_na <- na.omit(dia_all[3:93])
valves <- select(na.omit(dia_all), 94)
d_a <- select(na.omit(dia_all), 1:2)
```

```{r}
library(janitor)
perc_all <- adorn_percentages(dia_na, denominator = "row", na.rm = TRUE, 1:91) #na.rm = TRUE,
perc_all_100 <- perc_all*100

# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_all_100) # Calculate column sums
dia_all_per <- perc_all_100[, which(ma.sum > 10)] # Subset data


dia_all_per_d <- cbind(d_a, dia_all_per, valves)
```

```{r}
valve <- ggplot(dia_all, aes(y=dia_all$`Valves/gr`, x=dia_all$mean_VT22))+
  #geom_point(colour = "blue", size= 0.01)+
  geom_bar(stat = "identity", width = 25, fill = "blue")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=.8, size=0.2, colour = "blue") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "diatom \n valves/g", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Amph <- ggplot(dia_all_per_d , aes(y= `Amphora indistinta`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Amphora \n indistinta", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Astr <- ggplot(dia_all_per_d , aes(y=`Asterionella formosa`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
    theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Asterionella \nformosa", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Epit <- ggplot(dia_all_per_d , aes(y=`Epithemia sorex`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
    theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Epithemia \nsorex", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Pant <- ggplot(dia_all_per_d , aes(y=`Pantosekiella ocellata`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
    labs(title = " ", y = "Pantosekiella \nocellata", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Pseu <- ggplot(dia_all_per_d , aes(y=`Pseudostaurosira brevistriata`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
    theme_classic() +
      scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y= "Pseudostaurosira \nbrevistriata", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

PseuR <- ggplot(dia_all_per_d , aes(y=`Pseudostaurosira robusta`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
      scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
   theme_classic() +
  labs(title = " ", y = "Pseudostaurosira \nrobusta", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Stau <- ggplot(dia_all_per_d , aes(y=`Staurosira construens`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
    scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
    theme_classic() +
  labs(title = " ", y = "Staurosira \nconstruens", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

StauS <-ggplot(dia_all_per_d , aes(y=`Staurosirella lapponica`, x= mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
    theme_classic() +
      scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Staurosirella \nlapponica", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


#here are total count" change to% dia  - add the time scale
Steph <-ggplot(dia_all_per_d , aes(y=`Stephanodiscus minutulus`, x= mean_VT22))+
  geom_bar(stat = "identity", width = 25, fill = "green")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
    scale_fill_viridis_d(option = "H")+
  coord_flip()+ 
    labs(title = " ", y = "Stephanodiscus \nminutulus", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


# Pseq_per <- ggplot(P_stacked_per, aes(y = value, x = depth, fill = factor(variable))) +  # Group by data values
#   geom_bar(stat = "identity") +  # Preserve data values
#   scale_x_reverse(limits = c(13.822, 10.898), breaks = round(seq(10.898, 13.822, by = 0.1), digits = 2)) +
#   labs(title = "",  # Adjust title (optional)
#        x = "Depth",  # Adjust x-axis label
#        y = "Data") +  # Adjust y-axis label
#   theme_classic() + # Set theme (optional)
#   scale_fill_viridis_d(option = "H")+
# coord_flip()+ 
#   labs(title = "", y = "P Fraction %", x = "") + 
#   theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

### Pigments

```{r}
Fuco <- ggplot(LR_all , aes(y= LR_all$Fuco, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "orange")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Fucoxantine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Alpha <- ggplot(LR_all , aes(y= LR_all$Alfa_car, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G1: Alphacarotene (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Echi <- ggplot(LR_all , aes(y= LR_all$Echi1, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G1: Echinenone (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Diato <- ggplot(LR_all , aes(y= LR_all$Diadi, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G2: Diadionxanthine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Canta <- ggplot(LR_all , aes(y= LR_all$Canta, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G2: Canthaxantine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Zea <- ggplot(LR_all , aes(y= LR_all$Zeax, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: Zeaxanthine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Lut <- ggplot(LR_all , aes(y= LR_all$Lute, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  #  scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: Lutein (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

OHSp <- ggplot(LR_all , aes(y= LR_all$OH_Sph.ene, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: OH-Spheroidene (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Iso <- ggplot(LR_all , aes(y= LR_all$Isore, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G4: Isorenoretine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Chlx <- ggplot(LR_all , aes(y= LR_all$Chloxant, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G4: Chloraxantine (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Oke <- ggplot(LR_all , aes(y= LR_all$Okenone, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "purple")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G5: Okenone (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


Phph<- ggplot(LR_all , aes(y= LR_all$PhphT, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophytines (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Phide<- ggplot(LR_all , aes(y= LR_all$PhideT, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophorbides (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

```

### Pigments bulk

```{r}
FucoB <- ggplot(data= pig_all_bulk , aes(y= pig_all_bulk$Fuco, x= pig_all_bulk$mean_VT22),)+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar( stat = "identity", width = 25, fill = "orange")+
 #stat = "identity", width = 25,
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  scale_y_continuous(position = 'right')+ 
  labs(title = " ", y = "Fucoxantine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
  # Fuco <- ggplot(LR_all , aes(y= LR_all$Fuco, x= LR_all$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
#   geom_bar(data= LR_all , aes(y= LR_all$Fuco, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = "orange3")+
#   scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
#  #geom_path(alpha=1, size=1, colour = "green") +
#   theme_classic() +
#         scale_fill_viridis_d(option = "H")+
#    coord_flip()+ 
#   labs(title = " ", y = "Fucoxantine (nmol/g (OM))", x = " ") +
# theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


AlphaB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Alfa_car, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G1: Alphacarotene (nmol/g (wet sed.))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

EchiB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Echi1, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "goldenrod")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G1: Echinenone (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

DiatoB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Diadi, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G2: Diadionxanthine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

CantaB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Canta, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "orangered")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G2: Canthaxantine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

ZeaB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Zeax, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: Zeaxanthine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


LutB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Lute, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  #  scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: Lutein (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

OHSpB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$OH_Sph.ene, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgoldenrod3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G3: OH-Spheroidene (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

IsoB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Isore, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G4: Isorenoretine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

ChlxB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Chloxant, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "lightgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G4: Chloraxantine (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

OkeB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$Okenone, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "purple")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "G5: Okenone (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())


PhphB <- ggplot(pig_all_bulk , aes(y= pig_all_bulk$PhphT, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophytines (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

PhideB<- ggplot(pig_all_bulk , aes(y= pig_all_bulk$PhideT, x= pig_all_bulk$mean_VT22))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "darkgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_viridis_d(option = "H")+
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophorbides (nmol/g (wet sed))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

### Pollen

#### Meerfeldemaar

```{r}
library(reshape2)

pollen_stacked_per <- melt(pollen, var.names = names(pollen)[2:ncol(pollen)], measure.vars = names(pollen)[2:ncol(pollen)])


poll<- ggplot(pollen_stacked_per, aes(y= value, x=pollen_stacked_per$`Age varve BC`, fill=factor(variable)))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25)+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_manual(values = c("goldenrod", "springgreen2"))+
   coord_flip()+ 
  labs(title = " ", y = "AP and NAP (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Pinus <- ggplot(Pin , aes(y= `Pinus%`, x= age_pin))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "forestgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Pinus (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

Betula <- ggplot(Bet , aes(y= `Betula%`, x= age_bet))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "burlywood3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Betula (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

```

#### Holzmaar 1992

```{r}
Pollen_HZM<- read_excel("data/HZM92-A_B_pollen_edit.xlsx")

pol <- as.data.frame(select(Pollen_HZM, 8,9,11:84))
pol_cl <- as.data.frame(select(Pollen_HZM, 11:66))
pol_age <- as.data.frame(select(Pollen_HZM, 9,11:84))
pol_d <- as.data.frame(select(Pollen_HZM,8, 11:84))
N_AP <- as.data.frame(select(Pollen_HZM, 9,87,86))
```

```{r}
library(reshape2)
# 
pollen_stacked_per2 <- melt(N_AP, var.names = names(N_AP)[2:ncol(N_AP)], measure.vars = names(N_AP)[2:ncol(N_AP)])
# 
# 
pollHZM<- ggplot(pollen_stacked_per2, aes(y= value, x=VT_age, fill=factor(variable)))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25)+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_manual(values = c("goldenrod", "springgreen2"))+
   coord_flip()+
  labs(title = " ", y = "AP and NAP (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

PinusHZM <- ggplot(pol_age , aes(y= Pinus, x= VT_age))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "forestgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Pinus (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

BetulaHZM <- ggplot(pol_age , aes(y= Betula, x= VT_age))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "burlywood3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Betula (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())
```

```{r}
pollHZM
```

```{r}
# Define colour scheme
p.col <- c(rep("forestgreen", times=4), rep("gold2", times=18), rep("blue", times=39))
# Define y axis 
y.scale <- pol$VT_age

# Cluster analysis

ma.dist <- vegdist(pol_cl, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")

polHZM.plot <- strat.plot(pol_cl, yvar=y.scale, y.tks=y.scale,  ylim= c(11750, 15342), y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar diatoms LG Samples", ylabel="Sample", clust=ma.chclust)

# Add cluster zones
addClustZone(polHZM.plot, ma.chclust, nZone=6, lwd=1.5, lty=2, col="grey25")

```

#### Pollen Litt-Leroy stacked

```{r}
Pollen_stacked<- read_excel("data/pollen_stacked.xlsx")

pol_sta <- as.data.frame(select(Pollen_stacked, 1:119))
pol_sta_cl <- as.data.frame(select(Pollen_stacked, 3:59, 61:115))
pol_sta_age <- as.data.frame(select(Pollen_stacked, 1 ))
# pol_d <- as.data.frame(select(Pollen_HZM,7, 10:83))
N_AP_sta <- as.data.frame(select(Pollen_stacked, 1,117,118))
```

```{r}
library(reshape2)
# 
pollen_stacked_perSTA <- melt(N_AP_sta, var.names = names(N_AP_sta)[2:ncol(N_AP_sta)], measure.vars = names(N_AP_sta)[2:ncol(N_AP_sta)])
# 
# 
pollHZM_sta<- ggplot(pollen_stacked_perSTA, aes(y= value, x=Age_merged, fill=factor(variable)))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25)+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
        scale_fill_manual(values = c("goldenrod", "springgreen2"))+
   coord_flip()+
  labs(title = " ", y = "AP and NAP (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

PinusHZM_sta <- ggplot(pol_sta , aes(y= Pinus, x= Age_merged))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "forestgreen")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    #scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Pinus (%)", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank())

BetulaHZM_sta <- ggplot(pol_sta , aes(y= Betula, x= Age_merged))+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(stat = "identity", width = 25, fill = "burlywood3")+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2), guide = guide_axis(position = "right")) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+
  labs(title = " ", y = "Betula (%)", x = " ") +
theme(legend.position="none")#, axis.ticks.x = "" #axis.ticks.y= element_blank(), axis.line.y = element_blank(), #axis.text.y = element_blank())
```

```{r}
pollHZM_sta
```

```{r}
BetulaHZM_sta
```

### Sequential extraction

```{r}
P_tot_pre <- as.data.frame(select(LR_all, 8:11))
#P_tot <- bind_cols(depth= depth_LR, P_tot_pre)
P_tot <- bind_cols(age= LR_all$mean_VT22, P_tot_pre)

P_stacked <- melt(P_tot, var.names = names(P_tot)[2:ncol(P_tot)], measure.vars = names(P_tot)[2:ncol(P_tot)])

Fe_tot_pre <- as.data.frame(select(LR_all, 16:19))
#Fe_tot <- bind_cols(depth= depth_LR, Fe_tot_pre)
Fe_tot <- bind_cols(age= LR_all$mean_VT22, Fe_tot_pre)

Fe_stacked <- melt(Fe_tot, var.names = names(Fe_tot)[2:ncol(Fe_tot)], measure.vars = names(Fe_tot)[2:ncol(Fe_tot)])

Mn_tot_pre <- as.data.frame(select(LR_all, 12:15))
#Mn_tot <- bind_cols(depth= depth_LR, Mn_tot_pre)
Mn_tot <- bind_cols(age= LR_all$mean_VT22, Mn_tot_pre)

Mn_stacked <- melt(Mn_tot, var.names = names(Mn_tot)[2:ncol(Mn_tot)], measure.vars = names(Mn_tot)[2:ncol(Mn_tot)])

#colorsAll <- c("red", "green", "blue", "purple")
```

##### Percentage contributions

Calculation percentage contribution of the fractions

```{r}
P_tot$sum <- rowSums(P_tot[2:5])
P_tot$F1 <- (P_tot$`P-F1`/P_tot$sum)*100
P_tot$F2 <- (P_tot$`P-F2`/P_tot$sum)*100
P_tot$F3 <- (P_tot$`P-F3`/P_tot$sum)*100
P_tot$F4 <- (P_tot$`P-F4`/P_tot$sum)*100

P_stacked_per <- melt(P_tot, var.names = names(P_tot)[7:ncol(P_tot)], measure.vars = names(P_tot)[7:ncol(P_tot)])


Mn_tot$sum <- rowSums(Mn_tot[2:5])
Mn_tot$F1 <- (Mn_tot$`Mn-F1`/Mn_tot$sum)*100
Mn_tot$F2 <- (Mn_tot$`Mn-F2`/Mn_tot$sum)*100
Mn_tot$F3 <- (Mn_tot$`Mn-F3`/Mn_tot$sum)*100
Mn_tot$F4 <- (Mn_tot$`Mn-F4`/Mn_tot$sum)*100

Mn_stacked_per <- melt(Mn_tot, var.names = names(Mn_tot)[7:ncol(Mn_tot)], measure.vars = names(Mn_tot)[7:ncol(Mn_tot)])

Fe_tot$sum <- rowSums(Fe_tot[2:5])
Fe_tot$F1 <- (Fe_tot$`Fe-F1`/Fe_tot$sum)*100
Fe_tot$F2 <- (Fe_tot$`Fe-F2`/Fe_tot$sum)*100
Fe_tot$F3 <- (Fe_tot$`Fe-F3`/Fe_tot$sum)*100
Fe_tot$F4 <- (Fe_tot$`Fe-F4`/Fe_tot$sum)*100

Fe_stacked_per <- melt(Fe_tot, var.names = names(Fe_tot)[7:ncol(Fe_tot)], measure.vars = names(Fe_tot)[7:ncol(Fe_tot)])
```

##### Plots of the Sum fractions

```{r}
# Create the plot
Pseq_per <- ggplot(P_stacked_per, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("seagreen4", "turquoise3", "seagreen2", "turquoise"))+
coord_flip()+ 
  labs(title = "", y = "P Fraction \n (%)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

Mnseq_per <- ggplot(Mn_stacked_per, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25,position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  #scale_fill_viridis_d(option = "H")+
  scale_fill_manual(values = c("dodgerblue1", "darkorchid2", "dodgerblue3", "darkorchid4"))+
coord_flip()+ 
  labs(title = "", y = "Mn Fraction \n (%)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())


Feseq_per <-  ggplot(Fe_stacked_per, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  #scale_fill_viridis_d(option = "H")+
  scale_fill_manual(values = c("coral3", "goldenrod1", "coral1", "goldenrod4"))+
coord_flip()+ 
  labs(title = "", y = "Fe Fraction \n(%)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())
```

##### Total amounts

Plots of absolute numbers for of fractions

```{r}
# Create the plot
Pseq <- ggplot(P_stacked, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("seagreen4", "turquoise3", "seagreen2", "turquoise"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "P \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

# Create the plot
Mnseq <- ggplot(Mn_stacked, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("dodgerblue1", "darkorchid2", "dodgerblue3", "darkorchid4"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "Mn \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())

# Create the plot
Feseq <- ggplot(Fe_stacked, aes(y = value, x = age, fill = factor(variable))) +  # Group by data values
  geom_bar(stat = "identity",width = 25, position = position_stack(reverse = TRUE)) +  # Preserve data values
scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  labs(title = "",  # Adjust title (optional)
       x = "Age",  # Adjust x-axis label
       y = "Data") +  # Adjust y-axis label
  theme_classic() + # Set theme (optional)
  scale_fill_manual(values = c("coral3", "goldenrod1", "coral1", "goldenrod4"))+
coord_flip()+ 
      scale_y_continuous(position = 'right') +
  labs(title = "", y = "Fe \n (ug/g)", x = "") + 
  theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank()) #legend.position = "top", legend.title = element_blank())
```

### Pigments overlay

```{r}
Fuco2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk, aes(y= pig_all_bulk$Fuco*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = alpha("darkorange",1))+
  geom_col(data= LR_all , aes(y= LR_all$Fuco, x= LR_all$mean_VT22),  stat = "identity", width = 25,fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /100, name = "Fucoxantine \n(nmol/g (wet sed))")) +
  labs(title = " ", y = "Fucoxantine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange1"), axis.title.x.top = element_text(color = "gold3"))

Alpha2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+  
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Alfa_car*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "firebrick4")+
    geom_bar(data= LR_all , aes(y= LR_all$Alfa_car, x= LR_all$mean_VT22),  stat = "identity", width = 25, fill = alpha("coral1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right') +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /100, name = "G1: Alphacarotene \n(nmol/g (wet sed))")) +
  labs(title = " ", y = "G1: Alphacarotene \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "firebrick4"), axis.title.x.top = element_text(color = "coral2"))


Echi2 <- ggplot()+
  geom_bar(data=pig_all_bulk , aes(y= pig_all_bulk$Echi1*10, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "firebrick4") +
geom_bar(data= LR_all, aes(y= LR_all$Echi1/15, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("coral1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /10, name = "G1: Echinenone \n(nmol/g (wet sed))")) +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G1: Echinenone \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "firebrick4"), axis.title.x.top = element_text(color = "coral2"))

Diato2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Diadi*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = alpha("orangered",1))+
    geom_bar(data= LR_all, aes(y= LR_all$Diadi, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("yellow2", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G2: Diadionxanthine \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G2: Diadionxanthine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "orangered3"), axis.title.x.top = element_text(color = "yellow3"))

Canta2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Canta*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "orangered")+
    geom_bar(data= LR_all, aes(y= LR_all$Canta, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("yellow", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G2: Canthaxantine \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G2: Canthaxantine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "orangered3"), axis.title.x.top = element_text(color = "yellow3"))

Zea2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Zeax*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
    geom_bar(data= LR_all, aes(y= LR_all$Zeax, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G3: Zeaxanthine \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G3: Zeaxanthine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))


Lut2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Lute*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
  geom_bar(data= LR_all, aes(y= LR_all$Lute, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G3: Lutein \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G3: Lutein \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))

OHSp2 <- ggplot()+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$OH_Sph.ene*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "darkorange2")+
   geom_bar(data= LR_all, aes(y= LR_all$OH_Sph.ene, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("gold1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /100, name = "G3: OH-Spheroidene \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G3: OH-Spheroidene \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(),axis.title.x.bottom = element_text(color = "darkorange2"), axis.title.x.top = element_text(color = "gold3"))

Iso2 <- ggplot()+
  #geom_point(colour = "green", size= 0.02)+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Isore*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = alpha("springgreen3", 1))+
  geom_bar(data= LR_all, aes(y= LR_all$Isore, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("olivedrab1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G4: Isorenoretine \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G4: Isorenoretine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "Springgreen3"), axis.title.x.top = element_text(color = "olivedrab3"))

Chlx2 <- ggplot()+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Chloxant*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "springgreen3")+
    geom_bar(data= LR_all, aes(y= LR_all$Chloxant, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("olivedrab1", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G4: Chloraxantine \n(nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G4: Chloraxantine \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "springgreen3"), axis.title.x.top = element_text(color = "olivedrab3"))

Oke2 <- ggplot()+
  geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$Okenone*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "purple3")+
    geom_bar(data= LR_all, aes(y= LR_all$Okenone, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("violet", 0.5))+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
  scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "G5: Okenone \n (nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "G5: Okenone \n(nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "purple3"), axis.title.x.top = element_text(color = "violet"))


Phph2 <- ggplot()+
    geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$PhphT*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "darkgreen")+
  geom_bar(data= LR_all, aes(y= LR_all$PhphT, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("green3", 0.5))+
  #geom_point(colour = "green", size= 0.02)+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
   scale_y_continuous(position = 'right',  sec.axis = sec_axis(~ . /100, name = "Total Pheophytines \n (nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophytines \n (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkgreen"), axis.title.x.top = element_text(color = "green3"))

Phide2<- ggplot()+
    geom_bar(data= pig_all_bulk , aes(y= pig_all_bulk$PhideT*100, x= pig_all_bulk$mean_VT22), stat = "identity", width = 25, fill = "darkgreen")+
  geom_bar(data= LR_all, aes(y= LR_all$PhideT, x= LR_all$mean_VT22), stat = "identity", width = 25, fill = alpha("green3", 0.5))+
  #geom_point(colour = "green", size= 0.02)+
  scale_x_reverse(limits = c(15342, 11750), breaks = round(seq(11750, 15342, by = 100), digits = 2)) +
    scale_y_continuous(position = 'right', sec.axis = sec_axis(~ . /100, name = "Total Pheophorbides \n (nmol/g (wet sed))")) +
 #geom_path(alpha=1, size=1, colour = "green") +
  theme_classic() +
   coord_flip()+ 
  labs(title = " ", y = "Total Pheophorbides \n (nmol/g (OM))", x = " ") +
theme(legend.position="none", axis.ticks.y= element_blank(), axis.line.y = element_blank(), axis.text.y = element_blank(), axis.title.x.bottom = element_text(color = "darkgreen"), axis.title.x.top = element_text(color = "green3"))
```

## Plot

```{r}
png(filename = "fig/big_plot_start_t6.png", width = 4000, height = 1500, units = "px", pointsize = 12)
plot_grid(Cya, Chl, GSB, PSB, Ti, Si, SiTi, valve, TOC, Pseq, Pseq_per, Mnseq, Mnseq_per,  Feseq, Feseq_per, Amph, Astr, Epit, Pant, Pseu, PseuR, Stau, StauS, Steph, Fuco2, Alpha2, Echi2, Diato2, Canta2, Zea2, Lut2, OHSp2, Iso2, Chlx2, Oke2, Phph2, Phide2, poll, pollHZM, pollHZM_sta, Pinus, PinusHZM,PinusHZM_sta, Betula, BetulaHZM,BetulaHZM_sta, nrow = 1, align = "h",  rel_widths = c(0.5,0.5))#labels = c('A', 'B'), , label_size = 12
#multiplot(pig.plot,dia.plot)#p3, Pseq, Mnseq, Feseq, p4, p5, p6, Chl, Ti, Fe, cols=13)
dev.off()

```

### 

### Bulk sediment pigments %

```{r}
# pig_na <- na.omit(pig_all_bulk)
# PhideT <- pig_na[,91]
# PhphT <- pig_na[,90]
# pig_age <- pig_na[,2]

pig_na <- na.omit(pig_all_bulk[3:89])
PhphT <- select(na.omit(pig_all_bulk), 91)
PhideT <- select(na.omit(pig_all_bulk), 92)
CarT <- select(na.omit(pig_all_bulk), 90)
d_a_pig <- select(na.omit(pig_all_bulk), 1:2)
```

```{r}
# note that it multiplies the firsi collumn by 100, and the first collumn is depth, it doesn't use this column for % calc.!!!

library(janitor)
perc_pig <- adorn_percentages(pig_na, denominator = "row", na.rm = TRUE) #na.rm = TRUE,
perc_pig_100 <- perc_pig*100

# Remove pollen types where total abundanace is less than 10%
ma.sum <- colSums(perc_pig_100) # Calculate column sums
pig_all_per <- perc_pig_100[, which(ma.sum > 50)] # Subset data


pig_all_per_d <- cbind(d_a_pig, pig_all_per, CarT, PhphT, PhideT)
```

```{r}
data <- as.matrix(pig_all_per)
# Standardise data for CLUSTER analysis
data_standardized <- data
for(i in 1:ncol(data_standardized) ) {
  data_standardized[ , i] <- data_standardized[,i]/sd(data_standardized[,i])
}

ma.dist <- vegdist(data_standardized, method="euclidean", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE) 
ma.chclust <- chclust(ma.dist, method="coniss")
```

### Plot

```{r}
# Define colour scheme
p.col <- c(rep("orange", times=4), rep("gold2", times=8), rep("purple", times=5) ,rep("goldenrod", times=5), rep("green", times=19))
# Define y axis 
y.scale <- pig_all_per_d$mean_VT22

# Filled plot with cluster analysis
png(filename = "fig/pig_50perc_cut_percentages_age_CONISSclu6.png", width = 1754, height = 1240, units = "px", pointsize = 12)
pig.plot2 <- strat.plot(pig_all_per, yvar=y.scale, y.tks=y.scale,  y.rev=TRUE, plot.line=FALSE, plot.poly=TRUE, plot.bar=TRUE, col.bar="black", col.poly=p.col, col.poly.line="black", scale.percent=TRUE, xSpace=0.01, x.pc.inc=5, x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2, col.exag="auto", exag.mult=3, exag.alpha=0.50, title="Holzmaar pigments LG ", ylabel="Depth", clust = ma.chclust)

# Add cluster zones
addClustZone(pig.plot2, ma.chclust, nZone=6, lwd=1.5, lty=2, col="grey25")
dev.off()
```
