---
title: "Sequential Extraction Plots - Fe, Mn, P"
author: "Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Libraries

```{r libraries}
library(readxl)
library(tidyverse)
library(patchwork)
library(scales)
```

## Load Data

```{r load_data}
# Load the LR_data (contains Okenone and sequential extraction)
LR_data <- read_excel("../data/SourceData/LR_data.xlsx", sheet = "Redox_Dia_PigOMug")

# Load HSI-XRF data (contains RABD845calib and mean_VT22)
HSI_data <- read_excel("../data/SourceData/HZM_HSI-XRF_1mm_calibrated_napari_2025-01-13_V3.xlsx")

# Check the data structure
glimpse(LR_data)
glimpse(HSI_data)
```

## Prepare Sequential Extraction Data

```{r prepare_data}
# Select relevant columns for Fe, Mn, and P
seq_extract <- LR_data %>%
  select(comp_depth_mm, mean_VT22, 
         `P-F1`, `P-F2`, `P-F3`, `P-F4`,
         `Mn-F1`, `Mn-F2`, `Mn-F3`, `Mn-F4`,
         `Fe-F1`, `Fe-F2`, `Fe-F3`, `Fe-F4`,
         Okenone) %>%
  drop_na(mean_VT22)  # Remove rows without age data

# Create a complete dataset with no missing fractions
seq_extract_complete <- seq_extract %>%
  filter(!is.na(`P-F1`) & !is.na(`P-F2`) & !is.na(`P-F3`) & !is.na(`P-F4`) &
         !is.na(`Mn-F1`) & !is.na(`Mn-F2`) & !is.na(`Mn-F3`) & !is.na(`Mn-F4`) &
         !is.na(`Fe-F1`) & !is.na(`Fe-F2`) & !is.na(`Fe-F3`) & !is.na(`Fe-F4`))

cat("Original dataset rows:", nrow(seq_extract), "\n")
cat("Complete dataset rows (no missing fractions):", nrow(seq_extract_complete), "\n")
cat("Rows removed:", nrow(seq_extract) - nrow(seq_extract_complete), "\n")

cat("Original dataset rows:", nrow(seq_extract), "\n")
cat("Complete dataset rows (no missing fractions):", nrow(seq_extract_complete), "\n")
cat("Rows removed:", nrow(seq_extract) - nrow(seq_extract_complete), "\n")

# Reshape COMPLETE data for plotting
seq_extract_long <- seq_extract_complete %>%
  pivot_longer(cols = matches("F[1-4]$"), 
               names_to = c("Element", "Fraction"), 
               names_pattern = "(.+)-(F[1-4])",
               values_to = "Value") %>%
  mutate(Fraction = factor(Fraction, levels = c("F1", "F2", "F3", "F4")),
         Element = factor(Element, levels = c("P", "Mn", "Fe")))

# Calculate percentages for 100% stacked plot
seq_extract_pct <- seq_extract_long %>%
  group_by(comp_depth_mm, mean_VT22, Element) %>%
  mutate(Total = sum(Value, na.rm = FALSE),  # Don't use na.rm since we filtered complete data
         Percentage = if_else(Total > 0, (Value / Total) * 100, 0)) %>%
  ungroup()

# Verify percentages sum to 100 for each sample/element combination
pct_check <- seq_extract_pct %>%
  group_by(comp_depth_mm, mean_VT22, Element) %>%
  summarise(pct_sum = sum(Percentage), .groups = 'drop')

cat("\nPercentage sums (should all be ~100):\n")
cat("Min:", min(pct_check$pct_sum), "\n")
cat("Max:", max(pct_check$pct_sum), "\n")
cat("Mean:", mean(pct_check$pct_sum), "\n")

# Summary statistics
summary(seq_extract_long)

# Calculate Fe/P molar ratio for Fraction 2
# Atomic weights: Fe = 55.845 g/mol, P = 30.974 g/mol
fe_p_ratio <- seq_extract_complete %>%
  mutate(Fe_F2_mol = `Fe-F2` / 55.845,  # Convert to moles
         P_F2_mol = `P-F2` / 30.974,     # Convert to moles
         Fe_P_molar_ratio = Fe_F2_mol / P_F2_mol) %>%
  select(comp_depth_mm, mean_VT22, `Fe-F2`, `P-F2`, Fe_F2_mol, P_F2_mol, Fe_P_molar_ratio) %>%
  filter(!is.na(Fe_P_molar_ratio), !is.infinite(Fe_P_molar_ratio))

cat("\nFe/P Molar Ratio Statistics (Fraction 2):\n")
cat("Min:", min(fe_p_ratio$Fe_P_molar_ratio, na.rm = TRUE), "\n")
cat("Max:", max(fe_p_ratio$Fe_P_molar_ratio, na.rm = TRUE), "\n")
cat("Mean:", mean(fe_p_ratio$Fe_P_molar_ratio, na.rm = TRUE), "\n")
cat("Median:", median(fe_p_ratio$Fe_P_molar_ratio, na.rm = TRUE), "\n")
```

## Define Color Palette

```{r colors}
# Define a nice color palette for the 4 fractions
fraction_colors <- c("F1" = "#2166AC", 
                     "F2" = "#4393C3", 
                     "F3" = "#FDAE61", 
                     "F4" = "#D73027")

# Alternative palettes to try:
# viridis: c("#440154", "#31688E", "#35B779", "#FDE724")
# earth tones: c("#8C510A", "#D8B365", "#5AB4AC", "#01665E")
```

## Stacked Bar Plots (Absolute Values)

```{r stacked_absolute, fig.width=12, fig.height=8}
# Create stacked bar plot with absolute values
p_stacked <- ggplot(seq_extract_long, 
                    aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors,
                    labels = c("Fraction 1", "Fraction 2", "Fraction 3", "Fraction 4")) +
  facet_wrap(~ Element, scales = "free_y", ncol = 1,
             labeller = labeller(Element = c("P" = "Phosphorus (P)",
                                             "Mn" = "Manganese (Mn)",
                                             "Fe" = "Iron (Fe)"))) +
  labs(x = "Age (years BP)", 
       y = "Concentration",
       title = "Sequential Extraction - Absolute Values",
       subtitle = "Fe, Mn, and P fractions over time",
       fill = "Extraction Fraction") +
  theme_bw() +
  theme(legend.position = "right",
        strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(size = 11, face = "bold"),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)) +
  scale_x_reverse()  # Reverse x-axis for stratigraphic convention

print(p_stacked)
```

## 100% Stacked Bar Plots (Relative Proportions)

```{r stacked_percentage, fig.width=12, fig.height=8}
# Create 100% stacked bar plot
# Use position = "fill" to ensure proper 100% stacking
p_stacked_pct <- ggplot(seq_extract_long, 
                        aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "fill", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors,
                    labels = c("Fraction 1", "Fraction 2", "Fraction 3", "Fraction 4")) +
  facet_wrap(~ Element, scales = "fixed", ncol = 1,
             labeller = labeller(Element = c("P" = "Phosphorus (P)",
                                             "Mn" = "Manganese (Mn)",
                                             "Fe" = "Iron (Fe)"))) +
  labs(x = "Age (years BP)", 
       y = "Relative Proportion (%)",
       title = "Sequential Extraction - Relative Proportions",
       subtitle = "Fe, Mn, and P fractions over time (100% stacked)",
       fill = "Extraction Fraction") +
  theme_bw() +
  theme(legend.position = "right",
        strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(size = 11, face = "bold"),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)) +
  scale_x_reverse() +  # Reverse x-axis for stratigraphic convention
  scale_y_continuous(labels = scales::percent_format())

print(p_stacked_pct)
```

## Alternative Visualization - Side by Side Comparison

```{r side_by_side, fig.width=14, fig.height=10}
# Create individual plots for each element with both absolute and percentage
create_element_plot <- function(element_name, display_name) {
  
  # Absolute plot
  p_abs <- seq_extract_long %>%
    filter(Element == element_name) %>%
    ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
    geom_col(position = "stack", alpha = 0.8, width = 20) +
    scale_fill_manual(values = fraction_colors) +
    labs(x = "Age (years BP)", 
         y = "Concentration",
         title = paste(display_name, "- Absolute")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(size = 11, face = "bold")) +
    scale_x_reverse()
  
  # Percentage plot - use position = "fill" for proper 100% stacking
  p_pct <- seq_extract_long %>%
    filter(Element == element_name) %>%
    ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
    geom_col(position = "fill", alpha = 0.8, width = 20) +
    scale_fill_manual(values = fraction_colors,
                      labels = c("Fraction 1", "Fraction 2", "Fraction 3", "Fraction 4")) +
    labs(x = "Age (years BP)", 
         y = "Proportion (%)",
         title = paste(display_name, "- Relative (100%)")) +
    theme_bw() +
    theme(legend.position = "right",
          plot.title = element_text(size = 11, face = "bold")) +
    scale_x_reverse() +
    scale_y_continuous(labels = scales::percent_format())
  
  return(p_abs + p_pct)
}

# Create plots for each element
p_P <- create_element_plot("P", "Phosphorus (P)")
p_Mn <- create_element_plot("Mn", "Manganese (Mn)")
p_Fe <- create_element_plot("Fe", "Iron (Fe)")

# Combine all plots
combined_plot <- (p_P / p_Mn / p_Fe) + 
  plot_layout(guides = 'collect') +
  plot_annotation(
    title = 'Sequential Extraction Results: Fe, Mn, and P',
    subtitle = 'Absolute concentrations (left) and relative proportions (right)',
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 12))
  )

print(combined_plot)
```

## Bar Chart Version (Alternative Style)

```{r bar_chart, fig.width=12, fig.height=8}
# For a more discrete look, we can use geom_col with smaller bin widths
# First create bins
seq_extract_binned <- seq_extract_long %>%
  mutate(age_bin = cut(mean_VT22, breaks = 30))

# Stacked bar chart
p_bar_stacked <- ggplot(seq_extract_long, 
                        aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "stack", width = 20, alpha = 0.8) +
  scale_fill_manual(values = fraction_colors,
                    labels = c("Fraction 1", "Fraction 2", "Fraction 3", "Fraction 4")) +
  facet_wrap(~ Element, scales = "free_y", ncol = 1,
             labeller = labeller(Element = c("P" = "Phosphorus (P)",
                                             "Mn" = "Manganese (Mn)",
                                             "Fe" = "Iron (Fe)"))) +
  labs(x = "Age (years BP)", 
       y = "Concentration",
       title = "Sequential Extraction - Bar Chart (Absolute Values)",
       fill = "Extraction Fraction") +
  theme_bw() +
  theme(legend.position = "right",
        strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(size = 11, face = "bold")) +
  scale_x_reverse()

print(p_bar_stacked)
```

## Combined Plot with Okenone and Bacteriopheophytin-a

```{r combined_with_pigments, fig.width=14, fig.height=12}
# Prepare Okenone data from LR_data
okenone_data <- LR_data %>%
  select(comp_depth_mm, mean_VT22, Okenone) %>%
  filter(!is.na(mean_VT22), !is.na(Okenone))

# Prepare HSI data - use mean_VT22 and RABD845calib directly
hsi_pigment <- HSI_data %>%
  select(mean_VT22, RABD845calib) %>%
  filter(!is.na(mean_VT22), !is.na(RABD845calib))

# Determine common x-axis range for all plots
age_min <- min(c(seq_extract_long$mean_VT22, okenone_data$mean_VT22, hsi_pigment$mean_VT22, fe_p_ratio$mean_VT22), na.rm = TRUE)
age_max <- max(c(seq_extract_long$mean_VT22, okenone_data$mean_VT22, hsi_pigment$mean_VT22, fe_p_ratio$mean_VT22), na.rm = TRUE)

cat("Common age range for all plots:", age_min, "to", age_max, "\n")

# Create individual plots for each element with pigments
# P with Okenone
p_P_okenone <- ggplot() +
  geom_col(data = seq_extract_long %>% filter(Element == "P"),
           aes(x = mean_VT22, y = Value, fill = Fraction),
           position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors, name = "P Fraction") +
  scale_x_reverse(limits = c(age_max, age_min)) +
  labs(y = "P Concentration", x = NULL) +
  theme_bw() +
  theme(legend.position = "right") +
  ggtitle("Phosphorus (P)")

p_okenone <- ggplot(okenone_data, aes(x = mean_VT22, y = Okenone)) +
  geom_line(color = "#E41A1C", size = 1) +
  geom_point(color = "#E41A1C", size = 2, alpha = 0.6) +
  scale_x_reverse(limits = c(age_max, age_min)) +
  labs(y = "Okenone", x = NULL) +
  theme_bw()

# Mn with Bacteriopheophytin-a (RABD845)
p_Mn_bphe <- ggplot() +
  geom_col(data = seq_extract_long %>% filter(Element == "Mn"),
           aes(x = mean_VT22, y = Value, fill = Fraction),
           position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors, name = "Mn Fraction") +
  scale_x_reverse(limits = c(age_max, age_min)) +
  labs(y = "Mn Concentration", x = NULL) +
  theme_bw() +
  theme(legend.position = "right") +
  ggtitle("Manganese (Mn)")

p_bphe <- ggplot(hsi_pigment, aes(x = mean_VT22, y = RABD845calib)) +
  geom_line(color = "#377EB8", size = 1, alpha = 0.7) +
  scale_x_reverse(limits = c(age_max, age_min)) +
  labs(y = "Bacteriopheophytin-a\n(RABD845 calibrated)", x = NULL) +
  theme_bw()

# Fe alone
p_Fe_alone <- ggplot() +
  geom_col(data = seq_extract_long %>% filter(Element == "Fe"),
           aes(x = mean_VT22, y = Value, fill = Fraction),
           position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors, name = "Fe Fraction") +
  scale_x_reverse(limits = c(age_max, age_min)) +
  labs(y = "Fe Concentration", x = NULL) +
  theme_bw() +
  theme(legend.position = "right") +
  ggtitle("Iron (Fe)")

# Fe/P Molar Ratio (Fraction 2)
p_fe_p_ratio <- ggplot(fe_p_ratio, aes(x = mean_VT22, y = Fe_P_molar_ratio)) +
  geom_line(color = "#4DAF4A", size = 1) +
  geom_point(color = "#4DAF4A", size = 2, alpha = 0.6) +
  scale_x_reverse(limits = c(age_max, age_min)) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red", size = 0.8) +
  annotate("text", x = age_min + (age_max - age_min) * 0.1, y = 2.3, 
           label = "Fe/P = 2", color = "red", size = 4) +
  labs(y = "Fe/P Molar Ratio\n(Fraction 2)", x = "Age (years BP)") +
  theme_bw()

# Combine all plots
combined_pigments <- (p_P_okenone / p_okenone / p_Mn_bphe / p_bphe / p_Fe_alone / p_fe_p_ratio) +
  plot_layout(heights = c(2, 1, 2, 1, 2, 1)) +
  plot_annotation(
    title = 'Sequential Extraction with Pigment Proxies and Fe/P Ratio',
    subtitle = 'Fe, Mn, and P fractions with Okenone, Bacteriopheophytin-a, and Fe/P molar ratio (F2)',
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 12))
  )

print(combined_pigments)
```

## Alternative: Side-by-side with Pigments

```{r side_by_side_pigments, fig.width=16, fig.height=10}
# Create a more compact side-by-side layout
# Absolute values on left, pigments on right

# P and Mn absolute + 100% stacked
p_P_abs <- seq_extract_long %>%
  filter(Element == "P") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors) +
  labs(x = NULL, y = "P Concentration", title = "Phosphorus") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_reverse()

p_P_rel <- seq_extract_long %>%
  filter(Element == "P") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "fill", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors) +
  labs(x = NULL, y = "P Proportion", title = "P (100%)") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_reverse() +
  scale_y_continuous(labels = scales::percent_format())

p_okenone_plot <- ggplot(okenone_data, aes(x = mean_VT22, y = Okenone)) +
  geom_line(color = "#E41A1C", size = 1) +
  geom_point(color = "#E41A1C", size = 1.5, alpha = 0.6) +
  labs(x = NULL, y = "Okenone", title = "Okenone") +
  theme_bw() +
  scale_x_reverse()

p_Mn_abs <- seq_extract_long %>%
  filter(Element == "Mn") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors) +
  labs(x = NULL, y = "Mn Concentration", title = "Manganese") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_reverse()

p_Mn_rel <- seq_extract_long %>%
  filter(Element == "Mn") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "fill", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors) +
  labs(x = NULL, y = "Mn Proportion", title = "Mn (100%)") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_reverse() +
  scale_y_continuous(labels = scales::percent_format())

p_bphe_plot <- ggplot(hsi_pigment, aes(x = mean_VT22, y = RABD845calib)) +
  geom_line(color = "#377EB8", size = 1, alpha = 0.7) +
  labs(x = NULL, y = "Bphe-a", title = "Bacteriopheophytin-a") +
  theme_bw() +
  scale_x_reverse()

p_Fe_abs <- seq_extract_long %>%
  filter(Element == "Fe") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "stack", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors, 
                    labels = c("Fraction 1", "Fraction 2", "Fraction 3", "Fraction 4")) +
  labs(x = "Age (years BP)", y = "Fe Concentration", title = "Iron") +
  theme_bw() +
  theme(legend.position = "right") +
  scale_x_reverse()

p_Fe_rel <- seq_extract_long %>%
  filter(Element == "Fe") %>%
  ggplot(aes(x = mean_VT22, y = Value, fill = Fraction)) +
  geom_col(position = "fill", alpha = 0.8, width = 20) +
  scale_fill_manual(values = fraction_colors) +
  labs(x = "Age (years BP)", y = "Fe Proportion", title = "Fe (100%)") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_reverse() +
  scale_y_continuous(labels = scales::percent_format())

# Fe/P Molar Ratio plot for grid
p_fe_p_ratio_grid <- ggplot(fe_p_ratio, aes(x = mean_VT22, y = Fe_P_molar_ratio)) +
  geom_line(color = "#4DAF4A", size = 1) +
  geom_point(color = "#4DAF4A", size = 1.5, alpha = 0.6) +
  geom_hline(yintercept = 1.5, linetype = "dashed", color = "red", size = 0.8) +
  annotate("text", x = max(fe_p_ratio$mean_VT22) * 0.85, y = 2.3, 
           label = "Fe/P = 2", color = "red", size = 3.5) +
  labs(x = "Age (years BP)", y = "Fe/P Ratio", title = "Fe/P Molar Ratio (F2)") +
  theme_bw() +
  scale_x_reverse()

# Combine in grid
combined_grid <- (p_P_abs | p_P_rel | p_okenone_plot) /
                 (p_Mn_abs | p_Mn_rel | p_bphe_plot) /
                 (p_Fe_abs | p_Fe_rel | p_fe_p_ratio_grid) +
  plot_layout(guides = 'collect') +
  plot_annotation(
    title = 'Sequential Extraction with Pigment Proxies - Comprehensive View',
    subtitle = 'Absolute concentrations, relative proportions, pigment indicators, and Fe/P molar ratio',
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 12))
  )

print(combined_grid)
```

## Correlation Analysis: Sequential Extraction vs Pigment Proxies

```{r correlation_analysis, fig.width=12, fig.height=10}
# Load correlation library
library(corrplot)

# Prepare data for correlation: merge sequential extraction with pigment data
# First, create totals and fractions for each element
seq_extract_totals <- seq_extract_complete %>%
  mutate(P_total = `P-F1` + `P-F2` + `P-F3` + `P-F4`,
         Mn_total = `Mn-F1` + `Mn-F2` + `Mn-F3` + `Mn-F4`,
         Fe_total = `Fe-F1` + `Fe-F2` + `Fe-F3` + `Fe-F4`) %>%
  select(comp_depth_mm, mean_VT22,
         P_F1 = `P-F1`, P_F2 = `P-F2`, P_F3 = `P-F3`, P_F4 = `P-F4`, P_total,
         Mn_F1 = `Mn-F1`, Mn_F2 = `Mn-F2`, Mn_F3 = `Mn-F3`, Mn_F4 = `Mn-F4`, Mn_total,
         Fe_F1 = `Fe-F1`, Fe_F2 = `Fe-F2`, Fe_F3 = `Fe-F3`, Fe_F4 = `Fe-F4`, Fe_total,
         Okenone) %>%
  drop_na(Okenone)  # Remove rows without Okenone data

# Merge with HSI pigment data (RABD845)
# Use mean_VT22 as the merge key - this is the age scale
# Read RABD845 from the main HSI Excel file, not the interpolated CSV
HSI_data_raw <- read_excel("../data/SourceData/HZM_HSI-XRF_1mm_calibrated_napari_2025-01-13_V3.xlsx") %>%
  select(mean_VT22, RABD845calib) %>%
  drop_na(mean_VT22, RABD845calib) %>%
  rename(RABD845 = RABD845calib)

# Round mean_VT22 for merging to handle slight differences in decimal places
hsi_rabd <- HSI_data_raw %>%
  mutate(mean_VT22_rounded = round(mean_VT22, 0))

# Prepare seq_extract_totals with rounded mean_VT22
seq_extract_for_merge <- seq_extract_totals %>%
  mutate(mean_VT22_rounded = round(mean_VT22, 0))

# Merge the datasets by age (mean_VT22)
correlation_data <- seq_extract_for_merge %>%
  left_join(hsi_rabd %>% select(mean_VT22_rounded, RABD845), 
            by = "mean_VT22_rounded") %>%
  drop_na(RABD845, Okenone)  # Keep only rows with both sequential extraction, Okenone, and RABD845

cat("Samples with complete correlation data:", nrow(correlation_data), "\n")

# Check for any remaining NAs
cat("\nMissing values in correlation_data:\n")
print(colSums(is.na(correlation_data)))

# Remove any rows with NaN or infinite values
correlation_data <- correlation_data %>%
  filter_all(all_vars(is.finite(.)))

# Calculate Pearson correlations for P fractions with Okenone and RABD845
cat("\n=== PHOSPHORUS (P) ===\n")
cat("P correlations with Okenone:\n")

# Helper function to safely calculate correlations
safe_cor_test <- function(x, y, name1, name2) {
  tryCatch({
    # Count finite observations
    valid_obs <- sum(is.finite(x) & is.finite(y))
    if (valid_obs < 3) {
      cat(name1, "vs", name2, ": Not enough finite observations (n =", valid_obs, ")\n")
      return(NULL)
    }
    result <- cor.test(x, y, method = "pearson")
    cat(name1, "vs", name2, ": r =", round(result$estimate, 3), 
        ", p =", format(result$p.value, scientific = TRUE, digits = 3), 
        "(n =", valid_obs, ")\n")
    return(result)
  }, error = function(e) {
    cat(name1, "vs", name2, ": Error -", e$message, "\n")
    return(NULL)
  })
}

p_okenone_f1 <- safe_cor_test(correlation_data$P_F1, correlation_data$Okenone, "P-F1", "Okenone")
p_okenone_f2 <- safe_cor_test(correlation_data$P_F2, correlation_data$Okenone, "P-F2", "Okenone")
p_okenone_f3 <- safe_cor_test(correlation_data$P_F3, correlation_data$Okenone, "P-F3", "Okenone")
p_okenone_f4 <- safe_cor_test(correlation_data$P_F4, correlation_data$Okenone, "P-F4", "Okenone")
p_okenone_total <- safe_cor_test(correlation_data$P_total, correlation_data$Okenone, "P-Total", "Okenone")

cat("\nP correlations with RABD845:\n")
p_rabd_f1 <- safe_cor_test(correlation_data$P_F1, correlation_data$RABD845, "P-F1", "RABD845")
p_rabd_f2 <- safe_cor_test(correlation_data$P_F2, correlation_data$RABD845, "P-F2", "RABD845")
p_rabd_f3 <- safe_cor_test(correlation_data$P_F3, correlation_data$RABD845, "P-F3", "RABD845")
p_rabd_f4 <- safe_cor_test(correlation_data$P_F4, correlation_data$RABD845, "P-F4", "RABD845")
p_rabd_total <- safe_cor_test(correlation_data$P_total, correlation_data$RABD845, "P-Total", "RABD845")

# Calculate correlations for Mn fractions
cat("\n=== MANGANESE (Mn) ===\n")
cat("Mn correlations with Okenone:\n")
mn_okenone_f1 <- safe_cor_test(correlation_data$Mn_F1, correlation_data$Okenone, "Mn-F1", "Okenone")
mn_okenone_f2 <- safe_cor_test(correlation_data$Mn_F2, correlation_data$Okenone, "Mn-F2", "Okenone")
mn_okenone_f3 <- safe_cor_test(correlation_data$Mn_F3, correlation_data$Okenone, "Mn-F3", "Okenone")
mn_okenone_f4 <- safe_cor_test(correlation_data$Mn_F4, correlation_data$Okenone, "Mn-F4", "Okenone")
mn_okenone_total <- safe_cor_test(correlation_data$Mn_total, correlation_data$Okenone, "Mn-Total", "Okenone")

cat("\nMn correlations with RABD845:\n")
mn_rabd_f1 <- safe_cor_test(correlation_data$Mn_F1, correlation_data$RABD845, "Mn-F1", "RABD845")
mn_rabd_f2 <- safe_cor_test(correlation_data$Mn_F2, correlation_data$RABD845, "Mn-F2", "RABD845")
mn_rabd_f3 <- safe_cor_test(correlation_data$Mn_F3, correlation_data$RABD845, "Mn-F3", "RABD845")
mn_rabd_f4 <- safe_cor_test(correlation_data$Mn_F4, correlation_data$RABD845, "Mn-F4", "RABD845")
mn_rabd_total <- safe_cor_test(correlation_data$Mn_total, correlation_data$RABD845, "Mn-Total", "RABD845")

# Calculate correlations for Fe fractions
cat("\n=== IRON (Fe) ===\n")
cat("Fe correlations with Okenone:\n")
fe_okenone_f1 <- safe_cor_test(correlation_data$Fe_F1, correlation_data$Okenone, "Fe-F1", "Okenone")
fe_okenone_f2 <- safe_cor_test(correlation_data$Fe_F2, correlation_data$Okenone, "Fe-F2", "Okenone")
fe_okenone_f3 <- safe_cor_test(correlation_data$Fe_F3, correlation_data$Okenone, "Fe-F3", "Okenone")
fe_okenone_f4 <- safe_cor_test(correlation_data$Fe_F4, correlation_data$Okenone, "Fe-F4", "Okenone")
fe_okenone_total <- safe_cor_test(correlation_data$Fe_total, correlation_data$Okenone, "Fe-Total", "Okenone")

cat("\nFe correlations with RABD845:\n")
fe_rabd_f1 <- safe_cor_test(correlation_data$Fe_F1, correlation_data$RABD845, "Fe-F1", "RABD845")
fe_rabd_f2 <- safe_cor_test(correlation_data$Fe_F2, correlation_data$RABD845, "Fe-F2", "RABD845")
fe_rabd_f3 <- safe_cor_test(correlation_data$Fe_F3, correlation_data$RABD845, "Fe-F3", "RABD845")
fe_rabd_f4 <- safe_cor_test(correlation_data$Fe_F4, correlation_data$RABD845, "Fe-F4", "RABD845")
fe_rabd_total <- safe_cor_test(correlation_data$Fe_total, correlation_data$RABD845, "Fe-Total", "RABD845")
```

## Correlation Matrices Visualization

```{r correlation_matrices, fig.width=14, fig.height=6}
cat("\nDiagnostic information for correlation matrices:\n")
cat("Correlation data dimensions:", dim(correlation_data), "\n")
cat("Number of complete cases (all non-NA):", sum(complete.cases(correlation_data)), "\n")

# If we have data, create correlation matrices
if (nrow(correlation_data) > 0) {
  
  # Try to create correlation matrix with Okenone
  if (sum(!is.na(correlation_data$Okenone)) > 2) {
    cat("\nCreating Okenone correlation matrix...\n")
    tryCatch({
      corr_matrix_okenone <- correlation_data %>%
        select(P_F1, P_F2, P_F3, P_F4, P_total,
               Mn_F1, Mn_F2, Mn_F3, Mn_F4, Mn_total,
               Fe_F1, Fe_F2, Fe_F3, Fe_F4, Fe_total,
               Okenone) %>%
        cor(use = "pairwise.complete.obs")
      
      # Plot Okenone correlation matrix
      par(mar = c(0, 0, 2, 0))
      corrplot(corr_matrix_okenone, 
               method = "color",
               type = "full",
               addCoef.col = "black",
               number.cex = 0.7,
               diag = TRUE,
               title = "Correlation with Okenone",
               tl.cex = 0.8)
    }, error = function(e) {
      cat("Could not create Okenone correlation matrix:", e$message, "\n")
    })
  } else {
    cat("Not enough Okenone data for correlation matrix\n")
  }
  
  # Try to create correlation matrix with RABD845
  if (sum(!is.na(correlation_data$RABD845)) > 2) {
    cat("Creating RABD845 correlation matrix...\n")
    tryCatch({
      corr_matrix_rabd <- correlation_data %>%
        select(P_F1, P_F2, P_F3, P_F4, P_total,
               Mn_F1, Mn_F2, Mn_F3, Mn_F4, Mn_total,
               Fe_F1, Fe_F2, Fe_F3, Fe_F4, Fe_total,
               RABD845) %>%
        cor(use = "pairwise.complete.obs")
      
      # Plot RABD845 correlation matrix
      par(mar = c(0, 0, 2, 0))
      corrplot(corr_matrix_rabd, 
               method = "color",
               type = "full",
               addCoef.col = "black",
               number.cex = 0.7,
               diag = TRUE,
               title = "Correlation with RABD845",
               tl.cex = 0.8)
    }, error = function(e) {
      cat("Could not create RABD845 correlation matrix:", e$message, "\n")
    })
  } else {
    cat("Not enough RABD845 data for correlation matrix\n")
  }
  
  par(mfrow = c(1, 1))
} else {
  cat("\nERROR: correlation_data is empty!\n")
  cat("This means there are no rows with both Okenone and RABD845 values.\n")
  cat("Check if the data merging was successful and both pigment columns have data.\n")
}
```

## Scatter Plots: Selected Correlations

```{r scatter_correlations, fig.width=14, fig.height=10}
# Extract correlation coefficients and p-values for plotting
create_scatter_plot <- function(x_data, y_data, x_label, y_label, title) {
  cor_test <- cor.test(x_data, y_data, method = "pearson")
  r <- cor_test$estimate
  p_val <- cor_test$p.value
  
  # Determine significance
  sig_label <- if_else(p_val < 0.001, "***", 
                       if_else(p_val < 0.01, "**", 
                              if_else(p_val < 0.05, "*", "ns")))
  
  plot_data <- tibble(x = x_data, y = y_data)
  
  ggplot(plot_data, aes(x = x, y = y)) +
    geom_point(size = 3, alpha = 0.6, color = "#2166AC") +
    geom_smooth(method = "lm", se = TRUE, color = "#D73027", alpha = 0.2) +
    labs(x = x_label, 
         y = y_label,
         title = title,
         subtitle = paste0("r = ", round(r, 3), " (p ", sig_label, ")")) +
    theme_bw() +
    theme(plot.title = element_text(size = 11, face = "bold"),
          plot.subtitle = element_text(size = 10))
}

# Create scatter plots for significant or interesting correlations
p1 <- create_scatter_plot(correlation_data$P_F2, correlation_data$Okenone, 
                         "P-F2 Concentration", "Okenone", "P-F2 vs Okenone")
p2 <- create_scatter_plot(correlation_data$P_total, correlation_data$RABD845, 
                         "P-Total Concentration", "RABD845", "P-Total vs RABD845")

p3 <- create_scatter_plot(correlation_data$Mn_F2, correlation_data$Okenone, 
                         "Mn-F2 Concentration", "Okenone", "Mn-F2 vs Okenone")
p4 <- create_scatter_plot(correlation_data$Mn_total, correlation_data$RABD845, 
                         "Mn-Total Concentration", "RABD845", "Mn-Total vs RABD845")

p5 <- create_scatter_plot(correlation_data$Fe_F2, correlation_data$Okenone, 
                         "Fe-F2 Concentration", "Okenone", "Fe-F2 vs Okenone")
p6 <- create_scatter_plot(correlation_data$Fe_total, correlation_data$RABD845, 
                         "Fe-Total Concentration", "RABD845", "Fe-Total vs RABD845")

# Combine scatter plots
scatter_combined <- (p1 | p2) / (p3 | p4) / (p5 | p6) +
  plot_annotation(
    title = 'Scatter Plots: Sequential Extraction vs Pigment Proxies',
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(scatter_combined)
```

## Correlation Summary Table

```{r correlation_summary}
# Create a summary table of all correlations
correlation_summary <- tibble(
  Element = c(rep("P", 5), rep("P", 5), rep("Mn", 5), rep("Mn", 5), rep("Fe", 5), rep("Fe", 5)),
  Pigment = c(rep("Okenone", 10), rep("Okenone", 10), rep("Okenone", 10)),
  Fraction = c("F1", "F2", "F3", "F4", "Total", "F1", "F2", "F3", "F4", "Total",
               "F1", "F2", "F3", "F4", "Total", "F1", "F2", "F3", "F4", "Total",
               "F1", "F2", "F3", "F4", "Total", "F1", "F2", "F3", "F4", "Total"),
  r_value = c(
    # P vs Okenone
    p_okenone_f1$estimate, p_okenone_f2$estimate, p_okenone_f3$estimate, p_okenone_f4$estimate, p_okenone_total$estimate,
    # P vs RABD845
    p_rabd_f1$estimate, p_rabd_f2$estimate, p_rabd_f3$estimate, p_rabd_f4$estimate, p_rabd_total$estimate,
    # Mn vs Okenone
    mn_okenone_f1$estimate, mn_okenone_f2$estimate, mn_okenone_f3$estimate, mn_okenone_f4$estimate, mn_okenone_total$estimate,
    # Mn vs RABD845
    mn_rabd_f1$estimate, mn_rabd_f2$estimate, mn_rabd_f3$estimate, mn_rabd_f4$estimate, mn_rabd_total$estimate,
    # Fe vs Okenone
    fe_okenone_f1$estimate, fe_okenone_f2$estimate, fe_okenone_f3$estimate, fe_okenone_f4$estimate, fe_okenone_total$estimate,
    # Fe vs RABD845
    fe_rabd_f1$estimate, fe_rabd_f2$estimate, fe_rabd_f3$estimate, fe_rabd_f4$estimate, fe_rabd_total$estimate
  ),
  p_value = c(
    # P vs Okenone
    p_okenone_f1$p.value, p_okenone_f2$p.value, p_okenone_f3$p.value, p_okenone_f4$p.value, p_okenone_total$p.value,
    # P vs RABD845
    p_rabd_f1$p.value, p_rabd_f2$p.value, p_rabd_f3$p.value, p_rabd_f4$p.value, p_rabd_total$p.value,
    # Mn vs Okenone
    mn_okenone_f1$p.value, mn_okenone_f2$p.value, mn_okenone_f3$p.value, mn_okenone_f4$p.value, mn_okenone_total$p.value,
    # Mn vs RABD845
    mn_rabd_f1$p.value, mn_rabd_f2$p.value, mn_rabd_f3$p.value, mn_rabd_f4$p.value, mn_rabd_total$p.value,
    # Fe vs Okenone
    fe_okenone_f1$p.value, fe_okenone_f2$p.value, fe_okenone_f3$p.value, fe_okenone_f4$p.value, fe_okenone_total$p.value,
    # Fe vs RABD845
    fe_rabd_f1$p.value, fe_rabd_f2$p.value, fe_rabd_f3$p.value, fe_rabd_f4$p.value, fe_rabd_total$p.value
  )
) %>%
  mutate(
    Pigment = if_else(row_number() <= 10, "Okenone", "RABD845"),
    Significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    ),
    r_rounded = round(r_value, 3),
    p_rounded = format(p_value, scientific = TRUE, digits = 2)
  ) %>%
  select(Element, Pigment, Fraction, r_rounded, Significance, p_rounded) %>%
  rename(r = r_rounded, p_value = p_rounded)

# Display the table
knitr::kable(correlation_summary, 
             caption = "Correlation Summary: Sequential Extraction Fractions vs Pigment Proxies")

# Save the table
write.csv(correlation_summary, "../data/Correlation_Summary.csv", row.names = FALSE)
cat("Correlation summary table saved to ../data/Correlation_Summary.csv\n")
```

## Save Plots

```{r save_plots}
# Save the main plots
ggsave("../figs/Sequential_Extraction_Absolute.pdf", 
       plot = p_stacked, 
       width = 12, height = 8, units = "in")

ggsave("../figs/Sequential_Extraction_Relative.pdf", 
       plot = p_stacked_pct, 
       width = 12, height = 8, units = "in")

ggsave("../figs/Sequential_Extraction_Combined.pdf", 
       plot = combined_plot, 
       width = 14, height = 10, units = "in")

# Save the pigment-combined plots
ggsave("../figs/Sequential_Extraction_with_Pigments.pdf", 
       plot = combined_pigments, 
       width = 14, height = 12, units = "in")

ggsave("../figs/Sequential_Extraction_Grid_with_Pigments.pdf", 
       plot = combined_grid, 
       width = 16, height = 10, units = "in")

# Save correlation visualization plots
ggsave("../figs/Correlation_Matrices.pdf", 
       width = 14, height = 6, units = "in")

ggsave("../figs/Correlation_Scatter_Plots.pdf", 
       plot = scatter_combined, 
       width = 14, height = 10, units = "in")

# Also save as PNG for easy viewing
ggsave("../figs/Sequential_Extraction_Absolute.png", 
       plot = p_stacked, 
       width = 12, height = 8, units = "in", dpi = 300)

ggsave("../figs/Sequential_Extraction_Relative.png", 
       plot = p_stacked_pct, 
       width = 12, height = 8, units = "in", dpi = 300)

ggsave("../figs/Sequential_Extraction_Combined.png", 
       plot = combined_plot, 
       width = 14, height = 10, units = "in", dpi = 300)

ggsave("../figs/Sequential_Extraction_with_Pigments.png", 
       plot = combined_pigments, 
       width = 14, height = 12, units = "in", dpi = 300)

ggsave("../figs/Sequential_Extraction_Grid_with_Pigments.png", 
       plot = combined_grid, 
       width = 16, height = 10, units = "in", dpi = 300)

ggsave("../figs/Correlation_Scatter_Plots.png", 
       plot = scatter_combined, 
       width = 14, height = 10, units = "in", dpi = 300)

print("Plots saved successfully!")
```

## Session Info

```{r session_info}
sessionInfo()
```
